### 2024-11-15 paper_figures.Rmd
initial commit

---
title: "paper figures"
author: "Benedikt Obermayer"
date: "2024/11/12"
output: 
  html_document:
    toc: true
    df_print: paged
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache=TRUE, cache.lazy=FALSE, message=FALSE, warning=FALSE,dev=c('pdf'))
library(ggpubr)
library(stats)
library(dplyr)
library(ggplot2)
library(tidyverse)
library(caret)
library(scales)
library(cowplot)
library(ComplexHeatmap)
library(circlize)
library(ggrepel)
library(car)
library(glmnet)
library(pROC)
library(readxl)
library(variancePartition)
library(RColorBrewer)
library(gtools)
library(grid)
library(DescTools)
library(lme4)
```

setup

```{r definitions}
fig_scale <- 1.
columns <- list('diversity'=c('nclusters_final_downsampled',
                              'nclusters_eff_downsampled',
                              'mean_cluster_size_downsampled',
                              'std_cluster_size_downsampled',
                              'cluster_gini_downsampled',
                              'cluster_entropy_downsampled',
                              'cluster_inverse_simpson_downsampled',
                              'top_clone_occupancy_downsampled',
                              'big_clones_occupancy_downsampled'),
                'isotypes'=c('alpha_ratio_clusters',
                             'frac_clusters_SA1','frac_clusters_SA2','frac_clusters_SE',
                             'frac_clusters_SG1','frac_clusters_SG2','frac_clusters_SG3',
                             'frac_clusters_SG4','frac_clusters_SM',
                             'frac_clusters_Sa','frac_clusters_Sg3',
                             'frac_clusters_Sg2b','frac_clusters_Sg2c',
                             'frac_clusters_Sm','frac_clusters_Sg1'),
                'structural'=c('insert_frequency',
                               'frac_breaks_inversions','frac_breaks_duplications',
                               'median_duplication_size','median_inversion_size'),
                'context'=c('homology_fw','homology_rv',
                            'donor_score_S', 'donor_score_W', 
                            'donor_score_WGCW', 'donor_score_GAGCT',
                            'donor_score_GGGST',
                            'donor_complexity',
                            'acceptor_score_S', 'acceptor_score_W', 
                            'acceptor_score_WGCW', 'acceptor_score_GAGCT',
                            'acceptor_score_GGGST',
                            'acceptor_complexity',
                            'n_homology_switch', 
                            'n_untemplated_switch',
                            'frac_blunt'),
                'breaks'=c('mean_length',
                           'std_length',
                           'mean_GC',
                           'std_GC',
                           'mean_length_SA',
                           'mean_length_SG',
                           'std_length_SA',
                           'std_length_SG',
                           'frac_breaks_single',
                           'frac_breaks_sequential',
                           'frac_breaks_intra',
                           'mean_intra_break_size',
                           'std_intra_break_size',
                           'spread_SM','spread_SA','spread_SG',
                           'spread_SA1','spread_SA2',
                           'spread_SG1','spread_SG2',
                           'spread_SG2B','spread_SG2C',
                           'spread_SG3','spread_SG4',
                           'frac_breaks_SM_SM','frac_breaks_SM_SG',
                           'frac_breaks_SG_SG','frac_breaks_SM_SA',
                           'frac_breaks_SG_SA','frac_breaks_SA_SA'),
                'context_isotype'=c('homology_fw_SM_SA','homology_rv_SM_SA',
                                    'homology_fw_SM_SG','homology_rv_SM_SG',
                                    'donor_score_GAGCT_SM_SG','donor_score_GGGST_SM_SA',
                                    'donor_score_GGGST_SM_SG','donor_score_S_SM_SA',
                                    'donor_score_S_SM_SG','donor_score_W_SM_SA',
                                    'donor_score_W_SM_SG','donor_score_WGCW_SM_SA',
                                    'donor_score_WGCW_SM_SG','donor_score_GAGCT_SM_SA',
                                    'donor_complexity_SM_SG','donor_complexity_SM_SA',
                                    'acceptor_score_GAGCT_SM_SA','acceptor_score_GAGCT_SM_SG',
                                    'acceptor_score_GGGST_SM_SA','acceptor_score_GGGST_SM_SG',
                                    'acceptor_score_S_SM_SA','acceptor_score_WGCW_SM_SG',  
                                    'acceptor_score_S_SM_SG','acceptor_score_W_SM_SA',
                                    'acceptor_score_W_SM_SG','acceptor_score_WGCW_SM_SA',
                                    'acceptor_complexity_SM_SA',
                                    'acceptor_complexity_SM_SG',
                                    'n_homology_switch_SA', 
                                    'n_homology_switch_SG', 
                                    'n_untemplated_switch_SA',
                                    'n_untemplated_switch_SG',
                                    'frac_blunt_SA',
                                    'frac_blunt_SG'),
                'diversity_raw'=c('nclusters_final',
                                  'nclusters_eff',
                                  'mean_cluster_size',
                                  'std_cluster_size',
                                  'cluster_gini',
                                  'cluster_entropy',
                                  'cluster_inverse_simpson',
                                  'top_clone_occupancy',
                                  'big_clones_occupancy'),
                'variants'=c('frac_variants_germline',
                             'frac_variants_transitions',
                             'num_variants')) %>%
  unlist() %>%
  data.frame(col=.) %>%
  tibble::rownames_to_column('category') %>%
  dplyr::mutate(category=factor(gsub('[0-9]*$','',category), 
                                levels=c('library',
                                         'isotypes',
                                         'diversity',
                                         'diversity_raw',
                                         'structural',
                                         'context',
                                         'context_isotype',
                                         'breaks',
                                         'variants')))

cat_cols <- c('structural'='#FD9E40',
              'context'='#FED6AC',
              'context_isotype'='#FED6AC',
              'breaks'='#1E8AC4',
              'isotypes'='#AEF58E',
              'diversity_raw'='#3A7A2C',
              'diversity'='#3A7A2C',
              'variants'='lightpink3',
              'library'='gray40')

genotype_colors <- c('WT'='gray40','Xlf'='#e1c2f0','Shld1'='#b870dc','Shld1_Xlf'='#a447d2',
                     'Trp53bp1 KO'='darksalmon','Brca1 KO'='darkkhaki','Rif1 KO'='skyblue1','Lig4 KO'='mediumpurple2')

tissue_colors <- c('splenocytes'='orange2','lymph node'='orangered3','CH12'='forestgreen')
var_colors <- c('batch'='palegreen2','diagnosis'='tan2','donor'='firebrick4','sequencing'='slategray1','unexplained'='gray80',
                'sex'='steelblue2','age'='darkseagreen4')

diagnosis_colors <- c('control'='gray20','HD'='gray20',
                      'subclass'='lightslategray','CVID'='coral1',
                      'ICOS'='#8DD3C7','Kabuki'='#BEBADA',
                      'IgG4def'='burlywood4','TNFRSF13B'='palegreen4',
                      'other'='#BC80BD','AID'='#FDB462',
                      'ATM'='#A65628','LIG4'='mediumpurple2',
                      'NIPBL'='#FCCDE5','BRCA1'='darkkhaki',
                      'NFKB2'='#FFED6F','PI3K'='#80B1D3')

batch_colors <- setNames(c(RColorBrewer::brewer.pal('Set1',n=9),RColorBrewer::brewer.pal('Set1',n=6)),
                         c('20230728','20230731','20230904','20230908','20231025',
                           '20231030','20240603','20240711','20240725','20220425',
                           '20221118','20221212','20230120','20230213','20240830'))

group_colors <- c('SmB+CD21n'='gray20',
                  'SmB-CD21lo'='orange3',
                  'SmB-CD21n'='lightseagreen')

col_cols <- columns %>% 
  dplyr::mutate(color=cat_cols[as.character(category)]) %>%
  dplyr::mutate(col=gsub('_downsampled|_switch','',col)) %>%
  dplyr::pull(color, col)

my_theme <- function() {
  theme_classic(base_size=7) +
  theme(strip.background=element_blank(),
        strip.placement='outside',
        strip.text = element_text(size=7),
        legend.key.size=unit(2.5, 'mm'),
        legend.spacing=unit(1,'mm'),
        legend.text = element_text(size=7),
        axis.text = element_text(size=7),
        axis.title = element_text(size=7))
}
```

<!-- # get data -->

<!-- ```{r get_data} -->
<!-- # data from Vincendeau for Fig. 2 -->
<!-- # mouse tissue data for Figs. 2 + 3  -->
<!-- # CH12 data for Fig. 3 -->
<!-- # patient data for Fig. 4 -->
<!-- FB.data <- read.csv('data/FB_data.csv',header=1, row.names=1) -->
<!-- FB.data.pooled <- read.csv('data/FB_data_pooled.csv',header=1, row.names=1) -->
<!-- QP.data <- read.csv('data/QP_data.csv',header=1, row.names=1) -->
<!-- ``` -->

# Figure 1 + supplements

plot # clusters vs. # clones

```{r Fig1D, fig.width=1.4 * fig_scale, fig.height=1.4 * fig_scale}
# simulation data for Fig. 1
synthetic.data <- read.csv('data/synthetic_data.csv',header=1, row.names=1)
ggplot(synthetic.data %>% 
         dplyr::mutate(nreads=factor(nreads, levels=c('10k','20k','50k','100k'))), 
       aes(x=nclones, y=nclusters_final, color=nreads)) + 
  geom_abline(slope=1,intercept=0, size=.5, color='gray80') +
  geom_point(size=.5) + 
  geom_smooth(method='lm', alpha=.2, size=.5, se=FALSE) +
  scale_x_log10() + 
  scale_y_log10() +
  my_theme() +
  scale_color_brewer(palette='Set2') +
  stat_cor(method = "pearson", aes(label=..rr.label..), size=2, show.legend=FALSE) +
  theme(legend.position=c(.85,.35),
        legend.background = element_rect(fill=NA)) +
  labs(x='# clones', y='# filtered clusters', color='# reads')
```

cell switch: plot # clusters vs. # cells

```{r Fig1E, fig.width=1.4 * fig_scale, fig.height=1.4 * fig_scale}
# cell number experiment for  Fig. 1
cellswitch.data <- read.csv('data/cellswitch_data.csv',header=1, row.names=1)

ggplot(cellswitch.data %>%
         dplyr::mutate(donor=plyr::revalue(donor,
                                           c('21084'='A',
                                             '21085'='B',
                                             '21086'='C'))),
       aes(x=ncells, y=nclusters_final)) + 
  geom_abline(slope=1,intercept=0, size=.5, color='gray80') +
  geom_point(aes(color=donor),size=.5) + 
  geom_smooth(size=.5, method='lm', alpha=.2, color='black') +
  stat_cor(method = "pearson", aes(label=..rr.label..), size=2, show.legend=FALSE) +
  scale_x_log10() + 
  scale_y_log10() +
  my_theme() +
  scale_color_brewer(palette='Set1') +
  theme(legend.position=c(.8,.25),
        legend.background = element_rect(fill=NA)) +
  labs(x='# input cells', y='# filtered clusters', color='donor')
```

show VDJ results: # clones depending on # of input cells

```{r Fig1F, fig.width=1.4 * fig_scale, fig.height=1.4 * fig_scale}
# bulk VDJ results for Fig. 1
VDJ_results <- read.csv('data/VDJ_results.csv',header=1, row.names=1)

vdj_stats <- VDJ_results %>%
  dplyr::mutate(clone=paste0(gsub('\\([0-9]*\\)$','',allVHitsWithScore,'_',str_length(aaSeqCDR3)))) %>%
  dplyr::group_by(donor,ncells,sample) %>%
  dplyr::summarise(nclusters_final=n_distinct(clone),
                   alpha_ratio_clusters=mean(grepl('IGHA',allCHitsWithScore))) %>%
  dplyr::mutate(method='VDJ')

swibrid_stats <- cellswitch.data %>%
  dplyr::select(donor,ncells,
                nclusters_final,
                alpha_ratio_clusters) %>%
  dplyr::filter(ncells %in% c(50000,100000)) %>%
  dplyr::mutate(method='swibrid')

vdj_cols <- intersect(colnames(swibrid_stats),colnames(vdj_stats))

vdj_df <- rbind(swibrid_stats[,vdj_cols],vdj_stats[,vdj_cols]) %>%
  gather(metric,value, c(nclusters_final,alpha_ratio_clusters)) %>%
  dplyr::mutate(ncells=factor(as.character(as.integer(ncells)), levels=c('50000','100000')))

vdj_df %>% 
  dplyr::filter(metric=='nclusters_final') %>%
  ggplot(aes(x=factor(gsub('000$','',ncells), levels=c('50','100')), y=value)) +
  geom_boxplot(outlier.shape=NA, lwd=.5) +
  geom_point(aes(color=gsub('donor ','',donor)),size=.5) +
  facet_wrap(~method, ncol=2, scales='free_y') + 
  my_theme() +
  scale_color_brewer(palette='Set1') +
  theme(strip.clip='off',
        legend.position='none',
        legend.background =element_rect(fill=NA)) +
  labs(y='# clones',x='# cells / 1000',color='')
```

isotype ratios in VDJ vs. SWIBRID

```{r Fig1G, fig.width=1.4 * fig_scale,fig.height=1.4 * fig_scale}
vdj_df %>% 
  dplyr::mutate(ncells=factor(gsub('000$','k',ncells), levels=c('50k','100k')),
                donor=gsub('donor ','',donor)) %>%
  dplyr::filter(metric=='alpha_ratio_clusters') %>%
  dplyr::group_by(ncells, donor, method) %>%
  dplyr::summarise(mean_alpha_ratio=mean(value),
                   se_alpha_ratio=sd(value)/sqrt(length(value))) %>%
  pivot_wider(id_cols=c(ncells,donor),names_from=method,values_from=c('mean_alpha_ratio','se_alpha_ratio')) %>%
  ggplot(aes(x=mean_alpha_ratio_VDJ, y=mean_alpha_ratio_swibrid,
             xmin=mean_alpha_ratio_VDJ-se_alpha_ratio_VDJ,
             xmax=mean_alpha_ratio_VDJ+se_alpha_ratio_VDJ,
             ymin=mean_alpha_ratio_swibrid-se_alpha_ratio_swibrid,
             ymax=mean_alpha_ratio_swibrid+se_alpha_ratio_swibrid)) +
  geom_point(aes(color=donor,shape=ncells),size=1) +
  geom_errorbar(linewidth=.25) +
  geom_errorbarh(linewidth=.25) +
  scale_x_continuous(breaks=c(.6,.65)) +
  my_theme() +
  scale_color_brewer(palette='Set1') +
  theme(strip.background=element_blank()) +
  labs(y='fraction IGHA swibrid',x='fraction IGHA VDJ',shape='# cells')
```

# Figure 2 + supplements

fraction of explained variance for different feature groups in HD cohort

```{r Fig2B,fig.width=7 * fig_scale,fig.height=1.4 * fig_scale}
my_fitExtractVarPartModel <- function(formula, data) {
  # replace fitExtractVarPartModel from variancePartition package by simple function to run over single features one at a time
  # extract relevant parts of code from calcVarPart.R
  # https://github.com/GabrielHoffman/variancePartition/blob/b578069c0e5dab6234516b12e624e3ce995f850d/R/calcVarPart.R#L292-L363

  vars <- c(all.vars(formula),'Residuals')
  res <- setNames(rep(NA, length(vars)-1),sort(vars[2:length(vars)]))

  try({
    fit <- lmer(formula, data, REML=FALSE)
    w <- weights(fit)
    if (is.null(w)) {
      w <- rep(1, nrow(fit$model))
    }
    varComp <- lapply(lme4::VarCorr(fit), function(fit) attr(fit, "stddev")^2)
    varComp <- varComp[order(names(varComp))]
    idx <- which(colnames(fit@pp$X) != "(Intercept)")
    if (length(idx) > 0) {
      fxeff <- sapply(idx, function(i) {
        fit@pp$X[, i] * lme4::fixef(fit)[i]
      })
      colnames(fxeff) <- colnames(fit@pp$X)[idx]
      N <- nrow(fxeff)
      fixedVar <- apply(fxeff, 2, function(x) {
        weighted.var(x, w) * (N - 1) / N
      })
      varFixedTotal <- weighted.var(rowSums(fxeff), w) * (N - 1) / N
      for (key in names(fixedVar)) {
        varComp[[key]] <- as.array(fixedVar[[key]] / sum(fixedVar) *
                                     varFixedTotal)
      }
    }
    varComp$Residuals <- sigma(fit)^2
    varComp <- unlist(varComp)
    names(varComp) <- gsub("\\.\\(Intercept\\)", "", names(varComp))
    res <- varComp/sum(varComp)
  })
  res
}

# HD data for Fig. 2
HD.data <- read.csv('data/HD_data.csv', header=1, row.names=1)

HD.scaled <- HD.data %>%
  dplyr::filter(cohort=='C1') %>%
  dplyr::group_by(feature) %>%
  dplyr::mutate(value=as.vector(scale(value))) %>%
  dplyr::filter(!is.na(value)) %>%
  dplyr::ungroup()

HD.var <- sapply(unique(HD.scaled$feature), 
                 function(ftr)
                   tryCatch(my_fitExtractVarPartModel(value ~ (1 | donor) + (1 | sex) + (1 | age) + (1 | sequencing) + (1 | batch),
                                                      HD.scaled %>%
                                                        dplyr::filter(feature==ftr)),
                            error=setNames(rep(NA,5), c('donor','sex','age','sequencing','batch')))) %>%
                   t() %>%
  as.data.frame() %>%
  tibble::rownames_to_column('col') %>%
  gather(variable,variance,-col)

cols.here <- c('cluster_entropy_downsampled',
               'cluster_gini_downsampled',
               'mean_cluster_size_downsampled',
               'std_cluster_size_downsampled',
               'top_clone_occupancy_downsampled',
               'alpha_ratio_clusters',
               'frac_clusters_SA1',
               'frac_clusters_SA2',
               'frac_clusters_SG1',
               'frac_clusters_SG2',
               'insert_frequency',
               'frac_breaks_inversions',
               'frac_breaks_duplications',
               'median_duplication_size',
               'median_inversion_size',
               'homology_fw',
               'donor_score_WGCW',
               'acceptor_complexity',
               'n_homology_switch',
               'n_untemplated_switch',
               'frac_breaks_single',
               'frac_breaks_sequential',
               'frac_breaks_intra',
               'spread_SM',
               'spread_SA')

HD.var %>%  dplyr::left_join(columns, by='col') %>%
  dplyr::mutate(variable=factor(gsub('Residuals','unexplained',variable),
                                levels=rev(c('donor','sex','age','batch','sequencing','unexplained'))),
                category=factor(category, levels=c('structural','context','breaks','diversity','isotypes'))) %>%
  dplyr::filter(col %in% cols.here) %>%
  dplyr::mutate(col=factor(gsub('_downsampled|_switch','',col), levels=gsub('_downsampled|_switch','',cols.here))) %>%
  ggplot(aes(x=col, y=100*variance, fill=variable)) + 
  geom_col(width=.8) +
  my_theme() +
  facet_wrap(~category,scales='free_y', ncol=5) +
  theme(strip.clip='off',
        legend.position='bottom',
        legend.box='horizontal') +
  guides(fill=guide_legend(ncol=6)) +
  coord_flip() +
  scale_y_continuous(breaks=c(0,50,100)) +
  scale_fill_manual(values=var_colors,
                    breaks=c('donor','sex','age','batch','sequencing','unexplained')) +
  labs(x='', y='% explained variance', fill='')
```

```{r Fig2D, fig.width=1.75 * fig_scale,fig.height=1.7 * fig_scale}
donor_cutoff <- .25
batch_cutoff <- .35
sequencing_cutoff <- .2
selected.cols <- HD.var %>%
  dplyr::left_join(columns, by='col') %>%
  spread(variable, variance) %>%
  dplyr::filter(!(category %in% c('diversity_raw','context_isotype','library')), 
                !(col %in% c('log10_nreads','mean_GC')), 
                donor >= donor_cutoff, 
                sequencing < sequencing_cutoff, 
                batch < batch_cutoff) %>%
  dplyr::arrange(category) %>%
  dplyr::pull(col)

HD.var %>%  dplyr::left_join(columns, by='col') %>%
  dplyr::mutate(variable=factor(gsub('Residuals','unexplained',variable),
                                levels=c('sequencing','batch','donor','sex','age'))) %>%
  dplyr::filter(variable!='unexplained', category %in% c('library','isotypes','diversity','structural','context','breaks','variants')) %>%
  ggplot(aes(x=variable, y=100*variance)) + 
  geom_boxplot(aes(fill=variable), outlier.shape=NA, lwd=.5, legend=FALSE) + 
  geom_jitter(aes(color=ifelse(col %in% selected.cols, 'selected','unused')), position=position_jitter(width=.25), size=.25) +
  annotate("rect", xmin = .5, xmax = 1.5, ymax = 100*sequencing_cutoff, ymin = 0, fill='orangered', linetype=1, size=.25, alpha=.25) +
  annotate("rect", xmin = 1.5, xmax = 2.5, ymax = 100*batch_cutoff, ymin = 0, fill='orangered', linetype=1, size=.25, alpha=.25) +
  annotate("rect", xmin = 2.5, xmax = 3.5, ymax = Inf, ymin = 100*donor_cutoff, fill='orangered', linetype=1, size=.25, alpha=.25) +
  annotate('text',.6,78,label='unused',color='black',size=2, hjust=0) +
  annotate('text',.6,73,label='selected',color='orangered',size=2, hjust=0) +
  my_theme() +
  theme(axis.text.x=element_text(angle=45, hjust=1, vjust=1),
        strip.clip='off',
        legend.position='none',
        legend.box='vertical') +
  scale_fill_manual(values=var_colors, guide='none') +
  scale_color_manual(values=c('unused'='black','selected'='orangered')) +
  labs(x='', y='% explained variance', fill='', color='')
```

```{r FigS3A,fig.width=3.5 * fig_scale,fig.height=1.7 * fig_scale}
HD.var %>%  
  dplyr::left_join(columns, by='col') %>%
  dplyr::mutate(variable=factor(variable, levels=c('donor','sex','age','batch','sequencing','Residuals'))) %>%
  dplyr::filter(variable!='Residuals', category %in% c('library','isotypes','diversity','structural','context','breaks','variants')) %>%
  ggplot(aes(x=category, y=100*variance, fill=variable)) + 
  geom_boxplot(position=position_dodge(width=.8), outlier.shape=NA, lwd=.5) + 
  geom_jitter(position=position_jitterdodge(jitter.width=.3,dodge.width=.8), size=.25) +
  my_theme() +
  theme(axis.text.x=element_text(angle=45, hjust=1, vjust=1),
        strip.clip='off',
        legend.position='top',
        legend.box='horizontal') +
  scale_fill_manual(values=var_colors) +
  labs(x='', y='% explained variance', fill='')
```

confirm that downsampled diversity measures are better

```{r FigS3D_1, fig.width=1.4 * fig_scale, fig.height=1.7 * fig_scale}
tmp <- HD.var %>%  
  dplyr::left_join(columns, by='col') %>%
  dplyr::mutate(variable=factor(gsub('log10_nreads','# reads',variable),
                                levels=rev(c('donor','cohort','batch','sequencing','# reads','Residuals')))) %>%
  dplyr::filter(variable!='Residuals', category %in% c('diversity','diversity_raw'), variable!='cohort')

pvals <- tmp %>%
  dplyr::group_by(variable) %>%
  do(wc=wilcox.test(variance ~ category, data=.)) %>%
  summarise(variable, p=wc$p.value)

ggplot(tmp,aes(x=variable, y=100*variance, fill=category)) + 
  geom_boxplot(position=position_dodge(width=.8), outlier.shape=NA, lwd=.5) + 
  geom_jitter(position=position_jitterdodge(jitter.width=.3,dodge.width=.8), size=.5) +
  geom_text(data=pvals, aes(x=variable, label=stars.pval(p)), y=Inf, inherit.aes=FALSE, size=5) +
  my_theme() +
  theme(axis.text.x=element_text(angle=45, hjust=1, vjust=1),
        strip.clip='off',
        legend.position='top',
        legend.box='horizontal') +
  guides(fill=guide_legend(title.position='top')) +
  coord_cartesian(clip='off') +
  scale_fill_manual(values=c('diversity'='blue3','diversity_raw'='lightblue1'),
                    labels=c('downsampled','raw')) +
  labs(x='', y='% explained variance', fill='diversity measures')
```

confirm that using isotype-specific context features doesn't help

```{r FigS3D_2, fig.width=1.4 * fig_scale, fig.height=1.7 * fig_scale}
tmp <- HD.var %>%  
  dplyr::left_join(columns, by='col') %>%
  dplyr::mutate(variable=factor(gsub('log10_nreads','# reads',variable),
                                levels=rev(c('donor','cohort','batch','sequencing','# reads','Residuals')))) %>%
  dplyr::filter(variable!='Residuals', category %in% c('context','context_isotype'), variable!='cohort')

pvals <- tmp %>% 
  dplyr::group_by(variable) %>%
  do(wc=wilcox.test(variance ~ category, data=., exact=FALSE)) %>%
  summarise(variable, p=wc$p.value)

ggplot(tmp, aes(x=variable, y=100*variance, fill=category)) + 
  geom_boxplot(position=position_dodge(width=.8), outlier.shape=NA, lwd=.5) + 
  geom_jitter(position=position_jitterdodge(jitter.width=.3,dodge.width=.8), size=.5) +
  geom_text(data=pvals, aes(x=variable, label=stars.pval(p)), y=Inf, inherit.aes=FALSE, size=5) +
  my_theme() +
  guides(fill=guide_legend(title.position='top')) +
  theme(axis.text.x=element_text(angle=45, hjust=1, vjust=1),
        strip.clip='off',
        legend.position='top',
        legend.box='horizontal') +
  coord_cartesian(clip='off') +
  scale_fill_manual(values=c('context'='darkorchid3','context_isotype'='orchid1'),
                    labels=c('global','per isotype')) +
  labs(x='', y='% explained variance', fill='context measures')
```

test the effects of averaging over clusters or reads

```{r FigS3D_3, fig.width=1.4 * fig_scale,fig.height=1.7 * fig_scale}
HD.data.br <- read.csv('data/HD_data_by_reads.csv', header=1, row.names=1)

HD.scaled.br <- HD.data.br %>%
  dplyr::filter(cohort=='C1') %>%
  dplyr::group_by(feature) %>%
  dplyr::mutate(value=as.vector(scale(value))) %>%
  dplyr::filter(!is.na(value)) %>%
  dplyr::ungroup()

HD.var.br <- sapply(unique(HD.scaled.br$feature), 
                    function(ftr)
                      my_fitExtractVarPartModel(value ~ (1 | donor) + (1 | sex) + (1 | age) + (1 | sequencing) + (1 | batch),
                                                HD.scaled.br %>%
                                                  dplyr::filter(feature==ftr))) %>%
  t() %>%
  as.data.frame() %>%
  tibble::rownames_to_column('col') %>%
  gather(variable,variance,-col)

fts <- intersect(HD.var$col, HD.var.br$col)

tmp <- rbind(HD.var %>% 
               dplyr::mutate(averaging='clusters'), 
             HD.var.br %>%
               dplyr::mutate(averaging='reads')) %>% 
  dplyr::filter(col %in% fts) %>%
  dplyr::left_join(columns, by='col') %>%
  dplyr::mutate(variable=factor(variable, levels=rev(c('donor','sex','age','batch','sequencing','Residuals')))) %>%
  dplyr::filter(variable!='Residuals', 
                category %in% c("context","breaks"),
                variable %in% c("donor", "batch", "sequencing")) %>% 
  dplyr::arrange(col,averaging,variable)

pvals <- tmp %>%
  dplyr::group_by(variable) %>%
  do(wc=wilcox.test(variance ~ averaging, paired=TRUE, data=., exact=FALSE)) %>%
  summarise(variable, p=wc$p.value)

ggplot(tmp, aes(x=variable, y=100*variance, fill=averaging)) + 
  geom_boxplot(position=position_dodge(width=.8), outlier.shape=NA, lwd=.5) + 
  geom_jitter(position=position_jitterdodge(jitter.width=.1,dodge.width=.8), size=.5) +
  geom_text(data=pvals, aes(x=variable, label=stars.pval(p)), y=Inf, inherit.aes=FALSE, size=5) +
  my_theme() +
  coord_cartesian(clip='off') +
  guides(fill=guide_legend(title.position='top')) +
  theme(axis.text.x=element_text(angle=45, hjust=1, vjust=1),
        legend.position='top',
        legend.box='horizontal') +
  scale_fill_manual(values=c('clusters'='aquamarine4','reads'='darkseagreen1')) +
  labs(x='', y='% explained variance', fill='averages over')
```

show top examples for features related to age or sex

```{r Fig2E, fig.width=.9 * fig_scale, fig.height=1.7 * fig_scale}
age.cols <- HD.var %>%
  dplyr::left_join(columns, by='col') %>%
  spread(variable, variance) %>%
  dplyr::filter(!(category %in% c('library','diversity_raw','context_isotype'))) %>%
  dplyr::slice_max(age, n=1) %>%
  dplyr::pull(col)
age.cols <- 'top_clone_occupancy_downsampled'

sex.cols <- HD.var %>%
  dplyr::left_join(columns, by='col') %>%
  spread(variable, variance) %>%
  dplyr::filter(!(category %in% c('library','diversity_raw','context_isotype'))) %>%
  dplyr::slice_max(sex, n=1) %>%
  dplyr::pull(col)
sex.cols <- 'spread_SG1'

HD.data %>%
  dplyr::filter(feature %in% sex.cols, cohort=='C1') %>%
  dplyr::mutate(feature=factor(gsub('_downsampled|_switch','',feature),
                               levels=gsub('_downsampled|_switch','',sex.cols))) %>% 
  ggplot(aes(x=sex, y=value)) + 
  geom_boxplot(outlier.shape=NA, lwd=.5) +
  geom_jitter(width=.2, size=.1) + 
  stat_compare_means(comparisons=list(c('male','female')),
                     method='wilcox.test',label = "p.signif", hide.ns=TRUE, size=2.5) +
  facet_wrap(~feature, scales='free', ncol=8, switch='y') + 
  my_theme() + 
  coord_cartesian(clip='off') +
  theme(legend.position='none',
        axis.text.x=element_text(angle=45, hjust=1, vjust=1)) +
  labs(x='',y='')

HD.data %>%
  dplyr::filter(feature %in% age.cols, cohort=='C1') %>%
  dplyr::mutate(feature=factor(gsub('_downsampled|_switch','',feature),
                               levels=gsub('_downsampled|_switch','',age.cols))) %>% 
  ggplot(aes(x=age, y=value)) + 
  geom_boxplot(outlier.shape=NA, lwd=.5) +
  geom_jitter(width=.2, size=.1) + 
  stat_compare_means(comparisons=list(c('<50','>=50')),
                     method='wilcox.test',label = "p.signif", hide.ns=TRUE, size=2.5) +
  facet_wrap(~feature, scales='free', ncol=8, switch='y') + 
  my_theme() +
  coord_cartesian(clip='off') +
  theme(legend.position='none',
        axis.text.x=element_text(angle=45, hjust=1, vjust=1)) +
  labs(x='',y='')
```

HD feature correlation (on FB data)

```{r FigS3B, fig.width=4.25 * fig_scale, fig.height=4.25 * fig_scale}
library(dendextend)
FB.X <- HD.data %>%
  dplyr::filter(cohort=='C2') %>%
  dplyr::select(sample, feature, value) %>%
  spread(feature, value) %>%
  dplyr::mutate_if(is.numeric, scale) %>%
  tibble::column_to_rownames('sample')

R <- cor(FB.X[,selected.cols], use='na.or.complete')
io <- gsub('_downsampled|_switch','',rownames(R)[order.hclust(hclust(dist(R)))])

tmp <- as.data.frame(R) %>% 
  tibble::rownames_to_column('var2') %>%
  gather(var1, R, -var2) %>%
  dplyr::left_join(columns, by=c('var1'='col')) %>%
  dplyr::mutate(cat1=category) %>%
  dplyr::select(-category) %>%
  dplyr::left_join(columns, by=c('var2'='col')) %>%
  dplyr::mutate(cat2=category) %>%
  dplyr::select(-category)  %>%
  dplyr::mutate(var1=factor(gsub('_downsampled|_switch','',var1), levels=io),
                var2=factor(gsub('_downsampled|_switch','',var2), levels=io),
                i1=as.vector(sapply(var1, function(x) which(io==x))),
                i2=as.vector(sapply(var2, function(x) which(io==x))))

anno_grob_y <- grid::rectGrob(x=1:length(io), y=0, gp=gpar(col=NA, fill=col_cols[io], alpha=.5))
anno_grob_x <- grid::rectGrob(y=1:length(io), x=0, gp=gpar(col=NA, fill=col_cols[io], alpha=.5))

ggplot(tmp %>%
  dplyr::filter(i1 <= i2),
  aes(x=var1,y=var2,fill=R)) + 
  geom_tile() +
  scale_fill_gradient2(low='blue3',high='red3',mid='gray',limits=c(-1,1)) + 
  my_theme() +
  scale_x_discrete(position = "top") +
  coord_cartesian(clip='off') +
  theme(axis.text.x.top=element_text(angle=90, hjust=0., vjust=.5), #, colour=col_cols[io]),
        legend.position=c(-.1,1.2)) +
  labs(x='',y='') +
  annotation_custom(grob=anno_grob_y, xmin = 0, xmax = 1, ymin = length(io) + 10, ymax = length(io) + 28) +
  annotation_custom(grob=anno_grob_x, xmin = -9, xmax = 8, ymin = 0, ymax = 1)

```

distributions of features in the two HD cohorts

```{r Fig2C, fig.width=4.25 * fig_scale, fig.height=1.5 * fig_scale}
library(gtools)
cols.here.human <- c('insert_frequency',
                     'n_untemplated_switch',
                     'frac_breaks_sequential',
                     'frac_breaks_intra',
                     'cluster_gini_downsampled',
                     'frac_clusters_SG1')

cohort.data <- HD.data %>%
  dplyr::filter(sequencing=='minION') %>% 
  dplyr::filter(feature %in% cols.here.human) %>%
  dplyr::left_join(columns, by=c('feature'='col')) %>%
  dplyr::mutate(feature=factor(gsub('_downsampled|_switch','',feature), 
                               levels=gsub('_downsampled|_switch','',cols.here.human))) %>%
  dplyr::filter(category %in% c('diversity','isotypes','breaks','context','structural'))

ggplot(cohort.data, aes(x=cohort, y=value)) + 
  geom_boxplot(aes(group=cohort), position=position_dodge(width=.75), width=.7, outlier.shape=NA, lwd=.25, fill='gray80') + 
  geom_pointrange(data=cohort.data %>%
                    dplyr::group_by(donor,feature,category,cohort) %>%
                    dplyr::summarise(se=sd(value)/sqrt(n()),
                                     value=mean(value),
                                     n=n()) %>% dplyr::filter(n > 1),
                  aes(fill=cohort, ymin=value-se,ymax=value+se, group=cohort),
                  position=position_jitterdodge(jitter.width=.5, dodge.width=.75, jitter.height = 0), 
                  color='black',size=1,stroke=0.1,
                  linewidth=.1,fatten=1) + 
  my_theme() +
  coord_cartesian(clip='off') +
  facet_wrap(~feature, switch='y', scales='free_y', ncol=7) +
  theme(legend.position='none',
        axis.text.x=element_text(angle=90, hjust=1, vjust=.5)) +
  labs(x='',y='')
```

same thing for mouse tissue and Vincendeau

```{r Fig2F, fig.width=4. * fig_scale, fig.height=1.75 * fig_scale}
library(gtools)
mouse.data <- read.csv('data/mouse_data.csv',header=1, row.names=1)  %>%
  tibble::column_to_rownames('sample')
vincendeau.data <- read.csv('data/vincendeau_data.csv', header=1, row.names=1)  %>%
  tibble::column_to_rownames('sample')

cols.here.mouse <- c('frac_breaks_inversions',
                     'n_homology_switch',
                     'mean_cluster_size_downsampled')

cohort.data <- rbind(mouse.data %>%
                       dplyr::mutate(species='mouse tissue',
                                     cohort=material) %>%
                       dplyr::select(species, cohort, cols.here.mouse),
                     vincendeau.data %>%
                       dplyr::mutate(species='Vincendeau',
                                     cohort=genotype) %>%
                       dplyr::select(species, cohort, cols.here.mouse)) %>%
  dplyr::mutate(cohort=factor(cohort, levels=c('lymph node','splenocytes','WT','Xlf','Shld1','Shld1_Xlf'))) %>%
  gather(variable, value, cols.here.mouse) %>%
  dplyr::select(variable, value, species, cohort) %>%
  dplyr::left_join(columns, by=c('variable'='col')) %>%
  dplyr::mutate(variable=factor(gsub('_downsampled|_switch','',variable), 
                                levels=gsub('_downsampled|_switch','',cols.here.mouse))) %>%
  dplyr::filter(category %in% c('diversity','isotypes','breaks','context','structural'))

cohort.data %>%
  ggplot(aes(x=cohort, y=value, fill=cohort)) + 
  geom_boxplot(position=position_dodge(width=.75), width=.7, outlier.shape=NA, lwd=.25, fill='gray80') + 
  geom_point(position=position_jitterdodge(jitter.width=.2, dodge.width=.75, jitter.height = 0), size=.5) + 
  my_theme() +
  coord_cartesian(clip='off') +
  #annotate('text',x=5.5,y=-2,label='Vincendeau et al.', size=2) +
  facet_wrap(~variable, ncol=3, switch='y', scales='free_y') +
  theme(axis.text.x=element_text(angle=45, hjust=1, vjust=1),
        legend.position='none') +
  scale_x_discrete(breaks=c('lymph node','splenocytes','WT','Xlf','Shld1','Shld1_Xlf'),
                   limits=c('lymph node','splenocytes','','WT','Xlf','Shld1','Shld1_Xlf'), drop=FALSE) +
  labs(x='',y='', fill='') 
```

check length dependence of Vincendeau data

```{r FigS3E, fig.width=1.5 * fig_scale, fig.height=1.75 * fig_scale}
cols.here.mouse <- c('mean_length_SG')
cohort.data <- rbind(mouse.data %>%
                       dplyr::mutate(species='mouse tissue',
                                     cohort=material) %>%
                       dplyr::select(species, cohort, cols.here.mouse),
                     vincendeau.data %>%
                       dplyr::mutate(species='Vincendeau',
                                     cohort=genotype) %>%
                       dplyr::select(species, cohort, cols.here.mouse)) %>%
  dplyr::mutate(cohort=factor(cohort, levels=c('lymph node','splenocytes','WT','Xlf','Shld1','Shld1_Xlf'))) %>%
  gather(variable, value, cols.here.mouse) %>%
  dplyr::select(variable, value, species, cohort) %>%
  dplyr::left_join(columns, by=c('variable'='col')) %>%
  dplyr::mutate(variable=factor(gsub('_downsampled|_switch','',variable), 
                                levels=gsub('_downsampled|_switch','',cols.here.mouse))) %>%
  dplyr::filter(category %in% c('diversity','isotypes','breaks','context','structural'))

cohort.data %>%
  ggplot(aes(x=cohort, y=value, fill=cohort)) + 
  geom_boxplot(position=position_dodge(width=.75), width=.7, outlier.shape=NA, lwd=.25, fill='gray80') + 
  geom_point(position=position_jitterdodge(jitter.width=.2, dodge.width=.75, jitter.height = 0), size=.5) + 
  my_theme() +
  coord_cartesian(clip='off') +
  #annotate('text',x=5.5,y=-2,label='Vincendeau et al.', size=2) +
  facet_wrap(~variable, ncol=3, switch='y', scales='free_y') +
  theme(axis.text.x=element_text(angle=45, hjust=1, vjust=1),
        legend.position='none') +
  scale_x_discrete(breaks=c('lymph node','splenocytes','WT','Xlf','Shld1','Shld1_Xlf'),
                   limits=c('lymph node','splenocytes','','WT','Xlf','Shld1','Shld1_Xlf'), drop=FALSE) +
  labs(x='',y='', fill='')
```

compare minION against pacBio in HD cohort for selected features

```{r FigS3C, fig.width=2.75 * fig_scale, fig.height=5 * fig_scale}
tmp <- HD.scaled %>%
  dplyr::filter(feature %in% selected.cols, batch=='20230313') %>%
  dplyr::select(feature,value,sequencing,donor) %>%
  spread(sequencing,value) %>% 
  dplyr::left_join(columns,by=c('feature'='col')) %>% 
  dplyr::group_by(feature) %>%
  dplyr::mutate(R=cor(minION,pacBio)) %>%
  dplyr::mutate(feature=factor(feature, levels=columns %>% 
                                 dplyr::filter(col %in% selected.cols) %>% 
                                 dplyr::pull(col)),
                feature_name=gsub('_downsampled|_switch','',feature)) %>% #,' (R=',signif(R,2),')')) %>%
  dplyr::arrange(feature)

cols_shapes <- tmp %>% 
  dplyr::distinct(category, feature, feature_name) %>%
  dplyr::mutate(var_col=cat_cols[as.character(category)]) %>%
  dplyr::group_by(category) %>%
  dplyr::mutate(var_ind=1:n())

ggplot(tmp,aes(x=minION,y=pacBio)) +
  geom_abline(slope=1,intercept=0,color='gray',linewidth=.5) +
  geom_point(aes(color=feature,shape=feature),size=1) + 
  my_theme() +
  stat_cor(aes(label = ..r.label..), size=2.5) +
  labs(color='',shape='') +
  scale_color_manual(values=cols_shapes %>%
                       dplyr::pull(var_col,feature),
                     labels=cols_shapes %>%
                       dplyr::pull(feature_name)) +
  scale_shape_manual(values=cols_shapes %>%
                       dplyr::pull(var_ind,feature),
                     labels=cols_shapes %>%
                       dplyr::pull(feature_name)) +
  guides(color=guide_legend(ncol=2),
         shape=guide_legend(ncol=2))  +
  theme(legend.position='bottom',
        legend.key.size=unit(1.5,'mm'),
        legend.key.spacing.y=unit(0, 'mm'))
```

show a PCA of the Vincendeau data

```{r Fig2G, fig.width=2.5 * fig_scale, fig.height=1.7 * fig_scale}
selected.cols.mouse <- gsub('clusters_SG1','clusters_Sg2b',
                            gsub('clusters_SG2','clusters_Sg2c',
                                 gsub('clusters_SG3','clusters_Sg3',
                                      gsub('clusters_SA1','clusters_Sa',selected.cols))))
vincendeau.cols <- intersect(colnames(vincendeau.data),selected.cols.mouse)
vincendeau.X <- vincendeau.data[,vincendeau.cols] %>%
  dplyr::mutate_if(is.numeric, scale) 
vincendeau.X <- vincendeau.X[,colSums(is.na(vincendeau.X))==0]
vincendeau.meta <- vincendeau.data %>%
  dplyr::mutate(genotype=factor(genotype,levels=c('WT','Xlf','Shld1','Shld1_Xlf'))) %>%
  dplyr::select(genotype)

pca <- prcomp(vincendeau.X)
pcadat <- cbind(pca$x,vincendeau.meta)
pcarot <- data.frame(coef=pca$rot[,'PC1'],
                     col=gsub('_downsampled|_switch','',row.names(pca$rot)))  %>%
  dplyr::arrange(coef) %>%
  dplyr::mutate(col=factor(col, levels=col))

ggplot(pcadat,aes(x=PC1,y=PC2, color=genotype)) +
  geom_point(size=1) +
  geom_polygon(data = pcadat %>%
                 dplyr::group_by(genotype) %>%
                 dplyr::slice(chull(PC1, PC2)), alpha = 0.25, aes(fill=genotype)) +
  my_theme() +
  theme(aspect.ratio=pca$sdev[2]**2/pca$sdev[1]**2,
        legend.position=c(.15,.9),
        legend.background = element_blank()) +
  labs(x=paste0('PC 1 (',100*signif(pca$sdev[1]**2/sum(pca$sdev**2),3),'%)'),
       y=paste0('PC 2 (',100*signif(pca$sdev[2]**2/sum(pca$sdev**2),3),'%)'),
       color='',shape='',fill='')  +
  scale_color_manual(values=genotype_colors) +
  scale_fill_manual(values=genotype_colors)
```

show mean scaled values of Vincendeau in a scatter plot (big dots are significant, small ones are not)

```{r Fig2H, fig.width=7 * fig_scale, fig.height=2.25 * fig_scale}
tmp <- vincendeau.data %>%
  gather(variable, value, vincendeau.cols) %>%
  dplyr::group_by(variable) %>%
  dplyr::mutate(value=scale(value)) %>%
  dplyr::group_by(genotype, variable) %>%
  dplyr::summarise(mean=mean(value),
                   se=sd(value)/sqrt(n())) %>%
  tidyr::pivot_wider(id_cols=c('variable'),names_from='genotype',values_from=c('mean','se')) %>%
  dplyr::left_join(columns,by=c('variable'='col')) %>%
  dplyr::group_by(variable) %>%
  dplyr::mutate(variable=factor(variable, levels=columns %>% 
                                  dplyr::filter(col %in% vincendeau.cols) %>% 
                                  dplyr::pull(col)),
                var_col=cat_cols[category]) %>%
  dplyr::arrange(variable)

cols_shapes <- tmp %>% 
  dplyr::distinct(category, variable) %>%
  dplyr::mutate(var_col=cat_cols[as.character(category)],
                variable_name=gsub('_downsampled','',variable)) %>%
  dplyr::group_by(category) %>%
  dplyr::mutate(var_ind=1:n())

comparisons <- list(c('WT','Xlf'),
                    c('WT','Shld1'),
                    c('WT','Shld1_Xlf'),
                    c('Shld1','Xlf'))

plots <- lapply(comparisons, function(x) 
  tmp %>%
    dplyr::mutate_(mean1=paste0('mean_',x[[1]]),
                   mean2=paste0('mean_',x[[2]]),
                   se1=paste0('se_',x[[1]]),
                   se2=paste0('se_',x[[2]])) %>%
    dplyr::mutate(tstat=abs(mean2-mean1)/(sqrt(se2**2+se1**2)),
                  pval=dt(tstat,4)) %>%
    ggplot(aes(x=mean1,y=mean2)) +
    geom_abline(slope=1,intercept=0,color='gray',linewidth=.5) +
    geom_point(aes(color=variable,shape=variable,size=ifelse(pval < .05,'sig','ns'))) + 
    stat_cor(method = "pearson", aes(label=..r.label..), size=2.5, show.legend=FALSE) +
    geom_errorbar(aes(ymin=mean2-se2,
                      ymax=mean2+se2,
                      color=variable),
                  size=.25) +
    geom_errorbarh(aes(xmin=mean1-se1,
                       xmax=mean1+se1,
                       color=variable),
                   size=.25) +
    my_theme() +
    labs(color='',shape='') +
    scale_color_manual(values=cols_shapes %>% 
                         dplyr::pull(var_col,variable),
                       labels=cols_shapes %>%
                         dplyr::pull(variable_name)) +
    scale_shape_manual(values=cols_shapes %>%
                         dplyr::pull(var_ind,variable),
                       labels=cols_shapes %>%
                         dplyr::pull(variable_name)) +
    scale_size_manual(values=c('sig'=2,'ns'=.5),guide='none') +
    guides(color=guide_legend(ncol=5),
           shape=guide_legend(ncol=5)) +
    labs(x=x[[1]],y=x[[2]],fill='',color='',size=''))

plot_grid(plot_grid(plotlist=lapply(plots, function(x) x + theme(legend.position='none')), 
                    ncol=4, align='vh'),
          get_legend(plots[[1]] + theme(legend.margin=margin(0,0,0,0,'mm'))), ncol=1, rel_heights=c(.65,.35))
```

# Figure 3 + supplement (CH12)

show a PCA of spleen and lymph node data together with PP and CH12 using all switch regions or only Sm+Sa

```{r Fig3B, fig.width=3.5 * fig_scale, fig.height=1.75 * fig_scale}
CH12.data <- read.csv('data/CH12_data.csv',header=1, row.names=1) %>%
  tibble::column_to_rownames('sample')
CH12.data.noSg <- read.csv('data/CH12_data_noSg.csv', header=1, row.names=1) %>%
  tibble::column_to_rownames('sample')
mouse.data.noSg <- read.csv('data/mouse_data_noSg.csv', header=1, row.names=1)

mouse.cols <-  intersect(selected.cols.mouse, intersect(colnames(CH12.data),
                                                        colnames(mouse.data)))

CH12.take <- CH12.data$batch != '20240830'
mouse.take <- rep(TRUE, dim(mouse.data)[1]) 
mouse.X <- rbind(CH12.data[CH12.take,mouse.cols],
                 mouse.data[mouse.take,mouse.cols]) %>%
  dplyr::mutate_if(is.numeric, scale)
mouse.cols <- mouse.cols[colSums(is.na(mouse.X))==0]

pca <- prcomp(mouse.X[,mouse.cols])
pcadat <- cbind(pca$x,data.frame(material=c(rep('CH12',sum(CH12.take)),
                                            mouse.data$material[mouse.take]),
                                 batch=c(CH12.data$batch[CH12.take],
                                         mouse.data$batch[mouse.take]),
                                 genotype=c(CH12.data$genotype[CH12.take],
                                            rep('WT',sum(mouse.take)))))

p1 <- ggplot(pcadat,aes(x=PC1,y=PC2, color=material)) +
  geom_point(aes(shape=genotype),size=1) +
  geom_polygon(data = pcadat %>%
                 dplyr::group_by(material) %>%
                 dplyr::slice(chull(PC1, PC2)), alpha = 0.25, aes(fill=material)) +
  my_theme() +
  scale_fill_manual(values=tissue_colors) +
  scale_color_manual(values=tissue_colors) +
  labs(x=paste0('PC 1 (',100*signif(pca$sdev[1]**2/sum(pca$sdev**2),3),'%)'),
       y=paste0('PC 2 (',100*signif(pca$sdev[2]**2/sum(pca$sdev**2),3),'%)'),
       color='',shape='',fill='', title='all switch regions')

CH12.take.noSg <- rep(TRUE, dim(CH12.data.noSg)[1]) 
mouse.take.noSg <- rep(TRUE, dim(mouse.data.noSg)[1]) 
mouse.cols.noSg <-  grep('SG|Sg|sequential',mouse.cols,
                         value=TRUE,invert=TRUE)
mouse.X.noSg <- rbind(CH12.data.noSg[CH12.take.noSg,mouse.cols.noSg],
                     mouse.data.noSg[mouse.take.noSg,mouse.cols.noSg]) %>%
  dplyr::mutate_if(is.numeric, scale)

mouse.cols.noSg <- mouse.cols.noSg[colSums(is.na(mouse.X.noSg))==0]

pca.noSg <- prcomp(mouse.X.noSg[,mouse.cols.noSg])
pcadat.noSg <- cbind(pca.noSg$x,data.frame(material=c(rep('CH12',sum(CH12.take.noSg)),
                                                      mouse.data.noSg$material[mouse.take.noSg]),
                                           batch=c(CH12.data.noSg$batch[CH12.take.noSg],
                                                   mouse.data.noSg$batch[mouse.take.noSg]),
                                           genotype=c(CH12.data.noSg$genotype[CH12.take.noSg],
                                                      rep('WT',sum(mouse.take.noSg)))))

p2 <- ggplot(pcadat.noSg,aes(x=PC1,y=PC2, color=material)) +
  geom_point(aes(shape=genotype), size=1) +
  geom_polygon(data = pcadat.noSg %>%
                 dplyr::group_by(material) %>%
                 dplyr::slice(chull(PC1, PC2)), alpha = 0.25, aes(fill=material)) +
  my_theme() +
  scale_fill_manual(values=tissue_colors) +
  scale_color_manual(values=tissue_colors) +
  labs(x=paste0('PC 1 (',100*signif(pca.noSg$sdev[1]**2/sum(pca.noSg$sdev**2),3),'%)'),
       y=paste0('PC 2 (',100*signif(pca.noSg$sdev[2]**2/sum(pca.noSg$sdev**2),3),'%)'),
       color='',shape='',fill='', title='only Sm + Sa')

plot_grid(p1 + theme(legend.position='none'),
          p2 + theme(legend.position='none'),
          get_legend(p1 + theme(legend.box='vertical')),
          rel_widths=c(.4,.4,.2),
          ncol=3)

# p3 <- ggplot(pcadat,aes(x=PC3,y=PC4, color=material)) +
#   geom_point(aes(shape=genotype),size=1) +
#   geom_polygon(data = pcadat %>%
#                  dplyr::group_by(material) %>%
#                  dplyr::slice(chull(PC3, PC4)), alpha = 0.25, aes(fill=material)) +
#   scale_fill_manual(values=tissue_colors) +
#   scale_color_manual(values=tissue_colors) +
#   my_theme() +
#   labs(x=paste0('PC 3 (',100*signif(pca$sdev[3]**2/sum(pca$sdev**2),3),'%)'),
#        y=paste0('PC 4 (',100*signif(pca$sdev[4]**2/sum(pca$sdev**2),3),'%)'),
#        color='',shape='',fill='', title='all switch regions') 
# 
# p4 <- ggplot(pcadat.noSg,aes(x=PC3,y=PC4, color=material)) +
#   geom_point(aes(shape=genotype),size=1) +
#   geom_polygon(data = pcadat.noSg %>%
#                  dplyr::group_by(material) %>%
#                  dplyr::slice(chull(PC3, PC4)), alpha = 0.25, aes(fill=material)) +
#   my_theme() +
#   scale_fill_manual(values=tissue_colors) +
#   scale_color_manual(values=tissue_colors) +
#   labs(x=paste0('PC 3 (',100*signif(pca.noSg$sdev[3]**2/sum(pca.noSg$sdev**2),3),'%)'),
#        y=paste0('PC 4 (',100*signif(pca.noSg$sdev[4]**2/sum(pca.noSg$sdev**2),3),'%)'),
#        color='',shape='',fill='', title='only Sm + Sa') 
# 
# plot_grid(p3 + theme(legend.position='none'),
#           p4 + theme(legend.position='none'),
#           get_legend(p3 + theme(legend.box='vertical')),
#           rel_widths=c(.4,.4,.2),
#           ncol=3)
```

selected features in the CH12 data (2023 batches)

```{r Fig3D,fig.width=3.5 * fig_scale,fig.height=2.5 * fig_scale}
CH12.features <- c('frac_breaks_single',
                   'mean_length',
                   'frac_breaks_duplications',
                   'frac_breaks_inversions',
                   'insert_frequency',
                   'mean_cluster_size_downsampled')

CH12.genotypes <- c('WT','Trp53bp1 KO','Rif1 KO','Lig4 KO','Brca1 KO')

CH12.meta <- CH12.data %>%
  dplyr::mutate(batch=as.character(batch),
                genotype=factor(genotype,levels=CH12.genotypes)) %>%
  dplyr::select(batch,genotype)

tmp <- CH12.data %>% 
  dplyr::filter(batch != '20240830') %>%
  dplyr::select(genotype,CH12.features) %>%
  dplyr::mutate(genotype=factor(genotype,levels=CH12.genotypes)) %>%
  dplyr::group_by(genotype) %>%
  dplyr::mutate(ng=n(),
                genotype_new=paste0(genotype,' (n=',ng,')')) %>%
  gather(col,value,-c(genotype,ng,genotype_new)) %>% 
  dplyr::mutate(col=factor(gsub("_downsampled|_switch",'',col),
                           levels=gsub("_downsampled|_switch","",CH12.features)))

ggplot(tmp, aes(x=genotype,y=value,fill=genotype)) + 
  geom_boxplot(outlier.shape=NA, lwd=.5) + 
  geom_jitter(width=.1,height=0,size=.5) + 
  stat_compare_means(ref.group='WT',method='wilcox.test',label = "p.signif", hide.ns=TRUE, size=2) +
  facet_wrap(~col,ncol=3,scales='free_y') +
  my_theme() +
  coord_cartesian(clip='off') +
  theme(legend.position='none',
        strip.clip='off',
        axis.text.x=element_text(angle=90,hjust=1,vjust=.5)) +
  scale_fill_manual(values=genotype_colors) +
  scale_x_discrete(breaks=tmp$genotype,labels=tmp$genotype_new) +
  labs(x='',y='', color='# reads')
```

a PCA with all selected features

```{r Fig3E, fig.width=2. * fig_scale, fig.height=1.7 * fig_scale}
CH12.samples <- row.names(CH12.data)[CH12.data$batch != '20240830']
CH12.cols <- setdiff(grep('SG|Sg',intersect(selected.cols.mouse,colnames(CH12.data)),value=TRUE,invert=TRUE),
                     c('frac_breaks_sequential','frac_clusters_Sm'))
CH12.X <- CH12.data[CH12.samples,CH12.cols] %>%
  dplyr::mutate_if(is.numeric, scale)
CH12.cols <- CH12.cols[colSums(is.na(CH12.X))==0]

pca <- prcomp(CH12.X[CH12.samples,CH12.cols])
pcadat <- cbind(pca$x,CH12.meta[CH12.samples,])

ggplot(pcadat,aes(x=PC1,y=PC2, color=genotype)) +
  geom_point(size=2) +
  geom_polygon(data = pcadat %>%
                 dplyr::group_by(genotype) %>%
                 dplyr::slice(chull(PC1, PC2)), alpha = 0.25, aes(fill=genotype)) +
  my_theme() +
  theme(aspect.ratio=pca$sdev[2]**2/pca$sdev[1]**2,
        legend.position='none') +
  labs(x=paste0('PC 1 (',100*signif(pca$sdev[1]**2/sum(pca$sdev**2),3),'%)'),
       y=paste0('PC 2 (',100*signif(pca$sdev[2]**2/sum(pca$sdev**2),3),'%)'),
       color='',shape='',fill='') +
  scale_fill_manual(values=genotype_colors) +
  scale_color_manual(values=genotype_colors)

# ggplot(pcadat,aes(x=PC3,y=PC4, color=genotype)) +
#   geom_point(size=2) +
#   geom_polygon(data = pcadat %>%
#                  dplyr::group_by(genotype) %>%
#                  dplyr::slice(chull(PC3, PC4)), alpha = 0.25, aes(fill=genotype)) +
#   my_theme() +
#   theme(aspect.ratio=pca$sdev[4]**2/pca$sdev[3]**2) +
#   labs(x=paste0('PC 3 (',100*signif(pca$sdev[3]**2/sum(pca$sdev**2),3),'%)'),
#        y=paste0('PC 4 (',100*signif(pca$sdev[4]**2/sum(pca$sdev**2),3),'%)'),
#        color='',shape='',fill='') +
#   scale_fill_manual(values=genotype_colors) +
#   scale_color_manual(values=genotype_colors)
# 
# ggplot(pcadat,aes(x=PC1,y=PC2, color=batch)) +
#   geom_point(size=2) +
#   my_theme() +
#   theme(aspect.ratio=pca$sdev[2]**2/pca$sdev[1]**2) +
#   labs(x=paste0('PC 1 (',100*signif(pca$sdev[1]**2/sum(pca$sdev**2),3),'%)'),
#        y=paste0('PC 2 (',100*signif(pca$sdev[2]**2/sum(pca$sdev**2),3),'%)'),
#        color='',shape='')

```

heatmap of all selected features

```{r Fig3C, fig.width=3.5 * fig_scale, fig.height=3.75 * fig_scale}
breaks <- seq(-2,2,.2)
samples <- CH12.meta %>%
  tibble::rownames_to_column('sample') %>%
  dplyr::filter(batch != '20240830') %>%
  dplyr::arrange(genotype) %>%
  dplyr::pull(sample)
colnames(CH12.X) <- gsub('_switch|_downsampled','',colnames(CH12.X))

dend <- hclust(dist(CH12.X[samples,]))
dend <- reorder(as.dendrogram(dend),wts=order(match(CH12.meta %>%
                                                      tibble::rownames_to_column('sample') %>%
                                                      dplyr::mutate(genotype=factor(genotype, levels=unique(genotype))) %>%
                                                      dplyr::arrange(genotype) %>%
                                                      dplyr::pull(sample),
                                                    samples)),agglo.FUN=mean)

hm1 <- Heatmap(t(CH12.X[samples,gsub('_switch|_downsampled','',CH12.cols)]),
               col=colorRamp2(c(-2,0, 2), c("blue",'gray90', "red")),
               cluster_columns=dend,
               cluster_rows=TRUE,
               show_row_names=TRUE,
               show_column_names=FALSE,
               show_row_dend=TRUE,
               column_dend_height = unit(5, "mm"),
               top_annotation=HeatmapAnnotation(df=CH12.meta[samples,c('genotype','batch')],
                                                show_annotation_name=TRUE,
                                                show_legend=c(TRUE,FALSE),
                                                simple_anno_size=unit(2,'mm'),
                                                annotation_name_gp=gpar(fontsize=8),
                                                annotation_legend_param=list(labels_gp=gpar(fontsize=7),
                                                                             title_gp=gpar(fontsize=8),
                                                                             legend_width=unit(2,'mm'),
                                                                             legend_height=unit(2,'mm'),
                                                                             legend_direction="horizontal",
                                                                             nrow=1),
                                                col=list('genotype'=genotype_colors[levels(CH12.meta$genotype)],
                                                         'batch'=batch_colors[unique(CH12.meta[samples,'batch'])])),
               column_names_gp=gpar(fontsize=7),
               row_names_gp =gpar(fontsize = 7),
               row_split = columns %>% tibble::column_to_rownames('col') %>% .[CH12.cols, 'category'], 
               heatmap_legend_param=list(labels_gp=gpar(fontsize=7),
                                         title_gp=gpar(fontsize=8),
                                         legend_direction = "horizontal", 
                                         legend_width=unit(20,'mm'),
                                         legend_height=unit(2,'mm')),
               row_title=NULL,
               name='z-score')

draw(hm1, heatmap_legend_side = "bottom", annotation_legend_side = "bottom")#, legend_grouping="original")

```

do ridge regression

```{r Fig3G, fig.width=1.25 * fig_scale, fig.height=3.5 * fig_scale}
train_batches <- c('20220425','20221118','20221212','20230120','20230213')
test_batches <- '20240830'
train <- CH12.meta$batch %in% train_batches
test <- CH12.meta$batch %in% test_batches

CH12.X <- CH12.data[, CH12.cols] %>%
  # dplyr::mutate_if(is.numeric, function(x) (x-mean(x[train]))/sd(x[train])) 
  dplyr::mutate_if(is.numeric, function(x) { 
    s <- numeric(sum(train | test)); 
    s[train] <- (x[train]-mean(x[train]))/sd(x[train]);
    s[test] <- (x[test]-mean(x[test]))/sd(x[test]);
    s
  }) 
#CH12.X.test <- CH12.data[test, CH12.cols] %>%
#  dplyr::mutate_if(is.numeric, scale) #function(x) (x-mean(x[train]))/sd(x[train]))

set.seed(3)
cv_model <- cv.glmnet(as.matrix(CH12.X[train,]), CH12.meta$genotype[train], alpha = 0, family='multinomial')

mod <- glmnet(as.matrix(CH12.X[train,]), CH12.meta$genotype[train], alpha = 0, family = "multinomial",
              lambda = cv_model$lambda.min)

tmp <- as.data.frame(predict(mod, type = "response", 
                              newx=as.matrix(CH12.X[test,]))) %>%
  tibble::rownames_to_column('sample') %>% 
  gather(gt,p,-sample) %>% 
  dplyr::mutate(prediction=factor(gsub('.s0','',gt),levels=CH12.genotypes),
                p=as.numeric(p)) %>%
  dplyr::left_join(CH12.meta[test,] %>% 
                     tibble::rownames_to_column('sample') %>% 
                     dplyr::select(sample,genotype),by='sample') %>%
  dplyr::group_by(sample) %>%
  dplyr::mutate(c=ifelse(p == max(p),ifelse(genotype==prediction, 'green','red'),'black'),
                genotype=factor(genotype,levels=CH12.genotypes),
                sample=gsub('[0-9]*_(BC[0-9]*)_switch','\\1',sample)) %>%
  dplyr::ungroup() %>%
  dplyr::arrange(genotype) %>%
  dplyr::mutate(sample=factor(gsub('20240830_','',sample),
                              levels=gsub('20240830_','',unique(sample)))) 
ggplot(tmp, aes(x=prediction,y=sample,
                label=round(p,2),fill=p,color=c)) + 
  geom_tile() + 
  geom_text(data=. %>% dplyr::filter(p > .25),
            size=2) +
  scale_fill_gradient(low='white',high='steelblue3') +
  scale_color_manual(values=c(black='black',green='darkgreen',red='red3')) +
  my_theme() +
  scale_y_discrete(labels=ifelse(grepl('\\.20|\\.27|\\.14|\\.01|\\.04',tmp$sample),"*","")) +
  theme(legend.position='none',
        axis.ticks.y = element_blank(),
        #axis.text.y = element_blank(),
        axis.text.x=element_text(angle=90, hjust=1, vjust=.5)) +
  labs(x='',y='36 new samples (* 25ng IL4)')
```

show the ridge coefficients

```{r Fig3F, fig.width=2. * fig_scale, fig.height=2.5 * fig_scale}
tmp <- do.call(cbind,lapply(coef(mod), function(x) x[,'s0'])) %>%
  as.data.frame() %>%
  dplyr::filter(rowSums(abs(.)) > 0) %>%
  dplyr::slice_max(rowSums(abs(.)),n=20) %>%
  dplyr::arrange(rowSums(abs(.))) %>%
  tibble::rownames_to_column('col') %>%
  dplyr::filter(col %in% columns$col)

col.order <- gsub('_downsampled|_switch','',tmp$col[order.hclust(hclust(dist(tmp[,2:6])))])

tmp <- tmp %>%
  dplyr::mutate(col=gsub('_downsampled|_switch','',col),
                col=factor(col, levels=col.order))

anno_grob_y <- grid::rectGrob(y=1:length(tmp$col), x=0, gp=gpar(col=NA, fill=col_cols[rev(levels(tmp$col))], alpha=.5))

tmp %>%
  gather(genotype,coef,-col) %>%
  dplyr::mutate(genotype=factor(genotype, levels=CH12.genotypes),
                col=factor(col, levels=rev(levels(tmp$col)))) %>%
  ggplot(aes(x=genotype, y=col, fill=coef, label=round(coef,2))) + 
  geom_tile() +
  scale_fill_gradient2(low='pink2',high='steelblue3', mid='white', breaks=c(-.3, 0, .3)) +
  my_theme() +
  coord_cartesian(clip='off') +
  theme(axis.text.x=element_text(angle=45, hjust=1, vjust=1),
        legend.position=c(-1,-.15),
        legend.text=element_text(angle=90, hjust=.5, vjust=1),
        legend.direction='horizontal') +
  annotation_custom(grob=anno_grob_y, ymin = 0, ymax = 1, xmin = -4, xmax = 4.5) +
  labs(x='',y='') 
```

evaluate ridge regression by cross-validation

```{r FigS3G, fig.width=1. * fig_scale, fig.height=1.75 * fig_scale}
batch.CV <- c()
for (batch in unique(CH12.meta$batch)) {
  train <- CH12.meta$batch != batch
  test <- CH12.meta$batch == batch
  X <- CH12.data[, CH12.cols] %>%
    dplyr::mutate_if(is.numeric, function(x) { 
      s <- numeric(sum(train | test)); 
      s[train] <- (x[train]-mean(x[train]))/sd(x[train]);
      s[test] <- (x[test]-mean(x[test]))/sd(x[test]);
      s
      }) %>%
    as.matrix()

  cv_model <- cv.glmnet(X[train,], CH12.meta$genotype[train], alpha = 0, family='multinomial')
  mod <- glmnet(X[train,], CH12.meta$genotype[train], alpha = 0, family = "multinomial",
                lambda = cv_model$lambda.min)
  batch.CV[[batch]] <- mean(predict(mod, type = "class", newx=X[test,])[,'s0'] == CH12.meta[test,'genotype'])
}

set.seed(4)
sample.CV <- c()
for (k in seq(1,10)) {
  train <- runif(dim(CH12.meta)[1]) < .75
  test <- !train
  X <- CH12.data[, CH12.cols] %>%
    dplyr::mutate_if(is.numeric, function(x) { 
      s <- numeric(sum(train | test)); 
      s[train] <- (x[train]-mean(x[train]))/sd(x[train]);
      s[test] <- (x[test]-mean(x[test]))/sd(x[test]);
      s
      }) %>%
    as.matrix()

  cv_model <- cv.glmnet(X[train,], CH12.meta$genotype[train], alpha = 0, family='multinomial')
  mod <- glmnet(X[train,], CH12.meta$genotype[train], alpha = 0, family = "multinomial",
                lambda = cv_model$lambda.min)
  sample.CV[[as.character(k)]] <- mean(predict(mod, type = "class", newx=X[test,])[,'s0'] == CH12.meta[test,'genotype'])
}

data.frame(acc=c(unlist(batch.CV),unlist(sample.CV)),
           fold=c(names(batch.CV),names(sample.CV))) %>%
  dplyr::filter(!is.na(acc)) %>%
  dplyr::mutate(method=ifelse(fold %in% CH12.meta$batch,'batch','sample')) %>%
  dplyr::group_by(method) %>%
  dplyr::mutate(mean_acc=mean(acc)) %>%
  ggplot(aes(x=method,y=acc)) +
  geom_boxplot(outlier.shape=NA, lwd=.5) + 
  geom_jitter(width=.1, height=0) + 
  scale_y_continuous(limits=c(0,1)) +
  my_theme() +
  theme(axis.text.x=element_text(angle=90, hjust=1, vjust=.5)) +
  labs(x='',y='accuracy')
```

# Figure 4 + supplement (FB data)

explained variance for FB cohort (make FB_13/21 control)

```{r Fig4B,fig.width=1 * fig_scale, fig.height=1.7 * fig_scale}
FB.data <- read.csv('data/FB_data.csv', header=1, row.names=1) %>%
  dplyr::mutate(batch=as.character(batch),
                diagnosis = ifelse(donor %in% c('FB_13','FB_21'), 'control', diagnosis)) %>%
  tibble::column_to_rownames('sample')
FB.X <- FB.data[grepl('2023',row.names(FB.data)),intersect(colnames(FB.data),columns$col)] %>%
  dplyr::mutate_if(is.numeric, scale)
FB.meta <- FB.data %>%
  dplyr::filter(grepl('2023',batch)) %>%
  dplyr::select(donor,batch,sex,age,diagnosis,timepoint)

FB.var <- fitExtractVarPartModel(t(FB.X[,colSums(is.na(FB.X))==0]), 
                                 ~ (1 | donor) + (1 | diagnosis) + (1 | sex) + age + (1 | batch), FB.meta) %>%
  as.data.frame() %>%
  tibble::rownames_to_column('col') %>%
  gather(variable,variance,-col)

FB.var %>%  dplyr::left_join(columns, by='col') %>%
  dplyr::mutate(variable=factor(variable,levels=c('donor','diagnosis','sex','age','batch','Residuals'))) %>%
  dplyr::filter(variable!='Residuals', category %in% c('library','isotypes','diversity','structural','context','breaks','variants')) %>%
  ggplot(aes(x=variable, y=100*variance)) + 
  geom_boxplot(aes(fill=variable), outlier.shape=NA, lwd=.5, legend=FALSE) + 
  geom_jitter(position=position_jitter(width=.25), size=.25) +
  my_theme() +
  theme(axis.text.x=element_text(angle=45, hjust=1, vjust=1),
        strip.clip='off',
        legend.position='none',
        legend.box='vertical') +
  scale_fill_manual(values=var_colors, guide='none') +
  labs(x='', y='% explained variance', fill='', color='')
```

```{r FigS4A,fig.width=3.5 * fig_scale,fig.height=1.7 * fig_scale}
FB.var %>%  dplyr::left_join(columns, by='col') %>%
  dplyr::mutate(variable=factor(variable, levels=c('donor','diagnosis','sex','age','batch','Residuals'))) %>%
  dplyr::filter(variable!='Residuals', category %in% c('library','isotypes','diversity','structural','context','breaks','variants')) %>%
  ggplot(aes(x=category, y=100*variance, fill=variable)) + 
  geom_boxplot(position=position_dodge(width=.8), outlier.shape=NA, lwd=.5) + 
  geom_jitter(position=position_jitterdodge(jitter.width=.3,dodge.width=.8), size=.5) +
  my_theme() +
  theme(axis.text.x=element_text(angle=45, hjust=1, vjust=1),
        legend.position='top',
        legend.box='horizontal') +
  guides(fill=guide_legend(override.aes = list(shape=22))) +
  scale_fill_manual(values=var_colors) +
  labs(x='', y='% explained variance', fill='')
```

a PCA with all selected features on pooled FB data from 2023

```{r Fig4D, fig.width=2.25 * fig_scale, fig.height=2 * fig_scale} 
FB.data.pooled <- read.csv('data/FB_data_pooled.csv', header=1, row.names=1) %>%
  dplyr::mutate(num_variants_somatic = num_variants * (1 - frac_variants_germline),
                diagnosis = ifelse(donor %in% c('FB_13','FB_21'), 'control', diagnosis),
                status=ifelse(diagnosis %in% c('control','subclass'), 'control', 'CVID')) %>%
  tibble::column_to_rownames('sample')
QP.data <- read.csv('data/QP_data.csv', header=1, row.names=1) %>%
  tibble::column_to_rownames('sample')

FB_train_donors <- paste0('FB_',sprintf('%0.2d', 1:35))
human.meta <- rbind(FB.data.pooled %>%
                      dplyr::mutate(dataset=ifelse(grepl('FB',donor),
                                                   ifelse(sample_donor %in% FB_train_donors,'training','testing'),
                                                   'validation')) %>%
                      dplyr::select(donor,diagnosis,status,timepoint, dataset,sex,age,complication, EUROclass, pct_B, pct_IgA, pct_IgG),
                    QP.data %>%
                      dplyr::mutate(dataset='validation',
                                    timepoint='t1',
                                    complication=NA,
                                    EUROclass=NA,
                                    pct_B=NA,
                                    pct_IgA=NA,
                                    pct_IgG=NA,
                                    status=ifelse(diagnosis=='control','control','CVID')) %>%
                      dplyr::select(donor,diagnosis,status,timepoint,dataset,sex,age,complication, EUROclass, pct_B, pct_IgA, pct_IgG)) %>%
  dplyr::mutate(diagnosis=factor(diagnosis,levels=c('control','subclass','CVID','ICOS','Kabuki','IgG4def',
                                                    'AID','TNFRSF13B',
                                                    'NFKB2','PI3K','ATM','BRCA1','LIG4','NIPBL')),
                status=factor(status, levels=c('control','CVID')))

human.X <- cbind(rbind(FB.data.pooled[,selected.cols],
                       QP.data[,selected.cols]),
                 human.meta[,'dataset',drop=FALSE]) %>%
  tibble::rownames_to_column('sample') %>%
  dplyr::mutate(sample=gsub('[0-9]*_|pooled_','',sample)) %>%
  dplyr::group_by(dataset) %>%
  dplyr::mutate_if(is.numeric,scale) %>%
  dplyr::ungroup() %>%
  dplyr::select(c('sample',selected.cols)) %>%
  tibble::column_to_rownames('sample')

train <- human.meta$dataset=='training'
test <- human.meta$dataset=='testing'
validation <- human.meta$dataset=='validation'
pca <- prcomp(human.X[train,])
pcadat <- cbind(pca$x,human.meta[train,])

ggplot(pcadat,aes(x=PC1,y=PC2)) +
  geom_point(aes(color=diagnosis),size=1) +
  geom_line(aes(group=donor, color=diagnosis)) +
  geom_polygon(data = pcadat %>%
                 dplyr::group_by(status) %>%
                 dplyr::slice(chull(PC1, PC2)), alpha = 0.1, aes(fill=status), legend=FALSE) +
  my_theme() +
  #geom_label_repel(aes(label=gsub('FB_','',donor)),size=2, segment.size=.25, label.padding = unit(.5,'mm'), force=20, show.legend=FALSE) +
  scale_fill_manual(values=diagnosis_colors, guide="none") +
  scale_color_manual(values=diagnosis_colors) +
  guides(color=guide_legend(override.aes = list(shape=15,size=2), ncol=3)) +
  coord_cartesian(xlim=c(-8,10),
                  ylim=c(-5,8),
                  clip='off') +
  theme(aspect.ratio=pca$sdev[2]**2/pca$sdev[1]**2,
        legend.position='bottom',
        legend.box='horizontal') +
  labs(x=paste0('PC 1 (',100*signif(pca$sdev[1]**2/sum(pca$sdev**2),3),'%)'),
       y=paste0('PC 2 (',100*signif(pca$sdev[2]**2/sum(pca$sdev**2),3),'%)'),
       color='',shape='',fill='')
```

another PCA with only subclass features

```{r FigS4X1, fig.width=2.25 * fig_scale, fig.height=2 * fig_scale}
FB.cols.subclass <- grep('frac_clusters_S[MGA]+',colnames(FB.data.pooled),value=TRUE)
FB.X.subclass <- FB.data.pooled[FB.data.pooled$sample_donor %in% FB_train_donors,FB.cols.subclass] %>%
  dplyr::mutate_if(is.numeric,scale)

pca.subclass <- prcomp(FB.X.subclass)
pcadat.subclass <- cbind(pca.subclass$x,human.meta[human.meta$dataset=='training',])

ggplot(pcadat.subclass,aes(x=PC1,y=PC2)) +
  geom_point(aes(color=diagnosis),size=1) +
  geom_line(aes(group=donor, color=diagnosis)) +
  geom_polygon(data = pcadat.subclass %>%
                 dplyr::mutate(status=ifelse(diagnosis=='subclass', 'subclass', as.character(status))) %>%
                 dplyr::group_by(status) %>%
                 dplyr::slice(chull(PC1, PC2)), alpha = 0.25, aes(fill=status)) +
  my_theme() +
  #geom_label_repel(data = pcadat.subclass %>% 
  #                   dplyr::filter(diagnosis!='control'),
  #                 aes(label=gsub('FB_','',donor)),size=2, segment.size=.25, label.padding = unit(.5,'mm'), force=20, show.legend=FALSE) +
  scale_fill_manual(values=diagnosis_colors, guide='none') +
  scale_color_manual(values=diagnosis_colors) +
  guides(color=guide_legend(override.aes = list(shape=15,size=2), ncol=3)) +
  theme(aspect.ratio=pca$sdev[2]**2/pca$sdev[1]**2,
        legend.position='bottom',
        legend.box='horizontal') +
  labs(x=paste0('PC 1 (',100*signif(pca$sdev[1]**2/sum(pca$sdev**2),3),'%)'),
       y=paste0('PC 2 (',100*signif(pca$sdev[2]**2/sum(pca$sdev**2),3),'%)'),
       color='',shape='',fill='')
```

show subclass features for HD and subclass-deficient people

```{r FigS4G, fig.width=2 * fig_scale, fig.height=1.7 * fig_scale}
subclass.cols <- c("frac_clusters_SA1","frac_clusters_SA2",
                   "frac_clusters_SG1","frac_clusters_SG2","frac_clusters_SG3","frac_clusters_SG4")

tmp <- FB.data.pooled %>%
  dplyr::filter(sample_donor %in% FB_train_donors, diagnosis %in% c('control','subclass')) %>%
  gather(feature, frac_clusters, subclass.cols) %>%
  dplyr::mutate(feature=factor(gsub('frac_clusters_','',feature),
                               levels=gsub('frac_clusters_','',subclass.cols)))
ggplot(tmp, aes(x=feature, y=frac_clusters, color=diagnosis)) +
  geom_boxplot(position=position_dodge(width=.8),width=.7,outlier.shape=NA) +
  geom_point(position=position_dodge(width=.8), size=.5) +
  geom_line(data=tmp %>% dplyr::filter(diagnosis=='control'),
            aes(group=donor), position=position_nudge(x=-.2), alpha=.25, size=.25) +
  geom_line(data=tmp %>% dplyr::filter(diagnosis=='subclass'),
            aes(group=donor), position=position_nudge(x=+.2), alpha=.25, size=.25) +
  scale_color_manual(values=diagnosis_colors) +
  my_theme() +
  theme(legend.position=c(.8,.8)) +
  labs(color='',x='', y='fraction clusters')
```

correlate FACS against SWIBRID

```{r FigS4X2, fig.width=2.5 * fig_scale, fig.height=2 * fig_scale}
FB.data.pooled %>%
  dplyr::mutate(alpha_ratio_FACS=pct_IgA/(pct_IgA + pct_IgG)) %>%
  ggplot(aes(x=alpha_ratio_FACS, y=alpha_ratio_clusters, color=status)) + 
  geom_abline(slope=1, intercept=0, size=.5) +
  stat_cor(method = "pearson", aes(label=..r.label..), size=2.5, show.legend=FALSE, label.x=.75, label.y=c(.4, .33)) +
  geom_point() + 
  my_theme() +
  scale_color_manual(values=diagnosis_colors) +
  labs(x='FACS', y='SWIBRID', title='IgA / (IgA + IgG)')
```

box plots for selected features

```{r Fig4C, fig.width=4 * fig_scale, fig.height=2.7 * fig_scale}
FB.features <- c('frac_clusters_SG2',
                 'frac_clusters_SG3',
                 'mean_length',
                 'frac_breaks_single',
                 'n_homology_switch',
                 'frac_breaks_inversions',
                 'mean_cluster_size_downsampled',
                 'num_variants_somatic')

tmp <- FB.data.pooled[FB.data.pooled$sample_donor %in% FB_train_donors,] %>% 
  dplyr::select(donor,diagnosis,FB.features) %>%
  dplyr::mutate(diagnosis_simple=factor(ifelse(diagnosis %in% c('CVID','control','subclass'), diagnosis, 'other'), 
                                        levels=c('control','subclass','CVID','other'))) %>%
  dplyr::group_by(diagnosis_simple) %>%
  dplyr::mutate(ng=n_distinct(donor),
                diagnosis_new=paste0(diagnosis_simple,' (n=',ng,')')) %>%
  gather(col,value,-c(diagnosis_simple,ng,diagnosis_new,donor,diagnosis)) %>% 
  dplyr::mutate(col=factor(gsub("_downsampled|_switch",'',col),
                           levels=gsub("_downsampled|_switch","",FB.features)))

ggplot(tmp, aes(x=diagnosis_simple,y=value,fill=diagnosis_simple)) + 
  geom_boxplot(outlier.shape=NA, lwd=.5) + 
  geom_jitter(width=.1,height=0,size=.5) + 
  stat_compare_means(ref.group='control',method='wilcox.test',label = "p.signif", hide.ns=TRUE, size=2) +
  facet_wrap(~col,ncol=4,scales='free_y') +
  my_theme() +
  coord_cartesian(clip='off') +
  scale_fill_manual(values=diagnosis_colors) +
  theme(legend.position='none',
        axis.text.x=element_text(angle=90,hjust=1,vjust=.5)) +
  scale_x_discrete(breaks=tmp$diagnosis_simple,labels=tmp$diagnosis_new) +
  labs(x='',y='', color='# reads')
```

project new data into the old PCA

```{r Fig4E, fig.width=2.25 * fig_scale, fig.height=2 * fig_scale}
pcadat.combined <- rbind(cbind(as.matrix(human.X[train,]) %*% pca$rotation, human.meta[train,]),
                         cbind(as.matrix(human.X[test,]) %*% pca$rotation, human.meta[test,]),
                         cbind(as.matrix(human.X[validation,]) %*% pca$rotation, human.meta[validation,])) %>%
  dplyr::mutate(dataset=factor(dataset, levels=c('training','testing','validation')))

ggplot(pcadat.combined %>% dplyr::filter(dataset!='validation'),
       aes(x=PC1,y=PC2)) +
  geom_polygon(data = pcadat.combined %>% dplyr::filter(dataset=='training') %>%
                 dplyr::group_by(status) %>%
                 dplyr::slice(chull(PC1, PC2)), alpha = 0.1, color=NA, aes(fill=status), show.legend=FALSE) +
  geom_point(aes(color=diagnosis,alpha=dataset, shape=dataset), size=1) +
  geom_line(aes(group=donor, color=diagnosis,alpha=dataset, shape=dataset)) +
  #geom_label_repel(data=pcadat.combined %>% dplyr::filter(dataset=='testing') %>% dplyr::group_by(donor) %>% dplyr::slice_head(),
  #                 aes(label=gsub("FB_","",donor)),size=2, segment.size=.25, label.padding = unit(.5,'mm'), show.legend=FALSE) +
  my_theme() +
  scale_fill_manual(values=diagnosis_colors, guide='none') +
  scale_color_manual(values=diagnosis_colors) +
  scale_alpha_manual(values=c('testing'=1,'training'=.25)) +
  scale_shape_manual(values=c('testing'=15,'training'=16)) +
  guides(color=guide_legend(override.aes = list(shape=15,size=2),ncol=3)) +
  theme(legend.key.height=unit(2,'mm'),
        aspect.ratio=pca$sdev[2]**2/pca$sdev[1]**2,
        legend.position='bottom',
        legend.box='horizontal') +
  coord_cartesian(xlim=c(-8,10),
                  ylim=c(-5,8),
                  clip='off') +
  labs(x=paste0('PC 1 (',100*signif(pca$sdev[1]**2/sum(pca$sdev**2),3),'%)'),
       y=paste0('PC 2 (',100*signif(pca$sdev[2]**2/sum(pca$sdev**2),3),'%)'),
       color='',shape='',fill='',alpha='')
```

```{r Fig4G, fig.width=2.25 * fig_scale, fig.height=2 * fig_scale}
ggplot(pcadat.combined %>% dplyr::filter(dataset!='testing'),
       aes(x=PC1,y=PC2)) +
  geom_polygon(data = pcadat.combined %>% dplyr::filter(dataset=='training') %>%
                 dplyr::group_by(status) %>%
                 dplyr::slice(chull(PC1, PC2)), alpha = 0.1, color=NA, aes(fill=status), show.legend=FALSE) +
  geom_point(aes(color=diagnosis,alpha=dataset, shape=dataset), size=1) +
  geom_line(aes(group=donor, color=diagnosis,alpha=dataset, shape=dataset)) +
  # geom_label_repel(data=pcadat.combined %>% dplyr::filter(dataset=='validation') %>% dplyr::group_by(donor) %>% dplyr::slice_head(),
  #                 aes(label=donor),size=2, segment.size=.25, label.padding = unit(.5,'mm'), show.legend=FALSE) +
  my_theme() +
  scale_fill_manual(values=diagnosis_colors, guide='none') +
  scale_color_manual(values=diagnosis_colors) +
  scale_alpha_manual(values=c('training'=.25,'validation'=1)) +
  scale_shape_manual(values=c('training'=16,'validation'=15)) +
  guides(color=guide_legend(override.aes = list(shape=15,size=2), ncol=3)) +
  theme(legend.key.height=unit(2,'mm'),
        aspect.ratio=pca$sdev[2]**2/pca$sdev[1]**2,
        legend.position='bottom',
        legend.box='horizontal') +
  coord_cartesian(xlim=c(-8,10),
                  ylim=c(-5,8),
                  clip='off') +
  labs(x=paste0('PC 1 (',100*signif(pca$sdev[1]**2/sum(pca$sdev**2),3),'%)'),
       y=paste0('PC 2 (',100*signif(pca$sdev[2]**2/sum(pca$sdev**2),3),'%)'),
       color='',shape='',fill='',alpha='')
```

do ridge regression from training to testing or validation data (take out subclass def.)

```{r Fig4H, fig.width=2.5 * fig_scale, fig.height=1.5 * fig_scale}
set.seed(1)
train.here <- train & human.meta[train,'diagnosis'] != 'subclass'
cv_model <- cv.glmnet(as.matrix(human.X[train.here,]), human.meta[train.here,'status']!='control', alpha = 0, family='binomial')
mod <- glmnet(as.matrix(human.X[train.here,]), human.meta[train.here,'status']!='control', alpha = 0, family = "binomial",
                lambda = cv_model$lambda.min)
pred_1 <- predict(mod, type = "response", newx=as.matrix(human.X[test,]))
ROC_1 <- roc(human.meta[test,'status']!='control', pred_1)

pred_2 <- predict(mod, type = "response", newx=as.matrix(human.X[validation,]))
ROC_2 <- roc(human.meta[validation,'status']!='control', pred_2)

FB_res <- data.frame(specificity=c(ROC_1$specificities,ROC_2$specificities),
                     sensitivity=c(ROC_1$sensitivities,ROC_2$sensitivities),
                     thresholds=c(ROC_1$thresholds,ROC_2$thresholds),
                     model=c(rep(paste0('training -> testing; AUC=',signif(auc(ROC_1),3)),length(ROC_1$specificities)),
                             rep(paste0('training -> validation; AUC=',signif(auc(ROC_2),3)),length(ROC_2$specificities)))) %>%
  dplyr::mutate(model=factor(model, levels=unique(model))) %>%
  dplyr::arrange(model,sensitivity)

ggplot(FB_res, aes(x=specificity,y=sensitivity,color=model)) + 
  geom_step() +
  scale_x_reverse() +  
  geom_abline(slope=1,intercept=1,linewidth=.25,linetype='dashed') +
  my_theme() +
  guides(color=guide_legend(ncol=1)) +
  scale_color_manual(values=c('black','blue3')) +
  theme(legend.position=c(.5,.2)) +
  labs(color='',linetype='')
```

evaluate ridge regression by cross-validation

```{r FigS4X3, fig.width=.8 * fig_scale, fig.height=1 * fig_scale}
set.seed(4)
sample.CV <- c()
FB.samples <- row.names(FB.data.pooled)[FB.data.pooled$diagnosis != 'subclass']
for (k in seq(1,20)) {
  tr <- runif(length(FB.samples)) < .75
  te <- !tr
  X <- FB.data.pooled[FB.samples, selected.cols] %>%
    dplyr::mutate_if(is.numeric, function(x) {
      s <- numeric(sum(tr | te));
      s[tr] <- (x[tr]-mean(x[tr]))/sd(x[tr]);
      s[te] <- (x[te]-mean(x[te]))/sd(x[te]);
      s
      }) %>%
    #dplyr::mutate_if(is.numeric, scale) %>%
    as.matrix()

  y <- FB.data.pooled[FB.samples, 'status'] != 'control'
  cv_model <- cv.glmnet(X[tr,], y[tr], alpha = 0, family='binomial')
  mod <- glmnet(X[tr,], y[tr], alpha = 0, family = "binomial",lambda = cv_model$lambda.min)
  sample.CV[[as.character(k)]] <- mean(predict(mod, type = "class", newx=X[te,])[,'s0'] == y[te])
}

data.frame(acc=unlist(sample.CV),
           fold=names(sample.CV)) %>%
  dplyr::filter(!is.na(acc)) %>%
  dplyr::mutate(mean_acc=mean(acc)) %>%
  ggplot(aes(x=1,y=acc)) +
  #geom_boxplot(outlier.shape=NA, lwd=.5) + 
  geom_jitter(width=.1, height=0, size=.25) + 
  scale_y_continuous(limits=c(0,1)) +
  scale_x_continuous(limits=c(.85,1.15)) +
  my_theme() +
  theme(axis.text.x=element_blank(),
        axis.ticks = element_blank()) +
  labs(x='',y='accuracy')
```

heatmap

```{r Fig4F, fig.width=5.5 * fig_scale, fig.height=5.5 * fig_scale}
FB.samples <- row.names(FB.data.pooled)[grepl('FB',FB.data.pooled$donor)]
FB.X <- FB.data.pooled[FB.samples, selected.cols] %>%
  dplyr::mutate_if(is.numeric, scale)

row.names(FB.X) <- FB.data.pooled[FB.samples, c('donor','timepoint')] %>% 
  dplyr::mutate(name=gsub('t','',gsub('FB_','',paste0(donor,'_',timepoint)))) %>% 
  dplyr::pull(name)
colnames(FB.X) <- gsub('_downsampled|_switch','',colnames(FB.X))
pca_hm <- pcadat.combined[gsub('pooled_(FB)_([0-9]*)','\\1\\2',FB.samples), c('PC1','PC2','dataset','donor')]
row.names(pca_hm) <- row.names(FB.X)

FB.meta <- FB.data.pooled[FB.samples,] %>%
  dplyr::mutate(sample_donor=gsub("FB_","FB",sample_donor),
                dataset=ifelse(grepl('FB',donor),
                               ifelse(sample_donor %in% FB_train_donors,'training','testing'),
                               'validation')) %>%
  dplyr::select(donor,sample_donor,diagnosis,dataset,sex,age,timepoint,
                splenomegaly,lymphoproliferation,pct_B,pct_IgA,pct_IgG,complication)

dend <- hclust(dist(FB.X))
# dend <- reorder(as.dendrogram(dend),wts=order(match(FB.meta %>%
#                                                       dplyr::arrange(diagnosis) %>%
#                                                       dplyr::mutate(sample_donor=gsub('pooled_FB_','',sample_donor)) %>%
#                                                       dplyr::pull(sample_donor),
#                                                     row.names(FB.X))),agglo.FUN=mean)

class_df <- data.frame(class=plyr::revalue(as.character(cutree(dend, k = 4)),
                                           c('2'='cluster3',
                                             '3'='cluster1',
                                             '1'='cluster2',
                                             '4'='cluster2')),
                       name=row.names(FB.X))
class_colors <- c('cluster3'=diagnosis_colors[['control']],
                  'cluster1'=cat_cols[['structural']],
                  'cluster2'=cat_cols[['diversity']])

hm1 <- Heatmap(t(FB.X),
               col=colorRamp2(c(-2,0, 2), c("blue",'gray90', "red")),
               cluster_columns=dend,
               cluster_rows=TRUE,
               show_row_names=TRUE,
               show_column_names=TRUE,
               show_row_dend=TRUE,
               top_annotation=HeatmapAnnotation(df=FB.meta[,c('diagnosis','sex','age',#'pct_B','pct_IgA','pct_IgG',
                                                              #'splenomegaly','lymphoproliferation',
                                                              #'lung','PnPS','Tetanus','Diphteria',
                                                              #'complication','EUROclass',
                                                              'donor','timepoint'),drop=FALSE] %>% 
                                                  tibble::remove_rownames() %>%
                                                  dplyr::mutate(name=gsub('t','',gsub('FB_','',paste0(donor,'_',timepoint)))) %>%
                                                  dplyr::left_join(class_df,by='name') %>%
                                                  dplyr::select(-c(donor,timepoint)) %>%
                                                  tibble::column_to_rownames('name'),
                                                show_annotation_name=TRUE,
                                                show_legend=TRUE,
                                                simple_anno_size=unit(2,'mm'),
                                                annotation_legend_param=list(title_gp=gpar(fontsize=7),
                                                                             labels_gp=gpar(fontsize=7),
                                                                             grid_height = unit(3, "mm"),
                                                                             grid_width = unit(3, "mm")),
                                                annotation_name_gp=gpar(fontsize=8),
                                                col=list('diagnosis'=diagnosis_colors,
                                                         'dataset'=c('training'='gray60','testing'='gray20',
                                                                     'Charite'='red3','QP'='forestgreen'),
                                                         'class'=class_colors,
                                                         'EUROclass'=c('smB+21norm'='gray20',
                                                                       'smB+21low'='orange3',
                                                                       'smB-21norm'='lightblue3',
                                                                       'smB-21low'='darkseagreen2'),
                                                         'complication'=c('c'='coral3',
                                                                          'io'='olivedrab4'),
                                                         'age'=colorRamp2(c(20,50,80),c("yellow2","orange2","coral3")),
                                                         'sex'=c('male'='mistyrose2',
                                                                 'female'='lightseagreen'))),
               column_names_gp=gpar(fontsize=4),
               row_names_gp =gpar(fontsize = 7),
               row_dend_width = unit(7, "mm"),
               row_split = columns %>% tibble::column_to_rownames('col') %>% .[selected.cols, 'category'], 
               heatmap_legend_param=list(labels_gp=gpar(fontsize=7),
                                         title_gp=gpar(fontsize=8),
                                         legend_height=unit(10,'mm')),
               row_title=NULL,
               name='z-score')

hm3 <- Heatmap(t(pca_hm[,c('PC1','PC2')]),
               cluster_columns=dend,
               cluster_rows=TRUE,
               col=colorRamp2(c(-5, 0, 5), c("darkorchid3","gray90","darkseagreen3")),
               show_row_names=TRUE,
               show_column_names=TRUE,
               show_row_dend=TRUE,
               column_names_gp=gpar(fontsize=4),
               row_names_gp =gpar(fontsize=7),
               row_dend_width = unit(7, "mm"),
               heatmap_legend_param=list(labels_gp=gpar(fontsize=7),
                                         title_gp=gpar(fontsize=8),
                                         legend_height = unit(10, "mm")),
               row_title=NULL,
               name='PC')

hm1 %v% hm3
```

show PCA loadings per category

```{r FigS4C, fig.width=3 * fig_scale, fig.height=3 * fig_scale}
tmp <- data.frame(abs(pca$rotation)) %>%
  tibble::rownames_to_column('col') %>%
  dplyr::left_join(columns, by='col') %>%
  dplyr::mutate(col=factor(col, levels=columns %>% 
                             dplyr::filter(col %in% selected.cols) %>% 
                             dplyr::pull(col)),
                col_name=gsub("_downsampled|_switch",'',col))

cols_shapes <- tmp %>% 
  dplyr::distinct(category, col,col_name) %>%
  dplyr::mutate(var_col=cat_cols[as.character(category)]) %>%
  dplyr::group_by(category) %>%
  dplyr::mutate(var_ind=1:n())

ggplot(tmp,aes(x=PC1,y=PC2, color=col, shape=col)) +
  geom_point(size=1) +
  my_theme() +
  scale_color_manual(values=cols_shapes %>%
                       dplyr::pull(var_col,col),
                     labels=cols_shapes %>%
                       dplyr::pull(col_name)) +
  scale_shape_manual(values=cols_shapes %>%
                       dplyr::pull(var_ind,col),
                     labels=cols_shapes %>%
                       dplyr::pull(col_name)) +
  theme(legend.key.height=unit(3,'mm'),
        aspect.ratio=pca$sdev[2]**2/pca$sdev[1]**2,
        legend.spacing=unit(0,'mm'),
        legend.position='bottom',
        legend.box='horizontal') +
  labs(x='abs. PC 1', y='abs. PC 2',
       color='',shape='',fill='',alpha='')
```

categorize the samples

```{r FigS4D, fig.width=3 * fig_scale, fig.height=1.5 * fig_scale}
ggplot(pca_hm %>% tibble::rownames_to_column('name') %>%
         dplyr::filter(dataset!='validation') %>%
         dplyr::left_join(class_df, by='name'), 
       aes(x=PC1,y=PC2, color=class, shape=dataset)) +
  geom_point(size=1) +
  geom_line(aes(group=donor)) +
  my_theme() +
  scale_shape_manual(values=c('testing'=15,'training'=16,'validation'=18)) +
  scale_color_manual(values=class_colors) +
  theme(legend.key.height=unit(3,'mm'),
        aspect.ratio=pca$sdev[2]**2/pca$sdev[1]**2) +
  labs(x=paste0('PC 1 (',100*signif(pca$sdev[1]**2/sum(pca$sdev**2),3),'%)'),
       y=paste0('PC 2 (',100*signif(pca$sdev[2]**2/sum(pca$sdev**2),3),'%)'),
       color='class')
```

box plots for selected features

```{r Fig4J, fig.width=4.5 * fig_scale, fig.height=1.7 * fig_scale}
features.here <- c('frac_clusters_SG2',
                   'mean_cluster_size_downsampled',
                   'alpha_ratio_clusters',
                   'n_homology_switch',
                   'frac_breaks_duplications')

groups <- c('cluster1','cluster2','cluster3','control','BRCA1','ATM')

tmp <- cbind(rbind(FB.data.pooled[,features.here],
                   QP.data[,features.here]),
             human.meta[,c('diagnosis','dataset','donor','timepoint')]) %>%
  tibble::rownames_to_column('sample_donor') %>%
  dplyr::mutate(name=gsub('t','',gsub('FB_','',paste0(donor,'_',timepoint)))) %>%
  dplyr::left_join(class_df,by='name') %>%
  dplyr::mutate(group=ifelse(dataset=='validation',as.character(diagnosis),as.character(class))) %>%
  dplyr::filter(group %in% groups) %>%
  dplyr::mutate(group=factor(group,levels=groups)) %>%
  dplyr::group_by(group) %>%
  dplyr::mutate(ng=n_distinct(donor),
                group_new=paste0(group,' (n=',ng,')')) %>%
  gather(col,value,features.here) %>%
  dplyr::mutate(col=factor(gsub("_downsampled|_switch",'',col),
                           levels=gsub("_downsampled|_switch","",features.here)))

ggplot(tmp, aes(x=group,y=value,fill=group)) +
  geom_boxplot(outlier.shape=NA, lwd=.5) +
  geom_jitter(width=.1,height=0,size=.5) +
  stat_compare_means(comparisons=list(c('cluster1','cluster3'),
                                      c('cluster2','cluster3'),
                                      c('control','ATM'),
                                      c('control','BRCA1')),
                     method='wilcox.test',label = "p.signif", hide.ns=TRUE, size=2, vjust=.5) +
  facet_wrap(.~col,ncol=length(features.here),scales='free_y') +
  my_theme() +
  coord_cartesian(clip='off') +
  scale_fill_manual(values=c(diagnosis_colors,class_colors)) +
  theme(axis.text.x=element_text(angle=90,hjust=1,vjust=.5),
        legend.position='none') +
  scale_x_discrete(breaks=c('cluster1','cluster2','cluster3','control','ATM','BRCA1'),
                   labels=c('cluster1','cluster2','cluster3','control','ATM','BRCA1')) +
  labs(x='',y='', color='diagnosis')
```

check association of clinics with features

```{r clinical_features, fig.width=3 * fig_scale, fig.height=2.5 * fig_scale}
pvals <- list()
for (x in c("Tetanus",
            "Diphteria",
            "PnPS")) {
  for (f in selected.cols) {
    data <- FB.data.pooled[FB.samples,c(x,f)] %>%
      tidyr::drop_na()
    form <- as.formula(paste0(x,'=="negative" ~ ',f))
    res <- glm(form, family='binomial', data=data) %>% summary()
    pvals[[paste0(x,'_',f)]] <- data.frame(pval=coef(res)[2,4],
                                           vacc=x,
                                           col=f)
  }
}

pvals <- list()
for (x in c("splenomegaly",
            "lymphoproliferation",
            "lung")) {
  for (f in selected.cols) {
    data <- FB.data.pooled[FB.samples,c(x,f,'status')] %>%
      dplyr::filter(status!='control') %>%
      tidyr::drop_na()
    form <- as.formula(paste0(x,'!="" ~ ',f))
    res <- glm(form, family='binomial', data=data) %>% summary()
    pvals[[paste0(x,'_',f)]] <- data.frame(pval=coef(res)[2,4],
                                           vacc=x,
                                           col=f)
  }
}

FB.data.pooled[FB.samples, ] %>%
  dplyr::mutate(Tetanus=ifelse(status=='control','positive',Tetanus),
                Diphteria=ifelse(status=='control','positive',Diphteria),
                PnPS=ifelse(status=='control','positive',PnPS)) %>%
  gather(feature, value, c('frac_clusters_SA1','frac_clusters_SG3')) %>%
  gather(vacc, vacc_status, c('Tetanus','Diphteria','PnPS')) %>%
  dplyr::mutate(vacc_status=factor(ifelse(vacc_status=='','no data',vacc_status),
                                   c('negative','positive','no data'))) %>%
  ggplot(aes(x=vacc_status, y=value, fill=vacc_status)) +
  geom_boxplot(outlier.shape=NA) +
  geom_jitter(aes(color=status),width=.1, size=.5) +
  stat_compare_means(comparisons=list(c('negative','positive'),
                                      c('positive','no data')),
                     method='wilcox.test', label = "p.signif", hide.ns=TRUE, size=2, vjust=0) +
  facet_grid(feature ~ vacc, switch='y') + 
  my_theme() +
  scale_fill_manual(values=c('negative'='red3','positive'='gray')) +
  scale_color_manual(values=diagnosis_colors) +
  labs(x='',y='') +
  coord_cartesian(clip='off') +
  theme(legend.position='none',
        strip.clip='off',
        axis.text.x=element_text(angle=45, hjust=1, vjust=1))

FB.data.pooled[FB.samples, ] %>%
  gather(feature, value, c('frac_clusters_SG3','alpha_ratio_clusters')) %>%
  gather(disease, disease_status, c('splenomegaly','lymphoproliferation','lung')) %>%
  dplyr::mutate(disease_status=factor(ifelse(status=='control','control',
                                             ifelse(disease_status=='','absent','present')),
                                      levels=c('control','absent','present'))) %>%
  ggplot(aes(x=disease_status, y=value, fill=disease_status)) +
  geom_boxplot(outlier.shape=NA) +
  geom_jitter(aes(color=status),width=.1, size=.5) +
  stat_compare_means(comparisons=list(c('absent','present'),
                                      c('absent','control')),
                     method='wilcox.test', label = "p.signif", hide.ns=TRUE, size=2, vjust=0) +
  facet_grid(feature ~ disease, switch='y', scales='free_y') + 
  my_theme() +
  scale_fill_manual(values=c('present'='red3','absent'='gray')) +
  scale_color_manual(values=diagnosis_colors) +
  labs(x='',y='') +
  coord_cartesian(clip='off') +
  theme(legend.position='none',
        strip.clip='off',
        axis.text.x=element_text(angle=45, hjust=1, vjust=1))
```

check aggregated scores

```{r aggregated_FB_heatmap, fig.width=5.5 * fig_scale, fig.height=3.5 * fig_scale}
FB.X.aggregated <-  FB.data.pooled[FB.samples, selected.cols] %>%
  dplyr::mutate_if(is.numeric, scale) %>%
  tibble::rownames_to_column('sample') %>%
  gather(col, value, -sample) %>%
  dplyr::left_join(columns, by='col') %>%
  dplyr::group_by(sample, category) %>%
  dplyr::summarise(value=mean(abs(value))) %>%
  dplyr::left_join(FB.meta %>% tibble::rownames_to_column('sample'), by='sample') %>%
  #dplyr::filter(diagnosis!='control') %>%
  dplyr::select(sample,category,value) %>%
  spread(category,value) %>%
  tibble::column_to_rownames('sample')

hm2 <- Heatmap(t(FB.X.aggregated),
               #cluster_columns=dend,
               col=colorRamp2(c(0, 1.5), c("lightyellow2","steelblue3")),
               cluster_rows=TRUE,
               show_row_names=TRUE,
               show_column_names=TRUE,
               show_row_dend=TRUE,
               top_annotation=HeatmapAnnotation(df=FB.meta[row.names(FB.X.aggregated),c('diagnosis','sex','age',
                                                                                        'donor','timepoint'),drop=FALSE] %>% 
                                                  tibble::remove_rownames() %>%
                                                  dplyr::mutate(name=gsub('t','',gsub('FB_','',paste0(donor,'_',timepoint)))) %>%
                                                  dplyr::left_join(class_df,by='name') %>%
                                                  dplyr::select(-c(donor,timepoint)) %>%
                                                  tibble::column_to_rownames('name'),
                                                show_annotation_name=TRUE,
                                                show_legend=TRUE,
                                                simple_anno_size=unit(2,'mm'),
                                                annotation_legend_param=list(title_gp=gpar(fontsize=7),
                                                                             labels_gp=gpar(fontsize=7),
                                                                             grid_height = unit(3, "mm"),
                                                                             grid_width = unit(3, "mm")),
                                                annotation_name_gp=gpar(fontsize=8),
                                                col=list('diagnosis'=diagnosis_colors,
                                                         'dataset'=c('training'='gray60','testing'='gray20',
                                                                     'Charite'='red3','QP'='forestgreen'),
                                                         'class'=class_colors,
                                                         'EUROclass'=c('smB+21norm'='gray20',
                                                                       'smB+21low'='orange3',
                                                                       'smB-21norm'='lightblue3',
                                                                       'smB-21low'='darkseagreen2'),
                                                         'complication'=c('c'='coral3',
                                                                          'io'='olivedrab4'),
                                                         'age'=colorRamp2(c(20,50,80),c("yellow2","orange2","coral3")),
                                                         'sex'=c('male'='mistyrose2',
                                                                 'female'='lightseagreen'))),
               column_names_gp=gpar(fontsize=4),
               row_names_gp =gpar(fontsize = 7),
               row_dend_width = unit(7, "mm"),
               heatmap_legend_param=list(labels_gp=gpar(fontsize=7),
                                         title_gp=gpar(fontsize=8),
                                         legend_height=unit(10,'mm')),
               row_title=NULL,
               name='category score')

hm2
```

```{r aggregated_FB_boxplots, fig.width=4.5 * fig_scale, fig.height=1.7 * fig_scale}
tmp <- FB.X.aggregated %>%
  tibble::rownames_to_column('sample') %>%
  dplyr::left_join(FB.meta %>% tibble::rownames_to_column('sample'), by='sample') %>%
  dplyr::mutate(name=gsub('t','',gsub('FB_','',paste0(donor,'_',timepoint)))) %>%
  dplyr::left_join(class_df,by='name') %>%
  gather(col,value,c('diversity','isotypes','breaks','context','structural'))

ggplot(tmp, aes(x=class,y=value,fill=class)) +
  geom_boxplot(outlier.shape=NA, lwd=.5) +
  geom_jitter(width=.1,height=0,size=.5) +
  stat_compare_means(comparisons=list(c('cluster1','cluster3'),
                                      c('cluster2','cluster3'),
                                      c('cluster1','cluster2')),
                     method='wilcox.test',label = "p.signif", hide.ns=TRUE, size=2, vjust=.5) +
  facet_wrap(.~col,ncol=length(features.here),scales='free_y') +
  my_theme() +
  coord_cartesian(clip='off') +
  scale_fill_manual(values=c(diagnosis_colors,class_colors)) +
  theme(axis.text.x=element_text(angle=45,hjust=1,vjust=1),
        legend.position='none') +
  labs(x='',y='category score', color='diagnosis')
```

check correlation with measured B cell proportion in FACS

```{r pct_B_cor, fig.width=3 * fig_scale, fig.height=3 * fig_scale}
FB.data.pooled %>%
  gather(feature, value, selected.cols) %>%
  dplyr::group_by(feature) %>%
  dplyr::summarise(R=cor(value, pct_B, use='na.or.complete')) %>%
  dplyr::left_join(columns, by=c('feature'='col')) %>%
  dplyr::mutate(feature=gsub('_downsampled|_switch','',feature)) %>%
  dplyr::arrange(R) %>%
  dplyr::mutate(feature=factor(feature, levels=feature)) %>%
  ggplot(aes(y=R, x=feature, fill=category)) + 
  geom_col() +
  my_theme() +
  coord_flip() +
  scale_fill_manual(values=cat_cols) +
  labs(x='',y='correlation with % B cells') +
  theme(legend.position=c(.2,.9))

FB.data.pooled %>%
  gather(feature, value, selected.cols) %>%
  dplyr::group_by(feature) %>%
  dplyr::summarise(R=cor(value, pct_IgA, use='na.or.complete')) %>%
  dplyr::left_join(columns, by=c('feature'='col')) %>%
  dplyr::mutate(feature=gsub('_downsampled|_switch','',feature)) %>%
  dplyr::arrange(R) %>%
  dplyr::mutate(feature=factor(feature, levels=feature)) %>%
  ggplot(aes(y=R, x=feature, fill=category)) + 
  geom_col() +
  my_theme() +
  coord_flip() +
  scale_fill_manual(values=cat_cols) +
  labs(x='',y='correlation with % IgA cells') +
  theme(legend.position=c(.2,.9))

FB.data.pooled %>%
  gather(feature, value, selected.cols) %>%
  dplyr::group_by(feature) %>%
  dplyr::summarise(R=cor(value, pct_IgG, use='na.or.complete')) %>%
  dplyr::left_join(columns, by=c('feature'='col')) %>%
  dplyr::mutate(feature=gsub('_downsampled|_switch','',feature)) %>%
  dplyr::arrange(R) %>%
  dplyr::mutate(feature=factor(feature, levels=feature)) %>%
  ggplot(aes(y=R, x=feature, fill=category)) + 
  geom_col() +
  my_theme() +
  coord_flip() +
  scale_fill_manual(values=cat_cols) +
  labs(x='',y='correlation with % IgG cells') +
  theme(legend.position=c(.2,.9))

```

what drives clustering within clusters 1 + 2?

```{r cluster_drivers, fig.width=3 * fig_scale, fig.height=3 * fig_scale}
use <- class_df$class!='cluster3'
cluster.var <- fitExtractVarPartModel(t(FB.X[use,]), ~ (1 | class), data=class_df[use,'class',drop=FALSE]) %>%
  as.data.frame() %>%
  tibble::rownames_to_column('col') %>%
  gather(variable,variance,-col)

cluster.var %>%
  dplyr::filter(variable=='class') %>%
  dplyr::left_join(columns %>% dplyr::mutate(col=gsub('_downsampled|_switch','',col)) %>% 
                     dplyr::filter(category %in% c('context','breaks','isotypes','diversity','structural')), 
                   by='col') %>%
  dplyr::arrange(variance) %>%
  dplyr::mutate(col=factor(col, levels=col)) %>%
  ggplot(aes(y=variance, x=col, fill=category)) + 
  geom_col() +
  my_theme() +
  coord_flip() +
  scale_fill_manual(values=cat_cols) +
  labs(x='',y='explained variance between cluster1+2 by class') +
  theme(legend.position=c(.9,.1))
```

box plots for selected features

```{r Fig4K, fig.width=4. * fig_scale, fig.height=1.7 * fig_scale}
features.here <- c('frac_clusters_SG1',
                   'n_untemplated_switch',
                   'frac_breaks_intra',
                   'mean_length')

groups <- c('cluster1','cluster2','cluster3','control','BRCA1','ATM')

tmp <- cbind(rbind(FB.data.pooled[,features.here],
                   QP.data[,features.here]),
             human.meta[,c('diagnosis','dataset','donor','timepoint')]) %>%
  tibble::rownames_to_column('sample_donor') %>%
  dplyr::mutate(name=gsub('t','',gsub('FB_','',paste0(donor,'_',timepoint)))) %>%
  dplyr::left_join(class_df,by='name') %>%
  dplyr::mutate(group=ifelse(dataset=='validation',as.character(diagnosis),as.character(class))) %>%
  dplyr::filter(group %in% groups) %>%
  dplyr::mutate(group=factor(group,levels=groups)) %>%
  dplyr::group_by(group) %>%
  dplyr::mutate(ng=n_distinct(donor),
                group_new=paste0(group,' (n=',ng,')')) %>%
  gather(col,value,features.here) %>%
  dplyr::mutate(col=factor(gsub("_downsampled|_switch",'',col),
                           levels=gsub("_downsampled|_switch","",features.here)))

ggplot(tmp, aes(x=group,y=value,fill=group)) +
  geom_boxplot(outlier.shape=NA, lwd=.5) +
  geom_jitter(width=.1,height=0,size=.5) +
  stat_compare_means(comparisons=list(c('cluster1','cluster3'),
                                      c('cluster1','cluster2'),
                                      c('cluster2','cluster3'),
                                      c('control','ATM'),
                                      c('control','BRCA1')),
                     method='wilcox.test',label = "p.signif", hide.ns=TRUE, size=2, vjust=.5) +
  facet_wrap(.~col,ncol=length(features.here),scales='free_y') +
  my_theme() +
  coord_cartesian(clip='off') +
  scale_fill_manual(values=c(diagnosis_colors,class_colors)) +
  theme(axis.text.x=element_text(angle=90,hjust=1,vjust=.5),
        legend.position='none') +
  scale_x_discrete(breaks=c('cluster1','cluster2','cluster3','control','ATM','BRCA1'),
                   labels=c('cluster1','cluster2','cluster3','control','ATM','BRCA1')) +
  labs(x='',y='', color='diagnosis')
```

# session info

```{r sessioninfo}
sessionInfo()
```

### 2024-11-15 prepare_input.R

@@ -1,0 +1,397 @@
+library(dplyr)
+library(tidyverse)
+library(scales)
+
+columns <- list('diversity'=c('nclusters_final_downsampled',
+                              'nclusters_eff_downsampled',
+                              'mean_cluster_size_downsampled',
+                              'std_cluster_size_downsampled',
+                              'cluster_gini_downsampled',
+                              'cluster_entropy_downsampled',
+                              'cluster_inverse_simpson_downsampled',
+                              'top_clone_occupancy_downsampled',
+                              'big_clones_occupancy_downsampled'),
+                'isotypes'=c('alpha_ratio_clusters',
+                             'frac_clusters_SA1','frac_clusters_SA2','frac_clusters_SE',
+                             'frac_clusters_SG1','frac_clusters_SG2','frac_clusters_SG3',
+                             'frac_clusters_SG4','frac_clusters_SM',
+                             'frac_clusters_Sa','frac_clusters_Sg3',
+                             'frac_clusters_Sg2b','frac_clusters_Sg2c',
+                             'frac_clusters_Sm','frac_clusters_Sg1'),
+                'structural'=c('insert_frequency',
+                               'frac_breaks_inversions','frac_breaks_duplications',
+                               'median_duplication_size','median_inversion_size'),
+                'context'=c('homology_fw','homology_rv',
+                            'donor_score_S', 'donor_score_W', 
+                            'donor_score_WGCW', 'donor_score_GAGCT',
+                            'donor_score_GGGST',
+                            'donor_complexity',
+                            'acceptor_score_S', 'acceptor_score_W', 
+                            'acceptor_score_WGCW', 'acceptor_score_GAGCT',
+                            'acceptor_score_GGGST',
+                            'acceptor_complexity',
+                            'n_homology_switch', 
+                            'n_untemplated_switch',
+                            'frac_blunt'),
+                'breaks'=c('mean_length',
+                           'std_length',
+                           'mean_GC',
+                           'std_GC',
+                           'mean_length_SA',
+                           'mean_length_SG',
+                           'std_length_SA',
+                           'std_length_SG',
+                           'frac_breaks_single',
+                           'frac_breaks_sequential',
+                           'frac_breaks_intra',
[…]

### 2024-11-15 swibrid_runs/human/combine_replicates.sh

@@ -1,0 +1,10 @@
+#!/bin/bash -e
+
+donors=$(cut -f 1 ../../sodar/2023_FB/a_2023_FB.txt | grep -v Sample | sort -u )
+
+for donor in ${donors}; do
+	samples1=$(cut -f 1,17 ../../sodar/2023_FB/a_2023_FB.txt | grep -v Extract | grep ${donor} | cut -f 2 | sort -u | tr '\n' ',' | sed 's/,$//')
+	samples2=$(grep -v FAIL ../../sodar/2023_FB/a_2023_FB.txt | cut -f 1,17 | grep -v Extract | grep ${donor} | cut -f 2 | sort -u | tr '\n' ',' | sed 's/,$//')
+	[[ ${samples1} != ${samples2} ]] && echo ${samples1} ${samples2} #swibrid combine_replicates -s ${samples2} -c pooled_${donor} --nmax 50000
+	#[[ ! -s input/pooled_${donor}_info.csv ]] && echo ${donor} ${samples} && swibrid combine_replicates -s ${samples} -c pooled_${donor} --nmax 50000
+done

### 2024-11-15 swibrid_runs/human/plot_bars.py

@@ -1,0 +1,194 @@
+import os
+import sys
+import numpy as np
+import pandas as pd
+from argparse import ArgumentParser
+
+parser=ArgumentParser()
+
+parser.add_argument("--figure", dest="figure", help="""output figure""")
+parser.add_argument("--samples", dest="samples", help="""list of samples""")
+parser.add_argument("--color", dest="color", default='cluster', help="""what to color circles by [cluster]""")
+parser.add_argument(
+    "--fig_width",
+    dest="fig_width",
+    default=3,
+    type=float,
+    help="""fig width""",
+)
+parser.add_argument(
+    "--fig_height",
+    dest="fig_height",
+    type=float,
+    default=.75,
+    help="""figure height in inches""",
+)
+parser.add_argument(
+    "--dpi",
+    dest="dpi",
+    default=300,
+    type=int,
+    help="""dpi of output figure [300]""",
+)
+
+def rand_cmap(
+    nlabels,
+    type="bright",
+    first_color_black=False,
+    last_color_black=False,
+    verbose=False,
+    bad="gray",
+    under="gray",
+    over="gray",
+    seed=0,
+):
+    """
+    Creates a random colormap to be used together with matplotlib. Useful for segmentation tasks
+    :param nlabels: Number of labels (size of colormap)
+    :param type: 'bright' for strong colors, 'soft' for pastel colors
+    :param first_color_black: Option to use first color as black, True or False
+    :param last_color_black: Option to use last color as black, True or False
+    :param verbose: Prints the number of labels and shows the colormap. True or False
+    :return: colormap for matplotlib
+    (from https://stackoverflow.com/questions/14720331/how-to-generate-random-colors-in-matplotlib/14720445)
+    """
+    from matplotlib.colors import LinearSegmentedColormap
+    import colorsys
+    import numpy as np
+
+    np.random.seed(seed)
+
+    # Generate color map for bright colors, based on hsv
+    if type == "bright":
+        randHSVcolors = [
+            (
+                np.random.uniform(low=0.0, high=1),
+                np.random.uniform(low=0.2, high=1),
+                np.random.uniform(low=0.9, high=1),
+            )
+            for i in range(nlabels)
+        ]
+
+        # Convert HSV list to RGB
[…]

### 2024-11-15 swibrid_runs/human/plot_bars.sh

@@ -1,0 +1,4 @@
+#!/bin/bash -e
+
+python plot_bars.py --samples "20231211_rFP_01_downsampled,20231211_rFP_29_downsampled,20211122_85_n500_1_downsampled" --figure paper_figures/20240320_bar_plots_downsampled_isotype.pdf
+

### 2024-11-15 swibrid_runs/human/plot_breakpoint_stats.py

@@ -1,0 +1,255 @@
+
+
+def setup_argparse(parser):
+    parser.add_argument(
+        "-g",
+        "--gaps",
+        dest="gaps",
+        help="""required: file with gap positions (output of get_gaps.py)""",
+    )
+    parser.add_argument(
+        "-c",
+        "--clustering",
+        dest="clustering",
+        help="""file with clustering results""",
+    )
+    parser.add_argument(
+        "-a",
+        "--clustering_analysis",
+        dest="clustering_analysis",
+        help="""file with clustering analysis""",
+    )
+    parser.add_argument("-o", "--output", help="""output plot""")
+    parser.add_argument(
+        "-b",
+        "--binsize",
+        dest="binsize",
+        default=50,
+        type=int,
+        help="""binsize [50]""",
+    )
+    parser.add_argument(
+        "--scale_factor",
+        dest="scale_factor",
+        default=10,
+        type=int,
+        help="""factor to increase binsize for 2D histogram [10]""",
+    )
+    parser.add_argument(
+        "--max_gap",
+        dest="max_gap",
+        default=75,
+        type=int,
+        help="""max gap size to ignore [75]""",
+    )
+    parser.add_argument("--sample", dest="sample", help="""sample name (for figure title)""")
+    parser.add_argument(
+        "--switch_coords",
+        dest="switch_coords",
+        default="chr14:106050000-106337000:-",
+        help="""coordinates of switch region [chr14:106050000-106337000:-]""",
+    )
+    parser.add_argument(
+        "--switch_annotation",
+        dest="switch_annotation",
+        help="""bed file with switch annotation""",
+    )
+    parser.add_argument(
+        "--use_clones",
+        dest="use_clones",
+        help="""comma-separated list of clones to use or 'all' (default: filtered clusters, excluding singletons)""",
+    )
+    parser.add_argument(
+        "--weights",
+        dest="weights",
+        default="cluster",
+        help="""use different weights ("cluster" | "reads" | "adjusted") [cluster]""",
+    )
+
+
+if __name__ == '__main__':
+    import numpy as np
+    import pandas as pd
+    import scipy.sparse
+    import scipy.stats
+    import pysam
[…]

### 2024-11-15 swibrid_runs/human/plot_breakpoints.sh

@@ -1,0 +1,9 @@
+#!/bin/bash -e
+
+#sample=$1
+mkdir -p paper_figures
+
+#for sample in 20210422_plasmid_downsampled 20210618_FakePoly_downsampled 20211122_85_n500_1_downsampled 20220329_86_100000_1_downsampled 20231020_HD_K_downsampled 20231020_IRF4_downsampled 20231026_HD_K_downsampled 20231026_IRF4_downsampled 20231211_rFP_01_downsampled 20231211_rFP_04_downsampled 20231211_rFP_15_downsampled 20231211_rFP_22_downsampled 20231211_rFP_29_downsampled; do 
+for sample in 20231211_rFP_01_downsampled 20231211_rFP_29_downsampled 20211122_85_n500_1_downsampled; do
+	swibrid get_breakpoint_stats -g pipeline/${sample}/${sample}_gaps.npz -c pipeline/${sample}/${sample}_clustering.csv -a pipeline/${sample}/${sample}_cluster_analysis.csv -b 50 --max_gap 75 --switch_coords chr14:105583000-105872000:- --switch_annotation hg38_switch_regions.bed -o paper_figures/${sample}_breakpoints.csv -p paper_figures/${sample}_breakpoints.pdf --weights cluster
+done

### 2024-11-15 swibrid_runs/human/plot_circles.py

@@ -1,0 +1,257 @@
+import os
+import sys
+import numpy as np
+import pandas as pd
+from argparse import ArgumentParser
+
+parser=ArgumentParser()
+
+parser.add_argument("--figure", dest="figure", help="""output figure""")
+parser.add_argument("--samples", dest="samples", help="""list of samples""")
+parser.add_argument("--color", dest="color", default='cluster', help="""what to color circles by [cluster]""")
+parser.add_argument("--ncols", dest="ncols", default=3, type=int, help="""number of columns [3]""")
+parser.add_argument(
+    "--fig_width",
+    dest="fig_width",
+    default=1,
+    type=float,
+    help="""width of panels in inches [5]""",
+)
+parser.add_argument(
+    "--fig_height",
+    dest="fig_height",
+    type=float,
+    default=1,
+    help="""width of panels in inches [4.5]""",
+)
+parser.add_argument(
+    "--dpi",
+    dest="dpi",
+    default=300,
+    type=int,
+    help="""dpi of output figure [300]""",
+)
+parser.add_argument(
+    "--use_circlify",
+    dest="use_circlify",
+    default=False,
+    action="store_true",
+    help="""use circify package for circular packing""",
+)
+
+def rand_cmap(
+    nlabels,
+    type="bright",
+    first_color_black=False,
+    last_color_black=False,
+    verbose=False,
+    bad="gray",
+    under="gray",
+    over="gray",
+    seed=0,
+):
+    """
+    Creates a random colormap to be used together with matplotlib. Useful for segmentation tasks
+    :param nlabels: Number of labels (size of colormap)
+    :param type: 'bright' for strong colors, 'soft' for pastel colors
+    :param first_color_black: Option to use first color as black, True or False
+    :param last_color_black: Option to use last color as black, True or False
+    :param verbose: Prints the number of labels and shows the colormap. True or False
+    :return: colormap for matplotlib
+    (from https://stackoverflow.com/questions/14720331/how-to-generate-random-colors-in-matplotlib/14720445)
+    """
+    from matplotlib.colors import LinearSegmentedColormap
+    import colorsys
+    import numpy as np
+
+    np.random.seed(seed)
+
+    # Generate color map for bright colors, based on hsv
+    if type == "bright":
+        randHSVcolors = [
[…]

### 2024-11-15 swibrid_runs/human/plot_circles.sh

@@ -1,0 +1,5 @@
+#!/bin/bash -e
+
+python plot_circles.py --samples "20231211_rFP_01_downsampled,20231211_rFP_29_downsampled,20211122_85_n500_1_downsampled" --ncols 3 --figure paper_figures/20240320_bubble_plots_downsampled_cluster.pdf --color cluster
+#python plot_circles.py --samples "20231211_rFP_01_downsampled,20231211_rFP_29_downsampled,20211122_85_n500_1_downsampled" --ncols 3 --figure paper_figures/20240320_bubble_plots_downsampled_isotype.pdf --color isotype
+

### 2024-11-15 swibrid_runs/human/plot_clustering.sh

@@ -1,0 +1,13 @@
+#!/bin/bash -e
+
+#sample=$1
+mkdir -p paper_figures
+
+#for sample in 20210422_plasmid_downsampled 20210618_FakePoly_downsampled 20211122_85_n500_1_downsampled 20220329_86_100000_1_downsampled 20231020_HD_K_downsampled 20231020_IRF4_downsampled 20231026_HD_K_downsampled 20231026_IRF4_downsampled 20231211_rFP_01_downsampled 20231211_rFP_04_downsampled 20231211_rFP_15_downsampled 20231211_rFP_22_downsampled 20231211_rFP_29_downsampled; do 
+for sample in 20231211_rFP_01_downsampled 20231211_rFP_29_downsampled 20211122_85_n500_1_downsampled 20210422_plasmid_downsampled; do
+	swibrid plot_clustering --msa pipeline/${sample}/${sample}_msa.npz --figure paper_figures/${sample}.png --linkage pipeline/${sample}/${sample}_linkage.npz --info input/${sample}_info.csv  --clustering_results pipeline/${sample}/${sample}_clustering.csv --switch_annotation hg38_switch_regions.bed --switch_coords "chr14:105583000-105872000:-" --color_by cluster --clustering_stats pipeline/${sample}/${sample}_cluster_stats.csv --sample ${sample} --dpi 500 --fig_width 4 --fig_height 1.25 --cmax 0.5 --linkage_border .2 --cutoff 0.01 --cmax 0.1 --no_x_ticks --omit_scale_bar --dpi 600
+done
+
+for sample in 20211122_85_n500_1_downsampled; do
+	swibrid plot_clustering --msa pipeline/${sample}/${sample}_msa.npz --figure paper_figures/${sample}_cov.png --linkage pipeline/${sample}/${sample}_linkage.npz --info input/${sample}_info.csv  --clustering_results pipeline/${sample}/${sample}_clustering.csv --switch_annotation hg38_switch_regions.bed --switch_coords "chr14:105583000-105872000:-" --color_by coverage --clustering_stats pipeline/${sample}/${sample}_cluster_stats.csv --sample ${sample} --dpi 500 --fig_width 4 --fig_height 1.25 --cmax 0.5 --linkage_border .2 --cutoff 0.01 --cmax 0.1 --no_x_ticks --omit_scale_bar --dpi 600 --variants_matrix pipeline/${sample}/${sample}_variants.npz 
+done

### 2024-11-15 swibrid_runs/mouse/plot_clustering.sh

@@ -1,0 +1,9 @@
+#!/bin/bash -e
+
+#sample=$1
+mkdir -p paper_figures
+
+for sample in 20240817_S.A3_downsampled 20220425_WT_3_downsampled; do
+	swibrid plot_clustering --msa pipeline/${sample}/${sample}_msa.npz --figure paper_figures/${sample}.png --linkage pipeline/${sample}/${sample}_linkage.npz --info input/${sample}_info.csv  --clustering_results pipeline/${sample}/${sample}_clustering.csv --switch_annotation mm10_switch_regions.bed --switch_coords "chr12:113200000-113500000:-" --color_by cluster --clustering_stats pipeline/${sample}/${sample}_cluster_stats.csv --sample ${sample} --dpi 500 --fig_width 4 --fig_height 1.25 --cmax 0.5 --linkage_border .2 --cutoff 0.01 --cmax 0.1 --no_x_ticks --omit_scale_bar --dpi 600
+done
+

### 2024-12-19 paper_figures.Rmd
add data

@@
 author: "Benedikt Obermayer"
-date: "2024/11/12"
+date: "2024/12/10"
 output: 
@@
 library(grid)
-library(DescTools)
+#library(DescTools)
 library(lme4)
@@
                       'IgG4def'='burlywood4','TNFRSF13B'='palegreen4',
-                      'other'='#BC80BD','AID'='#FDB462',
+                      'other'='#BC80BD','AID'='darkgoldenrod3',
                       'ATM'='#A65628','LIG4'='mediumpurple2',
@@
   dplyr::mutate(variable=factor(gsub('log10_nreads','# reads',variable),
-                                levels=rev(c('donor','cohort','batch','sequencing','# reads','Residuals')))) %>%
-  dplyr::filter(variable!='Residuals', category %in% c('diversity','diversity_raw'), variable!='cohort')
+                                levels=rev(c('donor','batch','sequencing','# reads','Residuals')))) %>%
+  dplyr::filter(variable!='Residuals', category %in% c('diversity','diversity_raw'))
 
@@
   dplyr::group_by(variable) %>%
-  do(wc=wilcox.test(variance ~ category, data=.)) %>%
-  summarise(variable, p=wc$p.value)
+  dplyr::do(wc=wilcox.test(variance ~ category, data=.)) %>%
+  dplyr::summarise(variable, p=wc$p.value)
 
@@
   dplyr::group_by(variable) %>%
-  do(wc=wilcox.test(variance ~ category, data=., exact=FALSE)) %>%
-  summarise(variable, p=wc$p.value)
+  dplyr::do(wc=wilcox.test(variance ~ category, data=., exact=FALSE)) %>%
+  dplyr::summarise(variable, p=wc$p.value)
 
@@
   dplyr::group_by(variable) %>%
-  do(wc=wilcox.test(variance ~ averaging, paired=TRUE, data=., exact=FALSE)) %>%
-  summarise(variable, p=wc$p.value)
+  dplyr::do(wc=wilcox.test(variance ~ averaging, paired=TRUE, data=., exact=FALSE)) %>%
+  dplyr::summarise(variable, p=wc$p.value)
 
@@
 ```
+
+(this PCA uses `r dim(vincendeau.X)[2]` features)
 
@@
 
+(this PCA uses `r dim(mouse.X)[2]` and `r dim(mouse.X.noSg)[2]` features)
+
 selected features in the CH12 data (2023 batches)
@@
 ```
+
+(this PCA uses `r dim(CH12.X)[2]` features)
 
@@
         legend.position='bottom',
+        legend.key.spacing.y=unit(0,'mm'),
         legend.box='horizontal') +
@@
 ```
+
+(this PCA uses `r dim(FB.X.subclass)[2]` features)
 
@@
                                levels=gsub('frac_clusters_','',subclass.cols)))
-ggplot(tmp, aes(x=feature, y=frac_clusters, color=diagnosis)) +
-  geom_boxplot(position=position_dodge(width=.8),width=.7,outlier.shape=NA) +
-  geom_point(position=position_dodge(width=.8), size=.5) +
+
+pvals <- tmp %>%
+  dplyr::group_by(feature) %>%
+  dplyr::do(wc=wilcox.test(frac_clusters ~ diagnosis, data=.),
+            y=max(.$frac_clusters)) %>%
+  dplyr::summarise(feature, p.value=wc$p.value, y)
+
+ggplot(tmp, aes(x=feature, y=frac_clusters)) +
+  geom_boxplot(aes(color=diagnosis), position=position_dodge(width=.8),width=.7,outlier.shape=NA) +
+  geom_point(aes(color=diagnosis), position=position_dodge(width=.8), size=.5) +
   geom_line(data=tmp %>% dplyr::filter(diagnosis=='control'),
@@
             aes(group=donor), position=position_nudge(x=+.2), alpha=.25, size=.25) +
+  geom_text(data=pvals, aes(label=stars.pval(p.value), y=1.1*y)) +
   scale_color_manual(values=diagnosis_colors) +
   my_theme() +
+  coord_cartesian(clip='off') +
   theme(legend.position=c(.8,.8)) +
@@
 
-```{r FigS4X2, fig.width=2.5 * fig_scale, fig.height=2 * fig_scale}
-FB.data.pooled %>%
+```{r FigS4X2, fig.width=2.25 * fig_scale, fig.height=2 * fig_scale}
+tmp <- FB.data.pooled %>%
   dplyr::mutate(alpha_ratio_FACS=pct_IgA/(pct_IgA + pct_IgG)) %>%
-  ggplot(aes(x=alpha_ratio_FACS, y=alpha_ratio_clusters, color=status)) + 
+  dplyr::group_by(status) %>%
+  dplyr::mutate(status_R=paste0(status," (R=",signif(cor(alpha_ratio_FACS,alpha_ratio_clusters,use='na.or.complete'),2),")"))
+ggplot(tmp, aes(x=alpha_ratio_FACS, y=alpha_ratio_clusters, color=status)) + 
   geom_abline(slope=1, intercept=0, size=.5) +
-  stat_cor(method = "pearson", aes(label=..r.label..), size=2.5, show.legend=FALSE, label.x=.75, label.y=c(.4, .33)) +
+  #stat_cor(method = "pearson", aes(label=..r.label..), size=2.5, show.legend=FALSE, label.x=.75, label.y=c(.4, .33)) +
   geom_point() + 
   my_theme() +
-  scale_color_manual(values=diagnosis_colors) +
-  labs(x='FACS', y='SWIBRID', title='IgA / (IgA + IgG)')
+  scale_color_manual(values=diagnosis_colors, breaks=unique(tmp$status), labels=unique(tmp$status_R)) +
+  labs(x='FACS', y='SWIBRID', title='IgA / (IgA + IgG)') +
+  theme(legend.position=c(.8,.15),
+        legend.box=element_blank())
 ```
@@
 
-```{r Fig4H, fig.width=2.5 * fig_scale, fig.height=1.5 * fig_scale}
+```{r Fig4H, fig.width=2. * fig_scale, fig.height=1.5 * fig_scale}
 set.seed(1)
@@
                        name=row.names(FB.X))
+
 class_colors <- c('cluster3'=diagnosis_colors[['control']],
@@
 
-```{r FigS4C, fig.width=3 * fig_scale, fig.height=3 * fig_scale}
+```{r FigS4C, fig.width=3 * fig_scale, fig.height=4 * fig_scale}
 tmp <- data.frame(abs(pca$rotation)) %>%
@@
         legend.spacing=unit(0,'mm'),
+        legend.key.spacing.y=unit(0,'mm'),
         legend.position='bottom',
         legend.box='horizontal') +
+  guides(color=guide_legend(ncol=2),
+         shape=guide_legend(ncol=2)) +
   labs(x='abs. PC 1', y='abs. PC 2',
@@
 
-```{r FigS4D, fig.width=3 * fig_scale, fig.height=1.5 * fig_scale}
+```{r FigS4D, fig.width=2.5 * fig_scale, fig.height=2 * fig_scale}
 ggplot(pca_hm %>% tibble::rownames_to_column('name') %>%
@@
   geom_point(size=1) +
+  geom_point(data=pcadat.combined %>% dplyr::filter(dataset=='validation'),
+             aes(color=diagnosis), size=1) +
   geom_line(aes(group=donor)) +
   my_theme() +
-  scale_shape_manual(values=c('testing'=15,'training'=16,'validation'=18)) +
-  scale_color_manual(values=class_colors) +
+  scale_shape_manual(values=c('testing'=16,'training'=16,'validation'=15), guide='none') +
+  scale_color_manual(values=c(class_colors,diagnosis_colors), 
+                     breaks=c(names(class_colors),
+                              names(diagnosis_colors))) +
+  guides(color=guide_legend(override.aes = list(shape=c(rep(16,3),rep(15,7)),size=2), ncol=4)) +
   theme(legend.key.height=unit(3,'mm'),
+        legend.position='bottom',
+        legend.key.spacing.y=unit(0,'mm'),
         aspect.ratio=pca$sdev[2]**2/pca$sdev[1]**2) +
@@
        y=paste0('PC 2 (',100*signif(pca$sdev[2]**2/sum(pca$sdev**2),3),'%)'),
-       color='class')
+       color='')
+```
+
+what drives clustering within clusters 1 + 2?
+
+```{r cluster_drivers, fig.width=3 * fig_scale, fig.height=3 * fig_scale}
+use <- class_df$class!='cluster3'
+cluster.var <- fitExtractVarPartModel(t(FB.X[use,]), ~ (1 | class), data=class_df[use,'class',drop=FALSE]) %>%
+  as.data.frame() %>%
+  tibble::rownames_to_column('col') %>%
+  gather(variable,variance,-col)
+
+cluster.var %>%
+  dplyr::filter(variable=='class') %>%
+  dplyr::left_join(columns %>% dplyr::mutate(col=gsub('_downsampled|_switch','',col)) %>% 
+                     dplyr::filter(category %in% c('context','breaks','isotypes','diversity','structural')), 
+                   by='col') %>%
+  dplyr::arrange(variance) %>%
+  dplyr::mutate(col=factor(col, levels=col)) %>%
+  ggplot(aes(y=variance, x=col, fill=category)) + 
+  geom_col() +
+  my_theme() +
+  coord_flip() +
+  scale_fill_manual(values=cat_cols) +
+  labs(x='',y='explained variance between cluster1+2 by class') +
+  theme(legend.position=c(.9,.1))
 ```
@@
 
-```{r Fig4J, fig.width=4.5 * fig_scale, fig.height=1.7 * fig_scale}
-features.here <- c('frac_clusters_SG2',
+```{r Fig4J, fig.width=5.25 * fig_scale, fig.height=1.7 * fig_scale}
+features.here <- c('frac_clusters_SG1',
+                   'frac_clusters_SG2',
                    'mean_cluster_size_downsampled',
-                   'alpha_ratio_clusters',
                    'n_homology_switch',
+                   'n_untemplated_switch',
+                   'frac_breaks_intra',
                    'frac_breaks_duplications')
@@
   stat_compare_means(comparisons=list(c('cluster1','cluster3'),
+                                      c('cluster1','cluster2'),
                                       c('cluster2','cluster3'),
@@
 
-```{r clinical_features, fig.width=3 * fig_scale, fig.height=2.5 * fig_scale}
+```{r FigS4X4, fig.width=3 * fig_scale, fig.height=2.5 * fig_scale}
 pvals <- list()
@@
 
-pvals <- list()
 for (x in c("splenomegaly",
@@
 
-```{r aggregated_FB_heatmap, fig.width=5.5 * fig_scale, fig.height=3.5 * fig_scale}
+```{r FigS4X5, fig.width=5.5 * fig_scale, fig.height=3.5 * fig_scale}
 FB.X.aggregated <-  FB.data.pooled[FB.samples, selected.cols] %>%
@@
   #dplyr::filter(diagnosis!='control') %>%
-  dplyr::select(sample,category,value) %>%
+  dplyr::group_by(category) %>%
+  dplyr::mutate(#value=(value-mean(value[diagnosis=='control'])/sd(value[diagnosis=='control'])),
+                name=gsub('t','',gsub('FB_','',paste0(donor,'_',timepoint)))) %>%
+  dplyr::select(name,category,value) %>%
   spread(category,value) %>%
-  tibble::column_to_rownames('sample')
-
-hm2 <- Heatmap(t(FB.X.aggregated),
-               #cluster_columns=dend,
-               col=colorRamp2(c(0, 1.5), c("lightyellow2","steelblue3")),
+  tibble::column_to_rownames('name')
+
+FB.meta.aggregated <- FB.meta[,c('diagnosis','sex','age',
+                                 'donor','timepoint'),drop=FALSE] %>% 
+  dplyr::filter(!(diagnosis %in% c('control','subclass'))) %>%
+  tibble::remove_rownames() %>%
+  dplyr::mutate(name=gsub('t','',gsub('FB_','',paste0(donor,'_',timepoint)))) %>%
+  dplyr::left_join(class_df,by='name') %>%
+  dplyr::select(-c(donor,timepoint)) %>%
+  tibble::column_to_rownames('name')
+
+samples <- row.names(FB.meta.aggregated)
+dend <- hclust(dist(FB.X.aggregated[samples,]))
+dend <- reorder(as.dendrogram(dend),wts=order(match(FB.meta.aggregated %>%
+                                                      tibble::rownames_to_column('sample') %>%
+                                                      dplyr::arrange(class) %>%
+                                                      dplyr::pull(sample),
+                                                    samples)),agglo.FUN=mean)
+
+hm2 <- Heatmap(t(FB.X.aggregated[samples,]),
+               cluster_columns=dend,
+               col=colorRamp2(c(0, 2), c("lightyellow2","darkorchid3")),
                cluster_rows=TRUE,
@@
                show_row_dend=TRUE,
-               top_annotation=HeatmapAnnotation(df=FB.meta[row.names(FB.X.aggregated),c('diagnosis','sex','age',
-                                                                                        'donor','timepoint'),drop=FALSE] %>% 
-                                                  tibble::remove_rownames() %>%
-                                                  dplyr::mutate(name=gsub('t','',gsub('FB_','',paste0(donor,'_',timepoint)))) %>%
-                                                  dplyr::left_join(class_df,by='name') %>%
-                                                  dplyr::select(-c(donor,timepoint)) %>%
-                                                  tibble::column_to_rownames('name'),
+               top_annotation=HeatmapAnnotation(df=FB.meta.aggregated,
                                                 show_annotation_name=TRUE,
@@
 
-```{r aggregated_FB_boxplots, fig.width=4.5 * fig_scale, fig.height=1.7 * fig_scale}
-tmp <- FB.X.aggregated %>%
-  tibble::rownames_to_column('sample') %>%
-  dplyr::left_join(FB.meta %>% tibble::rownames_to_column('sample'), by='sample') %>%
-  dplyr::mutate(name=gsub('t','',gsub('FB_','',paste0(donor,'_',timepoint)))) %>%
-  dplyr::left_join(class_df,by='name') %>%
-  gather(col,value,c('diversity','isotypes','breaks','context','structural'))
-
-ggplot(tmp, aes(x=class,y=value,fill=class)) +
-  geom_boxplot(outlier.shape=NA, lwd=.5) +
-  geom_jitter(width=.1,height=0,size=.5) +
-  stat_compare_means(comparisons=list(c('cluster1','cluster3'),
-                                      c('cluster2','cluster3'),
-                                      c('cluster1','cluster2')),
-                     method='wilcox.test',label = "p.signif", hide.ns=TRUE, size=2, vjust=.5) +
-  facet_wrap(.~col,ncol=length(features.here),scales='free_y') +
-  my_theme() +
-  coord_cartesian(clip='off') +
-  scale_fill_manual(values=c(diagnosis_colors,class_colors)) +
-  theme(axis.text.x=element_text(angle=45,hjust=1,vjust=1),
-        legend.position='none') +
-  labs(x='',y='category score', color='diagnosis')
-```
+<!-- ```{r aggregated_FB_boxplots, fig.width=4.5 * fig_scale, fig.height=1.7 * fig_scale} -->
+<!-- tmp <- FB.X.aggregated %>% -->
+<!--   tibble::rownames_to_column('sample') %>% -->
+<!--   dplyr::left_join(FB.meta %>% tibble::rownames_to_column('sample'), by='sample') %>% -->
+<!--   dplyr::mutate(name=gsub('t','',gsub('FB_','',paste0(donor,'_',timepoint)))) %>% -->
+<!--   dplyr::left_join(class_df,by='name') %>% -->
+<!--   gather(col,value,c('diversity','isotypes','breaks','context','structural')) -->
+
+<!-- ggplot(tmp, aes(x=class,y=value,fill=class)) + -->
+<!--   geom_boxplot(outlier.shape=NA, lwd=.5) + -->
+<!--   geom_jitter(width=.1,height=0,size=.5) + -->
+<!--   stat_compare_means(comparisons=list(c('cluster1','cluster3'), -->
+<!--                                       c('cluster2','cluster3'), -->
+<!--                                       c('cluster1','cluster2')), -->
+<!--                      method='wilcox.test',label = "p.signif", hide.ns=TRUE, size=2, vjust=.5) + -->
+<!--   facet_wrap(.~col,ncol=length(features.here),scales='free_y') + -->
+<!--   my_theme() + -->
+<!--   coord_cartesian(clip='off') + -->
+<!--   scale_fill_manual(values=c(diagnosis_colors,class_colors)) + -->
+<!--   theme(axis.text.x=element_text(angle=45,hjust=1,vjust=1), -->
+<!--         legend.position='none') + -->
+<!--   labs(x='',y='category score', color='diagnosis') -->
+<!-- ``` -->
 
@@
 
-what drives clustering within clusters 1 + 2?
-
-```{r cluster_drivers, fig.width=3 * fig_scale, fig.height=3 * fig_scale}
-use <- class_df$class!='cluster3'
-cluster.var <- fitExtractVarPartModel(t(FB.X[use,]), ~ (1 | class), data=class_df[use,'class',drop=FALSE]) %>%
-  as.data.frame() %>%
-  tibble::rownames_to_column('col') %>%
-  gather(variable,variance,-col)
-
-cluster.var %>%
-  dplyr::filter(variable=='class') %>%
-  dplyr::left_join(columns %>% dplyr::mutate(col=gsub('_downsampled|_switch','',col)) %>% 
-                     dplyr::filter(category %in% c('context','breaks','isotypes','diversity','structural')), 
-                   by='col') %>%
-  dplyr::arrange(variance) %>%
-  dplyr::mutate(col=factor(col, levels=col)) %>%
-  ggplot(aes(y=variance, x=col, fill=category)) + 
-  geom_col() +
-  my_theme() +
-  coord_flip() +
-  scale_fill_manual(values=cat_cols) +
-  labs(x='',y='explained variance between cluster1+2 by class') +
-  theme(legend.position=c(.9,.1))
-```
-
-box plots for selected features
-
-```{r Fig4K, fig.width=4. * fig_scale, fig.height=1.7 * fig_scale}
-features.here <- c('frac_clusters_SG1',
-                   'n_untemplated_switch',
-                   'frac_breaks_intra',
-                   'mean_length')
-
-groups <- c('cluster1','cluster2','cluster3','control','BRCA1','ATM')
-
-tmp <- cbind(rbind(FB.data.pooled[,features.here],
-                   QP.data[,features.here]),
-             human.meta[,c('diagnosis','dataset','donor','timepoint')]) %>%
-  tibble::rownames_to_column('sample_donor') %>%
-  dplyr::mutate(name=gsub('t','',gsub('FB_','',paste0(donor,'_',timepoint)))) %>%
-  dplyr::left_join(class_df,by='name') %>%
-  dplyr::mutate(group=ifelse(dataset=='validation',as.character(diagnosis),as.character(class))) %>%
-  dplyr::filter(group %in% groups) %>%
-  dplyr::mutate(group=factor(group,levels=groups)) %>%
-  dplyr::group_by(group) %>%
-  dplyr::mutate(ng=n_distinct(donor),
-                group_new=paste0(group,' (n=',ng,')')) %>%
-  gather(col,value,features.here) %>%
-  dplyr::mutate(col=factor(gsub("_downsampled|_switch",'',col),
-                           levels=gsub("_downsampled|_switch","",features.here)))
-
-ggplot(tmp, aes(x=group,y=value,fill=group)) +
-  geom_boxplot(outlier.shape=NA, lwd=.5) +
-  geom_jitter(width=.1,height=0,size=.5) +
-  stat_compare_means(comparisons=list(c('cluster1','cluster3'),
-                                      c('cluster1','cluster2'),
-                                      c('cluster2','cluster3'),
-                                      c('control','ATM'),
-                                      c('control','BRCA1')),
-                     method='wilcox.test',label = "p.signif", hide.ns=TRUE, size=2, vjust=.5) +
-  facet_wrap(.~col,ncol=length(features.here),scales='free_y') +
-  my_theme() +
-  coord_cartesian(clip='off') +
-  scale_fill_manual(values=c(diagnosis_colors,class_colors)) +
-  theme(axis.text.x=element_text(angle=90,hjust=1,vjust=.5),
-        legend.position='none') +
-  scale_x_discrete(breaks=c('cluster1','cluster2','cluster3','control','ATM','BRCA1'),
-                   labels=c('cluster1','cluster2','cluster3','control','ATM','BRCA1')) +
-  labs(x='',y='', color='diagnosis')
-```
-
 # session info

### 2024-12-19 swibrid_runs/mouse/demultiplex_dataset.sh

@@ -1,11 +1,15 @@
 #!/bin/bash 
 
+mkdir -p demux
 cat index/barcodes_primers.fa | makeblastdb -in - -title "barcodes_primers" -out demux/barcodes_primers -input_type fasta -dbtype nucl
 
-gunzip -c raw_data/${dataset}.fastq.gz | awk '{if(NR%4==1) {printf(">%s\n",substr($0,2));} else if(NR%4==2) print;}' | blastn -db demux/barcodes_primers -query - -task blastn-short -max_target_seqs 50 -outfmt "6 saccver qaccver slen qlen pident length qstart qend evalue" -gapopen 5 -gapextend 2 -reward 3 -penalty -4 -evalue 1 -num_threads 1 -perc_identity 70 > demux/blast_output_${dataset}.txt
+for sample in $(grep SAMPLES config.yaml | tr ',' '\n' | sed 's/SAMPLES: //' | tr -d '"][")'); do
 
-mkdir -p demux/${dataset}
-swibrid demultiplex -i raw_data/${dataset}.fastq.gz -b demux/blast_output_${dataset}.txt -f demux/${dataset}_demultiplexing.png --collapse -o demux/${dataset} -r demux/${dataset}_demultiplexing_stats.csv --split-reads -s  sample_sheets/${dataset}_sample_sheet.tsv -c 70
+	date=$(echo ${sample} | cut -c 1-8)
+	grep ${sample} sample_sheets/${date}_sample_sheet.tsv > demux/sample_sheet.tsv
 
-mkdir -p input
-ln -rfs demux/${dataset}/* input
+	gunzip -c raw_data/${sample}.fastq.gz | awk '{if(NR%4==1) {printf(">%s\n",substr($0,2));} else if(NR%4==2) print;}' | blastn -db demux/barcodes_primers -query - -task blastn-short -max_target_seqs 50 -outfmt "6 saccver qaccver slen qlen pident length qstart qend evalue" -gapopen 5 -gapextend 2 -reward 3 -penalty -4 -evalue 1 -num_threads 1 -perc_identity 70 > demux/blast_output.txt
+
+	swibrid demultiplex -i raw_data/${sample}.fastq.gz -b demux/blast_output.txt --collapse -o input/ -s demux/sample_sheet.tsv -c 70
+done
+

### 2025-01-09 paper_figures.Rmd
add Vincendeau & HTGTS; re-number figures

@@
 author: "Benedikt Obermayer"
-date: "2024/12/10"
+date: "2024/12/19"
 output: 
@@
 genotype_colors <- c('WT'='gray40','Xlf'='#e1c2f0','Shld1'='#b870dc','Shld1_Xlf'='#a447d2',
-                     'Trp53bp1 KO'='darksalmon','Brca1 KO'='darkkhaki','Rif1 KO'='skyblue1','Lig4 KO'='mediumpurple2')
+                     'Trp53bp1 KO'='darksalmon','Brca1 KO'='darkkhaki','Rif1 KO'='skyblue1','Lig4 KO'='mediumpurple2', 
+                     'Trp53bp1'='darksalmon','Atm'='#A65628','Lig4'='mediumpurple2', 'Rif1'='skyblue1', 'Brca1'='darkkhaki')
 
@@
 ```
-
-<!-- # get data -->
-
-<!-- ```{r get_data} -->
-<!-- # data from Vincendeau for Fig. 2 -->
-<!-- # mouse tissue data for Figs. 2 + 3  -->
-<!-- # CH12 data for Fig. 3 -->
-<!-- # patient data for Fig. 4 -->
-<!-- FB.data <- read.csv('data/FB_data.csv',header=1, row.names=1) -->
-<!-- FB.data.pooled <- read.csv('data/FB_data_pooled.csv',header=1, row.names=1) -->
-<!-- QP.data <- read.csv('data/QP_data.csv',header=1, row.names=1) -->
-<!-- ``` -->
 
@@
 
-```{r FigS3A,fig.width=3.5 * fig_scale,fig.height=1.7 * fig_scale}
+we now have `r length(selected.cols)` robust features
+
+```{r FigS2F,fig.width=3.5 * fig_scale,fig.height=1.7 * fig_scale}
 HD.var %>%  
@@
 
-```{r FigS3D_1, fig.width=1.4 * fig_scale, fig.height=1.7 * fig_scale}
+```{r FigS2J_1, fig.width=1.4 * fig_scale, fig.height=1.7 * fig_scale}
 tmp <- HD.var %>%  
@@
 
-```{r FigS3D_2, fig.width=1.4 * fig_scale, fig.height=1.7 * fig_scale}
+```{r FigS2J_2, fig.width=1.4 * fig_scale, fig.height=1.7 * fig_scale}
 tmp <- HD.var %>%  
@@
 
-```{r FigS3D_3, fig.width=1.4 * fig_scale,fig.height=1.7 * fig_scale}
+```{r FigS2J_3, fig.width=1.4 * fig_scale,fig.height=1.7 * fig_scale}
 HD.data.br <- read.csv('data/HD_data_by_reads.csv', header=1, row.names=1)
@@
 
-```{r FigS3B, fig.width=4.25 * fig_scale, fig.height=4.25 * fig_scale}
+```{r FigS2H, fig.width=4.25 * fig_scale, fig.height=4.25 * fig_scale}
 library(dendextend)
@@
 
-```{r FigS3E, fig.width=1.5 * fig_scale, fig.height=1.75 * fig_scale}
+```{r FigS2K, fig.width=1.5 * fig_scale, fig.height=1.75 * fig_scale}
 cols.here.mouse <- c('mean_length_SG')
@@
 
-```{r FigS3C, fig.width=2.75 * fig_scale, fig.height=5 * fig_scale}
+```{r FigS2G, fig.width=2.75 * fig_scale, fig.height=5 * fig_scale}
 tmp <- HD.scaled %>%
@@
   dplyr::group_by(category) %>%
-  dplyr::mutate(var_ind=1:n())
+  dplyr::mutate(var_ind=1:n()) %>%
+  dplyr::mutate(feature=factor(feature, levels=levels(tmp$feature))) %>% 
+  dplyr::arrange(feature)
 
@@
   geom_point(aes(color=feature,shape=feature),size=1) + 
+  #geom_text_repel(aes(label=feature), size=1) +
   my_theme() +
@@
   dplyr::group_by(category) %>%
-  dplyr::mutate(var_ind=1:n())
+  dplyr::mutate(var_ind=1:n()) %>%
+  dplyr::mutate(variable=factor(variable, levels=levels(tmp$variable))) %>% 
+  dplyr::arrange(variable)
 
@@
 # Figure 3 + supplement (CH12)
+
+```{r FigS3A, fig.width=5.5 * fig_scale, fig.height=4.5 * fig_scale}
+HTGTS.data <- read.csv("data/HTGTS_data.csv", header=1, row.names=1) %>%
+  tibble::column_to_rownames('sample')
+
+HTGTS.cols <- intersect(colnames(HTGTS.data),selected.cols.mouse)
+
+tmp <- HTGTS.data %>%
+  gather(variable, value, HTGTS.cols) %>%
+  dplyr::group_by(variable, celltype) %>%
+  dplyr::mutate(value=scale(value)) %>%
+  dplyr::group_by(celltype, genotype, variable) %>%
+  dplyr::summarise(mean=mean(value),
+                   se=sd(value)/sqrt(n())) %>%
+  tidyr::pivot_wider(id_cols=c('variable'),names_from=c('celltype','genotype'),values_from=c('mean','se')) %>%
+  dplyr::left_join(columns,by=c('variable'='col')) %>%
+  dplyr::group_by(variable) %>%
+  dplyr::mutate(variable=factor(variable, levels=columns %>% 
+                                  dplyr::filter(col %in% HTGTS.cols) %>% 
+                                  dplyr::pull(col)),
+                var_col=cat_cols[category]) %>%
+  dplyr::arrange(variable)
+
+cols_shapes <- tmp %>% 
+  dplyr::distinct(category, variable) %>%
+  dplyr::mutate(var_col=cat_cols[as.character(category)],
+                variable_name=gsub('_downsampled','',variable)) %>%
+  dplyr::group_by(category) %>%
+  dplyr::mutate(var_ind=1:n()) %>%
+  dplyr::mutate(variable=factor(variable, levels=levels(tmp$variable))) %>% 
+  dplyr::arrange(variable)
+
+comparisons <- list(c('CH12','WT','Atm'),
+                    c('CH12','WT','Trp53bp1'),
+                    c('CH12','WT','Lig4'),
+                    c('spleen','WT','Atm'),
+                    c('spleen','WT','Trp53bp1'))
+
+plots <- lapply(comparisons, function(x) 
+  tmp %>%
+    dplyr::mutate_(mean1=paste0('mean_',x[[1]],'_',x[[2]]),
+                   mean2=paste0('mean_',x[[1]],'_',x[[3]]),
+                   se1=paste0('se_',x[[1]],'_',x[[2]]),
+                   se2=paste0('se_',x[[1]],'_',x[[3]])) %>%
+    dplyr::mutate(tstat=abs(mean2-mean1)/(sqrt(se2**2+se1**2)),
+                  pval=dt(tstat,4)) %>%
+    ggplot(aes(x=mean1,y=mean2)) +
+    geom_abline(slope=1,intercept=0,color='gray',linewidth=.5) +
+    geom_point(aes(color=variable,shape=variable,size=ifelse(pval < .05,'sig','ns'))) + 
+    stat_cor(method = "pearson", aes(label=..r.label..), size=2.5, show.legend=FALSE, label.y.npc = 'bottom') +
+    geom_errorbar(aes(ymin=mean2-se2,
+                      ymax=mean2+se2,
+                      color=variable),
+                  size=.25) +
+    geom_errorbarh(aes(xmin=mean1-se1,
+                       xmax=mean1+se1,
+                       color=variable),
+                   size=.25) +
+    my_theme() +
+    labs(color='',shape='') +
+    scale_color_manual(values=cols_shapes %>% 
+                         dplyr::pull(var_col,variable),
+                       labels=cols_shapes %>%
+                         dplyr::pull(variable_name)) +
+    scale_shape_manual(values=cols_shapes %>%
+                         dplyr::pull(var_ind,variable),
+                       labels=cols_shapes %>%
+                         dplyr::pull(variable_name)) +
+    scale_size_manual(values=c('sig'=2,'ns'=.5),guide='none') +
+    guides(color=guide_legend(ncol=4),
+           shape=guide_legend(ncol=4)) +
+    labs(x=x[[2]],y=x[[3]],title=x[[1]],fill='',color='',size=''))
+
+plot_grid(plot_grid(plotlist=lapply(plots, function(x) x + theme(legend.position='none')), 
+                    ncol=3, align='vh'),
+          get_legend(plots[[1]] + theme(legend.margin=margin(0,0,0,0,'mm'))), ncol=1, rel_heights=c(.65,.35))
+```
+
+```{r FigS3C, fig.width=3.75 * fig_scale, fig.height=2.5 * fig_scale}
+library(gtools)
+
+cols.here.HTGTS <- c('n_homology_switch',
+                     'frac_breaks_inversions',
+                     'cluster_entropy',
+                     'mean_cluster_size',
+                     'acceptor_score_W',
+                     'donor_score_W')
+
+HTGTS.data %>%
+  dplyr::mutate(celltype=factor(celltype, levels=c('CH12','spleen')),
+                genotype=factor(genotype, levels=c('WT','Atm','Lig4','Trp53bp1'))) %>%
+  gather(variable, value, cols.here.HTGTS) %>%
+  dplyr::select(variable, value, genotype, celltype) %>%
+  dplyr::left_join(columns, by=c('variable'='col')) %>%
+  dplyr::mutate(variable=factor(gsub('_downsampled|_switch','',variable), 
+                                levels=gsub('_downsampled|_switch','',cols.here.HTGTS))) %>%
+  dplyr::filter(category %in% c('diversity','isotypes','breaks','context','structural')) %>%
+  ggplot(aes(x=genotype, y=value, fill=genotype)) + 
+  geom_boxplot(position=position_dodge(width=.75), width=.7, outlier.shape=NA, lwd=.25) + 
+  geom_point(position=position_jitterdodge(jitter.width=.2, dodge.width=.75, jitter.height = 0), size=.5) + 
+  my_theme() +
+  coord_cartesian(clip='off') +
+  stat_compare_means(ref.group='WT',method='wilcox.test',label = "p.signif", hide.ns=TRUE, size=2) +
+  ggh4x::facet_grid2(celltype~variable, switch='y', scales='free_y', independent='y', strip=ggh4x::strip_vanilla(clip='off')) +
+  theme(axis.text.x=element_text(angle=45, hjust=1, vjust=1),
+        legend.position='none') +
+  scale_fill_manual(values=genotype_colors) +
+  #scale_x_discrete(breaks=c('lymph node','splenocytes','WT','Xlf','Shld1','Shld1_Xlf'),
+  #                 limits=c('lymph node','splenocytes','','WT','Xlf','Shld1','Shld1_Xlf'), drop=FALSE) +
+  labs(x='',y='', fill='') 
+```
+
+show homology values for HTGTS data
+
+```{r FigS3D, fig.width=3.25 * fig_scale, fig.height=1.25 * fig_scale}
+read.csv('data/HTGTS_homology.csv',header=1, row.names=1) %>%
+  dplyr::mutate(genotype=factor(genotype, levels=c('WT','Atm','Lig4','Trp53bp1'))) %>%
+  dplyr::filter(xor(isotype=='Sa', celltype=='spleen')) %>%
+  dplyr::group_by( celltype, genotype, isotype, n_homology) %>%
+  dplyr::summarise(mean_frac=mean(frac),
+                   se_frac=sd(frac)/sqrt(n())) %>%
+  ggplot(aes(x=n_homology, y=mean_frac, color=genotype)) +
+  geom_line(aes(group=genotype)) +
+  geom_errorbar(aes(ymin=mean_frac-se_frac,ymax=mean_frac+se_frac),width=0) +
+  ggh4x::facet_nested(. ~ celltype  + isotype , scales='free', space='free_x', nest_line = TRUE, switch='y') +
+  my_theme() +
+  scale_color_manual(values=genotype_colors) +
+  labs(color='', y='fraction junctions') +
+  coord_cartesian(clip='off') 
+```
+
+```{r FigS3B, fig.width=2.5 * fig_scale, fig.height=3. * fig_scale}
+HTGTS.CH12.X <- HTGTS.data[HTGTS.data$celltype=='CH12',HTGTS.cols] %>%
+  dplyr::mutate_if(is.numeric, scale) 
+HTGTS.spleen.X <- HTGTS.data[HTGTS.data$celltype=='spleen',HTGTS.cols] %>%
+  dplyr::mutate_if(is.numeric, scale)
+
+HTGTS.meta <- HTGTS.data %>%
+  dplyr::mutate(genotype=factor(genotype,levels=c('WT','Atm','Lig4','Trp53bp1'))) %>%
+  dplyr::select(celltype,genotype)
+
+pca.CH12 <- prcomp(HTGTS.CH12.X)
+pcadat.CH12 <- cbind(pca.CH12$x,HTGTS.meta[row.names(HTGTS.CH12.X),])
+pca.spleen <- prcomp(HTGTS.spleen.X)
+pcadat.spleen <- cbind(pca.spleen$x,HTGTS.meta[row.names(HTGTS.spleen.X),])
+
+p1 <- ggplot(pcadat.CH12,aes(x=PC1,y=PC2, color=genotype)) +
+  geom_point(size=1) +
+  geom_polygon(data = pcadat.CH12 %>%
+                 dplyr::group_by(genotype) %>%
+                 dplyr::slice(chull(PC1, PC2)), alpha = 0.25, aes(fill=genotype)) +
+  my_theme() +
+  theme(aspect.ratio=pca.CH12$sdev[2]**2/pca.CH12$sdev[1]**2,
+        legend.position=c(.92,.3),
+        legend.background = element_blank()) +
+  labs(x=paste0('PC 1 (',100*signif(pca.CH12$sdev[1]**2/sum(pca.CH12$sdev**2),3),'%)'),
+       y=paste0('PC 2 (',100*signif(pca.CH12$sdev[2]**2/sum(pca.CH12$sdev**2),3),'%)'),
+       color='',shape='',fill='', title='CH12')  +
+  scale_color_manual(values=genotype_colors) +
+  scale_fill_manual(values=genotype_colors)
+
+p2 <- ggplot(pcadat.spleen,aes(x=PC1,y=PC2, color=genotype)) +
+  geom_point(size=1) +
+  geom_polygon(data = pcadat.spleen %>%
+                 dplyr::group_by(genotype) %>%
+                 dplyr::slice(chull(PC1, PC2)), alpha = 0.25, aes(fill=genotype)) +
+  my_theme() +
+  theme(aspect.ratio=pca.spleen$sdev[2]**2/pca.spleen$sdev[1]**2,
+        legend.position='none',
+        legend.background = element_blank()) +
+  labs(x=paste0('PC 1 (',100*signif(pca.spleen$sdev[1]**2/sum(pca.spleen$sdev**2),3),'%)'),
+       y=paste0('PC 2 (',100*signif(pca.spleen$sdev[2]**2/sum(pca.spleen$sdev**2),3),'%)'),
+       color='',shape='',fill='', title='spleen')  +
+  scale_color_manual(values=genotype_colors) +
+  scale_fill_manual(values=genotype_colors)
+
+plot_grid(p1,p2, ncol=1, align='vh')
+```
+
+these PCAs use `r dim(HTGTS.spleen.X)[2]` (spleen) and `r dim(HTGTS.CH12.X)[2]` (CH12) features
 
@@
 
+```{r FigS3X1, fig.width=4 * fig_scale, fig.height=1.75 * fig_scale, include=FALSE, eval=FALSE}
+dualPCA <- function(X, Y) {
+  library(irlba)
+  # use CCA-like extension of dual PCA as described here: https://xinmingtu.cn/blog/2022/CCA_dual_PCA/
+  #sv <- irlba(as.matrix(X) %*% as.matrix(t(Y)), nv=n)
+  sv <- svd(as.matrix(X) %*% as.matrix(t(Y)))
+  x <- sv$u %*% diag(sqrt(sv$d))
+  row.names(x) <- row.names(X)
+  colnames(x) <- paste0('dPC',seq(1,dim(x)[2]))
+  y <- sv$v %*% diag(sqrt(sv$d))
+  row.names(y) <- row.names(Y)
+  colnames(y) <- paste0('dPC',seq(1,dim(y)[2]))
+  list(x=x, y=y, sdev=sqrt(sv$d/(dim(X)[2]-1)))
+}
+
+CH12.X <- CH12.data[CH12.take, mouse.cols] %>% 
+  dplyr::mutate_if(is.numeric, scale)
+mouse.X <- mouse.data[mouse.take, mouse.cols] %>% 
+  dplyr::mutate_if(is.numeric, scale)
+
+dpca <- dualPCA(CH12.X, mouse.X)
+dpcadat <- cbind(rbind(dpca$x, dpca$y),
+                 data.frame(group=c(CH12.data[row.names(dpca$x),'genotype'],
+                                    mouse.data$material)))
+
+ggplot(dpcadat,aes(x=dPC1,y=dPC2, color=group)) +
+  geom_point(size=1) +
+  geom_polygon(data = dpcadat %>%
+                 dplyr::group_by(group) %>%
+                 dplyr::slice(chull(dPC1, dPC2)), alpha = 0.1, aes(fill=group), linewidth=.1) +
+  my_theme() +
+  scale_fill_manual(values=c(genotype_colors,tissue_colors)) +
+  scale_color_manual(values=c(genotype_colors,tissue_colors)) +
+  theme(aspect.ratio=dpca$sdev[2]**2/dpca$sdev[1]**2,
+        legend.spacing=unit(0,'mm')) +
+  labs(x=paste0('dPC 1 (',100*signif(dpca$sdev[1]**2/sum(dpca$sdev**2),3),'%)'),
+       y=paste0('dPC 2 (',100*signif(dpca$sdev[2]**2/sum(dpca$sdev**2),3),'%)'),
+       color='',shape='',fill='')
+```
+
+```{r FigS3X2, fig.width=3 * fig_scale, fig.height=4 * fig_scale, include=FALSE, eval=FALSE}
+tmp <- t(t(t(as.matrix(rbind(CH12.X, mouse.X))) %*% as.matrix(dpcadat[,c('dPC1','dPC2')])) / colSums(dpcadat[,c('dPC1','dPC2')]**2)) %>%
+  as.data.frame() %>%
+  tibble::rownames_to_column('col') %>%
+  dplyr::left_join(columns, by='col') %>%
+  dplyr::mutate(col=factor(col, levels=columns %>% 
+                             dplyr::filter(col %in% mouse.cols) %>%
+                             dplyr::pull(col)),
+                col_name=gsub("_downsampled|_switch",'',col))
+
+cols_shapes <- tmp %>% 
+  dplyr::distinct(category, col,col_name) %>%
+  dplyr::mutate(var_col=cat_cols[as.character(category)]) %>%
+  dplyr::group_by(category) %>%
+  dplyr::mutate(var_ind=1:n()) %>%
+  dplyr::mutate(col=factor(col, levels=levels(tmp$col))) %>% 
+  dplyr::arrange(col)
+
+ggplot(tmp,aes(x=dPC1,y=dPC2, color=col, shape=col)) +
+  geom_point(size=1) +
+  #geom_text_repel(aes(label=col), size=1) +
+  my_theme() +
+  scale_color_manual(values=cols_shapes %>%
+                       dplyr::pull(var_col,col),
+                     labels=cols_shapes %>%
+                       dplyr::pull(col_name)) +
+  scale_shape_manual(values=cols_shapes %>%
+                       dplyr::pull(var_ind,col),
+                     labels=cols_shapes %>%
+                       dplyr::pull(col_name)) +
+  theme(legend.key.size=unit(1.5,'mm'),
+        aspect.ratio=dpca$sdev[2]**2/dpca$sdev[1]**2,
+        legend.spacing=unit(0,'mm'),
+        legend.key.spacing.y=unit(0,'mm'),
+        legend.position='bottom',
+        legend.box='horizontal') +
+  guides(color=guide_legend(ncol=2),
+         shape=guide_legend(ncol=2)) +
+  labs(x='dPC 1', y='dPC 2',
+       color='',shape='',fill='',alpha='')
+```
+
 selected features in the CH12 data (2023 batches)
@@
   scale_color_manual(values=genotype_colors) 
-
-# ggplot(pcadat,aes(x=PC3,y=PC4, color=genotype)) +
-#   geom_point(size=2) +
-#   geom_polygon(data = pcadat %>%
-#                  dplyr::group_by(genotype) %>%
-#                  dplyr::slice(chull(PC3, PC4)), alpha = 0.25, aes(fill=genotype)) +
-#   my_theme() +
-#   theme(aspect.ratio=pca$sdev[4]**2/pca$sdev[3]**2) +
-#   labs(x=paste0('PC 3 (',100*signif(pca$sdev[3]**2/sum(pca$sdev**2),3),'%)'),
-#        y=paste0('PC 4 (',100*signif(pca$sdev[4]**2/sum(pca$sdev**2),3),'%)'),
-#        color='',shape='',fill='') +
-#   scale_fill_manual(values=genotype_colors) +
-#   scale_color_manual(values=genotype_colors)
-# 
-# ggplot(pcadat,aes(x=PC1,y=PC2, color=batch)) +
-#   geom_point(size=2) +
-#   my_theme() +
-#   theme(aspect.ratio=pca$sdev[2]**2/pca$sdev[1]**2) +
-#   labs(x=paste0('PC 1 (',100*signif(pca$sdev[1]**2/sum(pca$sdev**2),3),'%)'),
-#        y=paste0('PC 2 (',100*signif(pca$sdev[2]**2/sum(pca$sdev**2),3),'%)'),
-#        color='',shape='')
-
 ```
@@
 
-```{r FigS3G, fig.width=1. * fig_scale, fig.height=1.75 * fig_scale}
+```{r FigS3X5, fig.width=1. * fig_scale, fig.height=1.75 * fig_scale, include=FALSE}
 batch.CV <- c()
@@
 ```
+
+(average accuracy is `r mean(unlist(batch.CV))` for cross-validation across batches and `r mean(unlist(sample.CV))` for cross-validation across samples)
 
@@
   theme(aspect.ratio=pca$sdev[2]**2/pca$sdev[1]**2,
+        legend.key.size=unit(1.5,'mm'),
+        legend.key.spacing.y=unit(0,'mm'),
         legend.position='bottom',
@@
 
+(this PCA uses `r dim(human.X)[2]` features)
+
 another PCA with only subclass features
 
-```{r FigS4X1, fig.width=2.25 * fig_scale, fig.height=2 * fig_scale}
+```{r FigS4C, fig.width=2.75 * fig_scale, fig.height=1.75 * fig_scale}
 FB.cols.subclass <- grep('frac_clusters_S[MGA]+',colnames(FB.data.pooled),value=TRUE)
@@
   scale_color_manual(values=diagnosis_colors) +
-  guides(color=guide_legend(override.aes = list(shape=15,size=2), ncol=3)) +
-  theme(aspect.ratio=pca$sdev[2]**2/pca$sdev[1]**2,
-        legend.position='bottom',
+  guides(color=guide_legend(override.aes = list(shape=15,size=2), ncol=1)) +
+  theme(aspect.ratio=pca.subclass$sdev[2]**2/pca.subclass$sdev[1]**2,
+        legend.position='right',
+        legend.key.size=unit(1.5,'mm'),
         legend.key.spacing.y=unit(0,'mm'),
         legend.box='horizontal') +
-  labs(x=paste0('PC 1 (',100*signif(pca$sdev[1]**2/sum(pca$sdev**2),3),'%)'),
-       y=paste0('PC 2 (',100*signif(pca$sdev[2]**2/sum(pca$sdev**2),3),'%)'),
+  labs(x=paste0('PC 1 (',100*signif(pca.subclass$sdev[1]**2/sum(pca.subclass$sdev**2),3),'%)'),
+       y=paste0('PC 2 (',100*signif(pca.subclass$sdev[2]**2/sum(pca.subclass$sdev**2),3),'%)'),
        color='',shape='',fill='')
@@
 
-```{r FigS4G, fig.width=2 * fig_scale, fig.height=1.7 * fig_scale}
+```{r FigS4B, fig.width=2 * fig_scale, fig.height=1.7 * fig_scale}
 subclass.cols <- c("frac_clusters_SA1","frac_clusters_SA2",
@@
 
-```{r FigS4X2, fig.width=2.25 * fig_scale, fig.height=2 * fig_scale}
+```{r FigS4H, fig.width=2.25 * fig_scale, fig.height=2 * fig_scale}
 tmp <- FB.data.pooled %>%
@@
 
-project new data into the old PCA
+project test data into the training data PCA
 
@@
   scale_shape_manual(values=c('testing'=15,'training'=16)) +
-  guides(color=guide_legend(override.aes = list(shape=15,size=2),ncol=3)) +
-  theme(legend.key.height=unit(2,'mm'),
+  guides(color=guide_legend(override.aes = list(shape=15,size=2), ncol=3),
+         shape=guide_legend(ncol=1),
+         alpha=guide_legend(ncol=1)) +
+  theme(legend.key.height=unit(1.5,'mm'),
+        legend.key.spacing=unit(0,'mm'),
         aspect.ratio=pca$sdev[2]**2/pca$sdev[1]**2,
@@
 
-```{r Fig4G, fig.width=2.25 * fig_scale, fig.height=2 * fig_scale}
+project validation data into training PCA
+
+```{r Fig4F, fig.width=2.25 * fig_scale, fig.height=2 * fig_scale}
 ggplot(pcadat.combined %>% dplyr::filter(dataset!='testing'),
@@
   scale_shape_manual(values=c('training'=16,'validation'=15)) +
-  guides(color=guide_legend(override.aes = list(shape=15,size=2), ncol=3)) +
-  theme(legend.key.height=unit(2,'mm'),
+  guides(color=guide_legend(override.aes = list(shape=15,size=2), ncol=3),
+         shape=guide_legend(ncol=1),
+         alpha=guide_legend(ncol=1)) +
+  theme(legend.key.height=unit(1.5,'mm'),
+        legend.key.spacing=unit(0,'mm'),
         aspect.ratio=pca$sdev[2]**2/pca$sdev[1]**2,
@@
                      thresholds=c(ROC_1$thresholds,ROC_2$thresholds),
-                     model=c(rep(paste0('training -> testing; AUC=',signif(auc(ROC_1),3)),length(ROC_1$specificities)),
-                             rep(paste0('training -> validation; AUC=',signif(auc(ROC_2),3)),length(ROC_2$specificities)))) %>%
+                     model=c(rep(paste0('training -> testing; AUC=',signif(ROC_1$auc,3)),length(ROC_1$specificities)),
+                             rep(paste0('training -> validation; AUC=',signif(ROC_2$auc,3)),length(ROC_2$specificities)))) %>%
   dplyr::mutate(model=factor(model, levels=unique(model))) %>%
@@
 
-```{r FigS4X3, fig.width=.8 * fig_scale, fig.height=1 * fig_scale}
+```{r FigS4X1, fig.width=.8 * fig_scale, fig.height=1 * fig_scale, include=FALSE}
 set.seed(4)
@@
 
+(average accuracy is `r mean(unlist(sample.CV))`)
+
 heatmap
 
-```{r Fig4F, fig.width=5.5 * fig_scale, fig.height=5.5 * fig_scale}
+```{r Fig4G, fig.width=5.5 * fig_scale, fig.height=5.5 * fig_scale}
 FB.samples <- row.names(FB.data.pooled)[grepl('FB',FB.data.pooled$donor)]
@@
 
-```{r FigS4C, fig.width=3 * fig_scale, fig.height=4 * fig_scale}
-tmp <- data.frame(abs(pca$rotation)) %>%
+```{r FigS4D, fig.width=3 * fig_scale, fig.height=4 * fig_scale}
+tmp <- data.frame(pca$rotation) %>%
   tibble::rownames_to_column('col') %>%
@@
   dplyr::group_by(category) %>%
-  dplyr::mutate(var_ind=1:n())
-
-ggplot(tmp,aes(x=PC1,y=PC2, color=col, shape=col)) +
+  dplyr::mutate(var_ind=1:n()) %>%
+  dplyr::mutate(col=factor(col, levels=levels(tmp$col))) %>% 
+  dplyr::arrange(col)
+
+ggplot(tmp,aes(x=abs(PC1),y=abs(PC2), color=col, shape=col)) +
   geom_point(size=1) +
+  #geom_text_repel(aes(label=col), size=1) +
   my_theme() +
@@
                        dplyr::pull(col_name)) +
-  theme(legend.key.height=unit(3,'mm'),
+  theme(legend.key.size=unit(1.5,'mm'),
         aspect.ratio=pca$sdev[2]**2/pca$sdev[1]**2,
@@
        color='',shape='',fill='',alpha='')
+
+# ggplot(tmp,aes(x=PC1,y=PC2, color=col, shape=col)) +
+#   geom_point(size=1) +
+#   #geom_text_repel(aes(label=col), size=1) +
+#   my_theme() +
+#   scale_color_manual(values=cols_shapes %>%
+#                        dplyr::pull(var_col,col),
+#                      labels=cols_shapes %>%
+#                        dplyr::pull(col_name)) +
+#   scale_shape_manual(values=cols_shapes %>%
+#                        dplyr::pull(var_ind,col),
+#                      labels=cols_shapes %>%
+#                        dplyr::pull(col_name)) +
+#   theme(legend.key.size=unit(1.5,'mm'),
+#         aspect.ratio=pca$sdev[2]**2/pca$sdev[1]**2,
+#         legend.spacing=unit(0,'mm'),
+#         legend.key.spacing.y=unit(0,'mm'),
+#         legend.position='bottom',
+#         legend.box='horizontal') +
+#   guides(color=guide_legend(ncol=2),
+#          shape=guide_legend(ncol=2)) +
+#   labs(x='PC 1', y='PC 2',
+#        color='',shape='',fill='',alpha='')
 ```
@@
 
-```{r FigS4D, fig.width=2.5 * fig_scale, fig.height=2 * fig_scale}
+```{r FigS4G, fig.width=2.5 * fig_scale, fig.height=2 * fig_scale}
 ggplot(pca_hm %>% tibble::rownames_to_column('name') %>%
@@
   guides(color=guide_legend(override.aes = list(shape=c(rep(16,3),rep(15,7)),size=2), ncol=4)) +
-  theme(legend.key.height=unit(3,'mm'),
+  theme(legend.key.size=unit(1.5,'mm'),
         legend.position='bottom',
@@
        color='')
-```
-
-what drives clustering within clusters 1 + 2?
-
-```{r cluster_drivers, fig.width=3 * fig_scale, fig.height=3 * fig_scale}
-use <- class_df$class!='cluster3'
-cluster.var <- fitExtractVarPartModel(t(FB.X[use,]), ~ (1 | class), data=class_df[use,'class',drop=FALSE]) %>%
-  as.data.frame() %>%
-  tibble::rownames_to_column('col') %>%
-  gather(variable,variance,-col)
-
-cluster.var %>%
-  dplyr::filter(variable=='class') %>%
-  dplyr::left_join(columns %>% dplyr::mutate(col=gsub('_downsampled|_switch','',col)) %>% 
-                     dplyr::filter(category %in% c('context','breaks','isotypes','diversity','structural')), 
-                   by='col') %>%
-  dplyr::arrange(variance) %>%
-  dplyr::mutate(col=factor(col, levels=col)) %>%
-  ggplot(aes(y=variance, x=col, fill=category)) + 
-  geom_col() +
-  my_theme() +
-  coord_flip() +
-  scale_fill_manual(values=cat_cols) +
-  labs(x='',y='explained variance between cluster1+2 by class') +
-  theme(legend.position=c(.9,.1))
 ```
@@
                                       c('cluster1','cluster2'),
-                                      c('cluster2','cluster3'),
-                                      c('control','ATM'),
+                                      c('cluster2','cluster3')),
+                     method='wilcox.test',label = "p.signif", hide.ns=TRUE, size=2, vjust=.5) +
+  stat_compare_means(comparisons=list(c('control','ATM'),
                                       c('control','BRCA1')),
@@
 
-```{r FigS4X4, fig.width=3 * fig_scale, fig.height=2.5 * fig_scale}
+```{r FigS4E, fig.width=3 * fig_scale, fig.height=2.5 * fig_scale}
 pvals <- list()
@@
     form <- as.formula(paste0(x,'=="negative" ~ ',f))
-    res <- glm(form, family='binomial', data=data) %>% summary()
-    pvals[[paste0(x,'_',f)]] <- data.frame(pval=coef(res)[2,4],
-                                           vacc=x,
-                                           col=f)
-  }
-}
-
-for (x in c("splenomegaly",
-            "lymphoproliferation",
-            "lung")) {
-  for (f in selected.cols) {
-    data <- FB.data.pooled[FB.samples,c(x,f,'status')] %>%
-      dplyr::filter(status!='control') %>%
-      tidyr::drop_na()
-    form <- as.formula(paste0(x,'!="" ~ ',f))
     res <- glm(form, family='binomial', data=data) %>% summary()
@@
 
-FB.data.pooled[FB.samples, ] %>%
-  gather(feature, value, c('frac_clusters_SG3','alpha_ratio_clusters')) %>%
-  gather(disease, disease_status, c('splenomegaly','lymphoproliferation','lung')) %>%
-  dplyr::mutate(disease_status=factor(ifelse(status=='control','control',
-                                             ifelse(disease_status=='','absent','present')),
-                                      levels=c('control','absent','present'))) %>%
-  ggplot(aes(x=disease_status, y=value, fill=disease_status)) +
-  geom_boxplot(outlier.shape=NA) +
-  geom_jitter(aes(color=status),width=.1, size=.5) +
-  stat_compare_means(comparisons=list(c('absent','present'),
-                                      c('absent','control')),
-                     method='wilcox.test', label = "p.signif", hide.ns=TRUE, size=2, vjust=0) +
-  facet_grid(feature ~ disease, switch='y', scales='free_y') + 
-  my_theme() +
-  scale_fill_manual(values=c('present'='red3','absent'='gray')) +
-  scale_color_manual(values=diagnosis_colors) +
-  labs(x='',y='') +
-  coord_cartesian(clip='off') +
-  theme(legend.position='none',
-        strip.clip='off',
-        axis.text.x=element_text(angle=45, hjust=1, vjust=1))
+# for (x in c("splenomegaly",
+#             "lymphoproliferation",
+#             "lung")) {
+#   for (f in selected.cols) {
+#     data <- FB.data.pooled[FB.samples,c(x,f,'status')] %>%
+#       dplyr::filter(status!='control') %>%
+#       tidyr::drop_na()
+#     form <- as.formula(paste0(x,'!="" ~ ',f))
+#     res <- glm(form, family='binomial', data=data) %>% summary()
+#     pvals[[paste0(x,'_',f)]] <- data.frame(pval=coef(res)[2,4],
+#                                            vacc=x,
+#                                            col=f)
+#   }
+# }
+
+# FB.data.pooled[FB.samples, ] %>%
+#   gather(feature, value, c('frac_clusters_SG3','alpha_ratio_clusters')) %>%
+#   gather(disease, disease_status, c('splenomegaly','lymphoproliferation','lung')) %>%
+#   dplyr::mutate(disease_status=factor(ifelse(status=='control','control',
+#                                              ifelse(disease_status=='','absent','present')),
+#                                       levels=c('control','absent','present'))) %>%
+#   ggplot(aes(x=disease_status, y=value, fill=disease_status)) +
+#   geom_boxplot(outlier.shape=NA) +
+#   geom_jitter(aes(color=status),width=.1, size=.5) +
+#   stat_compare_means(comparisons=list(c('absent','present'),
+#                                       c('absent','control')),
+#                      method='wilcox.test', label = "p.signif", hide.ns=TRUE, size=2, vjust=0) +
+#   facet_grid(feature ~ disease, switch='y', scales='free_y') + 
+#   my_theme() +
+#   scale_fill_manual(values=c('present'='red3','absent'='gray')) +
+#   scale_color_manual(values=diagnosis_colors) +
+#   labs(x='',y='') +
+#   coord_cartesian(clip='off') +
+#   theme(legend.position='none',
+#         strip.clip='off',
+#         axis.text.x=element_text(angle=45, hjust=1, vjust=1))
 ```
@@
 
-```{r FigS4X5, fig.width=5.5 * fig_scale, fig.height=3.5 * fig_scale}
+```{r FigS4F, fig.width=5.5 * fig_scale, fig.height=3.5 * fig_scale}
 FB.X.aggregated <-  FB.data.pooled[FB.samples, selected.cols] %>%
@@
 
+<!-- ```{r FigS4X2, fig.width=4 * fig_scale, fig.height=1.75 * fig_scale, include=FALSE, eval=FALSE} -->
+<!-- all.cols <- intersect(colnames(CH12.X), colnames(human.X)) -->
+<!-- dpca <- dualPCA(CH12.X[,all.cols], human.X[,all.cols]) -->
+
+<!-- dpcadat <- cbind(rbind(dpca$x,dpca$y), -->
+<!--                 data.frame(species=c(rep('mouse',dim(dpca$x)[1]), -->
+<!--                                      rep('human',dim(dpca$y)[1])), -->
+<!--                            genotype=c(CH12.meta[,'genotype'], -->
+<!--                                       human.meta[,'diagnosis']))) -->
+
+<!-- ggplot(dpcadat,aes(x=dPC1,y=dPC2, color=genotype)) + -->
+<!--   geom_point(aes(shape=species),size=.5) + -->
+<!--   geom_polygon(data = dpcadat %>% -->
+<!--                  dplyr::group_by(genotype) %>% -->
+<!--                  dplyr::slice(chull(dPC1, dPC2)), alpha = 0.1, aes(fill=genotype), linewidth=.1) + -->
+<!--   my_theme() + -->
+<!--   scale_fill_manual(values=c(diagnosis_colors,genotype_colors)) + -->
+<!--   scale_color_manual(values=c(diagnosis_colors,genotype_colors)) + -->
+<!--   scale_shape_manual(values=c('human'=15,'mouse'=16)) + -->
+<!--   guides(fill=guide_legend(ncol=2, override.aes = list(pch=c(rep(16,5),rep(15,14)))), -->
+<!--          color=guide_legend(ncol=2, override.aes = list(pch=c(rep(16,5),rep(15,14)))), -->
+<!--          shape='none') + -->
+<!--   labs(x=paste0('dPC 1 (',100*signif(dpca$sdev[1]**2/sum(dpca$sdev**2),3),'%)'), -->
+<!--        y=paste0('dPC 2 (',100*signif(dpca$sdev[2]**2/sum(dpca$sdev**2),3),'%)'), -->
+<!--        color='',shape='',fill='') -->
+
+<!-- ggplot(dpcadat,aes(x=dPC1,y=dPC2, color=genotype)) + -->
+<!--   geom_point(aes(shape=species),size=.5) + -->
+<!--   geom_polygon(data = dpcadat %>% -->
+<!--                  dplyr::group_by(genotype) %>% -->
+<!--                  dplyr::slice(chull(dPC1, dPC2)), alpha = 0.1, aes(fill=genotype), linewidth=.1) + -->
+<!--   my_theme() + -->
+<!--   facet_wrap(~species, ncol=2) + -->
+<!--   scale_fill_manual(values=c(diagnosis_colors,genotype_colors)) + -->
+<!--   scale_color_manual(values=c(diagnosis_colors,genotype_colors)) + -->
+<!--   scale_shape_manual(values=c('human'=15,'mouse'=16)) + -->
+<!--   theme(legend.key.size=unit(1.5,'mm'), -->
+<!--         aspect.ratio=dpca$sdev[2]**2/dpca$sdev[1]**2) + -->
+<!--   guides(fill=guide_legend(ncol=2, override.aes = list(pch=c(rep(16,5),rep(15,14)))), -->
+<!--          color=guide_legend(ncol=2, override.aes = list(pch=c(rep(16,5),rep(15,14)))), -->
+<!--          shape='none') + -->
+<!--   labs(x=paste0('dPC 1 (',100*signif(dpca$sdev[1]**2/sum(dpca$sdev**2),3),'%)'), -->
+<!--        y=paste0('dPC 2 (',100*signif(dpca$sdev[2]**2/sum(dpca$sdev**2),3),'%)'), -->
+<!--        color='',shape='',fill='') -->
+
+<!-- dpcadat[,'group'] <- as.character(dpcadat$genotype) -->
+<!-- dpcadat[FB.meta$sample_donor,'group'] <- as.character(class_df$class) -->
+<!-- dpcadat[,'dataset'] <- 'CH12' -->
+<!-- dpcadat[dpcadat$species=='human','dataset'] <- human.meta$dataset -->
+<!-- ggplot(dpcadat,# %>% dplyr::filter(dataset!='validation'), -->
+<!--        aes(x=dPC1,y=dPC2, color=group)) + -->
+<!--   geom_point(aes(shape=species),size=.5) + -->
+<!--   geom_polygon(data = dpcadat %>% -->
+<!--                  #dplyr::filter(dataset!='validation') %>% -->
+<!--                  dplyr::group_by(group) %>% -->
+<!--                  dplyr::slice(chull(dPC1, dPC2)), alpha = 0.1, aes(fill=group), linewidth=.1) + -->
+<!--   my_theme() + -->
+<!--   facet_wrap(~species, ncol=2) + -->
+<!--   scale_fill_manual(values=c(diagnosis_colors,genotype_colors,class_colors)) + -->
+<!--   scale_color_manual(values=c(diagnosis_colors,genotype_colors,class_colors)) + -->
+<!--   scale_shape_manual(values=c('human'=15,'mouse'=16)) + -->
+<!--   theme(legend.key.size=unit(1.5,'mm'), -->
+<!--         aspect.ratio=dpca$sdev[2]**2/dpca$sdev[1]**2) + -->
+<!--   guides(fill=guide_legend(ncol=2), #, override.aes = list(pch=c(rep(16,5),rep(15,10)))), -->
+<!--          color=guide_legend(ncol=2), #, override.aes = list(pch=c(rep(16,5),rep(15,10)))), -->
+<!--          shape='none') + -->
+<!--   labs(x=paste0('dPC 1 (',100*signif(dpca$sdev[1]**2/sum(dpca$sdev**2),3),'%)'), -->
+<!--        y=paste0('dPC 2 (',100*signif(dpca$sdev[2]**2/sum(dpca$sdev**2),3),'%)'), -->
+<!--        color='',shape='',fill='') -->
+<!-- ``` -->
+
+<!-- ```{r FigS4X3, fig.width=3 * fig_scale, fig.height=4 * fig_scale} -->
+<!-- tmp <- t(t(t(as.matrix(rbind(CH12.X[,all.cols], human.X[,all.cols]))) %*% as.matrix(dpcadat[,c('dPC1','dPC2')])) / colSums(dpcadat[,c('dPC1','dPC2')]**2)) %>% -->
+<!--   as.data.frame() %>% -->
+<!--   tibble::rownames_to_column('col') %>% -->
+<!--   dplyr::left_join(columns, by='col') %>% -->
+<!--   dplyr::mutate(col=factor(col, levels=columns %>%  -->
+<!--                              dplyr::filter(col %in% all.cols) %>% -->
+<!--                              dplyr::pull(col)), -->
+<!--                 col_name=gsub("_downsampled|_switch",'',col)) -->
+
+<!-- cols_shapes <- tmp %>%  -->
+<!--   dplyr::distinct(category, col, col_name) %>% -->
+<!--   dplyr::mutate(var_col=cat_cols[as.character(category)]) %>% -->
+<!--   dplyr::group_by(category) %>% -->
+<!--   dplyr::mutate(var_ind=1:n()) %>% -->
+<!--   dplyr::mutate(col=factor(col, levels=levels(tmp$col))) %>%  -->
+<!--   dplyr::arrange(col) -->
+
+<!-- ggplot(tmp,aes(x=dPC1,y=dPC2, color=col, shape=col)) + -->
+<!--   geom_point(size=1) + -->
+<!--   #geom_text_repel(aes(label=col), size=1) + -->
+<!--   my_theme() + -->
+<!--   scale_color_manual(values=cols_shapes %>% -->
+<!--                        dplyr::pull(var_col,col), -->
+<!--                      labels=cols_shapes %>% -->
+<!--                        dplyr::pull(col_name)) + -->
+<!--   scale_shape_manual(values=cols_shapes %>% -->
+<!--                        dplyr::pull(var_ind,col), -->
+<!--                      labels=cols_shapes %>% -->
+<!--                        dplyr::pull(col_name)) + -->
+<!--   theme(legend.key.size=unit(1.5,'mm'), -->
+<!--         aspect.ratio=dpca$sdev[2]**2/dpca$sdev[1]**2, -->
+<!--         legend.spacing=unit(0,'mm'), -->
+<!--         legend.key.spacing.y=unit(0,'mm'), -->
+<!--         legend.position='bottom', -->
+<!--         legend.box='horizontal') + -->
+<!--   guides(color=guide_legend(ncol=2), -->
+<!--          shape=guide_legend(ncol=2)) + -->
+<!--   labs(x='dPC 1', y='dPC 2', -->
+<!--        color='',shape='',fill='',alpha='') -->
+<!-- ``` -->
+
+<!-- ```{r FigS4X4, fig.width=6.75 * fig_scale, fig.height=3.5 * fig_scale} -->
+<!-- features.here <- c('cluster_entropy_downsampled', -->
+<!--                    'top_clone_occupancy_downsampled', -->
+<!--                    'donor_complexity', -->
+<!--                    'donor_score_W', -->
+<!--                    'donor_score_WGCW', -->
+<!--                    'homology_fw', -->
+<!--                    'n_homology_switch', -->
+<!--                    'mean_length_SA', -->
+<!--                    'frac_breaks_intra', -->
+<!--                    'frac_breaks_inversions') -->
+
+<!-- tmp <- cbind(rbind(CH12.data[,features.here], -->
+<!--                    FB.data.pooled[,features.here], -->
+<!--                    QP.data[,features.here]), -->
+<!--              rbind(CH12.meta %>%  -->
+<!--                      dplyr::mutate(diagnosis=genotype, -->
+<!--                                    dataset='CH12') %>% -->
+<!--                      dplyr::select(diagnosis, dataset), -->
+<!--                    human.meta[,c('diagnosis','dataset')])) %>% -->
+<!--   dplyr::filter(diagnosis %in% c('WT','Trp53bp1 KO','Brca1 KO','Rif1 KO','Lig4 KO','control','CVID','TNFRSF13B','ATM','BRCA1')) %>% -->
+<!--   tibble::rownames_to_column('sample') %>% -->
+<!--   gather(col,value,features.here) %>% -->
+<!--   dplyr::mutate(col=factor(gsub("_downsampled|_switch",'',col), -->
+<!--                            levels=gsub("_downsampled|_switch","",features.here)), -->
+<!--                 group=droplevels(factor(diagnosis, levels=c(names(genotype_colors), names(diagnosis_colors))))) -->
+
+<!-- ggplot(tmp, aes(x=ifelse(dataset=='CH12',as.numeric(group),as.numeric(group) + 1),y=value,fill=group)) + -->
+<!--   geom_boxplot(outlier.shape=NA, lwd=.5) + -->
+<!--   geom_jitter(width=.1,height=0,size=.1) + -->
+<!--   stat_compare_means(comparisons=list(c(1,2), -->
+<!--                                       c(1,3), -->
+<!--                                       c(1,4), -->
+<!--                                       c(1,5)), -->
+<!--                      method='wilcox.test',label = "p.signif", hide.ns=TRUE, size=2, vjust=.5) + -->
+<!--   stat_compare_means(comparisons=list(c(7,8), -->
+<!--                                       c(7,9), -->
+<!--                                       c(7,10), -->
+<!--                                       c(7,11)), -->
+<!--                      method='wilcox.test',label = "p.signif", hide.ns=TRUE, size=2, vjust=.5) + -->
+<!--   facet_wrap(.~col,ncol=5,scales='free_y') + -->
+<!--   my_theme() + -->
+<!--   coord_cartesian(clip='off') + -->
+<!--   scale_fill_manual(values=c(diagnosis_colors,genotype_colors)) + -->
+<!--   theme(axis.text.x=element_text(angle=90,hjust=1,vjust=.5), -->
+<!--         legend.position='none') + -->
+<!--   scale_x_continuous(breaks=c(seq(1,5), seq(1,5) + 6), -->
+<!--                      labels=levels(tmp$group)) + -->
+<!--   labs(x='',y='', color='diagnosis') -->
+<!-- ``` -->
+
+<!-- ```{r FigS4X5, fig.width=3.5 * fig_scale, fig.height=3 * fig_scale} -->
+<!-- OC_sample_sheet <- read.csv('../sodar/2023_OC/a_2023_OC.txt', -->
+<!--                             sep='\t',header=1) %>% -->
+<!--   dplyr::left_join(read.csv('../sodar/2023_OC/s_2023_OC.txt',sep='\t',header=1), -->
+<!--                    by='Sample.Name') %>% -->
+<!--   dplyr::mutate(donor=as.character(Source.Name), -->
+<!--                 sample=Extract.Name.1, -->
+<!--                 batch=Parameter.Value.Batch., -->
+<!--                 QC=Parameter.Value.QC., -->
+<!--                 sequencing=Parameter.Value.Sequencing.method.) %>% -->
+<!--   dplyr::distinct(donor) -->
+
+<!-- OC.data <-read.csv('../human_samples/20240307_hg38_results/output/summary/20240307_human_samples_stats.csv', -->
+<!--                    header=1,row.names=1) %>% -->
+<!--   tibble::rownames_to_column('sample') %>% -->
+<!--   dplyr::filter(grepl('pooled_',sample)) %>% -->
+<!--   dplyr::mutate(sample_donor=gsub('pooled_','',sample)) %>% -->
+<!--   dplyr::inner_join(OC_sample_sheet, by=c('sample_donor'='donor')) %>% -->
+<!--   dplyr::filter(nreads_final > 500) %>% -->
+<!--   dplyr::mutate_all(~replace_na(.,0)) %>% -->
+<!--   dplyr::mutate(log10_nreads=log10(nreads_final), -->
+<!--                 log10_nclusters=log10(nclusters_final)) %>% -->
+<!--   tibble::column_to_rownames('sample_donor') -->
+
+<!-- OC.meta <- OC.data %>% -->
+<!--   tibble::rownames_to_column('donor') %>% -->
+<!--   dplyr::mutate(status=ifelse(grepl('BIH',sample),'OC','control'), -->
+<!--                 sample_donor=donor, -->
+<!--                 diagnosis=ifelse(donor %in% c('BIH116','BIH141','BIH152'), 'BRCA1', status), -->
+<!--                 timepoint='t1', -->
+<!--                 dataset='OC', -->
+<!--                 sex=NA, -->
+<!--                 age=NA, -->
+<!--                 complication=NA, -->
+<!--                 EUROclass=NA, -->
+<!--                 pct_B=NA, -->
+<!--                 pct_IgA=NA, -->
+<!--                 pct_IgG=NA) %>% -->
+<!--   dplyr::select(sample_donor,donor,diagnosis,status,timepoint,dataset,sex,age,complication,EUROclass,pct_B,pct_IgA,pct_IgG) %>% -->
+<!--   tibble::column_to_rownames('sample_donor') -->
+
+<!-- OC.X <- OC.data[row.names(OC.meta), selected.cols] %>% -->
+<!--   dplyr::mutate_if(is.numeric, scale) -->
+
+<!-- pcadat.all <- rbind(cbind(as.matrix(human.X[train,]) %*% pca$rotation, human.meta[train,]), -->
+<!--                     cbind(as.matrix(human.X[test,]) %*% pca$rotation, human.meta[test,]), -->
+<!--                     cbind(as.matrix(human.X[validation,]) %*% pca$rotation, human.meta[validation,]), -->
+<!--                     cbind(as.matrix(OC.X) %*% pca$rotation, OC.meta)) %>% -->
+<!--   dplyr::mutate(dataset=factor(dataset, levels=c('training','testing','validation','OC'))) -->
+
+<!-- ggplot(pcadat.all, aes(x=PC1,y=PC2)) + -->
+<!--   geom_polygon(data = pcadat.all %>% dplyr::filter(dataset=='training') %>% -->
+<!--                  dplyr::group_by(status) %>% -->
+<!--                  dplyr::slice(chull(PC1, PC2)), alpha = 0.1, color=NA, aes(fill=status), show.legend=FALSE) + -->
+<!--   geom_point(aes(color=diagnosis,alpha=dataset, shape=dataset), size=1) + -->
+<!--   geom_text_repel(data=pcadat.all %>% dplyr::filter(dataset=='OC'), aes(label=donor,color=diagnosis), size=2, segment.size=.25) + -->
+<!--   geom_line(aes(group=donor, color=diagnosis,alpha=dataset, shape=dataset)) + -->
+<!--   my_theme() + -->
+<!--   scale_fill_manual(values=c(diagnosis_colors,'OC'='deeppink'), guide='none') + -->
+<!--   scale_color_manual(values=c(diagnosis_colors,'OC'='deeppink')) + -->
+<!--   scale_alpha_manual(values=c('testing'=.25,'training'=.25,'validation'=.5,'OC'=1)) + -->
+<!--   scale_shape_manual(values=c('testing'=15,'training'=16,'validation'=17,'OC'=18)) + -->
+<!--   guides(color=guide_legend(override.aes = list(shape=15,size=2),ncol=3), -->
+<!--          shape=guide_legend(ncol=1), -->
+<!--          alpha=guide_legend(ncol=1)) + -->
+<!--   theme(legend.key.height=unit(1.5,'mm'), -->
+<!--         legend.key.spacing=unit(0,'mm'), -->
+<!--         aspect.ratio=pca$sdev[2]**2/pca$sdev[1]**2, -->
+<!--         legend.position='bottom', -->
+<!--         legend.box='horizontal') + -->
+<!--   coord_cartesian(xlim=c(-8,10), -->
+<!--                   ylim=c(-5,8), -->
+<!--                   clip='off') + -->
+<!--   labs(x=paste0('PC 1 (',100*signif(pca$sdev[1]**2/sum(pca$sdev**2),3),'%)'), -->
+<!--        y=paste0('PC 2 (',100*signif(pca$sdev[2]**2/sum(pca$sdev**2),3),'%)'), -->
+<!--        color='',shape='',fill='',alpha='') -->
+
+<!-- ``` -->
+
+<!-- ```{r FigS4X6, fig.width=3.5 * fig_scale, fig.height=3 * fig_scale} -->
+<!-- dpca <- dualPCA(human.X, OC.X) -->
+<!-- dpcadat <- cbind(rbind(dpca$x,dpca$y), -->
+<!--                  pcadat.all) -->
+
+<!-- ggplot(dpcadat, aes(x=dPC1,y=dPC2)) + -->
+<!--   geom_polygon(data = dpcadat %>% dplyr::filter(dataset=='training') %>% -->
+<!--                  dplyr::group_by(status) %>% -->
+<!--                  dplyr::slice(chull(dPC1, dPC2)), alpha = 0.1, color=NA, aes(fill=status), show.legend=FALSE) + -->
+<!--   geom_point(aes(color=diagnosis,alpha=dataset, shape=dataset), size=1) + -->
+<!--   geom_line(aes(group=donor, color=diagnosis,alpha=dataset, shape=dataset)) + -->
+<!--   geom_text_repel(data=dpcadat %>% dplyr::filter(dataset=='OC'), aes(label=donor, color=diagnosis), size=2, segment.size=.25) + -->
+<!--   my_theme() + -->
+<!--   scale_fill_manual(values=c(diagnosis_colors,'OC'='deeppink'), guide='none') + -->
+<!--   scale_color_manual(values=c(diagnosis_colors,'OC'='deeppink')) + -->
+<!--   scale_alpha_manual(values=c('testing'=.25,'training'=.25,'validation'=.5,'OC'=1)) + -->
+<!--   scale_shape_manual(values=c('testing'=15,'training'=16,'validation'=17,'OC'=18)) + -->
+<!--   guides(color=guide_legend(override.aes = list(shape=15,size=2),ncol=3), -->
+<!--          shape=guide_legend(ncol=1), -->
+<!--          alpha=guide_legend(ncol=1)) + -->
+<!--   theme(legend.key.height=unit(1.5,'mm'), -->
+<!--         legend.key.spacing=unit(0,'mm'), -->
+<!--         #aspect.ratio=dpca$sdev[2]**2/dpca$sdev[1]**2, -->
+<!--         legend.position='bottom', -->
+<!--         legend.box='horizontal') + -->
+<!--   coord_cartesian(clip='off') + -->
+<!--   labs(x=paste0('dPC 1 (',100*signif(dpca$sdev[1]**2/sum(dpca$sdev**2),3),'%)'), -->
+<!--        y=paste0('dPC 2 (',100*signif(dpca$sdev[2]**2/sum(dpca$sdev**2),3),'%)'), -->
+<!--        color='',shape='',fill='',alpha='') -->
+
+<!-- ``` -->
+
+<!-- ```{r FigS4X7, fig.width=3.5 * fig_scale, fig.height=3 * fig_scale} -->
+<!-- LB_sample_sheet <- read.csv('../sodar/2023_LB/a_2023_LB.txt', -->
+<!--                             sep='\t',header=1) %>% -->
+<!--   dplyr::left_join(read.csv('../sodar/2023_LB/s_2023_LB.txt',sep='\t',header=1), -->
+<!--                    by='Sample.Name') %>% -->
+<!--   dplyr::mutate(donor=as.character(Source.Name), -->
+<!--                 sample=Extract.Name.1, -->
+<!--                 batch=Parameter.Value.Batch., -->
+<!--                 QC=Parameter.Value.QC., -->
+<!--                 diagnosis=Characteristics.Diagnosis., -->
+<!--                 sequencing=Parameter.Value.Sequencing.method.) %>% -->
+<!--   dplyr::distinct(donor,diagnosis) -->
+
+<!-- LB.data <- read.csv('../human_samples/20240307_hg38_results/output/summary/20240307_human_samples_stats.csv', -->
+<!--                    header=1,row.names=1) %>% -->
+<!--   tibble::rownames_to_column('sample') %>% -->
+<!--   dplyr::filter(grepl('pooled_',sample)) %>% -->
+<!--   dplyr::mutate(sample_donor=gsub('pooled_','',sample)) %>% -->
+<!--   dplyr::inner_join(LB_sample_sheet, by=c('sample_donor'='donor')) %>% -->
+<!--   dplyr::filter(nreads_final > 500) %>% -->
+<!--   dplyr::mutate_all(~replace_na(.,0)) %>% -->
+<!--   dplyr::mutate(log10_nreads=log10(nreads_final), -->
+<!--                 log10_nclusters=log10(nclusters_final)) %>% -->
+<!--   tibble::column_to_rownames('sample_donor') -->
+
+<!-- LB.meta <- LB.data %>% -->
+<!--   tibble::rownames_to_column('donor') %>% -->
+<!--   dplyr::mutate(sample_donor=donor, -->
+<!--                 status=diagnosis, -->
+<!--                 timepoint='t1', -->
+<!--                 dataset='LB', -->
+<!--                 sex=NA, -->
+<!--                 age=NA, -->
+<!--                 complication=NA, -->
+<!--                 EUROclass=NA, -->
+<!--                 pct_B=NA, -->
+<!--                 pct_IgA=NA, -->
+<!--                 pct_IgG=NA) %>% -->
+<!--   dplyr::select(sample_donor,donor,diagnosis,status,timepoint,dataset,sex,age,complication,EUROclass,pct_B,pct_IgA,pct_IgG) %>% -->
+<!--   tibble::column_to_rownames('sample_donor') -->
+
+<!-- LB.X <- LB.data[row.names(LB.meta), selected.cols] %>% -->
+<!--   dplyr::mutate_if(is.numeric, scale) -->
+
+<!-- pcadat.all <- rbind(cbind(as.matrix(human.X[train,]) %*% pca$rotation, human.meta[train,]), -->
+<!--                     cbind(as.matrix(human.X[test,]) %*% pca$rotation, human.meta[test,]), -->
+<!--                     cbind(as.matrix(human.X[validation,]) %*% pca$rotation, human.meta[validation,]), -->
+<!--                     cbind(as.matrix(LB.X) %*% pca$rotation, LB.meta)) %>% -->
+<!--   dplyr::mutate(dataset=factor(dataset, levels=c('training','testing','validation','LB'))) -->
+
+<!-- ggplot(pcadat.all, aes(x=PC1,y=PC2)) + -->
+<!--   geom_polygon(data = pcadat.all %>% dplyr::filter(dataset=='training') %>% -->
+<!--                  dplyr::group_by(status) %>% -->
+<!--                  dplyr::slice(chull(PC1, PC2)), alpha = 0.1, color=NA, aes(fill=status), show.legend=FALSE) + -->
+<!--   geom_point(aes(color=diagnosis,alpha=dataset, shape=dataset), size=1) + -->
+<!--   geom_text_repel(data=pcadat.all %>% dplyr::filter(dataset=='LB'), aes(label=donor,color=diagnosis), size=2, segment.size=.25) + -->
+<!--   geom_line(aes(group=donor, color=diagnosis,alpha=dataset, shape=dataset)) + -->
+<!--   my_theme() + -->
+<!--   scale_fill_manual(values=c(diagnosis_colors,'unknown'='gray'), guide='none') + -->
+<!--   scale_color_manual(values=c(diagnosis_colors,'unknown'='gray')) + -->
+<!--   scale_alpha_manual(values=c('testing'=.25,'training'=.25,'validation'=.5,'LB'=1)) + -->
+<!--   scale_shape_manual(values=c('testing'=15,'training'=16,'validation'=17,'LB'=18)) + -->
+<!--   guides(color=guide_legend(override.aes = list(shape=15,size=2),ncol=3), -->
+<!--          shape=guide_legend(ncol=1), -->
+<!--          alpha=guide_legend(ncol=1)) + -->
+<!--   theme(legend.key.height=unit(1.5,'mm'), -->
+<!--         legend.key.spacing=unit(0,'mm'), -->
+<!--         aspect.ratio=pca$sdev[2]**2/pca$sdev[1]**2, -->
+<!--         legend.position='bottom', -->
+<!--         legend.box='horizontal') + -->
+<!--   coord_cartesian(xlim=c(-8,10), -->
+<!--                   ylim=c(-5,8), -->
+<!--                   clip='off') + -->
+<!--   labs(x=paste0('PC 1 (',100*signif(pca$sdev[1]**2/sum(pca$sdev**2),3),'%)'), -->
+<!--        y=paste0('PC 2 (',100*signif(pca$sdev[2]**2/sum(pca$sdev**2),3),'%)'), -->
+<!--        color='',shape='',fill='',alpha='') -->
+
+<!-- ``` -->
+
+<!-- ```{r FigS4X8, fig.width=3.5 * fig_scale, fig.height=3 * fig_scale} -->
+<!-- dpca <- dualPCA(human.X, LB.X) -->
+<!-- dpcadat <- cbind(rbind(dpca$x,dpca$y), -->
+<!--                  pcadat.all) -->
+
+<!-- ggplot(dpcadat, aes(x=dPC1,y=dPC2)) + -->
+<!--   geom_polygon(data = dpcadat %>% dplyr::filter(dataset=='training') %>% -->
+<!--                  dplyr::group_by(status) %>% -->
+<!--                  dplyr::slice(chull(dPC1, dPC2)), alpha = 0.1, color=NA, aes(fill=status), show.legend=FALSE) + -->
+<!--   geom_point(aes(color=diagnosis,alpha=dataset, shape=dataset), size=1) + -->
+<!--   geom_line(aes(group=donor, color=diagnosis,alpha=dataset, shape=dataset)) + -->
+<!--   geom_text_repel(data=dpcadat %>% dplyr::filter(dataset=='LB'), aes(label=donor, color=diagnosis), size=2, segment.size=.25) + -->
+<!--   my_theme() + -->
+<!--   scale_fill_manual(values=c(diagnosis_colors,'unknown'='gray'), guide='none') + -->
+<!--   scale_color_manual(values=c(diagnosis_colors,'unknown'='gray')) + -->
+<!--   scale_alpha_manual(values=c('testing'=.25,'training'=.25,'validation'=.5,'LB'=1)) + -->
+<!--   scale_shape_manual(values=c('testing'=15,'training'=16,'validation'=17,'LB'=18)) + -->
+<!--   guides(color=guide_legend(override.aes = list(shape=15,size=2),ncol=3), -->
+<!--          shape=guide_legend(ncol=1), -->
+<!--          alpha=guide_legend(ncol=1)) + -->
+<!--   theme(legend.key.height=unit(1.5,'mm'), -->
+<!--         legend.key.spacing=unit(0,'mm'), -->
+<!--         #aspect.ratio=dpca$sdev[2]**2/dpca$sdev[1]**2, -->
+<!--         legend.position='bottom', -->
+<!--         legend.box='horizontal') + -->
+<!--   coord_cartesian(clip='off') + -->
+<!--   labs(x=paste0('dPC 1 (',100*signif(dpca$sdev[1]**2/sum(dpca$sdev**2),3),'%)'), -->
+<!--        y=paste0('dPC 2 (',100*signif(dpca$sdev[2]**2/sum(dpca$sdev**2),3),'%)'), -->
+<!--        color='',shape='',fill='',alpha='') -->
+
+<!-- ``` -->
+
 <!-- ```{r aggregated_FB_boxplots, fig.width=4.5 * fig_scale, fig.height=1.7 * fig_scale} -->
@@
 
-check correlation with measured B cell proportion in FACS
-
-```{r pct_B_cor, fig.width=3 * fig_scale, fig.height=3 * fig_scale}
-FB.data.pooled %>%
-  gather(feature, value, selected.cols) %>%
-  dplyr::group_by(feature) %>%
-  dplyr::summarise(R=cor(value, pct_B, use='na.or.complete')) %>%
-  dplyr::left_join(columns, by=c('feature'='col')) %>%
-  dplyr::mutate(feature=gsub('_downsampled|_switch','',feature)) %>%
-  dplyr::arrange(R) %>%
-  dplyr::mutate(feature=factor(feature, levels=feature)) %>%
-  ggplot(aes(y=R, x=feature, fill=category)) + 
-  geom_col() +
-  my_theme() +
-  coord_flip() +
-  scale_fill_manual(values=cat_cols) +
-  labs(x='',y='correlation with % B cells') +
-  theme(legend.position=c(.2,.9))
-
-FB.data.pooled %>%
-  gather(feature, value, selected.cols) %>%
-  dplyr::group_by(feature) %>%
-  dplyr::summarise(R=cor(value, pct_IgA, use='na.or.complete')) %>%
-  dplyr::left_join(columns, by=c('feature'='col')) %>%
-  dplyr::mutate(feature=gsub('_downsampled|_switch','',feature)) %>%
-  dplyr::arrange(R) %>%
-  dplyr::mutate(feature=factor(feature, levels=feature)) %>%
-  ggplot(aes(y=R, x=feature, fill=category)) + 
-  geom_col() +
-  my_theme() +
-  coord_flip() +
-  scale_fill_manual(values=cat_cols) +
-  labs(x='',y='correlation with % IgA cells') +
-  theme(legend.position=c(.2,.9))
-
-FB.data.pooled %>%
-  gather(feature, value, selected.cols) %>%
-  dplyr::group_by(feature) %>%
-  dplyr::summarise(R=cor(value, pct_IgG, use='na.or.complete')) %>%
-  dplyr::left_join(columns, by=c('feature'='col')) %>%
-  dplyr::mutate(feature=gsub('_downsampled|_switch','',feature)) %>%
-  dplyr::arrange(R) %>%
-  dplyr::mutate(feature=factor(feature, levels=feature)) %>%
-  ggplot(aes(y=R, x=feature, fill=category)) + 
-  geom_col() +
-  my_theme() +
-  coord_flip() +
-  scale_fill_manual(values=cat_cols) +
-  labs(x='',y='correlation with % IgG cells') +
-  theme(legend.position=c(.2,.9))
-
-```
+<!-- check correlation with measured B cell proportion in FACS -->
+
+<!-- ```{r pct_B_cor, fig.width=3 * fig_scale, fig.height=3 * fig_scale} -->
+<!-- FB.data.pooled %>% -->
+<!--   gather(feature, value, selected.cols) %>% -->
+<!--   dplyr::group_by(feature) %>% -->
+<!--   dplyr::summarise(R=cor(value, pct_B, use='na.or.complete')) %>% -->
+<!--   dplyr::left_join(columns, by=c('feature'='col')) %>% -->
+<!--   dplyr::mutate(feature=gsub('_downsampled|_switch','',feature)) %>% -->
+<!--   dplyr::arrange(R) %>% -->
+<!--   dplyr::mutate(feature=factor(feature, levels=feature)) %>% -->
+<!--   ggplot(aes(y=R, x=feature, fill=category)) +  -->
+<!--   geom_col() + -->
+<!--   my_theme() + -->
+<!--   coord_flip() + -->
+<!--   scale_fill_manual(values=cat_cols) + -->
+<!--   labs(x='',y='correlation with % B cells') + -->
+<!--   theme(legend.position=c(.2,.9)) -->
+
+<!-- FB.data.pooled %>% -->
+<!--   gather(feature, value, selected.cols) %>% -->
+<!--   dplyr::group_by(feature) %>% -->
+<!--   dplyr::summarise(R=cor(value, pct_IgA, use='na.or.complete')) %>% -->
+<!--   dplyr::left_join(columns, by=c('feature'='col')) %>% -->
+<!--   dplyr::mutate(feature=gsub('_downsampled|_switch','',feature)) %>% -->
+<!--   dplyr::arrange(R) %>% -->
+<!--   dplyr::mutate(feature=factor(feature, levels=feature)) %>% -->
+<!--   ggplot(aes(y=R, x=feature, fill=category)) +  -->
+<!--   geom_col() + -->
+<!--   my_theme() + -->
+<!--   coord_flip() + -->
+<!--   scale_fill_manual(values=cat_cols) + -->
+<!--   labs(x='',y='correlation with % IgA cells') + -->
+<!--   theme(legend.position=c(.2,.9)) -->
+
+<!-- FB.data.pooled %>% -->
+<!--   gather(feature, value, selected.cols) %>% -->
+<!--   dplyr::group_by(feature) %>% -->
+<!--   dplyr::summarise(R=cor(value, pct_IgG, use='na.or.complete')) %>% -->
+<!--   dplyr::left_join(columns, by=c('feature'='col')) %>% -->
+<!--   dplyr::mutate(feature=gsub('_downsampled|_switch','',feature)) %>% -->
+<!--   dplyr::arrange(R) %>% -->
+<!--   dplyr::mutate(feature=factor(feature, levels=feature)) %>% -->
+<!--   ggplot(aes(y=R, x=feature, fill=category)) +  -->
+<!--   geom_col() + -->
+<!--   my_theme() + -->
+<!--   coord_flip() + -->
+<!--   scale_fill_manual(values=cat_cols) + -->
+<!--   labs(x='',y='correlation with % IgG cells') + -->
+<!--   theme(legend.position=c(.2,.9)) -->
+
+<!-- ``` -->
 

### 2025-01-09 prepare_input.R

@@ -381,6 +381,44 @@ vincendeau.data <- read.csv('../mouse_samples/20240425_vincendeau/output/summary
   dplyr::select(tidyselect::any_of(c('genotype','sample',columns$col))) 
 write.csv(vincendeau.data, 'data/vincendeau_data.csv')
 
+HTGTS.samples <- c("SRR2104731","SRR2104732","SRR2104733","SRR2104734","SRR2104735","SRR2104736","SRR2104737","SRR2104738",
+                   "SRR2104739","SRR2104740","SRR2104741","SRR2104742","SRR2104743","SRR2104744",
+                   "SRR6293456","SRR6293457","SRR6293458","SRR6293459","SRR6293460","SRR6293461","SRR6293462","SRR6293463",
+                   "SRR6293464","SRR6293465","SRR6293466","SRR6293467","SRR6293468","SRR6293469","SRR6293470","SRR6293471",
+                   "SRR6293472","SRR6293473","SRR6293474","SRR6293475","SRR6293476","SRR6293477","SRR6293478","SRR6293479")
+
+HTGTS.meta <- data.frame(celltype=c(rep('spleen',14),rep('CH12',24)),
+                         genotype=c(rep('WT',5),rep('Atm',3),rep('Trp53bp1',6),
+                                    rep('WT',6),rep('Atm',6),rep('Trp53bp1',6),rep('Lig4',6)),
+                         sample=HTGTS.samples) %>%
+  dplyr::mutate(genotype=factor(genotype, levels=c('WT','Atm','Trp53bp1','Lig4')))
+
+HTGTS.data <- read.csv('../mouse_samples/20241204_HTGTS/output/summary/HTGTS_stats.csv',
+                       header=1,row.names=1) %>%
+  tibble::rownames_to_column('sample') %>%
+  dplyr::inner_join(HTGTS.meta, by='sample') %>%
+  dplyr::mutate_all(~replace_na(.,0)) %>%
+  dplyr::select(tidyselect::any_of(c('genotype','celltype','sample',columns$col))) 
+write.csv(HTGTS.data, 'data/HTGTS_data.csv')
+
+HTGTS.homology <- lapply(HTGTS.samples, function(smp) read.csv(paste0('../mouse_samples/20241204_HTGTS/pipeline/',smp,'/',smp,'_cluster_analysis.csv'), header=1, row.names=1) %>%
+                           tibble::rownames_to_column('cluster') %>%
[…]

### 2025-01-09 swibrid_runs/external/HTGTS/get_data.sh

@@ -1,0 +1,9 @@
+#!/bin/bash -e
+
+mkdir -p input tmp
+for acc in SRR21047{31..47} SRR62934{56..63}; do 
+        prefetch ${acc} --output-directory tmp
+	fasterq-dump -p --split-3 -O tmp tmp/${acc}/${acc}.sra
+	bbmerge.sh in1=tmp/${acc}_1.fastq in2=tmp/${acc}_2.fastq out=input/${acc}.fastq.gz
+	python make_info.py -i input/${acc}.fastq.gz -o input/${acc}_info.csv
+done

### 2025-01-09 swibrid_runs/external/HTGTS/make_info.py

@@ -1,0 +1,39 @@
+import os
+import sys
+from Bio import SeqIO
+import pandas as pd
+import gzip
+from argparse import ArgumentParser
+
+parser=ArgumentParser()
+parser.add_argument('-i','--input',dest='input',help="""Input fastq.gz file""")
+parser.add_argument('-o','--output',dest='output',help="""output info.csv file""")
+
+args=parser.parse_args()
+
+info={}
+read_id = ""
+with gzip.open(args.input,'rt') as inf:
+    reads=SeqIO.parse(inf,'fastq')
+    while True:
+        try:
+            read = next(reads)
+        except StopIteration:
+            break
+        if read.id == read_id:
+            info[read.id]['length'] += len(read)
+            info[read.id]['num_mates'] += 1
+        else:
+            info[read.id] = {
+                "length": len(read),
+                "num_mates": 1,
+                "barcodes": "BC01@1-30:+",
+                "primers": "primer_fw@0-20:+;primer_rv@{0}-{1}:+".format(
+                    max(0,len(read) - 20), len(read)
+                ),
+                "internal": "",
+            }
+        read_id = read.id
+
+pd.DataFrame.from_dict(info, orient="index").to_csv(args.output)
+

### 2025-01-09 swibrid_runs/external/vincendeau/make_info.py

@@ -1,0 +1,28 @@
+import os
+import sys
+from Bio import SeqIO
+import pandas as pd
+import gzip
+from argparse import ArgumentParser
+
+parser=ArgumentParser()
+parser.add_argument('-i','--input',dest='input',help="""Input fastq.gz file""")
+parser.add_argument('-o','--output',dest='output',help="""output info.csv file""")
+
+
+args=parser.parse_args()
+
+info={}
+with gzip.open(args.input,'rt') as inf:
+    for read in SeqIO.parse(inf,'fastq'):
+        info[read.id] = {
+            "length": len(read),
+            "barcodes": "BC01@1-30:+",
+            "primers": "primer_fw@0-50:+;primer_rv@{0}-{1}:+".format(
+                len(read) - 50, len(read)
+            ),
+            "internal": "",
+        }
+
+pd.DataFrame.from_dict(info, orient="index").to_csv(args.output)
+

### 2025-01-17 paper_figures.Rmd
capitalization

@@
 fig_scale <- 1.
-columns <- list('diversity'=c('nclusters_final_downsampled',
+columns <- list('Diversity'=c('nclusters_final_downsampled',
                               'nclusters_eff_downsampled',
@@
                               'big_clones_occupancy_downsampled'),
-                'isotypes'=c('alpha_ratio_clusters',
+                'Isotypes'=c('alpha_ratio_clusters',
                              'frac_clusters_SA1','frac_clusters_SA2','frac_clusters_SE',
@@
                              'frac_clusters_Sm','frac_clusters_Sg1'),
-                'structural'=c('insert_frequency',
+                'Structural'=c('insert_frequency',
                                'frac_breaks_inversions','frac_breaks_duplications',
                                'median_duplication_size','median_inversion_size'),
-                'context'=c('homology_fw','homology_rv',
+                'Context'=c('homology_fw','homology_rv',
                             'donor_score_S', 'donor_score_W', 
@@
                             'frac_blunt'),
-                'breaks'=c('mean_length',
+                'Breaks'=c('mean_length',
                            'std_length',
@@
                            'frac_breaks_SG_SA','frac_breaks_SA_SA'),
-                'context_isotype'=c('homology_fw_SM_SA','homology_rv_SM_SA',
+                'Context_isotype'=c('homology_fw_SM_SA','homology_rv_SM_SA',
                                     'homology_fw_SM_SG','homology_rv_SM_SG',
@@
                                     'frac_blunt_SG'),
-                'diversity_raw'=c('nclusters_final',
+                'Diversity_raw'=c('nclusters_final',
                                   'nclusters_eff',
@@
                                   'big_clones_occupancy'),
-                'variants'=c('frac_variants_germline',
+                'Variants'=c('frac_variants_germline',
                              'frac_variants_transitions',
@@
   dplyr::mutate(category=factor(gsub('[0-9]*$','',category), 
-                                levels=c('library',
-                                         'isotypes',
-                                         'diversity',
-                                         'diversity_raw',
-                                         'structural',
-                                         'context',
-                                         'context_isotype',
-                                         'breaks',
-                                         'variants')))
-
-cat_cols <- c('structural'='#FD9E40',
-              'context'='#FED6AC',
-              'context_isotype'='#FED6AC',
-              'breaks'='#1E8AC4',
-              'isotypes'='#AEF58E',
-              'diversity_raw'='#3A7A2C',
-              'diversity'='#3A7A2C',
-              'variants'='lightpink3',
-              'library'='gray40')
+                                levels=c('Library',
+                                         'Isotypes',
+                                         'Diversity',
+                                         'Diversity_raw',
+                                         'Structural',
+                                         'Context',
+                                         'Context_isotype',
+                                         'Breaks',
+                                         'Variants')))
+
+cat_cols <- c('Structural'='#FD9E40',
+              'Context'='#FED6AC',
+              'Context_isotype'='#FED6AC',
+              'Breaks'='#1E8AC4',
+              'Isotypes'='#AEF58E',
+              'Diversity_raw'='#3A7A2C',
+              'Diversity'='#3A7A2C',
+              'Variants'='lightpink3',
+              'Library'='gray40')
 
@@
 
-tissue_colors <- c('splenocytes'='orange2','lymph node'='orangered3','CH12'='forestgreen')
-var_colors <- c('batch'='palegreen2','diagnosis'='tan2','donor'='firebrick4','sequencing'='slategray1','unexplained'='gray80',
-                'sex'='steelblue2','age'='darkseagreen4')
-
-diagnosis_colors <- c('control'='gray20','HD'='gray20',
-                      'subclass'='lightslategray','CVID'='coral1',
+tissue_colors <- c('Splenocytes'='orange2','Lymph Node'='orangered3','CH12'='forestgreen')
+var_colors <- c('Batch'='palegreen2','Diagnosis'='tan2','Donor'='firebrick4','Sequencing'='slategray1','unexplained'='gray80',
+                'Sex'='steelblue2','Age'='darkseagreen4')
+
+diagnosis_colors <- c('Control'='gray20','HD'='gray20',
+                      'Subclass'='lightslategray','CVID'='coral1',
                       'ICOS'='#8DD3C7','Kabuki'='#BEBADA',
                       'IgG4def'='burlywood4','TNFRSF13B'='palegreen4',
-                      'other'='#BC80BD','AID'='darkgoldenrod3',
+                      'Atypical'='gray40',
+                      'CVID-like'='#BC80BD','AID'='darkgoldenrod3',
                       'ATM'='#A65628','LIG4'='mediumpurple2',
@@
         legend.background = element_rect(fill=NA)) +
-  labs(x='# clones', y='# filtered clusters', color='# reads')
+  labs(x='# Clones', y='# Filtered clusters', color='# Reads')
 ```
@@
 ggplot(cellswitch.data %>%
-         dplyr::mutate(donor=plyr::revalue(donor,
+         dplyr::mutate(Donor=plyr::revalue(Donor,
                                            c('21084'='A',
@@
   geom_abline(slope=1,intercept=0, size=.5, color='gray80') +
-  geom_point(aes(color=donor),size=.5) + 
+  geom_point(aes(color=Donor),size=.5) + 
   geom_smooth(size=.5, method='lm', alpha=.2, color='black') +
@@
         legend.background = element_rect(fill=NA)) +
-  labs(x='# input cells', y='# filtered clusters', color='donor')
+  labs(x='# Input cells', y='# Filtered clusters', color='Donor')
 ```
@@
   dplyr::mutate(clone=paste0(gsub('\\([0-9]*\\)$','',allVHitsWithScore,'_',str_length(aaSeqCDR3)))) %>%
-  dplyr::group_by(donor,ncells,sample) %>%
+  dplyr::group_by(Donor,ncells,Sample) %>%
   dplyr::summarise(nclusters_final=n_distinct(clone),
@@
 swibrid_stats <- cellswitch.data %>%
-  dplyr::select(donor,ncells,
+  dplyr::select(Donor,ncells,
                 nclusters_final,
@@
   dplyr::filter(ncells %in% c(50000,100000)) %>%
-  dplyr::mutate(method='swibrid')
+  dplyr::mutate(method='SWIBRID')
 
@@
   geom_boxplot(outlier.shape=NA, lwd=.5) +
-  geom_point(aes(color=gsub('donor ','',donor)),size=.5) +
+  geom_point(aes(color=gsub('Donor ','',Donor)),size=.5) +
   facet_wrap(~method, ncol=2, scales='free_y') + 
@@
         legend.background =element_rect(fill=NA)) +
-  labs(y='# clones',x='# cells / 1000',color='')
+  labs(y='# Clones',x='# Cells / 1000',color='')
 ```
@@
   dplyr::mutate(ncells=factor(gsub('000$','k',ncells), levels=c('50k','100k')),
-                donor=gsub('donor ','',donor)) %>%
+                Donor=gsub('Donor ','',Donor)) %>%
   dplyr::filter(metric=='alpha_ratio_clusters') %>%
-  dplyr::group_by(ncells, donor, method) %>%
+  dplyr::group_by(ncells, Donor, method) %>%
   dplyr::summarise(mean_alpha_ratio=mean(value),
                    se_alpha_ratio=sd(value)/sqrt(length(value))) %>%
-  pivot_wider(id_cols=c(ncells,donor),names_from=method,values_from=c('mean_alpha_ratio','se_alpha_ratio')) %>%
-  ggplot(aes(x=mean_alpha_ratio_VDJ, y=mean_alpha_ratio_swibrid,
+  pivot_wider(id_cols=c(ncells,Donor),names_from=method,values_from=c('mean_alpha_ratio','se_alpha_ratio')) %>%
+  ggplot(aes(x=mean_alpha_ratio_VDJ, y=mean_alpha_ratio_SWIBRID,
              xmin=mean_alpha_ratio_VDJ-se_alpha_ratio_VDJ,
              xmax=mean_alpha_ratio_VDJ+se_alpha_ratio_VDJ,
-             ymin=mean_alpha_ratio_swibrid-se_alpha_ratio_swibrid,
-             ymax=mean_alpha_ratio_swibrid+se_alpha_ratio_swibrid)) +
-  geom_point(aes(color=donor,shape=ncells),size=1) +
+             ymin=mean_alpha_ratio_SWIBRID-se_alpha_ratio_SWIBRID,
+             ymax=mean_alpha_ratio_SWIBRID+se_alpha_ratio_SWIBRID)) +
+  geom_point(aes(color=Donor,shape=ncells),size=1) +
   geom_errorbar(linewidth=.25) +
@@
   theme(strip.background=element_blank()) +
-  labs(y='fraction IGHA swibrid',x='fraction IGHA VDJ',shape='# cells')
+  labs(y='Fraction IGHA SWIBRID',x='Fraction IGHA VDJ',shape='# cells')
 ```
@@
 HD.scaled <- HD.data %>%
-  dplyr::filter(cohort=='C1') %>%
+  dplyr::filter(Cohort=='C1') %>%
   dplyr::group_by(feature) %>%
@@
                  function(ftr)
-                   tryCatch(my_fitExtractVarPartModel(value ~ (1 | donor) + (1 | sex) + (1 | age) + (1 | sequencing) + (1 | batch),
+                   tryCatch(my_fitExtractVarPartModel(value ~ (1 | Donor) + (1 | Sex) + (1 | Age) + (1 | Sequencing) + (1 | Batch),
                                                       HD.scaled %>%
                                                         dplyr::filter(feature==ftr)),
-                            error=setNames(rep(NA,5), c('donor','sex','age','sequencing','batch')))) %>%
+                            error=setNames(rep(NA,5), c('Donor','Sex','Age','Sequencing','Batch')))) %>%
                    t() %>%
@@
   dplyr::mutate(variable=factor(gsub('Residuals','unexplained',variable),
-                                levels=rev(c('donor','sex','age','batch','sequencing','unexplained'))),
-                category=factor(category, levels=c('structural','context','breaks','diversity','isotypes'))) %>%
+                                levels=rev(c('Donor','Sex','Age','Batch','Sequencing','unexplained'))),
+                category=factor(category, levels=c('Structural','Context','Breaks','Diversity','Isotypes'))) %>%
   dplyr::filter(col %in% cols.here) %>%
@@
   scale_fill_manual(values=var_colors,
-                    breaks=c('donor','sex','age','batch','sequencing','unexplained')) +
-  labs(x='', y='% explained variance', fill='')
+                    breaks=c('Donor','Sex','Age','Batch','Sequencing','unexplained')) +
+  labs(x='', y='% Explained variance', fill='')
 ```
@@
   spread(variable, variance) %>%
-  dplyr::filter(!(category %in% c('diversity_raw','context_isotype','library')), 
+  dplyr::filter(!(category %in% c('Diversity_raw','Context_isotype','library')), 
                 !(col %in% c('log10_nreads','mean_GC')), 
-                donor >= donor_cutoff, 
-                sequencing < sequencing_cutoff, 
-                batch < batch_cutoff) %>%
+                Donor >= donor_cutoff, 
+                Sequencing < sequencing_cutoff, 
+                Batch < batch_cutoff) %>%
   dplyr::arrange(category) %>%
@@
   dplyr::mutate(variable=factor(gsub('Residuals','unexplained',variable),
-                                levels=c('sequencing','batch','donor','sex','age'))) %>%
-  dplyr::filter(variable!='unexplained', category %in% c('library','isotypes','diversity','structural','context','breaks','variants')) %>%
+                                levels=c('Sequencing','Batch','Donor','Sex','Age'))) %>%
+  dplyr::filter(variable!='unexplained', category %in% c('Isotypes','Diversity','Structural','Context','Breaks','Variants')) %>%
   ggplot(aes(x=variable, y=100*variance)) + 
@@
   scale_color_manual(values=c('unused'='black','selected'='orangered')) +
-  labs(x='', y='% explained variance', fill='', color='')
+  labs(x='', y='% Explained variance', fill='', color='')
 ```
@@
   dplyr::left_join(columns, by='col') %>%
-  dplyr::mutate(variable=factor(variable, levels=c('donor','sex','age','batch','sequencing','Residuals'))) %>%
-  dplyr::filter(variable!='Residuals', category %in% c('library','isotypes','diversity','structural','context','breaks','variants')) %>%
+  dplyr::mutate(variable=factor(variable, 
+                                levels=c('Donor','Sex','Age','Batch','Sequencing','Residuals'))) %>%
+  dplyr::filter(variable!='Residuals', category %in% c('Isotypes','Diversity','Structural','Context','Breaks','Variants')) %>%
   ggplot(aes(x=category, y=100*variance, fill=variable)) + 
@@
   scale_fill_manual(values=var_colors) +
-  labs(x='', y='% explained variance', fill='')
+  labs(x='', y='% Explained variance', fill='')
 ```
@@
   dplyr::left_join(columns, by='col') %>%
-  dplyr::mutate(variable=factor(gsub('log10_nreads','# reads',variable),
-                                levels=rev(c('donor','batch','sequencing','# reads','Residuals')))) %>%
-  dplyr::filter(variable!='Residuals', category %in% c('diversity','diversity_raw'))
+  dplyr::mutate(variable=factor(variable,
+                                levels=rev(c('Donor','Batch','Sequencing','Residuals')))) %>%
+  dplyr::filter(variable!='Residuals', category %in% c('Diversity','Diversity_raw'))
 
@@
   coord_cartesian(clip='off') +
-  scale_fill_manual(values=c('diversity'='blue3','diversity_raw'='lightblue1'),
+  scale_fill_manual(values=c('Diversity'='blue3','Diversity_raw'='lightblue1'),
                     labels=c('downsampled','raw')) +
-  labs(x='', y='% explained variance', fill='diversity measures')
+  labs(x='', y='% Explained variance', fill='Diversity measures')
 ```
@@
   dplyr::left_join(columns, by='col') %>%
-  dplyr::mutate(variable=factor(gsub('log10_nreads','# reads',variable),
-                                levels=rev(c('donor','cohort','batch','sequencing','# reads','Residuals')))) %>%
-  dplyr::filter(variable!='Residuals', category %in% c('context','context_isotype'), variable!='cohort')
+  dplyr::mutate(variable=factor(variable,
+                                levels=rev(c('Donor','Batch','Sequencing','Residuals')))) %>%
+  dplyr::filter(variable!='Residuals', category %in% c('Context','Context_isotype'))
 
@@
   coord_cartesian(clip='off') +
-  scale_fill_manual(values=c('context'='darkorchid3','context_isotype'='orchid1'),
+  scale_fill_manual(values=c('Context'='darkorchid3','Context_isotype'='orchid1'),
                     labels=c('global','per isotype')) +
-  labs(x='', y='% explained variance', fill='context measures')
+  labs(x='', y='% Explained variance', fill='Context measures')
 ```
@@
 HD.scaled.br <- HD.data.br %>%
-  dplyr::filter(cohort=='C1') %>%
+  dplyr::filter(Cohort=='C1') %>%
   dplyr::group_by(feature) %>%
@@
                     function(ftr)
-                      my_fitExtractVarPartModel(value ~ (1 | donor) + (1 | sex) + (1 | age) + (1 | sequencing) + (1 | batch),
+                      my_fitExtractVarPartModel(value ~ (1 | Donor) + (1 | Sex) + (1 | Age) + (1 | Sequencing) + (1 | Batch),
                                                 HD.scaled.br %>%
@@
 tmp <- rbind(HD.var %>% 
-               dplyr::mutate(averaging='clusters'), 
+               dplyr::mutate(averaging='Clusters'), 
              HD.var.br %>%
-               dplyr::mutate(averaging='reads')) %>% 
+               dplyr::mutate(averaging='Reads')) %>% 
   dplyr::filter(col %in% fts) %>%
   dplyr::left_join(columns, by='col') %>%
-  dplyr::mutate(variable=factor(variable, levels=rev(c('donor','sex','age','batch','sequencing','Residuals')))) %>%
+  dplyr::mutate(variable=factor(variable, 
+                                levels=rev(c('Donor','Sex','Age','Batch','Sequencing','Residuals')))) %>%
   dplyr::filter(variable!='Residuals', 
-                category %in% c("context","breaks"),
-                variable %in% c("donor", "batch", "sequencing")) %>% 
+                category %in% c("Context","Breaks"),
+                variable %in% c("Donor", "Batch", "Sequencing")) %>% 
   dplyr::arrange(col,averaging,variable)
@@
         legend.box='horizontal') +
-  scale_fill_manual(values=c('clusters'='aquamarine4','reads'='darkseagreen1')) +
-  labs(x='', y='% explained variance', fill='averages over')
+  scale_fill_manual(values=c('Clusters'='aquamarine4','Reads'='darkseagreen1')) +
+  labs(x='', y='% Explained variance', fill='Averages over')
 ```
@@
   spread(variable, variance) %>%
-  dplyr::filter(!(category %in% c('library','diversity_raw','context_isotype'))) %>%
-  dplyr::slice_max(age, n=1) %>%
+  dplyr::filter(!(category %in% c('Library','Diversity_raw','Context_isotype'))) %>%
+  dplyr::slice_max(Age, n=1) %>%
   dplyr::pull(col)
@@
   spread(variable, variance) %>%
-  dplyr::filter(!(category %in% c('library','diversity_raw','context_isotype'))) %>%
-  dplyr::slice_max(sex, n=1) %>%
+  dplyr::filter(!(category %in% c('Library','Diversity_raw','Context_isotype'))) %>%
+  dplyr::slice_max(Sex, n=1) %>%
   dplyr::pull(col)
@@
 HD.data %>%
-  dplyr::filter(feature %in% sex.cols, cohort=='C1') %>%
+  dplyr::filter(feature %in% sex.cols, Cohort=='C1') %>%
   dplyr::mutate(feature=factor(gsub('_downsampled|_switch','',feature),
                                levels=gsub('_downsampled|_switch','',sex.cols))) %>% 
-  ggplot(aes(x=sex, y=value)) + 
+  ggplot(aes(x=str_to_title(Sex), y=value)) + 
   geom_boxplot(outlier.shape=NA, lwd=.5) +
   geom_jitter(width=.2, size=.1) + 
-  stat_compare_means(comparisons=list(c('male','female')),
+  stat_compare_means(comparisons=list(c('Male','Female')),
                      method='wilcox.test',label = "p.signif", hide.ns=TRUE, size=2.5) +
@@
 HD.data %>%
-  dplyr::filter(feature %in% age.cols, cohort=='C1') %>%
+  dplyr::filter(feature %in% age.cols, Cohort=='C1') %>%
   dplyr::mutate(feature=factor(gsub('_downsampled|_switch','',feature),
                                levels=gsub('_downsampled|_switch','',age.cols))) %>% 
-  ggplot(aes(x=age, y=value)) + 
+  ggplot(aes(x=Age, y=value)) + 
   geom_boxplot(outlier.shape=NA, lwd=.5) +
@@
 FB.X <- HD.data %>%
-  dplyr::filter(cohort=='C2') %>%
-  dplyr::select(sample, feature, value) %>%
+  dplyr::filter(Cohort=='C2') %>%
+  dplyr::select(Sample, feature, value) %>%
   spread(feature, value) %>%
   dplyr::mutate_if(is.numeric, scale) %>%
-  tibble::column_to_rownames('sample')
+  tibble::column_to_rownames('Sample')
 
@@
 cohort.data <- HD.data %>%
-  dplyr::filter(sequencing=='minION') %>% 
+  dplyr::filter(Sequencing=='minION') %>% 
   dplyr::filter(feature %in% cols.here.human) %>%
@@
                                levels=gsub('_downsampled|_switch','',cols.here.human))) %>%
-  dplyr::filter(category %in% c('diversity','isotypes','breaks','context','structural'))
-
-ggplot(cohort.data, aes(x=cohort, y=value)) + 
-  geom_boxplot(aes(group=cohort), position=position_dodge(width=.75), width=.7, outlier.shape=NA, lwd=.25, fill='gray80') + 
+  dplyr::filter(category %in% c('Diversity','Isotypes','Breaks','Context','Structural'))
+
+ggplot(cohort.data, aes(x=Cohort, y=value)) + 
+  geom_boxplot(aes(group=Cohort), position=position_dodge(width=.75), width=.7, outlier.shape=NA, lwd=.25, fill='gray80') + 
   geom_pointrange(data=cohort.data %>%
-                    dplyr::group_by(donor,feature,category,cohort) %>%
+                    dplyr::group_by(Donor,feature,category,Cohort) %>%
                     dplyr::summarise(se=sd(value)/sqrt(n()),
@@
                                      n=n()) %>% dplyr::filter(n > 1),
-                  aes(fill=cohort, ymin=value-se,ymax=value+se, group=cohort),
+                  aes(fill=Cohort, ymin=value-se,ymax=value+se, group=Cohort),
                   position=position_jitterdodge(jitter.width=.5, dodge.width=.75, jitter.height = 0), 
@@
 mouse.data <- read.csv('data/mouse_data.csv',header=1, row.names=1)  %>%
-  tibble::column_to_rownames('sample')
+  tibble::column_to_rownames('Sample')
 vincendeau.data <- read.csv('data/vincendeau_data.csv', header=1, row.names=1)  %>%
-  tibble::column_to_rownames('sample')
+  tibble::column_to_rownames('Sample')
 
@@
                        dplyr::mutate(species='mouse tissue',
-                                     cohort=material) %>%
-                       dplyr::select(species, cohort, cols.here.mouse),
+                                     Cohort=material) %>%
+                       dplyr::select(species, Cohort, cols.here.mouse),
                      vincendeau.data %>%
                        dplyr::mutate(species='Vincendeau',
-                                     cohort=genotype) %>%
-                       dplyr::select(species, cohort, cols.here.mouse)) %>%
-  dplyr::mutate(cohort=factor(cohort, levels=c('lymph node','splenocytes','WT','Xlf','Shld1','Shld1_Xlf'))) %>%
+                                     Cohort=Genotype) %>%
+                       dplyr::select(species, Cohort, cols.here.mouse)) %>%
+  dplyr::mutate(Cohort=factor(Cohort, levels=c('lymph node','splenocytes','WT','Xlf','Shld1','Shld1_Xlf'))) %>%
   gather(variable, value, cols.here.mouse) %>%
-  dplyr::select(variable, value, species, cohort) %>%
+  dplyr::select(variable, value, species, Cohort) %>%
   dplyr::left_join(columns, by=c('variable'='col')) %>%
@@
                                 levels=gsub('_downsampled|_switch','',cols.here.mouse))) %>%
-  dplyr::filter(category %in% c('diversity','isotypes','breaks','context','structural'))
+  dplyr::filter(category %in% c('Diversity','Isotypes','Breaks','Context','Structural'))
 
 cohort.data %>%
-  ggplot(aes(x=cohort, y=value, fill=cohort)) + 
+  ggplot(aes(x=Cohort, y=value, fill=Cohort)) + 
   geom_boxplot(position=position_dodge(width=.75), width=.7, outlier.shape=NA, lwd=.25, fill='gray80') + 
@@
   scale_x_discrete(breaks=c('lymph node','splenocytes','WT','Xlf','Shld1','Shld1_Xlf'),
+                   labels=c('Lymph node','Splenocytes','WT','Xlf','Shld1','Shld1_Xlf'),
                    limits=c('lymph node','splenocytes','','WT','Xlf','Shld1','Shld1_Xlf'), drop=FALSE) +
@@
 tmp <- HD.scaled %>%
-  dplyr::filter(feature %in% selected.cols, batch=='20230313') %>%
-  dplyr::select(feature,value,sequencing,donor) %>%
-  spread(sequencing,value) %>% 
+  dplyr::filter(feature %in% selected.cols, Batch=='20230313') %>%
+  dplyr::select(feature,value,Sequencing,Donor) %>%
+  spread(Sequencing,value) %>% 
   dplyr::left_join(columns,by=c('feature'='col')) %>% 
@@
 vincendeau.meta <- vincendeau.data %>%
-  dplyr::mutate(genotype=factor(genotype,levels=c('WT','Xlf','Shld1','Shld1_Xlf'))) %>%
-  dplyr::select(genotype)
+  dplyr::mutate(Genotype=factor(Genotype,levels=c('WT','Xlf','Shld1','Shld1_Xlf'))) %>%
+  dplyr::select(Genotype)
 
@@
 
-ggplot(pcadat,aes(x=PC1,y=PC2, color=genotype)) +
+ggplot(pcadat,aes(x=PC1,y=PC2, color=Genotype)) +
   geom_point(size=1) +
   geom_polygon(data = pcadat %>%
-                 dplyr::group_by(genotype) %>%
-                 dplyr::slice(chull(PC1, PC2)), alpha = 0.25, aes(fill=genotype)) +
+                 dplyr::group_by(Genotype) %>%
+                 dplyr::slice(chull(PC1, PC2)), alpha = 0.25, aes(fill=Genotype)) +
   my_theme() +
@@
   dplyr::mutate(value=scale(value)) %>%
-  dplyr::group_by(genotype, variable) %>%
+  dplyr::group_by(Genotype, variable) %>%
   dplyr::summarise(mean=mean(value),
                    se=sd(value)/sqrt(n())) %>%
-  tidyr::pivot_wider(id_cols=c('variable'),names_from='genotype',values_from=c('mean','se')) %>%
+  tidyr::pivot_wider(id_cols=c('variable'),names_from='Genotype',values_from=c('mean','se')) %>%
   dplyr::left_join(columns,by=c('variable'='col')) %>%
@@
 HTGTS.data <- read.csv("data/HTGTS_data.csv", header=1, row.names=1) %>%
-  tibble::column_to_rownames('sample')
+  tibble::column_to_rownames('Sample')
 
@@
 tmp <- HTGTS.data %>%
+  dplyr::mutate(celltype=gsub('spleen','Spleen',celltype)) %>%
   gather(variable, value, HTGTS.cols) %>%
@@
   dplyr::mutate(value=scale(value)) %>%
-  dplyr::group_by(celltype, genotype, variable) %>%
+  dplyr::group_by(celltype, Genotype, variable) %>%
   dplyr::summarise(mean=mean(value),
                    se=sd(value)/sqrt(n())) %>%
-  tidyr::pivot_wider(id_cols=c('variable'),names_from=c('celltype','genotype'),values_from=c('mean','se')) %>%
+  tidyr::pivot_wider(id_cols=c('variable'),names_from=c('celltype','Genotype'),values_from=c('mean','se')) %>%
   dplyr::left_join(columns,by=c('variable'='col')) %>%
@@
                     c('CH12','WT','Lig4'),
-                    c('spleen','WT','Atm'),
-                    c('spleen','WT','Trp53bp1'))
+                    c('Spleen','WT','Atm'),
+                    c('Spleen','WT','Trp53bp1'))
 
@@
 HTGTS.data %>%
-  dplyr::mutate(celltype=factor(celltype, levels=c('CH12','spleen')),
-                genotype=factor(genotype, levels=c('WT','Atm','Lig4','Trp53bp1'))) %>%
+  dplyr::mutate(celltype=factor(gsub('spleen','Spleen',celltype), levels=c('CH12','Spleen')),
+                Genotype=factor(Genotype, levels=c('WT','Atm','Lig4','Trp53bp1'))) %>%
   gather(variable, value, cols.here.HTGTS) %>%
-  dplyr::select(variable, value, genotype, celltype) %>%
+  dplyr::select(variable, value, Genotype, celltype) %>%
   dplyr::left_join(columns, by=c('variable'='col')) %>%
@@
                                 levels=gsub('_downsampled|_switch','',cols.here.HTGTS))) %>%
-  dplyr::filter(category %in% c('diversity','isotypes','breaks','context','structural')) %>%
-  ggplot(aes(x=genotype, y=value, fill=genotype)) + 
+  dplyr::filter(category %in% c('Diversity','Isotypes','Breaks','Context','Structural')) %>%
+  ggplot(aes(x=Genotype, y=value, fill=Genotype)) + 
   geom_boxplot(position=position_dodge(width=.75), width=.7, outlier.shape=NA, lwd=.25) + 
@@
 read.csv('data/HTGTS_homology.csv',header=1, row.names=1) %>%
-  dplyr::mutate(genotype=factor(genotype, levels=c('WT','Atm','Lig4','Trp53bp1'))) %>%
-  dplyr::filter(xor(isotype=='Sa', celltype=='spleen')) %>%
-  dplyr::group_by( celltype, genotype, isotype, n_homology) %>%
+  dplyr::mutate(Genotype=factor(Genotype, levels=c('WT','Atm','Lig4','Trp53bp1')),
+                celltype=gsub('spleen','Spleen',celltype)) %>%
+  dplyr::filter(xor(isotype=='Sa', celltype=='Spleen')) %>%
+  dplyr::group_by( celltype, Genotype, isotype, n_homology) %>%
   dplyr::summarise(mean_frac=mean(frac),
                    se_frac=sd(frac)/sqrt(n())) %>%
-  ggplot(aes(x=n_homology, y=mean_frac, color=genotype)) +
-  geom_line(aes(group=genotype)) +
+  ggplot(aes(x=n_homology, y=mean_frac, color=Genotype)) +
+  geom_line(aes(group=Genotype)) +
   geom_errorbar(aes(ymin=mean_frac-se_frac,ymax=mean_frac+se_frac),width=0) +
@@
   scale_color_manual(values=genotype_colors) +
-  labs(color='', y='fraction junctions') +
+  labs(color='', y='Fraction junctions') +
   coord_cartesian(clip='off') 
@@
 HTGTS.meta <- HTGTS.data %>%
-  dplyr::mutate(genotype=factor(genotype,levels=c('WT','Atm','Lig4','Trp53bp1'))) %>%
-  dplyr::select(celltype,genotype)
+  dplyr::mutate(Genotype=factor(Genotype,levels=c('WT','Atm','Lig4','Trp53bp1'))) %>%
+  dplyr::select(celltype,Genotype)
 
@@
 
-p1 <- ggplot(pcadat.CH12,aes(x=PC1,y=PC2, color=genotype)) +
+p1 <- ggplot(pcadat.CH12,aes(x=PC1,y=PC2, color=Genotype)) +
   geom_point(size=1) +
   geom_polygon(data = pcadat.CH12 %>%
-                 dplyr::group_by(genotype) %>%
-                 dplyr::slice(chull(PC1, PC2)), alpha = 0.25, aes(fill=genotype)) +
+                 dplyr::group_by(Genotype) %>%
+                 dplyr::slice(chull(PC1, PC2)), alpha = 0.25, aes(fill=Genotype)) +
   my_theme() +
@@
 
-p2 <- ggplot(pcadat.spleen,aes(x=PC1,y=PC2, color=genotype)) +
+p2 <- ggplot(pcadat.spleen,aes(x=PC1,y=PC2, color=Genotype)) +
   geom_point(size=1) +
   geom_polygon(data = pcadat.spleen %>%
-                 dplyr::group_by(genotype) %>%
-                 dplyr::slice(chull(PC1, PC2)), alpha = 0.25, aes(fill=genotype)) +
+                 dplyr::group_by(Genotype) %>%
+                 dplyr::slice(chull(PC1, PC2)), alpha = 0.25, aes(fill=Genotype)) +
   my_theme() +
@@
        y=paste0('PC 2 (',100*signif(pca.spleen$sdev[2]**2/sum(pca.spleen$sdev**2),3),'%)'),
-       color='',shape='',fill='', title='spleen')  +
+       color='',shape='',fill='', title='Spleen')  +
   scale_color_manual(values=genotype_colors) +
@@
 CH12.data <- read.csv('data/CH12_data.csv',header=1, row.names=1) %>%
-  tibble::column_to_rownames('sample')
+  tibble::column_to_rownames('Sample')
 CH12.data.noSg <- read.csv('data/CH12_data_noSg.csv', header=1, row.names=1) %>%
-  tibble::column_to_rownames('sample')
+  tibble::column_to_rownames('Sample')
 mouse.data.noSg <- read.csv('data/mouse_data_noSg.csv', header=1, row.names=1)
@@
 
-CH12.take <- CH12.data$batch != '20240830'
+CH12.take <- CH12.data$Batch != '20240830'
 mouse.take <- rep(TRUE, dim(mouse.data)[1]) 
@@
 pcadat <- cbind(pca$x,data.frame(material=c(rep('CH12',sum(CH12.take)),
-                                            mouse.data$material[mouse.take]),
-                                 batch=c(CH12.data$batch[CH12.take],
-                                         mouse.data$batch[mouse.take]),
-                                 genotype=c(CH12.data$genotype[CH12.take],
+                                            str_to_title(mouse.data$material[mouse.take])),
+                                 Batch=c(CH12.data$Batch[CH12.take],
+                                         mouse.data$Batch[mouse.take]),
+                                 Genotype=c(CH12.data$Genotype[CH12.take],
                                             rep('WT',sum(mouse.take)))))
@@
 p1 <- ggplot(pcadat,aes(x=PC1,y=PC2, color=material)) +
-  geom_point(aes(shape=genotype),size=1) +
+  geom_point(aes(shape=Genotype),size=1) +
   geom_polygon(data = pcadat %>%
@@
 pcadat.noSg <- cbind(pca.noSg$x,data.frame(material=c(rep('CH12',sum(CH12.take.noSg)),
-                                                      mouse.data.noSg$material[mouse.take.noSg]),
-                                           batch=c(CH12.data.noSg$batch[CH12.take.noSg],
-                                                   mouse.data.noSg$batch[mouse.take.noSg]),
-                                           genotype=c(CH12.data.noSg$genotype[CH12.take.noSg],
+                                                      str_to_title(mouse.data.noSg$material[mouse.take.noSg])),
+                                           Batch=c(CH12.data.noSg$Batch[CH12.take.noSg],
+                                                   mouse.data.noSg$Batch[mouse.take.noSg]),
+                                           Genotype=c(CH12.data.noSg$Genotype[CH12.take.noSg],
                                                       rep('WT',sum(mouse.take.noSg)))))
@@
 p2 <- ggplot(pcadat.noSg,aes(x=PC1,y=PC2, color=material)) +
-  geom_point(aes(shape=genotype), size=1) +
+  geom_point(aes(shape=Genotype), size=1) +
   geom_polygon(data = pcadat.noSg %>%
@@
 # p3 <- ggplot(pcadat,aes(x=PC3,y=PC4, color=material)) +
-#   geom_point(aes(shape=genotype),size=1) +
+#   geom_point(aes(shape=Genotype),size=1) +
 #   geom_polygon(data = pcadat %>%
@@
 # p4 <- ggplot(pcadat.noSg,aes(x=PC3,y=PC4, color=material)) +
-#   geom_point(aes(shape=genotype),size=1) +
+#   geom_point(aes(shape=Genotype),size=1) +
 #   geom_polygon(data = pcadat.noSg %>%
@@
 dpcadat <- cbind(rbind(dpca$x, dpca$y),
-                 data.frame(group=c(CH12.data[row.names(dpca$x),'genotype'],
+                 data.frame(group=c(CH12.data[row.names(dpca$x),'Genotype'],
                                     mouse.data$material)))
@@
 CH12.meta <- CH12.data %>%
-  dplyr::mutate(batch=as.character(batch),
-                genotype=factor(genotype,levels=CH12.genotypes)) %>%
-  dplyr::select(batch,genotype)
+  dplyr::mutate(Batch=as.character(Batch),
+                Genotype=factor(Genotype,levels=CH12.genotypes)) %>%
+  dplyr::select(Batch,Genotype)
 
 tmp <- CH12.data %>% 
-  dplyr::filter(batch != '20240830') %>%
-  dplyr::select(genotype,CH12.features) %>%
-  dplyr::mutate(genotype=factor(genotype,levels=CH12.genotypes)) %>%
-  dplyr::group_by(genotype) %>%
+  dplyr::filter(Batch != '20240830') %>%
+  dplyr::select(Genotype,CH12.features) %>%
+  dplyr::mutate(Genotype=factor(Genotype,levels=CH12.genotypes)) %>%
+  dplyr::group_by(Genotype) %>%
   dplyr::mutate(ng=n(),
-                genotype_new=paste0(genotype,' (n=',ng,')')) %>%
-  gather(col,value,-c(genotype,ng,genotype_new)) %>% 
+                Genotype_new=paste0(Genotype,' (n=',ng,')')) %>%
+  gather(col,value,-c(Genotype,ng,Genotype_new)) %>% 
   dplyr::mutate(col=factor(gsub("_downsampled|_switch",'',col),
@@
 
-ggplot(tmp, aes(x=genotype,y=value,fill=genotype)) + 
+ggplot(tmp, aes(x=Genotype,y=value,fill=Genotype)) + 
   geom_boxplot(outlier.shape=NA, lwd=.5) + 
@@
   scale_fill_manual(values=genotype_colors) +
-  scale_x_discrete(breaks=tmp$genotype,labels=tmp$genotype_new) +
-  labs(x='',y='', color='# reads')
+  scale_x_discrete(breaks=tmp$Genotype,labels=tmp$Genotype_new) +
+  labs(x='',y='', color='# Reads')
 ```
@@
 ```{r Fig3E, fig.width=2. * fig_scale, fig.height=1.7 * fig_scale}
-CH12.samples <- row.names(CH12.data)[CH12.data$batch != '20240830']
+CH12.samples <- row.names(CH12.data)[CH12.data$Batch != '20240830']
 CH12.cols <- setdiff(grep('SG|Sg',intersect(selected.cols.mouse,colnames(CH12.data)),value=TRUE,invert=TRUE),
@@
 
-ggplot(pcadat,aes(x=PC1,y=PC2, color=genotype)) +
+ggplot(pcadat,aes(x=PC1,y=PC2, color=Genotype)) +
   geom_point(size=2) +
   geom_polygon(data = pcadat %>%
-                 dplyr::group_by(genotype) %>%
-                 dplyr::slice(chull(PC1, PC2)), alpha = 0.25, aes(fill=genotype)) +
+                 dplyr::group_by(Genotype) %>%
+                 dplyr::slice(chull(PC1, PC2)), alpha = 0.25, aes(fill=Genotype)) +
   my_theme() +
@@
   tibble::rownames_to_column('sample') %>%
-  dplyr::filter(batch != '20240830') %>%
-  dplyr::arrange(genotype) %>%
+  dplyr::filter(Batch != '20240830') %>%
+  dplyr::arrange(Genotype) %>%
   dplyr::pull(sample)
@@
                                                       tibble::rownames_to_column('sample') %>%
-                                                      dplyr::mutate(genotype=factor(genotype, levels=unique(genotype))) %>%
-                                                      dplyr::arrange(genotype) %>%
+                                                      dplyr::mutate(Genotype=factor(Genotype, levels=unique(Genotype))) %>%
+                                                      dplyr::arrange(Genotype) %>%
                                                       dplyr::pull(sample),
@@
                column_dend_height = unit(5, "mm"),
-               top_annotation=HeatmapAnnotation(df=CH12.meta[samples,c('genotype','batch')],
+               top_annotation=HeatmapAnnotation(df=CH12.meta[samples,c('Genotype','Batch')],
                                                 show_annotation_name=TRUE,
@@
                                                                              nrow=1),
-                                                col=list('genotype'=genotype_colors[levels(CH12.meta$genotype)],
-                                                         'batch'=batch_colors[unique(CH12.meta[samples,'batch'])])),
+                                                col=list('Genotype'=genotype_colors[levels(CH12.meta$Genotype)],
+                                                         'Batch'=batch_colors[unique(CH12.meta[samples,'Batch'])])),
                column_names_gp=gpar(fontsize=7),
@@
 test_batches <- '20240830'
-train <- CH12.meta$batch %in% train_batches
-test <- CH12.meta$batch %in% test_batches
+train <- CH12.meta$Batch %in% train_batches
+test <- CH12.meta$Batch %in% test_batches
 
@@
 set.seed(3)
-cv_model <- cv.glmnet(as.matrix(CH12.X[train,]), CH12.meta$genotype[train], alpha = 0, family='multinomial')
-
-mod <- glmnet(as.matrix(CH12.X[train,]), CH12.meta$genotype[train], alpha = 0, family = "multinomial",
+cv_model <- cv.glmnet(as.matrix(CH12.X[train,]), CH12.meta$Genotype[train], alpha = 0, family='multinomial')
+
+mod <- glmnet(as.matrix(CH12.X[train,]), CH12.meta$Genotype[train], alpha = 0, family = "multinomial",
               lambda = cv_model$lambda.min)
@@
                      tibble::rownames_to_column('sample') %>% 
-                     dplyr::select(sample,genotype),by='sample') %>%
+                     dplyr::select(sample,Genotype),by='sample') %>%
   dplyr::group_by(sample) %>%
-  dplyr::mutate(c=ifelse(p == max(p),ifelse(genotype==prediction, 'green','red'),'black'),
-                genotype=factor(genotype,levels=CH12.genotypes),
+  dplyr::mutate(c=ifelse(p == max(p),ifelse(Genotype==prediction, 'green','red'),'black'),
+                Genotype=factor(Genotype,levels=CH12.genotypes),
                 sample=gsub('[0-9]*_(BC[0-9]*)_switch','\\1',sample)) %>%
   dplyr::ungroup() %>%
-  dplyr::arrange(genotype) %>%
+  dplyr::arrange(Genotype) %>%
   dplyr::mutate(sample=factor(gsub('20240830_','',sample),
@@
 tmp %>%
-  gather(genotype,coef,-col) %>%
-  dplyr::mutate(genotype=factor(genotype, levels=CH12.genotypes),
+  gather(Genotype,coef,-col) %>%
+  dplyr::mutate(Genotype=factor(Genotype, levels=CH12.genotypes),
                 col=factor(col, levels=rev(levels(tmp$col)))) %>%
-  ggplot(aes(x=genotype, y=col, fill=coef, label=round(coef,2))) + 
+  ggplot(aes(x=Genotype, y=col, fill=coef, label=round(coef,2))) + 
   geom_tile() +
@@
 batch.CV <- c()
-for (batch in unique(CH12.meta$batch)) {
-  train <- CH12.meta$batch != batch
-  test <- CH12.meta$batch == batch
+for (batch in unique(CH12.meta$Batch)) {
+  train <- CH12.meta$Batch != batch
+  test <- CH12.meta$Batch == batch
   X <- CH12.data[, CH12.cols] %>%
@@
 
-  cv_model <- cv.glmnet(X[train,], CH12.meta$genotype[train], alpha = 0, family='multinomial')
-  mod <- glmnet(X[train,], CH12.meta$genotype[train], alpha = 0, family = "multinomial",
+  cv_model <- cv.glmnet(X[train,], CH12.meta$Genotype[train], alpha = 0, family='multinomial')
+  mod <- glmnet(X[train,], CH12.meta$Genotype[train], alpha = 0, family = "multinomial",
                 lambda = cv_model$lambda.min)
-  batch.CV[[batch]] <- mean(predict(mod, type = "class", newx=X[test,])[,'s0'] == CH12.meta[test,'genotype'])
+  batch.CV[[batch]] <- mean(predict(mod, type = "class", newx=X[test,])[,'s0'] == CH12.meta[test,'Genotype'])
 }
@@
 
-  cv_model <- cv.glmnet(X[train,], CH12.meta$genotype[train], alpha = 0, family='multinomial')
-  mod <- glmnet(X[train,], CH12.meta$genotype[train], alpha = 0, family = "multinomial",
+  cv_model <- cv.glmnet(X[train,], CH12.meta$Genotype[train], alpha = 0, family='multinomial')
+  mod <- glmnet(X[train,], CH12.meta$Genotype[train], alpha = 0, family = "multinomial",
                 lambda = cv_model$lambda.min)
-  sample.CV[[as.character(k)]] <- mean(predict(mod, type = "class", newx=X[test,])[,'s0'] == CH12.meta[test,'genotype'])
+  sample.CV[[as.character(k)]] <- mean(predict(mod, type = "class", newx=X[test,])[,'s0'] == CH12.meta[test,'Genotype'])
 }
@@
   dplyr::filter(!is.na(acc)) %>%
-  dplyr::mutate(method=ifelse(fold %in% CH12.meta$batch,'batch','sample')) %>%
+  dplyr::mutate(method=ifelse(fold %in% CH12.meta$Batch,'Batch','Sample')) %>%
   dplyr::group_by(method) %>%
@@
   theme(axis.text.x=element_text(angle=90, hjust=1, vjust=.5)) +
-  labs(x='',y='accuracy')
+  labs(x='',y='Accuracy')
 ```
@@
 
-explained variance for FB cohort (make FB_13/21 control)
+explained variance for FB cohort (make FB_13/21 "Atypical")
 
@@
 FB.data <- read.csv('data/FB_data.csv', header=1, row.names=1) %>%
-  dplyr::mutate(batch=as.character(batch),
-                diagnosis = ifelse(donor %in% c('FB_13','FB_21'), 'control', diagnosis)) %>%
-  tibble::column_to_rownames('sample')
+  dplyr::mutate(Batch=as.character(Batch)) %>%
+  tibble::column_to_rownames('Sample')
 FB.X <- FB.data[grepl('2023',row.names(FB.data)),intersect(colnames(FB.data),columns$col)] %>%
@@
 FB.meta <- FB.data %>%
-  dplyr::filter(grepl('2023',batch)) %>%
-  dplyr::select(donor,batch,sex,age,diagnosis,timepoint)
+  dplyr::filter(grepl('2023',Batch)) %>%
+  dplyr::select(Donor,Batch,Sex,Age,Diagnosis,Timepoint)
 
 FB.var <- fitExtractVarPartModel(t(FB.X[,colSums(is.na(FB.X))==0]), 
-                                 ~ (1 | donor) + (1 | diagnosis) + (1 | sex) + age + (1 | batch), FB.meta) %>%
+                                 ~ (1 | Donor) + (1 | Diagnosis) + (1 | Sex) + Age + (1 | Batch), FB.meta) %>%
   as.data.frame() %>%
@@
 FB.var %>%  dplyr::left_join(columns, by='col') %>%
-  dplyr::mutate(variable=factor(variable,levels=c('donor','diagnosis','sex','age','batch','Residuals'))) %>%
-  dplyr::filter(variable!='Residuals', category %in% c('library','isotypes','diversity','structural','context','breaks','variants')) %>%
+  dplyr::mutate(variable=factor(variable,levels=c('Donor','Diagnosis','Sex','Age','Batch','Residuals'))) %>%
+  dplyr::filter(variable!='Residuals', category %in% c('Isotypes','Diversity','Structural','Context','Breaks','Variants')) %>%
   ggplot(aes(x=variable, y=100*variance)) + 
@@
   scale_fill_manual(values=var_colors, guide='none') +
-  labs(x='', y='% explained variance', fill='', color='')
+  labs(x='', y='% Explained variance', fill='', color='')
 ```
@@
 FB.var %>%  dplyr::left_join(columns, by='col') %>%
-  dplyr::mutate(variable=factor(variable, levels=c('donor','diagnosis','sex','age','batch','Residuals'))) %>%
-  dplyr::filter(variable!='Residuals', category %in% c('library','isotypes','diversity','structural','context','breaks','variants')) %>%
+  dplyr::mutate(variable=factor(variable, levels=c('Donor','Diagnosis','Sex','Age','Batch','Residuals'))) %>%
+  dplyr::filter(variable!='Residuals', category %in% c('Isotypes','Diversity','Structural','Context','Breaks','Variants')) %>%
   ggplot(aes(x=category, y=100*variance, fill=variable)) + 
@@
   dplyr::mutate(num_variants_somatic = num_variants * (1 - frac_variants_germline),
-                diagnosis = ifelse(donor %in% c('FB_13','FB_21'), 'control', diagnosis),
-                status=ifelse(diagnosis %in% c('control','subclass'), 'control', 'CVID')) %>%
-  tibble::column_to_rownames('sample')
+                status=ifelse(Diagnosis %in% c('Control','Subclass','Atypical'), 'Control', 'CVID')) %>%
+  tibble::column_to_rownames('Sample')
 QP.data <- read.csv('data/QP_data.csv', header=1, row.names=1) %>%
-  tibble::column_to_rownames('sample')
+  dplyr::mutate(Diagnosis=plyr::revalue(Diagnosis,c('control'='Control'))) %>%
+  tibble::column_to_rownames('Sample')
 
@@
 human.meta <- rbind(FB.data.pooled %>%
-                      dplyr::mutate(dataset=ifelse(grepl('FB',donor),
-                                                   ifelse(sample_donor %in% FB_train_donors,'training','testing'),
+                      dplyr::mutate(dataset=ifelse(grepl('FB',Donor),
+                                                   ifelse(Sample_donor %in% FB_train_donors,'training','testing'),
                                                    'validation')) %>%
-                      dplyr::select(donor,diagnosis,status,timepoint, dataset,sex,age,complication, EUROclass, pct_B, pct_IgA, pct_IgG),
+                      dplyr::select(Donor,Diagnosis,status,Timepoint, dataset,Sex,Age,complication, EUROclass, pct_B, pct_IgA, pct_IgG),
                     QP.data %>%
                       dplyr::mutate(dataset='validation',
-                                    timepoint='t1',
+                                    Timepoint='t1',
                                     complication=NA,
@@
                                     pct_IgG=NA,
-                                    status=ifelse(diagnosis=='control','control','CVID')) %>%
-                      dplyr::select(donor,diagnosis,status,timepoint,dataset,sex,age,complication, EUROclass, pct_B, pct_IgA, pct_IgG)) %>%
-  dplyr::mutate(diagnosis=factor(diagnosis,levels=c('control','subclass','CVID','ICOS','Kabuki','IgG4def',
+                                    status=ifelse(Diagnosis=='Control','Control','CVID')) %>%
+                      dplyr::select(Donor,Diagnosis,status,Timepoint,dataset,Sex,Age,complication, EUROclass, pct_B, pct_IgA, pct_IgG)) %>%
+  dplyr::mutate(Diagnosis=factor(Diagnosis,levels=c('Control','Subclass','Atypical','CVID','ICOS','Kabuki','IgG4def',
                                                     'AID','TNFRSF13B',
                                                     'NFKB2','PI3K','ATM','BRCA1','LIG4','NIPBL')),
-                status=factor(status, levels=c('control','CVID')))
+                status=factor(status, levels=c('Control','CVID')))
 
@@
 ggplot(pcadat,aes(x=PC1,y=PC2)) +
-  geom_point(aes(color=diagnosis),size=1) +
-  geom_line(aes(group=donor, color=diagnosis)) +
+  geom_point(aes(color=Diagnosis),size=1) +
+  geom_line(aes(group=Donor, color=Diagnosis)) +
   geom_polygon(data = pcadat %>%
@@
   my_theme() +
-  #geom_label_repel(aes(label=gsub('FB_','',donor)),size=2, segment.size=.25, label.padding = unit(.5,'mm'), force=20, show.legend=FALSE) +
   scale_fill_manual(values=diagnosis_colors, guide="none") +
@@
 FB.cols.subclass <- grep('frac_clusters_S[MGA]+',colnames(FB.data.pooled),value=TRUE)
-FB.X.subclass <- FB.data.pooled[FB.data.pooled$sample_donor %in% FB_train_donors,FB.cols.subclass] %>%
+FB.X.subclass <- FB.data.pooled[FB.data.pooled$Sample_donor %in% FB_train_donors,FB.cols.subclass] %>%
   dplyr::mutate_if(is.numeric,scale)
@@
 ggplot(pcadat.subclass,aes(x=PC1,y=PC2)) +
-  geom_point(aes(color=diagnosis),size=1) +
-  geom_line(aes(group=donor, color=diagnosis)) +
+  geom_point(aes(color=Diagnosis),size=1) +
+  geom_line(aes(group=Donor, color=Diagnosis)) +
   geom_polygon(data = pcadat.subclass %>%
-                 dplyr::mutate(status=ifelse(diagnosis=='subclass', 'subclass', as.character(status))) %>%
+                 dplyr::mutate(status=ifelse(Diagnosis=='Subclass', 'Subclass', as.character(status))) %>%
                  dplyr::group_by(status) %>%
@@
   my_theme() +
-  #geom_label_repel(data = pcadat.subclass %>% 
-  #                   dplyr::filter(diagnosis!='control'),
-  #                 aes(label=gsub('FB_','',donor)),size=2, segment.size=.25, label.padding = unit(.5,'mm'), force=20, show.legend=FALSE) +
   scale_fill_manual(values=diagnosis_colors, guide='none') +
@@
 tmp <- FB.data.pooled %>%
-  dplyr::filter(sample_donor %in% FB_train_donors, diagnosis %in% c('control','subclass')) %>%
+  dplyr::filter(Sample_donor %in% FB_train_donors, Diagnosis %in% c('Control','Subclass')) %>%
   gather(feature, frac_clusters, subclass.cols) %>%
@@
   dplyr::group_by(feature) %>%
-  dplyr::do(wc=wilcox.test(frac_clusters ~ diagnosis, data=.),
+  dplyr::do(wc=wilcox.test(frac_clusters ~ Diagnosis, data=.),
             y=max(.$frac_clusters)) %>%
@@
 ggplot(tmp, aes(x=feature, y=frac_clusters)) +
-  geom_boxplot(aes(color=diagnosis), position=position_dodge(width=.8),width=.7,outlier.shape=NA) +
-  geom_point(aes(color=diagnosis), position=position_dodge(width=.8), size=.5) +
-  geom_line(data=tmp %>% dplyr::filter(diagnosis=='control'),
-            aes(group=donor), position=position_nudge(x=-.2), alpha=.25, size=.25) +
-  geom_line(data=tmp %>% dplyr::filter(diagnosis=='subclass'),
-            aes(group=donor), position=position_nudge(x=+.2), alpha=.25, size=.25) +
+  geom_boxplot(aes(color=Diagnosis), position=position_dodge(width=.8),width=.7,outlier.shape=NA) +
+  geom_point(aes(color=Diagnosis), position=position_dodge(width=.8), size=.5) +
+  geom_line(data=tmp %>% dplyr::filter(Diagnosis=='Control'),
+            aes(group=Donor), position=position_nudge(x=-.2), alpha=.25, size=.25) +
+  geom_line(data=tmp %>% dplyr::filter(Diagnosis=='Ssubclass'),
+            aes(group=Donor), position=position_nudge(x=+.2), alpha=.25, size=.25) +
   geom_text(data=pvals, aes(label=stars.pval(p.value), y=1.1*y)) +
@@
   theme(legend.position=c(.8,.8)) +
-  labs(color='',x='', y='fraction clusters')
+  labs(color='',x='', y='Fraction clusters')
 ```
@@
 
-tmp <- FB.data.pooled[FB.data.pooled$sample_donor %in% FB_train_donors,] %>% 
-  dplyr::select(donor,diagnosis,FB.features) %>%
-  dplyr::mutate(diagnosis_simple=factor(ifelse(diagnosis %in% c('CVID','control','subclass'), diagnosis, 'other'), 
-                                        levels=c('control','subclass','CVID','other'))) %>%
-  dplyr::group_by(diagnosis_simple) %>%
-  dplyr::mutate(ng=n_distinct(donor),
-                diagnosis_new=paste0(diagnosis_simple,' (n=',ng,')')) %>%
-  gather(col,value,-c(diagnosis_simple,ng,diagnosis_new,donor,diagnosis)) %>% 
+tmp <- FB.data.pooled[FB.data.pooled$Sample_donor %in% FB_train_donors,] %>% 
+  dplyr::select(Donor,Diagnosis,FB.features) %>%
+  dplyr::mutate(Diagnosis_simple=factor(plyr::revalue(Diagnosis, c('AID'='CVID-like',
+                                                                   'ICOS'='CVID-like',
+                                                                   'IgG4def'='CVID-like',
+                                                                   'Kabuki'='CVID-like',
+                                                                   'NFKB2'='CVID-like',
+                                                                   'PI3K'='CVID-like',
+                                                                   'TNFRSF13B'='CVID-like')),
+                                        levels=c('Control','Subclass','Atypical','CVID','CVID-like'))) %>%
+  dplyr::group_by(Diagnosis_simple) %>%
+  dplyr::mutate(ng=n_distinct(Donor),
+                Diagnosis_new=paste0(Diagnosis_simple,' (n=',ng,')')) %>%
+  gather(col,value,-c(Diagnosis_simple,ng,Diagnosis_new,Donor,Diagnosis)) %>% 
   dplyr::mutate(col=factor(gsub("_downsampled|_switch",'',col),
@@
 
-ggplot(tmp, aes(x=diagnosis_simple,y=value,fill=diagnosis_simple)) + 
+ggplot(tmp, aes(x=Diagnosis_simple,y=value,fill=Diagnosis_simple)) + 
   geom_boxplot(outlier.shape=NA, lwd=.5) + 
   geom_jitter(width=.1,height=0,size=.5) + 
-  stat_compare_means(ref.group='control',method='wilcox.test',label = "p.signif", hide.ns=TRUE, size=2) +
+  stat_compare_means(ref.group='Control',method='wilcox.test',label = "p.signif", hide.ns=TRUE, size=2) +
   facet_wrap(~col,ncol=4,scales='free_y') +
@@
         axis.text.x=element_text(angle=90,hjust=1,vjust=.5)) +
-  scale_x_discrete(breaks=tmp$diagnosis_simple,labels=tmp$diagnosis_new) +
-  labs(x='',y='', color='# reads')
+  scale_x_discrete(breaks=tmp$Diagnosis_simple,labels=tmp$Diagnosis_new) +
+  labs(x='',y='', color='# Reads')
 ```
@@
                  dplyr::slice(chull(PC1, PC2)), alpha = 0.1, color=NA, aes(fill=status), show.legend=FALSE) +
-  geom_point(aes(color=diagnosis,alpha=dataset, shape=dataset), size=1) +
-  geom_line(aes(group=donor, color=diagnosis,alpha=dataset, shape=dataset)) +
-  #geom_label_repel(data=pcadat.combined %>% dplyr::filter(dataset=='testing') %>% dplyr::group_by(donor) %>% dplyr::slice_head(),
-  #                 aes(label=gsub("FB_","",donor)),size=2, segment.size=.25, label.padding = unit(.5,'mm'), show.legend=FALSE) +
+  geom_point(aes(color=Diagnosis,alpha=dataset, shape=dataset), size=1) +
+  geom_line(aes(group=Donor, color=Diagnosis,alpha=dataset, shape=dataset)) +
   my_theme() +
@@
                   clip='off') +
-  labs(x=paste0('PC 1 (',100*signif(pca$sdev[1]**2/sum(pca$sdev**2),3),'%)'),
-       y=paste0('PC 2 (',100*signif(pca$sdev[2]**2/sum(pca$sdev**2),3),'%)'),
+  labs(x='PC 1', #paste0('PC 1 (',100*signif(pca$sdev[1]**2/sum(pca$sdev**2),3),'%)'),
+       y='PC 2', #paste0('PC 2 (',100*signif(pca$sdev[2]**2/sum(pca$sdev**2),3),'%)'),
        color='',shape='',fill='',alpha='')
@@
                  dplyr::slice(chull(PC1, PC2)), alpha = 0.1, color=NA, aes(fill=status), show.legend=FALSE) +
-  geom_point(aes(color=diagnosis,alpha=dataset, shape=dataset), size=1) +
-  geom_line(aes(group=donor, color=diagnosis,alpha=dataset, shape=dataset)) +
-  # geom_label_repel(data=pcadat.combined %>% dplyr::filter(dataset=='validation') %>% dplyr::group_by(donor) %>% dplyr::slice_head(),
-  #                 aes(label=donor),size=2, segment.size=.25, label.padding = unit(.5,'mm'), show.legend=FALSE) +
+  geom_point(aes(color=Diagnosis,alpha=dataset, shape=dataset), size=1) +
+  geom_line(aes(group=Donor, color=Diagnosis,alpha=dataset, shape=dataset)) +
   my_theme() +
@@
 
-do ridge regression from training to testing or validation data (take out subclass def.)
-
-```{r Fig4H, fig.width=2. * fig_scale, fig.height=1.5 * fig_scale}
+do ridge regression from training to testing or validation data (take out subclass and atypical)
+
+```{r Fig4G, fig.width=2. * fig_scale, fig.height=1.5 * fig_scale}
 set.seed(1)
-train.here <- train & human.meta[train,'diagnosis'] != 'subclass'
-cv_model <- cv.glmnet(as.matrix(human.X[train.here,]), human.meta[train.here,'status']!='control', alpha = 0, family='binomial')
-mod <- glmnet(as.matrix(human.X[train.here,]), human.meta[train.here,'status']!='control', alpha = 0, family = "binomial",
+train.here <- train & !(human.meta[train,'Diagnosis'] %in% c('Subclass','Atypical'))
+cv_model <- cv.glmnet(as.matrix(human.X[train.here,]), human.meta[train.here,'status']!='Control', alpha = 0, family='binomial')
+mod <- glmnet(as.matrix(human.X[train.here,]), human.meta[train.here,'status']!='Control', alpha = 0, family = "binomial",
                 lambda = cv_model$lambda.min)
 pred_1 <- predict(mod, type = "response", newx=as.matrix(human.X[test,]))
-ROC_1 <- roc(human.meta[test,'status']!='control', pred_1)
+ROC_1 <- roc(human.meta[test,'status']!='Control', pred_1)
 
 pred_2 <- predict(mod, type = "response", newx=as.matrix(human.X[validation,]))
-ROC_2 <- roc(human.meta[validation,'status']!='control', pred_2)
+ROC_2 <- roc(human.meta[validation,'status']!='Control', pred_2)
 
@@
 sample.CV <- c()
-FB.samples <- row.names(FB.data.pooled)[FB.data.pooled$diagnosis != 'subclass']
+FB.samples <- row.names(FB.data.pooled)[!(FB.data.pooled$Diagnosis %in% c('Subclass','Atypical'))]
 for (k in seq(1,20)) {
@@
 
-  y <- FB.data.pooled[FB.samples, 'status'] != 'control'
+  y <- FB.data.pooled[FB.samples, 'status'] != 'Control'
   cv_model <- cv.glmnet(X[tr,], y[tr], alpha = 0, family='binomial')
@@
         axis.ticks = element_blank()) +
-  labs(x='',y='accuracy')
+  labs(x='',y='Accuracy')
 ```
@@
 
-```{r Fig4G, fig.width=5.5 * fig_scale, fig.height=5.5 * fig_scale}
-FB.samples <- row.names(FB.data.pooled)[grepl('FB',FB.data.pooled$donor)]
+```{r Fig4H, fig.width=5.5 * fig_scale, fig.height=5.5 * fig_scale}
+FB.samples <- row.names(FB.data.pooled)[grepl('FB',FB.data.pooled$Donor)]
 FB.X <- FB.data.pooled[FB.samples, selected.cols] %>%
@@
 
-row.names(FB.X) <- FB.data.pooled[FB.samples, c('donor','timepoint')] %>% 
-  dplyr::mutate(name=gsub('t','',gsub('FB_','',paste0(donor,'_',timepoint)))) %>% 
+row.names(FB.X) <- FB.data.pooled[FB.samples, c('Donor','Timepoint')] %>% 
+  dplyr::mutate(name=gsub('t','',gsub('FB_','',paste0(Donor,'_',Timepoint)))) %>% 
   dplyr::pull(name)
 colnames(FB.X) <- gsub('_downsampled|_switch','',colnames(FB.X))
-pca_hm <- pcadat.combined[gsub('pooled_(FB)_([0-9]*)','\\1\\2',FB.samples), c('PC1','PC2','dataset','donor')]
+pca_hm <- pcadat.combined[gsub('pooled_(FB)_([0-9]*)','\\1\\2',FB.samples), c('PC1','PC2','dataset','Donor')]
 row.names(pca_hm) <- row.names(FB.X)
@@
 FB.meta <- FB.data.pooled[FB.samples,] %>%
-  dplyr::mutate(sample_donor=gsub("FB_","FB",sample_donor),
-                dataset=ifelse(grepl('FB',donor),
-                               ifelse(sample_donor %in% FB_train_donors,'training','testing'),
+  dplyr::mutate(Sample_donor=gsub("FB_","FB",Sample_donor),
+                dataset=ifelse(grepl('FB',Donor),
+                               ifelse(Sample_donor %in% FB_train_donors,'training','testing'),
                                'validation')) %>%
-  dplyr::select(donor,sample_donor,diagnosis,dataset,sex,age,timepoint,
+  dplyr::select(Donor,Sample_donor,Diagnosis,dataset,Sex,Age,Timepoint,
                 splenomegaly,lymphoproliferation,pct_B,pct_IgA,pct_IgG,complication)
@@
 #                                                       dplyr::arrange(diagnosis) %>%
-#                                                       dplyr::mutate(sample_donor=gsub('pooled_FB_','',sample_donor)) %>%
-#                                                       dplyr::pull(sample_donor),
+#                                                       dplyr::mutate(Sample_donor=gsub('pooled_FB_','',Sample_donor)) %>%
+#                                                       dplyr::pull(Sample_donor),
 #                                                     row.names(FB.X))),agglo.FUN=mean)
 
-class_df <- data.frame(class=plyr::revalue(as.character(cutree(dend, k = 4)),
-                                           c('2'='cluster3',
-                                             '3'='cluster1',
-                                             '1'='cluster2',
-                                             '4'='cluster2')),
+class_df <- data.frame(Class=plyr::revalue(as.character(cutree(dend, k = 4)),
+                                           c('2'='Cluster3',
+                                             '3'='Cluster1',
+                                             '1'='Cluster2',
+                                             '4'='Cluster2')),
                        name=row.names(FB.X))
 
-class_colors <- c('cluster3'=diagnosis_colors[['control']],
-                  'cluster1'=cat_cols[['structural']],
-                  'cluster2'=cat_cols[['diversity']])
+class_colors <- c('Cluster3'=diagnosis_colors[['Control']],
+                  'Cluster1'=cat_cols[['Structural']],
+                  'Cluster2'=cat_cols[['Diversity']])
 
@@
                show_row_dend=TRUE,
-               top_annotation=HeatmapAnnotation(df=FB.meta[,c('diagnosis','sex','age',#'pct_B','pct_IgA','pct_IgG',
-                                                              #'splenomegaly','lymphoproliferation',
-                                                              #'lung','PnPS','Tetanus','Diphteria',
-                                                              #'complication','EUROclass',
-                                                              'donor','timepoint'),drop=FALSE] %>% 
+               top_annotation=HeatmapAnnotation(df=FB.meta[,c('Diagnosis','Sex','Age',#'pct_B','pct_IgA','pct_IgG',
+                                                              'Donor','Timepoint'),drop=FALSE] %>% 
                                                   tibble::remove_rownames() %>%
-                                                  dplyr::mutate(name=gsub('t','',gsub('FB_','',paste0(donor,'_',timepoint)))) %>%
+                                                  dplyr::mutate(name=gsub('t','',gsub('FB_','',paste0(Donor,'_',Timepoint))),
+                                                                Diagnosis=factor(Diagnosis,
+                                                                                 levels=c('Control','Subclass','Atypical',
+                                                                                          'CVID','ICOS','Kabuki',
+                                                                                          'AID','IgG4def','NFKB2','PI3K','TNFRSF13B'))) %>%
                                                   dplyr::left_join(class_df,by='name') %>%
-                                                  dplyr::select(-c(donor,timepoint)) %>%
+                                                  dplyr::select(-c(Donor,Timepoint)) %>%
                                                   tibble::column_to_rownames('name'),
@@
                                                 annotation_name_gp=gpar(fontsize=8),
-                                                col=list('diagnosis'=diagnosis_colors,
-                                                         'dataset'=c('training'='gray60','testing'='gray20',
-                                                                     'Charite'='red3','QP'='forestgreen'),
-                                                         'class'=class_colors,
-                                                         'EUROclass'=c('smB+21norm'='gray20',
-                                                                       'smB+21low'='orange3',
-                                                                       'smB-21norm'='lightblue3',
-                                                                       'smB-21low'='darkseagreen2'),
-                                                         'complication'=c('c'='coral3',
-                                                                          'io'='olivedrab4'),
-                                                         'age'=colorRamp2(c(20,50,80),c("yellow2","orange2","coral3")),
-                                                         'sex'=c('male'='mistyrose2',
+                                                col=list('Diagnosis'=diagnosis_colors,
+                                                         'Class'=class_colors,
+                                                         'Age'=colorRamp2(c(20,50,80),c("yellow2","orange2","coral3")),
+                                                         'Sex'=c('male'='mistyrose2',
                                                                  'female'='lightseagreen'))),
@@
          shape=guide_legend(ncol=2)) +
-  labs(x='abs. PC 1', y='abs. PC 2',
+  labs(x='Abs. PC 1', y='Abs. PC 2',
        color='',shape='',fill='',alpha='')
@@
          dplyr::left_join(class_df, by='name'), 
-       aes(x=PC1,y=PC2, color=class, shape=dataset)) +
+       aes(x=PC1,y=PC2, color=Class, shape=dataset)) +
   geom_point(size=1) +
   geom_point(data=pcadat.combined %>% dplyr::filter(dataset=='validation'),
-             aes(color=diagnosis), size=1) +
-  geom_line(aes(group=donor)) +
+             aes(color=Diagnosis), size=1) +
+  geom_line(aes(group=Donor)) +
   my_theme() +
@@
         aspect.ratio=pca$sdev[2]**2/pca$sdev[1]**2) +
-  labs(x=paste0('PC 1 (',100*signif(pca$sdev[1]**2/sum(pca$sdev**2),3),'%)'),
-       y=paste0('PC 2 (',100*signif(pca$sdev[2]**2/sum(pca$sdev**2),3),'%)'),
+  labs(x='PC 1', #paste0('PC 1 (',100*signif(pca$sdev[1]**2/sum(pca$sdev**2),3),'%)'),
+       y='PC 2', #paste0('PC 2 (',100*signif(pca$sdev[2]**2/sum(pca$sdev**2),3),'%)'),
        color='')
@@
 
-groups <- c('cluster1','cluster2','cluster3','control','BRCA1','ATM')
+groups <- c('Cluster1','Cluster2','Cluster3','Control','BRCA1','ATM')
 
@@
                    QP.data[,features.here]),
-             human.meta[,c('diagnosis','dataset','donor','timepoint')]) %>%
-  tibble::rownames_to_column('sample_donor') %>%
-  dplyr::mutate(name=gsub('t','',gsub('FB_','',paste0(donor,'_',timepoint)))) %>%
+             human.meta[,c('Diagnosis','dataset','Donor','Timepoint')]) %>%
+  tibble::rownames_to_column('Sample_donor') %>%
+  dplyr::mutate(name=gsub('t','',gsub('FB_','',paste0(Donor,'_',Timepoint)))) %>%
   dplyr::left_join(class_df,by='name') %>%
-  dplyr::mutate(group=ifelse(dataset=='validation',as.character(diagnosis),as.character(class))) %>%
+  dplyr::mutate(group=ifelse(dataset=='validation',as.character(Diagnosis),as.character(Class))) %>%
   dplyr::filter(group %in% groups) %>%
@@
   dplyr::group_by(group) %>%
-  dplyr::mutate(ng=n_distinct(donor),
+  dplyr::mutate(ng=n_distinct(Donor),
                 group_new=paste0(group,' (n=',ng,')')) %>%
@@
   geom_jitter(width=.1,height=0,size=.5) +
-  stat_compare_means(comparisons=list(c('cluster1','cluster3'),
-                                      c('cluster1','cluster2'),
-                                      c('cluster2','cluster3')),
+  stat_compare_means(comparisons=list(c('Cluster1','Cluster3'),
+                                      c('Cluster1','Cluster2'),
+                                      c('Cluster2','Cluster3')),
                      method='wilcox.test',label = "p.signif", hide.ns=TRUE, size=2, vjust=.5) +
-  stat_compare_means(comparisons=list(c('control','ATM'),
-                                      c('control','BRCA1')),
+  stat_compare_means(comparisons=list(c('Control','ATM'),
+                                      c('Control','BRCA1')),
                      method='wilcox.test',label = "p.signif", hide.ns=TRUE, size=2, vjust=.5) +
@@
         legend.position='none') +
-  scale_x_discrete(breaks=c('cluster1','cluster2','cluster3','control','ATM','BRCA1'),
-                   labels=c('cluster1','cluster2','cluster3','control','ATM','BRCA1')) +
-  labs(x='',y='', color='diagnosis')
+  scale_x_discrete(breaks=c('Cluster1','Cluster2','Cluster3','Control','ATM','BRCA1'),
+                   labels=c('Cluster1','Cluster2','Cluster3','Control','ATM','BRCA1')) +
+  labs(x='',y='', color='Diagnosis')
 ```
@@
 FB.data.pooled[FB.samples, ] %>%
-  dplyr::mutate(Tetanus=ifelse(status=='control','positive',Tetanus),
-                Diphteria=ifelse(status=='control','positive',Diphteria),
-                PnPS=ifelse(status=='control','positive',PnPS)) %>%
+  dplyr::mutate(Tetanus=ifelse(status=='Control','Positive',str_to_title(Tetanus)),
+                Diphteria=ifelse(status=='Control','Positive',str_to_title(Diphteria)),
+                PnPS=ifelse(status=='Control','Positive',str_to_title(PnPS))) %>%
   gather(feature, value, c('frac_clusters_SA1','frac_clusters_SG3')) %>%
   gather(vacc, vacc_status, c('Tetanus','Diphteria','PnPS')) %>%
-  dplyr::mutate(vacc_status=factor(ifelse(vacc_status=='','no data',vacc_status),
-                                   c('negative','positive','no data'))) %>%
+  dplyr::mutate(vacc_status=factor(ifelse(vacc_status=='','No data',vacc_status),
+                                   c('Negative','Positive','No data'))) %>%
   ggplot(aes(x=vacc_status, y=value, fill=vacc_status)) +
@@
   geom_jitter(aes(color=status),width=.1, size=.5) +
-  stat_compare_means(comparisons=list(c('negative','positive'),
-                                      c('positive','no data')),
+  stat_compare_means(comparisons=list(c('Negative','Positive'),
+                                      c('Positive','No data')),
                      method='wilcox.test', label = "p.signif", hide.ns=TRUE, size=2, vjust=0) +
@@
   my_theme() +
-  scale_fill_manual(values=c('negative'='red3','positive'='gray')) +
+  scale_fill_manual(values=c('Negative'='red3','Positive'='gray')) +
   scale_color_manual(values=diagnosis_colors) +
@@
   dplyr::left_join(FB.meta %>% tibble::rownames_to_column('sample'), by='sample') %>%
-  #dplyr::filter(diagnosis!='control') %>%
+  dplyr::filter(!(Diagnosis %in% c('Control','Subclass','Atypical'))) %>%
   dplyr::group_by(category) %>%
-  dplyr::mutate(#value=(value-mean(value[diagnosis=='control'])/sd(value[diagnosis=='control'])),
-                name=gsub('t','',gsub('FB_','',paste0(donor,'_',timepoint)))) %>%
+  dplyr::mutate(#value=(value-mean(value[Diagnosis=='control'])/sd(value[Diagnosis=='control'])),
+                name=gsub('t','',gsub('FB_','',paste0(Donor,'_',Timepoint)))) %>%
   dplyr::select(name,category,value) %>%
@@
 
-FB.meta.aggregated <- FB.meta[,c('diagnosis','sex','age',
-                                 'donor','timepoint'),drop=FALSE] %>% 
-  dplyr::filter(!(diagnosis %in% c('control','subclass'))) %>%
+FB.meta.aggregated <- FB.meta[,c('Diagnosis','Sex','Age',
+                                 'Donor','Timepoint'),drop=FALSE] %>% 
+  dplyr::filter(!(Diagnosis %in% c('control','subclass'))) %>%
   tibble::remove_rownames() %>%
-  dplyr::mutate(name=gsub('t','',gsub('FB_','',paste0(donor,'_',timepoint)))) %>%
+  dplyr::mutate(name=gsub('t','',gsub('FB_','',paste0(Donor,'_',Timepoint)))) %>%
   dplyr::left_join(class_df,by='name') %>%
-  dplyr::select(-c(donor,timepoint)) %>%
-  tibble::column_to_rownames('name')
-
-samples <- row.names(FB.meta.aggregated)
+  dplyr::select(-c(Donor,Timepoint)) %>%
+  tibble::column_to_rownames('name') %>%
+  dplyr::mutate(Diagnosis=factor(Diagnosis,
+                                 levels=c('Control','Subclass','Atypical',
+                                          'CVID','ICOS','Kabuki',
+                                          'AID','IgG4def','NFKB2','PI3K','TNFRSF13B')))
+
+samples <- row.names(FB.meta.aggregated[!(FB.meta.aggregated$Diagnosis  %in% c('Control','Subclass','Atypical')),])
 dend <- hclust(dist(FB.X.aggregated[samples,]))
@@
                                                       tibble::rownames_to_column('sample') %>%
-                                                      dplyr::arrange(class) %>%
+                                                      dplyr::arrange(Class) %>%
                                                       dplyr::pull(sample),
@@
                show_row_dend=TRUE,
-               top_annotation=HeatmapAnnotation(df=FB.meta.aggregated,
+               top_annotation=HeatmapAnnotation(df=FB.meta.aggregated[samples,],
                                                 show_annotation_name=TRUE,
@@
 <!--                                      rep('human',dim(dpca$y)[1])), -->
-<!--                            genotype=c(CH12.meta[,'genotype'], -->
+<!--                            Genotype=c(CH12.meta[,'Genotype'], -->
 <!--                                       human.meta[,'diagnosis']))) -->
 
-<!-- ggplot(dpcadat,aes(x=dPC1,y=dPC2, color=genotype)) + -->
+<!-- ggplot(dpcadat,aes(x=dPC1,y=dPC2, color=Genotype)) + -->
 <!--   geom_point(aes(shape=species),size=.5) + -->
 <!--   geom_polygon(data = dpcadat %>% -->
-<!--                  dplyr::group_by(genotype) %>% -->
-<!--                  dplyr::slice(chull(dPC1, dPC2)), alpha = 0.1, aes(fill=genotype), linewidth=.1) + -->
+<!--                  dplyr::group_by(Genotype) %>% -->
+<!--                  dplyr::slice(chull(dPC1, dPC2)), alpha = 0.1, aes(fill=Genotype), linewidth=.1) + -->
 <!--   my_theme() + -->
@@
 
-<!-- ggplot(dpcadat,aes(x=dPC1,y=dPC2, color=genotype)) + -->
+<!-- ggplot(dpcadat,aes(x=dPC1,y=dPC2, color=Genotype)) + -->
 <!--   geom_point(aes(shape=species),size=.5) + -->
 <!--   geom_polygon(data = dpcadat %>% -->
-<!--                  dplyr::group_by(genotype) %>% -->
-<!--                  dplyr::slice(chull(dPC1, dPC2)), alpha = 0.1, aes(fill=genotype), linewidth=.1) + -->
+<!--                  dplyr::group_by(Genotype) %>% -->
+<!--                  dplyr::slice(chull(dPC1, dPC2)), alpha = 0.1, aes(fill=Genotype), linewidth=.1) + -->
 <!--   my_theme() + -->
@@
 
-<!-- dpcadat[,'group'] <- as.character(dpcadat$genotype) -->
-<!-- dpcadat[FB.meta$sample_donor,'group'] <- as.character(class_df$class) -->
+<!-- dpcadat[,'group'] <- as.character(dpcadat$Genotype) -->
+<!-- dpcadat[FB.meta$Sample_donor,'group'] <- as.character(class_df$class) -->
 <!-- dpcadat[,'dataset'] <- 'CH12' -->
@@
 
-<!-- ```{r FigS4X4, fig.width=6.75 * fig_scale, fig.height=3.5 * fig_scale} -->
-<!-- features.here <- c('cluster_entropy_downsampled', -->
-<!--                    'top_clone_occupancy_downsampled', -->
-<!--                    'donor_complexity', -->
-<!--                    'donor_score_W', -->
-<!--                    'donor_score_WGCW', -->
-<!--                    'homology_fw', -->
-<!--                    'n_homology_switch', -->
-<!--                    'mean_length_SA', -->
-<!--                    'frac_breaks_intra', -->
-<!--                    'frac_breaks_inversions') -->
-
-<!-- tmp <- cbind(rbind(CH12.data[,features.here], -->
-<!--                    FB.data.pooled[,features.here], -->
-<!--                    QP.data[,features.here]), -->
-<!--              rbind(CH12.meta %>%  -->
-<!--                      dplyr::mutate(diagnosis=genotype, -->
-<!--                                    dataset='CH12') %>% -->
-<!--                      dplyr::select(diagnosis, dataset), -->
-<!--                    human.meta[,c('diagnosis','dataset')])) %>% -->
-<!--   dplyr::filter(diagnosis %in% c('WT','Trp53bp1 KO','Brca1 KO','Rif1 KO','Lig4 KO','control','CVID','TNFRSF13B','ATM','BRCA1')) %>% -->
-<!--   tibble::rownames_to_column('sample') %>% -->
-<!--   gather(col,value,features.here) %>% -->
-<!--   dplyr::mutate(col=factor(gsub("_downsampled|_switch",'',col), -->
-<!--                            levels=gsub("_downsampled|_switch","",features.here)), -->
-<!--                 group=droplevels(factor(diagnosis, levels=c(names(genotype_colors), names(diagnosis_colors))))) -->
-
-<!-- ggplot(tmp, aes(x=ifelse(dataset=='CH12',as.numeric(group),as.numeric(group) + 1),y=value,fill=group)) + -->
-<!--   geom_boxplot(outlier.shape=NA, lwd=.5) + -->
-<!--   geom_jitter(width=.1,height=0,size=.1) + -->
-<!--   stat_compare_means(comparisons=list(c(1,2), -->
-<!--                                       c(1,3), -->
-<!--                                       c(1,4), -->
-<!--                                       c(1,5)), -->
-<!--                      method='wilcox.test',label = "p.signif", hide.ns=TRUE, size=2, vjust=.5) + -->
-<!--   stat_compare_means(comparisons=list(c(7,8), -->
-<!--                                       c(7,9), -->
-<!--                                       c(7,10), -->
-<!--                                       c(7,11)), -->
-<!--                      method='wilcox.test',label = "p.signif", hide.ns=TRUE, size=2, vjust=.5) + -->
-<!--   facet_wrap(.~col,ncol=5,scales='free_y') + -->
-<!--   my_theme() + -->
-<!--   coord_cartesian(clip='off') + -->
-<!--   scale_fill_manual(values=c(diagnosis_colors,genotype_colors)) + -->
-<!--   theme(axis.text.x=element_text(angle=90,hjust=1,vjust=.5), -->
-<!--         legend.position='none') + -->
-<!--   scale_x_continuous(breaks=c(seq(1,5), seq(1,5) + 6), -->
-<!--                      labels=levels(tmp$group)) + -->
-<!--   labs(x='',y='', color='diagnosis') -->
-<!-- ``` -->
-
-<!-- ```{r FigS4X5, fig.width=3.5 * fig_scale, fig.height=3 * fig_scale} -->
-<!-- OC_sample_sheet <- read.csv('../sodar/2023_OC/a_2023_OC.txt', -->
-<!--                             sep='\t',header=1) %>% -->
-<!--   dplyr::left_join(read.csv('../sodar/2023_OC/s_2023_OC.txt',sep='\t',header=1), -->
-<!--                    by='Sample.Name') %>% -->
-<!--   dplyr::mutate(donor=as.character(Source.Name), -->
-<!--                 sample=Extract.Name.1, -->
-<!--                 batch=Parameter.Value.Batch., -->
-<!--                 QC=Parameter.Value.QC., -->
-<!--                 sequencing=Parameter.Value.Sequencing.method.) %>% -->
-<!--   dplyr::distinct(donor) -->
-
-<!-- OC.data <-read.csv('../human_samples/20240307_hg38_results/output/summary/20240307_human_samples_stats.csv', -->
-<!--                    header=1,row.names=1) %>% -->
-<!--   tibble::rownames_to_column('sample') %>% -->
-<!--   dplyr::filter(grepl('pooled_',sample)) %>% -->
-<!--   dplyr::mutate(sample_donor=gsub('pooled_','',sample)) %>% -->
-<!--   dplyr::inner_join(OC_sample_sheet, by=c('sample_donor'='donor')) %>% -->
-<!--   dplyr::filter(nreads_final > 500) %>% -->
-<!--   dplyr::mutate_all(~replace_na(.,0)) %>% -->
-<!--   dplyr::mutate(log10_nreads=log10(nreads_final), -->
-<!--                 log10_nclusters=log10(nclusters_final)) %>% -->
-<!--   tibble::column_to_rownames('sample_donor') -->
-
-<!-- OC.meta <- OC.data %>% -->
-<!--   tibble::rownames_to_column('donor') %>% -->
-<!--   dplyr::mutate(status=ifelse(grepl('BIH',sample),'OC','control'), -->
-<!--                 sample_donor=donor, -->
-<!--                 diagnosis=ifelse(donor %in% c('BIH116','BIH141','BIH152'), 'BRCA1', status), -->
-<!--                 timepoint='t1', -->
-<!--                 dataset='OC', -->
-<!--                 sex=NA, -->
-<!--                 age=NA, -->
-<!--                 complication=NA, -->
-<!--                 EUROclass=NA, -->
-<!--                 pct_B=NA, -->
-<!--                 pct_IgA=NA, -->
-<!--                 pct_IgG=NA) %>% -->
-<!--   dplyr::select(sample_donor,donor,diagnosis,status,timepoint,dataset,sex,age,complication,EUROclass,pct_B,pct_IgA,pct_IgG) %>% -->
-<!--   tibble::column_to_rownames('sample_donor') -->
-
-<!-- OC.X <- OC.data[row.names(OC.meta), selected.cols] %>% -->
-<!--   dplyr::mutate_if(is.numeric, scale) -->
-
-<!-- pcadat.all <- rbind(cbind(as.matrix(human.X[train,]) %*% pca$rotation, human.meta[train,]), -->
-<!--                     cbind(as.matrix(human.X[test,]) %*% pca$rotation, human.meta[test,]), -->
-<!--                     cbind(as.matrix(human.X[validation,]) %*% pca$rotation, human.meta[validation,]), -->
-<!--                     cbind(as.matrix(OC.X) %*% pca$rotation, OC.meta)) %>% -->
-<!--   dplyr::mutate(dataset=factor(dataset, levels=c('training','testing','validation','OC'))) -->
-
-<!-- ggplot(pcadat.all, aes(x=PC1,y=PC2)) + -->
-<!--   geom_polygon(data = pcadat.all %>% dplyr::filter(dataset=='training') %>% -->
-<!--                  dplyr::group_by(status) %>% -->
-<!--                  dplyr::slice(chull(PC1, PC2)), alpha = 0.1, color=NA, aes(fill=status), show.legend=FALSE) + -->
-<!--   geom_point(aes(color=diagnosis,alpha=dataset, shape=dataset), size=1) + -->
-<!--   geom_text_repel(data=pcadat.all %>% dplyr::filter(dataset=='OC'), aes(label=donor,color=diagnosis), size=2, segment.size=.25) + -->
-<!--   geom_line(aes(group=donor, color=diagnosis,alpha=dataset, shape=dataset)) + -->
-<!--   my_theme() + -->
-<!--   scale_fill_manual(values=c(diagnosis_colors,'OC'='deeppink'), guide='none') + -->
-<!--   scale_color_manual(values=c(diagnosis_colors,'OC'='deeppink')) + -->
-<!--   scale_alpha_manual(values=c('testing'=.25,'training'=.25,'validation'=.5,'OC'=1)) + -->
-<!--   scale_shape_manual(values=c('testing'=15,'training'=16,'validation'=17,'OC'=18)) + -->
-<!--   guides(color=guide_legend(override.aes = list(shape=15,size=2),ncol=3), -->
-<!--          shape=guide_legend(ncol=1), -->
-<!--          alpha=guide_legend(ncol=1)) + -->
-<!--   theme(legend.key.height=unit(1.5,'mm'), -->
-<!--         legend.key.spacing=unit(0,'mm'), -->
-<!--         aspect.ratio=pca$sdev[2]**2/pca$sdev[1]**2, -->
-<!--         legend.position='bottom', -->
-<!--         legend.box='horizontal') + -->
-<!--   coord_cartesian(xlim=c(-8,10), -->
-<!--                   ylim=c(-5,8), -->
-<!--                   clip='off') + -->
-<!--   labs(x=paste0('PC 1 (',100*signif(pca$sdev[1]**2/sum(pca$sdev**2),3),'%)'), -->
-<!--        y=paste0('PC 2 (',100*signif(pca$sdev[2]**2/sum(pca$sdev**2),3),'%)'), -->
-<!--        color='',shape='',fill='',alpha='') -->
-
-<!-- ``` -->
-
-<!-- ```{r FigS4X6, fig.width=3.5 * fig_scale, fig.height=3 * fig_scale} -->
-<!-- dpca <- dualPCA(human.X, OC.X) -->
-<!-- dpcadat <- cbind(rbind(dpca$x,dpca$y), -->
-<!--                  pcadat.all) -->
-
-<!-- ggplot(dpcadat, aes(x=dPC1,y=dPC2)) + -->
-<!--   geom_polygon(data = dpcadat %>% dplyr::filter(dataset=='training') %>% -->
-<!--                  dplyr::group_by(status) %>% -->
-<!--                  dplyr::slice(chull(dPC1, dPC2)), alpha = 0.1, color=NA, aes(fill=status), show.legend=FALSE) + -->
-<!--   geom_point(aes(color=diagnosis,alpha=dataset, shape=dataset), size=1) + -->
-<!--   geom_line(aes(group=donor, color=diagnosis,alpha=dataset, shape=dataset)) + -->
-<!--   geom_text_repel(data=dpcadat %>% dplyr::filter(dataset=='OC'), aes(label=donor, color=diagnosis), size=2, segment.size=.25) + -->
-<!--   my_theme() + -->
-<!--   scale_fill_manual(values=c(diagnosis_colors,'OC'='deeppink'), guide='none') + -->
-<!--   scale_color_manual(values=c(diagnosis_colors,'OC'='deeppink')) + -->
-<!--   scale_alpha_manual(values=c('testing'=.25,'training'=.25,'validation'=.5,'OC'=1)) + -->
-<!--   scale_shape_manual(values=c('testing'=15,'training'=16,'validation'=17,'OC'=18)) + -->
-<!--   guides(color=guide_legend(override.aes = list(shape=15,size=2),ncol=3), -->
-<!--          shape=guide_legend(ncol=1), -->
-<!--          alpha=guide_legend(ncol=1)) + -->
-<!--   theme(legend.key.height=unit(1.5,'mm'), -->
-<!--         legend.key.spacing=unit(0,'mm'), -->
-<!--         #aspect.ratio=dpca$sdev[2]**2/dpca$sdev[1]**2, -->
-<!--         legend.position='bottom', -->
-<!--         legend.box='horizontal') + -->
-<!--   coord_cartesian(clip='off') + -->
-<!--   labs(x=paste0('dPC 1 (',100*signif(dpca$sdev[1]**2/sum(dpca$sdev**2),3),'%)'), -->
-<!--        y=paste0('dPC 2 (',100*signif(dpca$sdev[2]**2/sum(dpca$sdev**2),3),'%)'), -->
-<!--        color='',shape='',fill='',alpha='') -->
-
-<!-- ``` -->
-
-<!-- ```{r FigS4X7, fig.width=3.5 * fig_scale, fig.height=3 * fig_scale} -->
-<!-- LB_sample_sheet <- read.csv('../sodar/2023_LB/a_2023_LB.txt', -->
-<!--                             sep='\t',header=1) %>% -->
-<!--   dplyr::left_join(read.csv('../sodar/2023_LB/s_2023_LB.txt',sep='\t',header=1), -->
-<!--                    by='Sample.Name') %>% -->
-<!--   dplyr::mutate(donor=as.character(Source.Name), -->
-<!--                 sample=Extract.Name.1, -->
-<!--                 batch=Parameter.Value.Batch., -->
-<!--                 QC=Parameter.Value.QC., -->
-<!--                 diagnosis=Characteristics.Diagnosis., -->
-<!--                 sequencing=Parameter.Value.Sequencing.method.) %>% -->
-<!--   dplyr::distinct(donor,diagnosis) -->
-
-<!-- LB.data <- read.csv('../human_samples/20240307_hg38_results/output/summary/20240307_human_samples_stats.csv', -->
-<!--                    header=1,row.names=1) %>% -->
-<!--   tibble::rownames_to_column('sample') %>% -->
-<!--   dplyr::filter(grepl('pooled_',sample)) %>% -->
-<!--   dplyr::mutate(sample_donor=gsub('pooled_','',sample)) %>% -->
-<!--   dplyr::inner_join(LB_sample_sheet, by=c('sample_donor'='donor')) %>% -->
-<!--   dplyr::filter(nreads_final > 500) %>% -->
-<!--   dplyr::mutate_all(~replace_na(.,0)) %>% -->
-<!--   dplyr::mutate(log10_nreads=log10(nreads_final), -->
-<!--                 log10_nclusters=log10(nclusters_final)) %>% -->
-<!--   tibble::column_to_rownames('sample_donor') -->
-
-<!-- LB.meta <- LB.data %>% -->
-<!--   tibble::rownames_to_column('donor') %>% -->
-<!--   dplyr::mutate(sample_donor=donor, -->
-<!--                 status=diagnosis, -->
-<!--                 timepoint='t1', -->
-<!--                 dataset='LB', -->
-<!--                 sex=NA, -->
-<!--                 age=NA, -->
-<!--                 complication=NA, -->
-<!--                 EUROclass=NA, -->
-<!--                 pct_B=NA, -->
-<!--                 pct_IgA=NA, -->
-<!--                 pct_IgG=NA) %>% -->
-<!--   dplyr::select(sample_donor,donor,diagnosis,status,timepoint,dataset,sex,age,complication,EUROclass,pct_B,pct_IgA,pct_IgG) %>% -->
-<!--   tibble::column_to_rownames('sample_donor') -->
-
-<!-- LB.X <- LB.data[row.names(LB.meta), selected.cols] %>% -->
-<!--   dplyr::mutate_if(is.numeric, scale) -->
-
-<!-- pcadat.all <- rbind(cbind(as.matrix(human.X[train,]) %*% pca$rotation, human.meta[train,]), -->
-<!--                     cbind(as.matrix(human.X[test,]) %*% pca$rotation, human.meta[test,]), -->
-<!--                     cbind(as.matrix(human.X[validation,]) %*% pca$rotation, human.meta[validation,]), -->
-<!--                     cbind(as.matrix(LB.X) %*% pca$rotation, LB.meta)) %>% -->
-<!--   dplyr::mutate(dataset=factor(dataset, levels=c('training','testing','validation','LB'))) -->
-
-<!-- ggplot(pcadat.all, aes(x=PC1,y=PC2)) + -->
-<!--   geom_polygon(data = pcadat.all %>% dplyr::filter(dataset=='training') %>% -->
-<!--                  dplyr::group_by(status) %>% -->
-<!--                  dplyr::slice(chull(PC1, PC2)), alpha = 0.1, color=NA, aes(fill=status), show.legend=FALSE) + -->
-<!--   geom_point(aes(color=diagnosis,alpha=dataset, shape=dataset), size=1) + -->
-<!--   geom_text_repel(data=pcadat.all %>% dplyr::filter(dataset=='LB'), aes(label=donor,color=diagnosis), size=2, segment.size=.25) + -->
-<!--   geom_line(aes(group=donor, color=diagnosis,alpha=dataset, shape=dataset)) + -->
-<!--   my_theme() + -->
-<!--   scale_fill_manual(values=c(diagnosis_colors,'unknown'='gray'), guide='none') + -->
-<!--   scale_color_manual(values=c(diagnosis_colors,'unknown'='gray')) + -->
-<!--   scale_alpha_manual(values=c('testing'=.25,'training'=.25,'validation'=.5,'LB'=1)) + -->
-<!--   scale_shape_manual(values=c('testing'=15,'training'=16,'validation'=17,'LB'=18)) + -->
-<!--   guides(color=guide_legend(override.aes = list(shape=15,size=2),ncol=3), -->
-<!--          shape=guide_legend(ncol=1), -->
-<!--          alpha=guide_legend(ncol=1)) + -->
-<!--   theme(legend.key.height=unit(1.5,'mm'), -->
-<!--         legend.key.spacing=unit(0,'mm'), -->
-<!--         aspect.ratio=pca$sdev[2]**2/pca$sdev[1]**2, -->
-<!--         legend.position='bottom', -->
-<!--         legend.box='horizontal') + -->
-<!--   coord_cartesian(xlim=c(-8,10), -->
-<!--                   ylim=c(-5,8), -->
-<!--                   clip='off') + -->
-<!--   labs(x=paste0('PC 1 (',100*signif(pca$sdev[1]**2/sum(pca$sdev**2),3),'%)'), -->
-<!--        y=paste0('PC 2 (',100*signif(pca$sdev[2]**2/sum(pca$sdev**2),3),'%)'), -->
-<!--        color='',shape='',fill='',alpha='') -->
-
-<!-- ``` -->
-
-<!-- ```{r FigS4X8, fig.width=3.5 * fig_scale, fig.height=3 * fig_scale} -->
-<!-- dpca <- dualPCA(human.X, LB.X) -->
-<!-- dpcadat <- cbind(rbind(dpca$x,dpca$y), -->
-<!--                  pcadat.all) -->
-
-<!-- ggplot(dpcadat, aes(x=dPC1,y=dPC2)) + -->
-<!--   geom_polygon(data = dpcadat %>% dplyr::filter(dataset=='training') %>% -->
-<!--                  dplyr::group_by(status) %>% -->
-<!--                  dplyr::slice(chull(dPC1, dPC2)), alpha = 0.1, color=NA, aes(fill=status), show.legend=FALSE) + -->
-<!--   geom_point(aes(color=diagnosis,alpha=dataset, shape=dataset), size=1) + -->
-<!--   geom_line(aes(group=donor, color=diagnosis,alpha=dataset, shape=dataset)) + -->
-<!--   geom_text_repel(data=dpcadat %>% dplyr::filter(dataset=='LB'), aes(label=donor, color=diagnosis), size=2, segment.size=.25) + -->
-<!--   my_theme() + -->
-<!--   scale_fill_manual(values=c(diagnosis_colors,'unknown'='gray'), guide='none') + -->
-<!--   scale_color_manual(values=c(diagnosis_colors,'unknown'='gray')) + -->
-<!--   scale_alpha_manual(values=c('testing'=.25,'training'=.25,'validation'=.5,'LB'=1)) + -->
-<!--   scale_shape_manual(values=c('testing'=15,'training'=16,'validation'=17,'LB'=18)) + -->
-<!--   guides(color=guide_legend(override.aes = list(shape=15,size=2),ncol=3), -->
-<!--          shape=guide_legend(ncol=1), -->
-<!--          alpha=guide_legend(ncol=1)) + -->
-<!--   theme(legend.key.height=unit(1.5,'mm'), -->
-<!--         legend.key.spacing=unit(0,'mm'), -->
-<!--         #aspect.ratio=dpca$sdev[2]**2/dpca$sdev[1]**2, -->
-<!--         legend.position='bottom', -->
-<!--         legend.box='horizontal') + -->
-<!--   coord_cartesian(clip='off') + -->
-<!--   labs(x=paste0('dPC 1 (',100*signif(dpca$sdev[1]**2/sum(dpca$sdev**2),3),'%)'), -->
-<!--        y=paste0('dPC 2 (',100*signif(dpca$sdev[2]**2/sum(dpca$sdev**2),3),'%)'), -->
-<!--        color='',shape='',fill='',alpha='') -->
-
-<!-- ``` -->
+```{r FigS4X4, fig.width=6.75 * fig_scale, fig.height=3.5 * fig_scale}
+features.here <- c('cluster_entropy_downsampled',
+                   'top_clone_occupancy_downsampled',
+                   'mean_cluster_size_downsampled',
+                   'donor_complexity',
+                   'donor_score_W',
+                   'donor_score_WGCW',
+                   'homology_fw',
+                   'n_homology_switch',
+                   'mean_length_SA',
+                   'frac_breaks_intra',
+                   'frac_breaks_inversions',
+                   'frac_breaks_duplications')
+
+tmp <- cbind(rbind(CH12.data[,features.here],
+                   FB.data.pooled[,features.here],
+                   QP.data[,features.here]),
+             rbind(CH12.meta %>%
+                     dplyr::mutate(Diagnosis=Genotype,
+                                   dataset='CH12') %>%
+                     dplyr::select(Diagnosis, dataset),
+                   human.meta[,c('Diagnosis','dataset')])) %>%
+  dplyr::filter(Diagnosis %in% c('WT','Trp53bp1 KO','Brca1 KO','Rif1 KO','Lig4 KO','Control','CVID','TNFRSF13B','ATM','BRCA1')) %>%
+  tibble::rownames_to_column('sample') %>%
+  gather(col,value,features.here) %>%
+  dplyr::mutate(col=factor(gsub("_downsampled|_switch",'',col),
+                           levels=gsub("_downsampled|_switch","",features.here)),
+                group=droplevels(factor(Diagnosis, levels=c(names(genotype_colors), names(diagnosis_colors)))))
+
+ggplot(tmp, aes(x=ifelse(dataset=='CH12',as.numeric(group),as.numeric(group) + 1),y=value,fill=group)) +
+  geom_boxplot(outlier.shape=NA, lwd=.5) +
+  geom_jitter(width=.1,height=0,size=.1) +
+  stat_compare_means(comparisons=list(c(1,2),
+                                      c(1,3),
+                                      c(1,4),
+                                      c(1,5)),
+                     method='wilcox.test',label = "p.signif", hide.ns=TRUE, size=2, vjust=.5) +
+  stat_compare_means(comparisons=list(c(7,8),
+                                      c(7,9),
+                                      c(7,10),
+                                      c(7,11)),
+                     method='wilcox.test',label = "p.signif", hide.ns=TRUE, size=2, vjust=.5) +
+  facet_wrap(.~col,ncol=6,scales='free_y') +
+  my_theme() +
+  coord_cartesian(clip='off') +
+  scale_fill_manual(values=c(diagnosis_colors,genotype_colors)) +
+  theme(axis.text.x=element_text(angle=90,hjust=1,vjust=.5),
+        legend.position='none') +
+  scale_x_continuous(breaks=c(seq(1,5), seq(1,5) + 6),
+                     labels=levels(tmp$group)) +
+  labs(x='',y='', color='Diagnosis')
+```
 
@@
 <!--   dplyr::left_join(FB.meta %>% tibble::rownames_to_column('sample'), by='sample') %>% -->
-<!--   dplyr::mutate(name=gsub('t','',gsub('FB_','',paste0(donor,'_',timepoint)))) %>% -->
+<!--   dplyr::mutate(name=gsub('t','',gsub('FB_','',paste0(Donor,'_',Timepoint)))) %>% -->
 <!--   dplyr::left_join(class_df,by='name') %>% -->
@@
 <!--         legend.position='none') + -->
-<!--   labs(x='',y='category score', color='diagnosis') -->
+<!--   labs(x='',y='category score', color='Diagnosis') -->
 <!-- ``` -->

### 2025-01-17 prepare_input.R

@@ -2,7 +2,7 @@ library(dplyr)
 library(tidyverse)
 library(scales)
 
-columns <- list('diversity'=c('nclusters_final_downsampled',
+columns <- list('Diversity'=c('nclusters_final_downsampled',
                               'nclusters_eff_downsampled',
                               'mean_cluster_size_downsampled',
                               'std_cluster_size_downsampled',
@@ -11,17 +11,17 @@ columns <- list('diversity'=c('nclusters_final_downsampled',
                               'cluster_inverse_simpson_downsampled',
                               'top_clone_occupancy_downsampled',
                               'big_clones_occupancy_downsampled'),
-                'isotypes'=c('alpha_ratio_clusters',
+                'Isotypes'=c('alpha_ratio_clusters',
                              'frac_clusters_SA1','frac_clusters_SA2','frac_clusters_SE',
                              'frac_clusters_SG1','frac_clusters_SG2','frac_clusters_SG3',
                              'frac_clusters_SG4','frac_clusters_SM',
                              'frac_clusters_Sa','frac_clusters_Sg3',
                              'frac_clusters_Sg2b','frac_clusters_Sg2c',
                              'frac_clusters_Sm','frac_clusters_Sg1'),
-                'structural'=c('insert_frequency',
+                'Structural'=c('insert_frequency',
                                'frac_breaks_inversions','frac_breaks_duplications',
                                'median_duplication_size','median_inversion_size'),
-                'context'=c('homology_fw','homology_rv',
+                'Context'=c('homology_fw','homology_rv',
                             'donor_score_S', 'donor_score_W', 
                             'donor_score_WGCW', 'donor_score_GAGCT',
                             'donor_score_GGGST',
@@ -33,7 +33,7 @@ columns <- list('diversity'=c('nclusters_final_downsampled',
                             'n_homology_switch', 
                             'n_untemplated_switch',
                             'frac_blunt'),
-                'breaks'=c('mean_length',
+                'Breaks'=c('mean_length',
                            'std_length',
                            'mean_GC',
                            'std_GC',
@@ -54,7 +54,7 @@ columns <- list('diversity'=c('nclusters_final_downsampled',
                            'frac_breaks_SM_SM','frac_breaks_SM_SG',
                            'frac_breaks_SG_SG','frac_breaks_SM_SA',
                            'frac_breaks_SG_SA','frac_breaks_SA_SA'),
[…]

### 2025-01-21 paper_figures.Rmd
update figures

@@
                       'IgG4def'='burlywood4','TNFRSF13B'='palegreen4',
-                      'Atypical'='gray40',
+                      'Atypical'='steelblue4',
                       'CVID-like'='#BC80BD','AID'='darkgoldenrod3',
@@
 
-show VDJ results: # clones depending on # of input cells
-
-```{r Fig1F, fig.width=1.4 * fig_scale, fig.height=1.4 * fig_scale}
+show VDJ results: isotype frequencies
+
+```{r Fig1F, fig.width=2.4 * fig_scale, fig.height=1.4 * fig_scale}
 # bulk VDJ results for Fig. 1
@@
 
-vdj_df %>% 
-  dplyr::filter(metric=='nclusters_final') %>%
-  ggplot(aes(x=factor(gsub('000$','',ncells), levels=c('50','100')), y=value)) +
-  geom_boxplot(outlier.shape=NA, lwd=.5) +
-  geom_point(aes(color=gsub('Donor ','',Donor)),size=.5) +
-  facet_wrap(~method, ncol=2, scales='free_y') + 
-  my_theme() +
-  scale_color_brewer(palette='Set1') +
-  theme(strip.clip='off',
-        legend.position='none',
-        legend.background =element_rect(fill=NA)) +
-  labs(y='# Clones',x='# Cells / 1000',color='')
-```
-
-isotype ratios in VDJ vs. SWIBRID
-
-```{r Fig1G, fig.width=1.4 * fig_scale,fig.height=1.4 * fig_scale}
+# vdj_df %>% 
+#   dplyr::filter(metric=='nclusters_final') %>%
+#   ggplot(aes(x=factor(gsub('000$','',ncells), levels=c('50','100')), y=value)) +
+#   geom_boxplot(outlier.shape=NA, lwd=.5) +
+#   geom_point(aes(color=gsub('Donor ','',Donor)),size=.5) +
+#   facet_wrap(~method, ncol=2, scales='free_y') + 
+#   my_theme() +
+#   scale_color_brewer(palette='Set1') +
+#   theme(strip.clip='off',
+#         legend.position='none',
+#         legend.background =element_rect(fill=NA)) +
+#   labs(y='# Clones',x='# Cells / 1000',color='')
+
 vdj_df %>% 
@@
 cols.here.mouse <- c('frac_breaks_inversions',
-                     'n_homology_switch',
+                     'frac_breaks_intra',
                      'mean_cluster_size_downsampled')
@@
 cohort.data %>%
-  ggplot(aes(x=Cohort, y=value, fill=Cohort)) + 
-  geom_boxplot(position=position_dodge(width=.75), width=.7, outlier.shape=NA, lwd=.25, fill='gray80') + 
+  ggplot(aes(x=Cohort, y=value, fill=species)) + 
+  geom_boxplot(position=position_dodge(width=.75), width=.7, outlier.shape=NA, lwd=.25) + 
   geom_point(position=position_jitterdodge(jitter.width=.2, dodge.width=.75, jitter.height = 0), size=.5) + 
@@
   coord_cartesian(clip='off') +
-  #annotate('text',x=5.5,y=-2,label='Vincendeau et al.', size=2) +
+  #annotate('text',x=1.5,y=Inf,label='SWIBRID', size=2) +
+  #annotate('text',x=5.5,y=Inf,label='Vincendeau et al.', size=2) +
   facet_wrap(~variable, ncol=3, switch='y', scales='free_y') +
   theme(axis.text.x=element_text(angle=45, hjust=1, vjust=1),
-        legend.position='none') +
+        legend.position=c(.2,.99)) +
+  scale_fill_manual(values=c('mouse tissue'='gray80','Vincendeau'='#cca9e6')) +
   scale_x_discrete(breaks=c('lymph node','splenocytes','WT','Xlf','Shld1','Shld1_Xlf'),
@@
                                                 simple_anno_size=unit(2,'mm'),
-                                                annotation_name_gp=gpar(fontsize=8),
+                                                annotation_name_gp=gpar(fontsize=7),
                                                 annotation_legend_param=list(labels_gp=gpar(fontsize=7),
-                                                                             title_gp=gpar(fontsize=8),
+                                                                             title_gp=gpar(fontsize=7),
                                                                              legend_width=unit(2,'mm'),
@@
                heatmap_legend_param=list(labels_gp=gpar(fontsize=7),
-                                         title_gp=gpar(fontsize=8),
+                                         title_gp=gpar(fontsize=7),
                                          legend_direction = "horizontal", 
@@
   my_theme() +
-  scale_y_discrete(labels=ifelse(grepl('\\.20|\\.27|\\.14|\\.01|\\.04',tmp$sample),"*","")) +
+  #scale_y_discrete(labels=ifelse(grepl('\\.20|\\.27|\\.14|\\.01|\\.04',tmp$sample),"*","")) +
   theme(legend.position='none',
         axis.ticks.y = element_blank(),
-        #axis.text.y = element_blank(),
+        axis.text.y = element_blank(),
         axis.text.x=element_text(angle=90, hjust=1, vjust=.5)) +
-  labs(x='',y='36 new samples (* 25ng IL4)')
+  labs(x='',y='36 new samples') # (* 25ng IL4)')
 ```
@@
 
-do ridge regression from training to testing or validation data (take out subclass and atypical)
+do ridge regression from training to testing or validation data (take out subclass and atypical, as well as biol replicates of training)
 
@@
 set.seed(1)
-train.here <- train & !(human.meta[train,'Diagnosis'] %in% c('Subclass','Atypical'))
+train.here <- train & !(human.meta[,'Diagnosis'] %in% c('Subclass','Atypical')) 
+test.here <- test & !(human.meta[, 'Donor'] %in% FB_train_donors)
 cv_model <- cv.glmnet(as.matrix(human.X[train.here,]), human.meta[train.here,'status']!='Control', alpha = 0, family='binomial')
@@
                 lambda = cv_model$lambda.min)
-pred_1 <- predict(mod, type = "response", newx=as.matrix(human.X[test,]))
-ROC_1 <- roc(human.meta[test,'status']!='Control', pred_1)
+pred_1 <- predict(mod, type = "response", newx=as.matrix(human.X[test.here,]))
+ROC_1 <- roc(human.meta[test.here,'status']!='Control', pred_1)
 
@@
   guides(color=guide_legend(ncol=1)) +
-  scale_color_manual(values=c('black','blue3')) +
+  scale_color_manual(values=c('steelblue4','lightsteelblue')) +
   theme(legend.position=c(.5,.2)) +
@@
 
-```{r Fig4H, fig.width=5.5 * fig_scale, fig.height=5.5 * fig_scale}
+```{r Fig4H, fig.width=5 * fig_scale, fig.height=5.5 * fig_scale}
 FB.samples <- row.names(FB.data.pooled)[grepl('FB',FB.data.pooled$Donor)]
@@
 class_df <- data.frame(Class=plyr::revalue(as.character(cutree(dend, k = 4)),
-                                           c('2'='Cluster3',
-                                             '3'='Cluster1',
-                                             '1'='Cluster2',
-                                             '4'='Cluster2')),
+                                           c('2'='Cluster1',
+                                             '3'='Cluster2',
+                                             '1'='Cluster3',
+                                             '4'='Cluster3')),
                        name=row.names(FB.X))
 
-class_colors <- c('Cluster3'=diagnosis_colors[['Control']],
-                  'Cluster1'=cat_cols[['Structural']],
-                  'Cluster2'=cat_cols[['Diversity']])
+class_colors <- c('Cluster1'=diagnosis_colors[['Control']],
+                  'Cluster2'=cat_cols[['Structural']],
+                  'Cluster3'=cat_cols[['Diversity']])
 
@@
                                                                              labels_gp=gpar(fontsize=7),
+                                                                             legend_height = unit(5, "mm"),
+                                                                             legend_width = unit(3, "mm"),
                                                                              grid_height = unit(3, "mm"),
                                                                              grid_width = unit(3, "mm")),
-                                                annotation_name_gp=gpar(fontsize=8),
+                                                annotation_name_gp=gpar(fontsize=7),
                                                 col=list('Diagnosis'=diagnosis_colors,
@@
                heatmap_legend_param=list(labels_gp=gpar(fontsize=7),
-                                         title_gp=gpar(fontsize=8),
-                                         legend_height=unit(10,'mm')),
+                                         title_gp=gpar(fontsize=7),
+                                         at=c(-2,0,2),
+                                         grid_height=unit(4, 'mm'),
+                                         grid_width=unit(3, 'mm'),
+                                         legend_height=unit(5,'mm'),
+                                         legend_width=unit(3,'mm')),
                row_title=NULL,
@@
                heatmap_legend_param=list(labels_gp=gpar(fontsize=7),
-                                         title_gp=gpar(fontsize=8),
-                                         legend_height = unit(10, "mm")),
+                                         title_gp=gpar(fontsize=7),
+                                         grid_height=unit(5, 'mm'),
+                                         grid_width=unit(3, 'mm'),
+                                         legend_height = unit(5, "mm"),
+                                         legend_width = unit(3, "mm")),
                row_title=NULL,
@@
 features.here <- c('frac_clusters_SG1',
-                   'frac_clusters_SG2',
                    'mean_cluster_size_downsampled',
@@
 
-ggplot(tmp, aes(x=group,y=value,fill=group)) +
+ggplot(tmp, aes(x=ifelse(grepl('Cluster',group),as.numeric(group),1+as.numeric(group)),
+                y=value,fill=group)) +
   geom_boxplot(outlier.shape=NA, lwd=.5) +
   geom_jitter(width=.1,height=0,size=.5) +
-  stat_compare_means(comparisons=list(c('Cluster1','Cluster3'),
-                                      c('Cluster1','Cluster2'),
-                                      c('Cluster2','Cluster3')),
-                     method='wilcox.test',label = "p.signif", hide.ns=TRUE, size=2, vjust=.5) +
-  stat_compare_means(comparisons=list(c('Control','ATM'),
-                                      c('Control','BRCA1')),
-                     method='wilcox.test',label = "p.signif", hide.ns=TRUE, size=2, vjust=.5) +
+  #annotate('text',x=2,y=Inf,label='', size=2) +
+  #annotate('text',x=6,y=Inf,label='validation', size=2) +
+  # stat_compare_means(comparisons=list(c('Cluster1','Cluster3'),
+  #                                     c('Cluster1','Cluster2'),
+  #                                     c('Cluster2','Cluster3')),
+  #                    method='wilcox.test',label = "p.signif", size=2, vjust=.5) +
+  # stat_compare_means(comparisons=list(c('Control','ATM'),
+  #                                     c('Control','BRCA1')),
+  #                    method='wilcox.test',label = "p.signif", size=2, vjust=.5) +
+  stat_compare_means(comparisons=list(c(1,2),c(1,3),c(2,3)),
+                     method='wilcox.test',label = "p.signif", size=2, vjust=.5) +
+  stat_compare_means(comparisons=list(c(5,6),c(5,7),c(6,7)),
+                     method='wilcox.test',label = "p.signif", size=2, vjust=.5) +
   facet_wrap(.~col,ncol=length(features.here),scales='free_y') +
@@
         legend.position='none') +
-  scale_x_discrete(breaks=c('Cluster1','Cluster2','Cluster3','Control','ATM','BRCA1'),
+  scale_x_continuous(breaks=c(1,2,3,5,6,7),
                      labels=c('Cluster1','Cluster2','Cluster3','Control','ATM','BRCA1')) +
@@
                                       c('Positive','No data')),
-                     method='wilcox.test', label = "p.signif", hide.ns=TRUE, size=2, vjust=0) +
+                     method='wilcox.test', label = "p.signif", size=2, vjust=0) +
   facet_grid(feature ~ vacc, switch='y') + 
@@
 
-```{r FigS4F, fig.width=5.5 * fig_scale, fig.height=3.5 * fig_scale}
+```{r FigS4F, fig.width=4.5 * fig_scale, fig.height=3.5 * fig_scale}
 FB.X.aggregated <-  FB.data.pooled[FB.samples, selected.cols] %>%
@@
                                  'Donor','Timepoint'),drop=FALSE] %>% 
-  dplyr::filter(!(Diagnosis %in% c('control','subclass'))) %>%
+  dplyr::filter(!(Diagnosis %in% c('Control','Subclass','Atypical'))) %>%
   tibble::remove_rownames() %>%
@@
   dplyr::mutate(Diagnosis=factor(Diagnosis,
-                                 levels=c('Control','Subclass','Atypical',
-                                          'CVID','ICOS','Kabuki',
+                                 levels=c('CVID','ICOS','Kabuki',
                                           'AID','IgG4def','NFKB2','PI3K','TNFRSF13B')))
@@
                                                                              labels_gp=gpar(fontsize=7),
+                                                                             legend_height = unit(5, "mm"),
+                                                                             legend_width = unit(3, "mm"),
+                                                                             title_position = "leftcenter-rot",
                                                                              grid_height = unit(3, "mm"),
                                                                              grid_width = unit(3, "mm")),
-                                                annotation_name_gp=gpar(fontsize=8),
+                                                annotation_name_gp=gpar(fontsize=7),
                                                 col=list('Diagnosis'=diagnosis_colors,
@@
                heatmap_legend_param=list(labels_gp=gpar(fontsize=7),
-                                         title_gp=gpar(fontsize=8),
-                                         legend_height=unit(10,'mm')),
+                                         title_gp=gpar(fontsize=7),
+                                         at=c(0,1,2),
+                                         title_position = "leftcenter-rot",
+                                         grid_height=unit(4, 'mm'),
+                                         grid_width=unit(3, 'mm'),
+                                         legend_height=unit(5,'mm'),
+                                         legend_width=unit(3,'mm')),
                row_title=NULL,
-               name='category score')
+               name='Category score')
 

### 2025-01-21 swibrid_runs/human/combine_replicates.sh

@@ -3,8 +3,6 @@
 donors=$(cut -f 1 ../../sodar/2023_FB/a_2023_FB.txt | grep -v Sample | sort -u )
 
 for donor in ${donors}; do
-	samples1=$(cut -f 1,17 ../../sodar/2023_FB/a_2023_FB.txt | grep -v Extract | grep ${donor} | cut -f 2 | sort -u | tr '\n' ',' | sed 's/,$//')
-	samples2=$(grep -v FAIL ../../sodar/2023_FB/a_2023_FB.txt | cut -f 1,17 | grep -v Extract | grep ${donor} | cut -f 2 | sort -u | tr '\n' ',' | sed 's/,$//')
-	[[ ${samples1} != ${samples2} ]] && echo ${samples1} ${samples2} #swibrid combine_replicates -s ${samples2} -c pooled_${donor} --nmax 50000
-	#[[ ! -s input/pooled_${donor}_info.csv ]] && echo ${donor} ${samples} && swibrid combine_replicates -s ${samples} -c pooled_${donor} --nmax 50000
+	samples=$(cut -f 1,17 ../../sodar/2023_FB/a_2023_FB.txt | grep -v Extract | grep ${donor} | cut -f 2 | sort -u | tr '\n' ',' | sed 's/,$//')
+	[[ ! -s input/pooled_${donor}_info.csv ]] && echo ${donor} ${samples} && swibrid combine_replicates -s ${samples} -c pooled_${donor} --nmax 50000
 done

### 2025-01-21 swibrid_runs/human/plot_bars.sh

@@ -1,4 +1,4 @@
 #!/bin/bash -e
 
-python plot_bars.py --samples "20231211_rFP_01_downsampled,20231211_rFP_29_downsampled,20211122_85_n500_1_downsampled" --figure paper_figures/20240320_bar_plots_downsampled_isotype.pdf
+python plot_bars.py --samples "20231211_rFP_01_downsampled,20250113_SKA1_downsampled,20211122_85_n500_1_downsampled"" --figure bar_plots_downsampled_isotype.pdf
 

### 2025-01-21 swibrid_runs/human/plot_breakpoints.sh

@@ -1,9 +1,5 @@
 #!/bin/bash -e
 
-#sample=$1
-mkdir -p paper_figures
-
-#for sample in 20210422_plasmid_downsampled 20210618_FakePoly_downsampled 20211122_85_n500_1_downsampled 20220329_86_100000_1_downsampled 20231020_HD_K_downsampled 20231020_IRF4_downsampled 20231026_HD_K_downsampled 20231026_IRF4_downsampled 20231211_rFP_01_downsampled 20231211_rFP_04_downsampled 20231211_rFP_15_downsampled 20231211_rFP_22_downsampled 20231211_rFP_29_downsampled; do 
-for sample in 20231211_rFP_01_downsampled 20231211_rFP_29_downsampled 20211122_85_n500_1_downsampled; do
-	swibrid get_breakpoint_stats -g pipeline/${sample}/${sample}_gaps.npz -c pipeline/${sample}/${sample}_clustering.csv -a pipeline/${sample}/${sample}_cluster_analysis.csv -b 50 --max_gap 75 --switch_coords chr14:105583000-105872000:- --switch_annotation hg38_switch_regions.bed -o paper_figures/${sample}_breakpoints.csv -p paper_figures/${sample}_breakpoints.pdf --weights cluster
+for sample in 20231211_rFP_01_downsampled 20250113_SKA1_downsampled 20211122_85_n500_1_downsampled; do
+	swibrid get_breakpoint_stats -g pipeline/${sample}/${sample}_gaps.npz -c pipeline/${sample}/${sample}_clustering.csv -a pipeline/${sample}/${sample}_cluster_analysis.csv -b 50 --max_gap 75 --switch_coords chr14:105583000-105872000:- --switch_annotation hg38_switch_regions.bed -o ${sample}_breakpoints.csv -p ${sample}_breakpoints.pdf --weights cluster
 done

### 2025-01-21 swibrid_runs/human/plot_circles.sh

@@ -1,5 +1,4 @@
 #!/bin/bash -e
 
-python plot_circles.py --samples "20231211_rFP_01_downsampled,20231211_rFP_29_downsampled,20211122_85_n500_1_downsampled" --ncols 3 --figure paper_figures/20240320_bubble_plots_downsampled_cluster.pdf --color cluster
-#python plot_circles.py --samples "20231211_rFP_01_downsampled,20231211_rFP_29_downsampled,20211122_85_n500_1_downsampled" --ncols 3 --figure paper_figures/20240320_bubble_plots_downsampled_isotype.pdf --color isotype
+python plot_circles.py --samples "20231211_rFP_01_downsampled,20250113_SKA1_downsampled,20211122_85_n500_1_downsampled" --ncols 3 --figure bubble_plots_downsampled_cluster.pdf --color cluster
 

### 2025-01-21 swibrid_runs/human/plot_clustering.sh

@@ -1,13 +1,9 @@
 #!/bin/bash -e
 
-#sample=$1
-mkdir -p paper_figures
-
-#for sample in 20210422_plasmid_downsampled 20210618_FakePoly_downsampled 20211122_85_n500_1_downsampled 20220329_86_100000_1_downsampled 20231020_HD_K_downsampled 20231020_IRF4_downsampled 20231026_HD_K_downsampled 20231026_IRF4_downsampled 20231211_rFP_01_downsampled 20231211_rFP_04_downsampled 20231211_rFP_15_downsampled 20231211_rFP_22_downsampled 20231211_rFP_29_downsampled; do 
-for sample in 20231211_rFP_01_downsampled 20231211_rFP_29_downsampled 20211122_85_n500_1_downsampled 20210422_plasmid_downsampled; do
-	swibrid plot_clustering --msa pipeline/${sample}/${sample}_msa.npz --figure paper_figures/${sample}.png --linkage pipeline/${sample}/${sample}_linkage.npz --info input/${sample}_info.csv  --clustering_results pipeline/${sample}/${sample}_clustering.csv --switch_annotation hg38_switch_regions.bed --switch_coords "chr14:105583000-105872000:-" --color_by cluster --clustering_stats pipeline/${sample}/${sample}_cluster_stats.csv --sample ${sample} --dpi 500 --fig_width 4 --fig_height 1.25 --cmax 0.5 --linkage_border .2 --cutoff 0.01 --cmax 0.1 --no_x_ticks --omit_scale_bar --dpi 600
+for sample in 20231211_rFP_01_downsampled 20250113_SKA1_downsampled 20211122_85_n500_1_downsampled; do
+	swibrid plot_clustering --msa pipeline/${sample}/${sample}_msa.npz --figure ${sample}_reads_by_cluster.png --linkage pipeline/${sample}/${sample}_linkage.npz --info input/${sample}_info.csv  --clustering_results pipeline/${sample}/${sample}_clustering.csv --switch_annotation hg38_switch_regions.bed --switch_coords "chr14:105583000-105872000:-" --color_by cluster --clustering_stats pipeline/${sample}/${sample}_cluster_stats.csv --sample ${sample} --dpi 500 --fig_width 4 --fig_height 1.25 --cmax 0.5 --linkage_border .2 --cutoff 0.01 --cmax 0.1 --no_x_ticks --omit_scale_bar --dpi 600
 done
 
[…]

### 2025-01-31 paper_figures.Rmd
some additional plots

@@
   dplyr::mutate(num_variants_somatic = num_variants * (1 - frac_variants_germline),
-                status=ifelse(Diagnosis %in% c('Control','Subclass','Atypical'), 'Control', 'CVID')) %>%
+                status=ifelse(Diagnosis %in% c('Control','Subclass','Atypical'), Diagnosis, 'CVID')) %>%
   tibble::column_to_rownames('Sample')
@@
                                                     'NFKB2','PI3K','ATM','BRCA1','LIG4','NIPBL')),
-                status=factor(status, levels=c('Control','CVID')))
+                status=factor(status, levels=c('Control','Subclass','Atypical','CVID')))
 
@@
 
-another PCA with only subclass features
-
-```{r FigS4C, fig.width=2.75 * fig_scale, fig.height=1.75 * fig_scale}
-FB.cols.subclass <- grep('frac_clusters_S[MGA]+',colnames(FB.data.pooled),value=TRUE)
-FB.X.subclass <- FB.data.pooled[FB.data.pooled$Sample_donor %in% FB_train_donors,FB.cols.subclass] %>%
-  dplyr::mutate_if(is.numeric,scale)
-
-pca.subclass <- prcomp(FB.X.subclass)
-pcadat.subclass <- cbind(pca.subclass$x,human.meta[human.meta$dataset=='training',])
-
-ggplot(pcadat.subclass,aes(x=PC1,y=PC2)) +
-  geom_point(aes(color=Diagnosis),size=1) +
-  geom_line(aes(group=Donor, color=Diagnosis)) +
-  geom_polygon(data = pcadat.subclass %>%
-                 dplyr::mutate(status=ifelse(Diagnosis=='Subclass', 'Subclass', as.character(status))) %>%
-                 dplyr::group_by(status) %>%
-                 dplyr::slice(chull(PC1, PC2)), alpha = 0.25, aes(fill=status)) +
-  my_theme() +
-  scale_fill_manual(values=diagnosis_colors, guide='none') +
-  scale_color_manual(values=diagnosis_colors) +
-  guides(color=guide_legend(override.aes = list(shape=15,size=2), ncol=1)) +
-  theme(aspect.ratio=pca.subclass$sdev[2]**2/pca.subclass$sdev[1]**2,
-        legend.position='right',
-        legend.key.size=unit(1.5,'mm'),
-        legend.key.spacing.y=unit(0,'mm'),
-        legend.box='horizontal') +
-  labs(x=paste0('PC 1 (',100*signif(pca.subclass$sdev[1]**2/sum(pca.subclass$sdev**2),3),'%)'),
-       y=paste0('PC 2 (',100*signif(pca.subclass$sdev[2]**2/sum(pca.subclass$sdev**2),3),'%)'),
-       color='',shape='',fill='')
-```
-
-(this PCA uses `r dim(FB.X.subclass)[2]` features)
+<!-- another PCA with only subclass features -->
+
+<!-- ```{r FigS4C, fig.width=2.75 * fig_scale, fig.height=1.75 * fig_scale} -->
+<!-- FB.cols.subclass <- grep('frac_clusters_S[MGA]+',colnames(FB.data.pooled),value=TRUE) -->
+
+<!-- human.X.subclass <- cbind(rbind(FB.data.pooled[,FB.cols.subclass], -->
+<!--                                 QP.data[,FB.cols.subclass]), -->
+<!--                           human.meta[,'dataset',drop=FALSE]) %>% -->
+<!--   tibble::rownames_to_column('sample') %>% -->
+<!--   dplyr::mutate(sample=gsub('[0-9]*_|pooled_','',sample)) %>% -->
+<!--   dplyr::group_by(dataset) %>% -->
+<!--   dplyr::mutate_if(is.numeric,scale) %>% -->
+<!--   dplyr::ungroup() %>% -->
+<!--   dplyr::select(c('sample',FB.cols.subclass)) %>% -->
+<!--   tibble::column_to_rownames('sample') -->
+
+<!-- pca.subclass <- prcomp(human.X.subclass[train,]) -->
+<!-- pcadat.subclass <- cbind(pca.subclass$x,human.meta[human.meta$dataset=='training',]) -->
+
+<!-- ggplot(pcadat.subclass,aes(x=PC1,y=PC2)) + -->
+<!--   geom_point(aes(color=Diagnosis),size=1) + -->
+<!--   geom_line(aes(group=Donor, color=Diagnosis)) + -->
+<!--   geom_polygon(data = pcadat.subclass %>% -->
+<!--                  #dplyr::mutate(status=ifelse(Diagnosis %in% c('Subclass','Atypical'), as.character(Diagnosis), as.character(status))) %>% -->
+<!--                  dplyr::group_by(status) %>% -->
+<!--                  dplyr::slice(chull(PC1, PC2)), alpha = 0.25, aes(fill=status)) + -->
+<!--   my_theme() + -->
+<!--   scale_fill_manual(values=diagnosis_colors, guide='none') + -->
+<!--   scale_color_manual(values=diagnosis_colors) + -->
+<!--   guides(color=guide_legend(override.aes = list(shape=15,size=2), ncol=1)) + -->
+<!--   theme(aspect.ratio=pca.subclass$sdev[2]**2/pca.subclass$sdev[1]**2, -->
+<!--         legend.position='right', -->
+<!--         legend.key.size=unit(1.5,'mm'), -->
+<!--         legend.key.spacing.y=unit(0,'mm'), -->
+<!--         legend.box='horizontal') + -->
+<!--   labs(x=paste0('PC 1 (',100*signif(pca.subclass$sdev[1]**2/sum(pca.subclass$sdev**2),3),'%)'), -->
+<!--        y=paste0('PC 2 (',100*signif(pca.subclass$sdev[2]**2/sum(pca.subclass$sdev**2),3),'%)'), -->
+<!--        color='',shape='',fill='') -->
+<!-- ``` -->
+
+<!-- (this PCA uses ` dim(human.X.subclass)[2]` features) -->
 
@@
 
-```{r FigS4B, fig.width=2 * fig_scale, fig.height=1.7 * fig_scale}
+```{r FigS4D, fig.width=2 * fig_scale, fig.height=1.7 * fig_scale}
 subclass.cols <- c("frac_clusters_SA1","frac_clusters_SA2",
@@
 
-```{r FigS4H, fig.width=2.25 * fig_scale, fig.height=2 * fig_scale}
+```{r FigS4E, fig.width=2.25 * fig_scale, fig.height=2 * fig_scale}
 tmp <- FB.data.pooled %>%
   dplyr::mutate(alpha_ratio_FACS=pct_IgA/(pct_IgA + pct_IgG)) %>%
+  dplyr::filter(Diagnosis %in% c('Control','CVID','CVID-like')) %>%
   dplyr::group_by(status) %>%
@@
   theme(legend.position=c(.8,.15),
-        legend.box=element_blank())
+        legend.box=element_blank(),
+        legend.title=element_blank())
 ```
@@
 
+<!-- project into PCA using only subclass features -->
+
+<!-- ```{r FigS4X5, fig.width=2.25 * fig_scale, fig.height=2 * fig_scale} -->
+<!-- pcadat.combined.subclass <- rbind(cbind(as.matrix(human.X.subclass[train,]) %*% pca.subclass$rotation, human.meta[train,]), -->
+<!--                                   cbind(as.matrix(human.X.subclass[test,]) %*% pca.subclass$rotation, human.meta[test,]), -->
+<!--                                   cbind(as.matrix(human.X.subclass[validation,]) %*% pca.subclass$rotation, human.meta[validation,])) %>% -->
+<!--   dplyr::mutate(dataset=factor(dataset, levels=c('training','testing','validation'))) -->
+
+<!-- ggplot(pcadat.combined.subclass, -->
+<!--        aes(x=PC1,y=PC2)) + -->
+<!--   geom_polygon(data = pcadat.combined.subclass %>% dplyr::filter(dataset=='training') %>% -->
+<!--                  dplyr::group_by(status) %>% -->
+<!--                  dplyr::slice(chull(PC1, PC2)), alpha = 0.1, color=NA, aes(fill=status), show.legend=FALSE) + -->
+<!--   geom_point(aes(color=Diagnosis, alpha=dataset, shape=dataset), size=1) + -->
+<!--   geom_line(aes(group=Donor, color=Diagnosis,alpha=dataset, shape=dataset)) + -->
+<!--   my_theme() + -->
+<!--   scale_fill_manual(values=diagnosis_colors, guide='none') + -->
+<!--   scale_color_manual(values=diagnosis_colors) + -->
+<!--   scale_alpha_manual(values=c('testing'=1,'training'=.25,'validation'=1)) + -->
+<!--   scale_shape_manual(values=c('testing'=15,'training'=16,'validation'=4)) + -->
+<!--   guides(color=guide_legend(override.aes = list(shape=15,size=2), ncol=3), -->
+<!--          shape=guide_legend(ncol=1), -->
+<!--          alpha=guide_legend(ncol=1)) + -->
+<!--   theme(legend.key.height=unit(1.5,'mm'), -->
+<!--         legend.key.spacing=unit(0,'mm'), -->
+<!--         aspect.ratio=pca$sdev[2]**2/pca$sdev[1]**2, -->
+<!--         legend.position='bottom', -->
+<!--         legend.box='horizontal') + -->
+<!--   coord_cartesian(xlim=c(-3,4), -->
+<!--                   ylim=c(-3,2), -->
+<!--                   clip='off') + -->
+<!--   labs(x='PC 1',y='PC 2', -->
+<!--        color='',shape='',fill='',alpha='') -->
+<!-- ``` -->
+
 project validation data into training PCA
@@
 (average accuracy is `r mean(unlist(sample.CV))`)
+
+<!-- regression using only subclass features -->
+
+<!-- ```{r FigS4X6, fig.width=2. * fig_scale, fig.height=1.5 * fig_scale, include=FALSE} -->
+<!-- set.seed(1) -->
+<!-- train.here <- train & !(human.meta[,'Diagnosis'] %in% c('Subclass','Atypical'))  -->
+<!-- test.here <- test & !(human.meta[, 'Donor'] %in% FB_train_donors) -->
+<!-- cv_model <- cv.glmnet(as.matrix(human.X.subclass[train.here,]), human.meta[train.here,'status']!='Control', alpha = 0, family='binomial') -->
+<!-- mod <- glmnet(as.matrix(human.X.subclass[train.here,]), human.meta[train.here,'status']!='Control', alpha = 0, family = "binomial", -->
+<!--                 lambda = cv_model$lambda.min) -->
+<!-- pred_1 <- predict(mod, type = "response", newx=as.matrix(human.X.subclass[test.here,])) -->
+<!-- ROC_1 <- roc(human.meta[test.here,'status']!='Control', pred_1) -->
+
+<!-- pred_2 <- predict(mod, type = "response", newx=as.matrix(human.X.subclass[validation,])) -->
+<!-- ROC_2 <- roc(human.meta[validation,'status']!='Control', pred_2) -->
+
+<!-- FB_res <- data.frame(specificity=c(ROC_1$specificities,ROC_2$specificities), -->
+<!--                      sensitivity=c(ROC_1$sensitivities,ROC_2$sensitivities), -->
+<!--                      thresholds=c(ROC_1$thresholds,ROC_2$thresholds), -->
+<!--                      model=c(rep(paste0('training -> testing; AUC=',signif(ROC_1$auc,3)),length(ROC_1$specificities)), -->
+<!--                              rep(paste0('training -> validation; AUC=',signif(ROC_2$auc,3)),length(ROC_2$specificities)))) %>% -->
+<!--   dplyr::mutate(model=factor(model, levels=unique(model))) %>% -->
+<!--   dplyr::arrange(model,sensitivity)  -->
+
+<!-- ggplot(FB_res, aes(x=specificity,y=sensitivity,color=model)) +  -->
+<!--   geom_step() + -->
+<!--   scale_x_reverse() +   -->
+<!--   geom_abline(slope=1,intercept=1,linewidth=.25,linetype='dashed') + -->
+<!--   my_theme() + -->
+<!--   guides(color=guide_legend(ncol=1)) + -->
+<!--   scale_color_manual(values=c('steelblue4','lightsteelblue')) + -->
+<!--   theme(legend.position=c(.5,.2)) + -->
+<!--   labs(color='',linetype='') -->
+<!-- ``` -->
 
@@
 
-```{r FigS4D, fig.width=3 * fig_scale, fig.height=4 * fig_scale}
+```{r FigS4C, fig.width=3 * fig_scale, fig.height=4 * fig_scale}
 tmp <- data.frame(pca$rotation) %>%
@@
                    'frac_breaks_intra',
-                   'frac_breaks_duplications')
+                   'insert_frequency')
 
@@
   coord_cartesian(clip='off') +
-  scale_fill_manual(values=c(diagnosis_colors,class_colors)) +
+  scale_fill_manual(values=setNames(c(rep('gray',3),rep('darkseagreen3',3)),groups)) +
   theme(axis.text.x=element_text(angle=90,hjust=1,vjust=.5),
@@
   scale_x_continuous(breaks=c(1,2,3,5,6,7),
-                     labels=c('Cluster1','Cluster2','Cluster3','Control','ATM','BRCA1')) +
+                     labels=groups) +
   labs(x='',y='', color='Diagnosis')
@@
 
-```{r FigS4E, fig.width=3 * fig_scale, fig.height=2.5 * fig_scale}
+```{r FigS4B, fig.width=3 * fig_scale, fig.height=2.5 * fig_scale}
 pvals <- list()
@@
 
+```{r inhibitors, fig.width=6.75 * fig_scale, fig.height=6.5 * fig_scale }
+inhibitors_sample_sheet <- read.csv('../sodar/2019_Inhibitors/a_2019_Inhibitors.txt',
+                                    sep='\t',header=1) %>%
+  dplyr::left_join(read.csv('../sodar/2019_Inhibitors/s_2019_Inhibitors.txt',sep='\t',header=1),
+                   by='Sample.Name') %>%
+  dplyr::mutate(Donor=Source.Name,
+                Treatment=Parameter.Value.Treatment.,
+                Dose=Parameter.Value.Dose.,
+                Batch=Parameter.Value.Batch.,
+                Sample=Comment.Extract.Name.) %>%
+  dplyr::distinct(Donor,Treatment,Dose,Sample,Batch,.keep_all = TRUE)
+
+inhibitors.data <- read.csv('../human_samples/20240307_hg38_results/output/summary/20240307_human_samples_stats.csv',
+                            header=1,row.names=1) %>%
+  tibble::rownames_to_column('Sample') %>%
+  dplyr::inner_join(inhibitors_sample_sheet %>% dplyr::distinct(Donor, Treatment, Sample, Batch), by='Sample') %>%
+  dplyr::filter(nreads_final >= 500, frac_clusters_SA1+frac_clusters_SA2 > .05,
+                frac_clusters_SG1 + frac_clusters_SG2 + frac_clusters_SG3 + frac_clusters_SG4 > .05) %>%
+  dplyr::mutate_all(~replace_na(.,0)) %>%
+  dplyr::mutate(Diagnosis=Treatment) %>%
+  dplyr::select(tidyselect::any_of(c('Donor','Diagnosis','Sample','Batch',columns$col))) %>%
+  dplyr::mutate(num_variants_somatic=num_variants * (1 - frac_variants_germline))
+
+# inhibitors.X <- inhibitors.data[,selected.cols] %>% dplyr::mutate_if(is.numeric, scale)
+# inhibitors.meta <- inhibitors.data[,c('Donor','Diagnosis')] %>%
+#   dplyr::mutate(Diagnosis=ifelse(grepl('BIBR',Diagnosis),
+#                                  'BIBR',
+#                                  ifelse(grepl('ATM',Diagnosis),'ATMi',
+#                                         ifelse(grepl('DMSO',Diagnosis),'DMSO',Diagnosis))))
+# 
+# pcadat <- cbind(prcomp(inhibitors.X)$x, inhibitors.meta)
+# 
+# ggplot(pcadat, aes(x=PC1,y=PC2)) +
+#   geom_point(aes(color=Diagnosis), size=2) +
+#   my_theme() +
+#   scale_color_manual(values=c(diagnosis_colors,c('DMSO'='gray40','HU'='orange1','Olaparib'='blue2','SRC7'='orchid1',
+#                                                  'DSMO'='gray40','SCR7'='orchid1','BIBR'='green3','ATMi'='brown4'))) +
+#   guides(color=guide_legend(override.aes = list(shape=15,size=2), ncol=3)) +
+#   theme(legend.key.height=unit(1.5,'mm'),
+#         legend.key.spacing=unit(0,'mm'),
+#         legend.position='bottom',
+#         legend.box='horizontal') +
+#   coord_cartesian(clip='off') +
+#   labs(x='PC 1', #paste0('PC 1 (',100*signif(pca$sdev[1]**2/sum(pca$sdev**2),3),'%)'),
+#        y='PC 2', #paste0('PC 2 (',100*signif(pca$sdev[2]**2/sum(pca$sdev**2),3),'%)'),
+#        color='',shape='',fill='',alpha='')
+# 
+# pcadat.test <- rbind(cbind(as.matrix(human.X[train,]) %*% pca$rotation, human.meta[train,c('Donor','Diagnosis')]),
+#                      cbind(as.matrix(human.X[test,]) %*% pca$rotation, human.meta[test,c('Donor','Diagnosis')]),
+#                      cbind(as.matrix(inhibitors.X) %*% pca$rotation,inhibitors.meta)) %>%
+#   dplyr::mutate(dataset=ifelse(grepl('FB',Donor),'FB','inhibitors'))
+# 
+# ggplot(pcadat.test, aes(x=PC1,y=PC2)) +
+#   geom_polygon(data = pcadat.test %>% dplyr::filter(dataset=='FB', Diagnosis %in% c('CVID','Control')) %>%
+#                  dplyr::group_by(Diagnosis) %>%
+#                  dplyr::slice(chull(PC1, PC2)), alpha = 0.1, color=NA, aes(fill=Diagnosis), show.legend=FALSE) +
+#   geom_point(aes(color=Diagnosis,alpha=dataset, shape=dataset), size=2) +
+#   my_theme() +
+#   scale_fill_manual(values=diagnosis_colors) +
+#   scale_color_manual(values=c(diagnosis_colors,c('DMSO'='gray40','HU'='orange1','Olaparib'='blue2','SRC7'='orchid1',
+#                                                  'DSMO'='gray40','SCR7'='orchid1','BIBR'='green3','ATMi'='brown4'))) +
+#   scale_alpha_manual(values=c('FB'=.25,'inhibitors'=1)) +
+#   scale_shape_manual(values=c('FB'=16,'inhibitors'=15)) +
+#   guides(color=guide_legend(override.aes = list(shape=15,size=2), ncol=3),
+#          shape=guide_legend(ncol=1),
+#          alpha=guide_legend(ncol=1)) +
+#   theme(legend.key.height=unit(1.5,'mm'),
+#         legend.key.spacing=unit(0,'mm'),
+#         aspect.ratio=pca$sdev[2]**2/pca$sdev[1]**2,
+#         legend.position='bottom',
+#         legend.box='horizontal') +
+#   coord_cartesian(xlim=c(-8,10),
+#                   ylim=c(-5,8),
+#                   clip='off') +
+#   labs(x='PC 1', #paste0('PC 1 (',100*signif(pca$sdev[1]**2/sum(pca$sdev**2),3),'%)'),
+#        y='PC 2', #paste0('PC 2 (',100*signif(pca$sdev[2]**2/sum(pca$sdev**2),3),'%)'),
+#        color='',shape='',fill='',alpha='')
+
+features.here <- c('top_clone_occupancy_downsampled',
+                   'mean_cluster_size_downsampled',
+                   'cluster_gini_downsampled',
+                   'mean_length',
+                   'donor_complexity',
+                   'donor_score_W',
+                   'donor_score_WGCW',
+                   'homology_fw',
+                   'n_homology_switch',
+                   'n_untemplated_switch',
+                   'frac_breaks_single',
+                   'frac_breaks_intra',
+                   'frac_breaks_sequential',
+                   'frac_breaks_inversions',
+                   'frac_breaks_duplications',
+                   'insert_frequency')
+
+tmp <- cbind(rbind(CH12.data[,features.here],
+                   FB.data.pooled[,features.here],
+                   QP.data[,features.here],
+                   inhibitors.data[,features.here]),
+             rbind(CH12.meta %>%
+                     dplyr::mutate(Diagnosis=Genotype,
+                                   dataset='CH12') %>%
+                     dplyr::select(Diagnosis, dataset),
+                   human.meta[,c('Diagnosis','dataset')],
+                   inhibitors.data[,c('Diagnosis'),drop=FALSE] %>%
+                     dplyr::mutate(dataset='inhibitors',
+                                   Diagnosis=ifelse(Diagnosis=='ATM','ATMi',Diagnosis)))) %>%
+  dplyr::filter(Diagnosis %in% c('WT','Trp53bp1 KO','Brca1 KO','Rif1 KO','Lig4 KO','Control','CVID','TNFRSF13B','ATM','BRCA1',
+                                 'DMSO','BIBR','ATMi')) %>%
+  tibble::rownames_to_column('sample') %>%
+  gather(col,value,features.here) %>%
+  dplyr::mutate(col=factor(gsub("_downsampled|_switch",'',col),
+                           levels=gsub("_downsampled|_switch","",features.here)),
+                group=droplevels(factor(Diagnosis, levels=c(names(genotype_colors), names(diagnosis_colors),
+                                                            c('DMSO','BIBR','ATMi')))))
+
+ggplot(tmp, aes(x=ifelse(dataset=='CH12',as.numeric(group),
+                         ifelse(dataset=='inhibitors',
+                                as.numeric(group) + 2,
+                                as.numeric(group) + 1)),
+                y=value,fill=group)) +
+  geom_boxplot(outlier.shape=NA, lwd=.5) +
+  geom_jitter(width=.1,height=0,size=.1) +
+  stat_compare_means(comparisons=list(c(1,2),
+                                      c(1,3),
+                                      c(1,4),
+                                      c(1,5)),
+                     method='wilcox.test',label = "p.signif", hide.ns=TRUE, size=2, vjust=.5) +
+  stat_compare_means(comparisons=list(c(7,8),
+                                      c(7,9),
+                                      c(7,10),
+                                      c(7,11)),
+                     method='wilcox.test',label = "p.signif", hide.ns=TRUE, size=2, vjust=.5) +
+  stat_compare_means(comparisons=list(c(13,14),
+                                      c(13,15)),
+                     method='wilcox.test',label = "p.signif", hide.ns=TRUE, size=2, vjust=.5) +
+  facet_wrap(.~col,ncol=4,scales='free_y') +
+  my_theme() +
+  coord_cartesian(clip='off') +
+  scale_fill_manual(values=c(diagnosis_colors,genotype_colors,
+                             c('DMSO'='gray40','HU'='orange1','Olaparib'='blue2','SCR7'='orchid1',
+                               'BIBR'='green3','ATMi'='brown4','Act'='yellow3'))) +
+  theme(axis.text.x=element_text(angle=90,hjust=1,vjust=.5),
+        legend.position='none') +
+  scale_x_continuous(breaks=c(seq(1,5), seq(1,5) + 6, seq(1,3) + 12),
+                     labels=levels(tmp$group)) +
+  labs(x='',y='', color='Diagnosis')
+```
+
 # session info

### 2025-01-31 prepare_input.R

@@ -370,8 +370,8 @@ FB.data.pooled <- human_results %>%
   dplyr::mutate_all(~replace_na(.,0)) %>%
   dplyr::mutate(num_variants_somatic=num_variants * (1 - frac_variants_germline)) %>%
   dplyr::select(tidyselect::any_of(c('Donor','Diagnosis','Timepoint','Sample_donor','Sex','Age','Sample',
-                                     'splenomegaly','lymphoproliferation','Tetanus','Diphteria','PnPS','lung',
-                                     'complication','EUROclass','pct_B','pct_IgA','pct_IgG',columns$col)))
+                                     'Tetanus','Diphteria','PnPS','lung',
+                                     'pct_B','pct_IgA','pct_IgG',columns$col)))
 write.csv(FB.data.pooled, 'data/FB_data_pooled.csv')
 
 vincendeau.data <- read.csv('../mouse_samples/20240425_vincendeau/output/summary/20240425_vincendeau_samples_stats.csv',

### 2025-01-31 paper_figures.Rmd
clean-up

--- 2025-01-21 paper_figures.Rmd
@@
 
-```{r FigS3X1, fig.width=4 * fig_scale, fig.height=1.75 * fig_scale, include=FALSE, eval=FALSE}
-dualPCA <- function(X, Y) {
-  library(irlba)
-  # use CCA-like extension of dual PCA as described here: https://xinmingtu.cn/blog/2022/CCA_dual_PCA/
-  #sv <- irlba(as.matrix(X) %*% as.matrix(t(Y)), nv=n)
-  sv <- svd(as.matrix(X) %*% as.matrix(t(Y)))
-  x <- sv$u %*% diag(sqrt(sv$d))
-  row.names(x) <- row.names(X)
-  colnames(x) <- paste0('dPC',seq(1,dim(x)[2]))
-  y <- sv$v %*% diag(sqrt(sv$d))
-  row.names(y) <- row.names(Y)
-  colnames(y) <- paste0('dPC',seq(1,dim(y)[2]))
-  list(x=x, y=y, sdev=sqrt(sv$d/(dim(X)[2]-1)))
-}
-
-CH12.X <- CH12.data[CH12.take, mouse.cols] %>% 
-  dplyr::mutate_if(is.numeric, scale)
-mouse.X <- mouse.data[mouse.take, mouse.cols] %>% 
-  dplyr::mutate_if(is.numeric, scale)
-
-dpca <- dualPCA(CH12.X, mouse.X)
-dpcadat <- cbind(rbind(dpca$x, dpca$y),
-                 data.frame(group=c(CH12.data[row.names(dpca$x),'Genotype'],
-                                    mouse.data$material)))
-
-ggplot(dpcadat,aes(x=dPC1,y=dPC2, color=group)) +
-  geom_point(size=1) +
-  geom_polygon(data = dpcadat %>%
-                 dplyr::group_by(group) %>%
-                 dplyr::slice(chull(dPC1, dPC2)), alpha = 0.1, aes(fill=group), linewidth=.1) +
-  my_theme() +
-  scale_fill_manual(values=c(genotype_colors,tissue_colors)) +
-  scale_color_manual(values=c(genotype_colors,tissue_colors)) +
-  theme(aspect.ratio=dpca$sdev[2]**2/dpca$sdev[1]**2,
-        legend.spacing=unit(0,'mm')) +
-  labs(x=paste0('dPC 1 (',100*signif(dpca$sdev[1]**2/sum(dpca$sdev**2),3),'%)'),
-       y=paste0('dPC 2 (',100*signif(dpca$sdev[2]**2/sum(dpca$sdev**2),3),'%)'),
-       color='',shape='',fill='')
-```
-
-```{r FigS3X2, fig.width=3 * fig_scale, fig.height=4 * fig_scale, include=FALSE, eval=FALSE}
-tmp <- t(t(t(as.matrix(rbind(CH12.X, mouse.X))) %*% as.matrix(dpcadat[,c('dPC1','dPC2')])) / colSums(dpcadat[,c('dPC1','dPC2')]**2)) %>%
-  as.data.frame() %>%
-  tibble::rownames_to_column('col') %>%
-  dplyr::left_join(columns, by='col') %>%
-  dplyr::mutate(col=factor(col, levels=columns %>% 
-                             dplyr::filter(col %in% mouse.cols) %>%
-                             dplyr::pull(col)),
-                col_name=gsub("_downsampled|_switch",'',col))
-
-cols_shapes <- tmp %>% 
-  dplyr::distinct(category, col,col_name) %>%
-  dplyr::mutate(var_col=cat_cols[as.character(category)]) %>%
-  dplyr::group_by(category) %>%
-  dplyr::mutate(var_ind=1:n()) %>%
-  dplyr::mutate(col=factor(col, levels=levels(tmp$col))) %>% 
-  dplyr::arrange(col)
-
-ggplot(tmp,aes(x=dPC1,y=dPC2, color=col, shape=col)) +
-  geom_point(size=1) +
-  #geom_text_repel(aes(label=col), size=1) +
-  my_theme() +
-  scale_color_manual(values=cols_shapes %>%
-                       dplyr::pull(var_col,col),
-                     labels=cols_shapes %>%
-                       dplyr::pull(col_name)) +
-  scale_shape_manual(values=cols_shapes %>%
-                       dplyr::pull(var_ind,col),
-                     labels=cols_shapes %>%
-                       dplyr::pull(col_name)) +
-  theme(legend.key.size=unit(1.5,'mm'),
-        aspect.ratio=dpca$sdev[2]**2/dpca$sdev[1]**2,
-        legend.spacing=unit(0,'mm'),
-        legend.key.spacing.y=unit(0,'mm'),
-        legend.position='bottom',
-        legend.box='horizontal') +
-  guides(color=guide_legend(ncol=2),
-         shape=guide_legend(ncol=2)) +
-  labs(x='dPC 1', y='dPC 2',
-       color='',shape='',fill='',alpha='')
-```
-
 selected features in the CH12 data (2023 batches)
@@
 
-```{r FigS3X5, fig.width=1. * fig_scale, fig.height=1.75 * fig_scale, include=FALSE}
+```{r FigS3X, fig.width=1. * fig_scale, fig.height=1.75 * fig_scale, include=FALSE}
 batch.CV <- c()
@@
                                                    'validation')) %>%
-                      dplyr::select(Donor,Diagnosis,status,Timepoint, dataset,Sex,Age,complication, EUROclass, pct_B, pct_IgA, pct_IgG),
+                      dplyr::select(Donor,Diagnosis,status,Timepoint, dataset,Sex,Age, pct_B, pct_IgA, pct_IgG),
                     QP.data %>%
@@
                                     Timepoint='t1',
-                                    complication=NA,
-                                    EUROclass=NA,
                                     pct_B=NA,
@@
                                     status=ifelse(Diagnosis=='Control','Control','CVID')) %>%
-                      dplyr::select(Donor,Diagnosis,status,Timepoint,dataset,Sex,Age,complication, EUROclass, pct_B, pct_IgA, pct_IgG)) %>%
+                      dplyr::select(Donor,Diagnosis,status,Timepoint,dataset,Sex,Age, pct_B, pct_IgA, pct_IgG)) %>%
   dplyr::mutate(Diagnosis=factor(Diagnosis,levels=c('Control','Subclass','Atypical','CVID','ICOS','Kabuki','IgG4def',
@@
 
-another PCA with only subclass features
-
-```{r FigS4C, fig.width=2.75 * fig_scale, fig.height=1.75 * fig_scale}
-FB.cols.subclass <- grep('frac_clusters_S[MGA]+',colnames(FB.data.pooled),value=TRUE)
-FB.X.subclass <- FB.data.pooled[FB.data.pooled$Sample_donor %in% FB_train_donors,FB.cols.subclass] %>%
-  dplyr::mutate_if(is.numeric,scale)
-
-pca.subclass <- prcomp(FB.X.subclass)
-pcadat.subclass <- cbind(pca.subclass$x,human.meta[human.meta$dataset=='training',])
-
-ggplot(pcadat.subclass,aes(x=PC1,y=PC2)) +
-  geom_point(aes(color=Diagnosis),size=1) +
-  geom_line(aes(group=Donor, color=Diagnosis)) +
-  geom_polygon(data = pcadat.subclass %>%
-                 dplyr::mutate(status=ifelse(Diagnosis=='Subclass', 'Subclass', as.character(status))) %>%
-                 dplyr::group_by(status) %>%
-                 dplyr::slice(chull(PC1, PC2)), alpha = 0.25, aes(fill=status)) +
-  my_theme() +
-  scale_fill_manual(values=diagnosis_colors, guide='none') +
-  scale_color_manual(values=diagnosis_colors) +
-  guides(color=guide_legend(override.aes = list(shape=15,size=2), ncol=1)) +
-  theme(aspect.ratio=pca.subclass$sdev[2]**2/pca.subclass$sdev[1]**2,
-        legend.position='right',
-        legend.key.size=unit(1.5,'mm'),
-        legend.key.spacing.y=unit(0,'mm'),
-        legend.box='horizontal') +
-  labs(x=paste0('PC 1 (',100*signif(pca.subclass$sdev[1]**2/sum(pca.subclass$sdev**2),3),'%)'),
-       y=paste0('PC 2 (',100*signif(pca.subclass$sdev[2]**2/sum(pca.subclass$sdev**2),3),'%)'),
-       color='',shape='',fill='')
-```
-
-(this PCA uses `r dim(FB.X.subclass)[2]` features)
-
 show subclass features for HD and subclass-deficient people
 
-```{r FigS4B, fig.width=2 * fig_scale, fig.height=1.7 * fig_scale}
+```{r FigS4D, fig.width=2 * fig_scale, fig.height=1.7 * fig_scale}
 subclass.cols <- c("frac_clusters_SA1","frac_clusters_SA2",
@@
 
-```{r FigS4X1, fig.width=.8 * fig_scale, fig.height=1 * fig_scale, include=FALSE}
+```{r FigS4X, fig.width=.8 * fig_scale, fig.height=1 * fig_scale, include=FALSE}
 set.seed(4)
@@
   dplyr::select(Donor,Sample_donor,Diagnosis,dataset,Sex,Age,Timepoint,
-                splenomegaly,lymphoproliferation,pct_B,pct_IgA,pct_IgG,complication)
+                pct_B,pct_IgA,pct_IgG)
 
@@
 
-<!-- ```{r FigS4X2, fig.width=4 * fig_scale, fig.height=1.75 * fig_scale, include=FALSE, eval=FALSE} -->
-<!-- all.cols <- intersect(colnames(CH12.X), colnames(human.X)) -->
-<!-- dpca <- dualPCA(CH12.X[,all.cols], human.X[,all.cols]) -->
-
-<!-- dpcadat <- cbind(rbind(dpca$x,dpca$y), -->
-<!--                 data.frame(species=c(rep('mouse',dim(dpca$x)[1]), -->
-<!--                                      rep('human',dim(dpca$y)[1])), -->
-<!--                            Genotype=c(CH12.meta[,'Genotype'], -->
-<!--                                       human.meta[,'diagnosis']))) -->
-
-<!-- ggplot(dpcadat,aes(x=dPC1,y=dPC2, color=Genotype)) + -->
-<!--   geom_point(aes(shape=species),size=.5) + -->
-<!--   geom_polygon(data = dpcadat %>% -->
-<!--                  dplyr::group_by(Genotype) %>% -->
-<!--                  dplyr::slice(chull(dPC1, dPC2)), alpha = 0.1, aes(fill=Genotype), linewidth=.1) + -->
-<!--   my_theme() + -->
-<!--   scale_fill_manual(values=c(diagnosis_colors,genotype_colors)) + -->
-<!--   scale_color_manual(values=c(diagnosis_colors,genotype_colors)) + -->
-<!--   scale_shape_manual(values=c('human'=15,'mouse'=16)) + -->
-<!--   guides(fill=guide_legend(ncol=2, override.aes = list(pch=c(rep(16,5),rep(15,14)))), -->
-<!--          color=guide_legend(ncol=2, override.aes = list(pch=c(rep(16,5),rep(15,14)))), -->
-<!--          shape='none') + -->
-<!--   labs(x=paste0('dPC 1 (',100*signif(dpca$sdev[1]**2/sum(dpca$sdev**2),3),'%)'), -->
-<!--        y=paste0('dPC 2 (',100*signif(dpca$sdev[2]**2/sum(dpca$sdev**2),3),'%)'), -->
-<!--        color='',shape='',fill='') -->
-
-<!-- ggplot(dpcadat,aes(x=dPC1,y=dPC2, color=Genotype)) + -->
-<!--   geom_point(aes(shape=species),size=.5) + -->
-<!--   geom_polygon(data = dpcadat %>% -->
-<!--                  dplyr::group_by(Genotype) %>% -->
-<!--                  dplyr::slice(chull(dPC1, dPC2)), alpha = 0.1, aes(fill=Genotype), linewidth=.1) + -->
-<!--   my_theme() + -->
-<!--   facet_wrap(~species, ncol=2) + -->
-<!--   scale_fill_manual(values=c(diagnosis_colors,genotype_colors)) + -->
-<!--   scale_color_manual(values=c(diagnosis_colors,genotype_colors)) + -->
-<!--   scale_shape_manual(values=c('human'=15,'mouse'=16)) + -->
-<!--   theme(legend.key.size=unit(1.5,'mm'), -->
-<!--         aspect.ratio=dpca$sdev[2]**2/dpca$sdev[1]**2) + -->
-<!--   guides(fill=guide_legend(ncol=2, override.aes = list(pch=c(rep(16,5),rep(15,14)))), -->
-<!--          color=guide_legend(ncol=2, override.aes = list(pch=c(rep(16,5),rep(15,14)))), -->
-<!--          shape='none') + -->
-<!--   labs(x=paste0('dPC 1 (',100*signif(dpca$sdev[1]**2/sum(dpca$sdev**2),3),'%)'), -->
-<!--        y=paste0('dPC 2 (',100*signif(dpca$sdev[2]**2/sum(dpca$sdev**2),3),'%)'), -->
-<!--        color='',shape='',fill='') -->
-
-<!-- dpcadat[,'group'] <- as.character(dpcadat$Genotype) -->
-<!-- dpcadat[FB.meta$Sample_donor,'group'] <- as.character(class_df$class) -->
-<!-- dpcadat[,'dataset'] <- 'CH12' -->
-<!-- dpcadat[dpcadat$species=='human','dataset'] <- human.meta$dataset -->
-<!-- ggplot(dpcadat,# %>% dplyr::filter(dataset!='validation'), -->
-<!--        aes(x=dPC1,y=dPC2, color=group)) + -->
-<!--   geom_point(aes(shape=species),size=.5) + -->
-<!--   geom_polygon(data = dpcadat %>% -->
-<!--                  #dplyr::filter(dataset!='validation') %>% -->
-<!--                  dplyr::group_by(group) %>% -->
-<!--                  dplyr::slice(chull(dPC1, dPC2)), alpha = 0.1, aes(fill=group), linewidth=.1) + -->
-<!--   my_theme() + -->
-<!--   facet_wrap(~species, ncol=2) + -->
-<!--   scale_fill_manual(values=c(diagnosis_colors,genotype_colors,class_colors)) + -->
-<!--   scale_color_manual(values=c(diagnosis_colors,genotype_colors,class_colors)) + -->
-<!--   scale_shape_manual(values=c('human'=15,'mouse'=16)) + -->
-<!--   theme(legend.key.size=unit(1.5,'mm'), -->
-<!--         aspect.ratio=dpca$sdev[2]**2/dpca$sdev[1]**2) + -->
-<!--   guides(fill=guide_legend(ncol=2), #, override.aes = list(pch=c(rep(16,5),rep(15,10)))), -->
-<!--          color=guide_legend(ncol=2), #, override.aes = list(pch=c(rep(16,5),rep(15,10)))), -->
-<!--          shape='none') + -->
-<!--   labs(x=paste0('dPC 1 (',100*signif(dpca$sdev[1]**2/sum(dpca$sdev**2),3),'%)'), -->
-<!--        y=paste0('dPC 2 (',100*signif(dpca$sdev[2]**2/sum(dpca$sdev**2),3),'%)'), -->
-<!--        color='',shape='',fill='') -->
-<!-- ``` -->
-
-<!-- ```{r FigS4X3, fig.width=3 * fig_scale, fig.height=4 * fig_scale} -->
-<!-- tmp <- t(t(t(as.matrix(rbind(CH12.X[,all.cols], human.X[,all.cols]))) %*% as.matrix(dpcadat[,c('dPC1','dPC2')])) / colSums(dpcadat[,c('dPC1','dPC2')]**2)) %>% -->
-<!--   as.data.frame() %>% -->
-<!--   tibble::rownames_to_column('col') %>% -->
-<!--   dplyr::left_join(columns, by='col') %>% -->
-<!--   dplyr::mutate(col=factor(col, levels=columns %>%  -->
-<!--                              dplyr::filter(col %in% all.cols) %>% -->
-<!--                              dplyr::pull(col)), -->
-<!--                 col_name=gsub("_downsampled|_switch",'',col)) -->
-
-<!-- cols_shapes <- tmp %>%  -->
-<!--   dplyr::distinct(category, col, col_name) %>% -->
-<!--   dplyr::mutate(var_col=cat_cols[as.character(category)]) %>% -->
-<!--   dplyr::group_by(category) %>% -->
-<!--   dplyr::mutate(var_ind=1:n()) %>% -->
-<!--   dplyr::mutate(col=factor(col, levels=levels(tmp$col))) %>%  -->
-<!--   dplyr::arrange(col) -->
-
-<!-- ggplot(tmp,aes(x=dPC1,y=dPC2, color=col, shape=col)) + -->
-<!--   geom_point(size=1) + -->
-<!--   #geom_text_repel(aes(label=col), size=1) + -->
-<!--   my_theme() + -->
-<!--   scale_color_manual(values=cols_shapes %>% -->
-<!--                        dplyr::pull(var_col,col), -->
-<!--                      labels=cols_shapes %>% -->
-<!--                        dplyr::pull(col_name)) + -->
-<!--   scale_shape_manual(values=cols_shapes %>% -->
-<!--                        dplyr::pull(var_ind,col), -->
-<!--                      labels=cols_shapes %>% -->
-<!--                        dplyr::pull(col_name)) + -->
-<!--   theme(legend.key.size=unit(1.5,'mm'), -->
-<!--         aspect.ratio=dpca$sdev[2]**2/dpca$sdev[1]**2, -->
-<!--         legend.spacing=unit(0,'mm'), -->
-<!--         legend.key.spacing.y=unit(0,'mm'), -->
-<!--         legend.position='bottom', -->
-<!--         legend.box='horizontal') + -->
-<!--   guides(color=guide_legend(ncol=2), -->
-<!--          shape=guide_legend(ncol=2)) + -->
-<!--   labs(x='dPC 1', y='dPC 2', -->
-<!--        color='',shape='',fill='',alpha='') -->
-<!-- ``` -->
-
-```{r FigS4X4, fig.width=6.75 * fig_scale, fig.height=3.5 * fig_scale}
-features.here <- c('cluster_entropy_downsampled',
-                   'top_clone_occupancy_downsampled',
-                   'mean_cluster_size_downsampled',
-                   'donor_complexity',
-                   'donor_score_W',
-                   'donor_score_WGCW',
-                   'homology_fw',
-                   'n_homology_switch',
-                   'mean_length_SA',
-                   'frac_breaks_intra',
-                   'frac_breaks_inversions',
-                   'frac_breaks_duplications')
-
-tmp <- cbind(rbind(CH12.data[,features.here],
-                   FB.data.pooled[,features.here],
-                   QP.data[,features.here]),
-             rbind(CH12.meta %>%
-                     dplyr::mutate(Diagnosis=Genotype,
-                                   dataset='CH12') %>%
-                     dplyr::select(Diagnosis, dataset),
-                   human.meta[,c('Diagnosis','dataset')])) %>%
-  dplyr::filter(Diagnosis %in% c('WT','Trp53bp1 KO','Brca1 KO','Rif1 KO','Lig4 KO','Control','CVID','TNFRSF13B','ATM','BRCA1')) %>%
-  tibble::rownames_to_column('sample') %>%
-  gather(col,value,features.here) %>%
-  dplyr::mutate(col=factor(gsub("_downsampled|_switch",'',col),
-                           levels=gsub("_downsampled|_switch","",features.here)),
-                group=droplevels(factor(Diagnosis, levels=c(names(genotype_colors), names(diagnosis_colors)))))
-
-ggplot(tmp, aes(x=ifelse(dataset=='CH12',as.numeric(group),as.numeric(group) + 1),y=value,fill=group)) +
-  geom_boxplot(outlier.shape=NA, lwd=.5) +
-  geom_jitter(width=.1,height=0,size=.1) +
-  stat_compare_means(comparisons=list(c(1,2),
-                                      c(1,3),
-                                      c(1,4),
-                                      c(1,5)),
-                     method='wilcox.test',label = "p.signif", hide.ns=TRUE, size=2, vjust=.5) +
-  stat_compare_means(comparisons=list(c(7,8),
-                                      c(7,9),
-                                      c(7,10),
-                                      c(7,11)),
-                     method='wilcox.test',label = "p.signif", hide.ns=TRUE, size=2, vjust=.5) +
-  facet_wrap(.~col,ncol=6,scales='free_y') +
-  my_theme() +
-  coord_cartesian(clip='off') +
-  scale_fill_manual(values=c(diagnosis_colors,genotype_colors)) +
-  theme(axis.text.x=element_text(angle=90,hjust=1,vjust=.5),
-        legend.position='none') +
-  scale_x_continuous(breaks=c(seq(1,5), seq(1,5) + 6),
-                     labels=levels(tmp$group)) +
-  labs(x='',y='', color='Diagnosis')
-```
-
-<!-- ```{r aggregated_FB_boxplots, fig.width=4.5 * fig_scale, fig.height=1.7 * fig_scale} -->
-<!-- tmp <- FB.X.aggregated %>% -->
-<!--   tibble::rownames_to_column('sample') %>% -->
-<!--   dplyr::left_join(FB.meta %>% tibble::rownames_to_column('sample'), by='sample') %>% -->
-<!--   dplyr::mutate(name=gsub('t','',gsub('FB_','',paste0(Donor,'_',Timepoint)))) %>% -->
-<!--   dplyr::left_join(class_df,by='name') %>% -->
-<!--   gather(col,value,c('diversity','isotypes','breaks','context','structural')) -->
-
-<!-- ggplot(tmp, aes(x=class,y=value,fill=class)) + -->
-<!--   geom_boxplot(outlier.shape=NA, lwd=.5) + -->
-<!--   geom_jitter(width=.1,height=0,size=.5) + -->
-<!--   stat_compare_means(comparisons=list(c('cluster1','cluster3'), -->
-<!--                                       c('cluster2','cluster3'), -->
-<!--                                       c('cluster1','cluster2')), -->
-<!--                      method='wilcox.test',label = "p.signif", hide.ns=TRUE, size=2, vjust=.5) + -->
-<!--   facet_wrap(.~col,ncol=length(features.here),scales='free_y') + -->
-<!--   my_theme() + -->
-<!--   coord_cartesian(clip='off') + -->
-<!--   scale_fill_manual(values=c(diagnosis_colors,class_colors)) + -->
-<!--   theme(axis.text.x=element_text(angle=45,hjust=1,vjust=1), -->
-<!--         legend.position='none') + -->
-<!--   labs(x='',y='category score', color='Diagnosis') -->
-<!-- ``` -->
-
-<!-- check correlation with measured B cell proportion in FACS -->
-
-<!-- ```{r pct_B_cor, fig.width=3 * fig_scale, fig.height=3 * fig_scale} -->
-<!-- FB.data.pooled %>% -->
-<!--   gather(feature, value, selected.cols) %>% -->
-<!--   dplyr::group_by(feature) %>% -->
-<!--   dplyr::summarise(R=cor(value, pct_B, use='na.or.complete')) %>% -->
-<!--   dplyr::left_join(columns, by=c('feature'='col')) %>% -->
-<!--   dplyr::mutate(feature=gsub('_downsampled|_switch','',feature)) %>% -->
-<!--   dplyr::arrange(R) %>% -->
-<!--   dplyr::mutate(feature=factor(feature, levels=feature)) %>% -->
-<!--   ggplot(aes(y=R, x=feature, fill=category)) +  -->
-<!--   geom_col() + -->
-<!--   my_theme() + -->
-<!--   coord_flip() + -->
-<!--   scale_fill_manual(values=cat_cols) + -->
-<!--   labs(x='',y='correlation with % B cells') + -->
-<!--   theme(legend.position=c(.2,.9)) -->
-
-<!-- FB.data.pooled %>% -->
-<!--   gather(feature, value, selected.cols) %>% -->
-<!--   dplyr::group_by(feature) %>% -->
-<!--   dplyr::summarise(R=cor(value, pct_IgA, use='na.or.complete')) %>% -->
-<!--   dplyr::left_join(columns, by=c('feature'='col')) %>% -->
-<!--   dplyr::mutate(feature=gsub('_downsampled|_switch','',feature)) %>% -->
-<!--   dplyr::arrange(R) %>% -->
-<!--   dplyr::mutate(feature=factor(feature, levels=feature)) %>% -->
-<!--   ggplot(aes(y=R, x=feature, fill=category)) +  -->
-<!--   geom_col() + -->
-<!--   my_theme() + -->
-<!--   coord_flip() + -->
-<!--   scale_fill_manual(values=cat_cols) + -->
-<!--   labs(x='',y='correlation with % IgA cells') + -->
-<!--   theme(legend.position=c(.2,.9)) -->
-
-<!-- FB.data.pooled %>% -->
-<!--   gather(feature, value, selected.cols) %>% -->
-<!--   dplyr::group_by(feature) %>% -->
-<!--   dplyr::summarise(R=cor(value, pct_IgG, use='na.or.complete')) %>% -->
-<!--   dplyr::left_join(columns, by=c('feature'='col')) %>% -->
-<!--   dplyr::mutate(feature=gsub('_downsampled|_switch','',feature)) %>% -->
-<!--   dplyr::arrange(R) %>% -->
-<!--   dplyr::mutate(feature=factor(feature, levels=feature)) %>% -->
-<!--   ggplot(aes(y=R, x=feature, fill=category)) +  -->
-<!--   geom_col() + -->
-<!--   my_theme() + -->
-<!--   coord_flip() + -->
-<!--   scale_fill_manual(values=cat_cols) + -->
-<!--   labs(x='',y='correlation with % IgG cells') + -->
-<!--   theme(legend.position=c(.2,.9)) -->
-
-<!-- ``` -->
-
 # session info

### 2025-01-31 prepare_input.R

@@ -153,7 +153,7 @@ HD_sample_sheet <- rbind(read.csv('../sodar/2023_OC/a_2023_OC.txt',
                            dplyr::left_join(read.csv('../sodar/2023_OC/s_2023_OC.txt',sep='\t',header=1),
                                             by='Sample.Name') %>%
                            dplyr::mutate(Donor=as.character(Source.Name),
-                                         Sample=Extract.Name.1,
+                                         Sample=Comment.Extract.Name.,
                                          Batch=as.character(Parameter.Value.Batch.),
                                          QC=Parameter.Value.QC.,
                                          Age=Characteristics.Age.,
@@ -169,7 +169,7 @@ HD_sample_sheet <- rbind(read.csv('../sodar/2023_OC/a_2023_OC.txt',
                            dplyr::mutate(Donor=Source.Name,
                                          Batch=as.character(Parameter.Value.Batch.),
                                          QC=Parameter.Value.QC.,
-                                         Sample=Extract.Name.1,
+                                         Sample=Comment.Extract.Name.,
                                          Age=Characteristics.Age.,
                                          Sex=Characteristics.Sex.,
                                          Cohort='C1',
@@ -200,7 +200,7 @@ HD_sample_sheet <- rbind(read.csv('../sodar/2023_OC/a_2023_OC.txt',
                                          Age=Parameter.Value.Age.,
                                          Sex=Characteristics.Sex.,
                                          Diagnosis=Characteristics.Diagnosis.,
-                                         Sample=Extract.Name.1,
+                                         Sample=Comment.Extract.Name.,
                                          Cohort='C2',
                                          Sequencing=Parameter.Value.Sequencing.method.) %>%
                            dplyr::filter(Diagnosis=='control',grepl('2023',Sample)) %>%
@@ -255,7 +255,7 @@ CH12_sample_sheet <- read.csv('../sodar/2022_CH12/a_2022_CH12.txt',
   dplyr::mutate(Genotype=Characteristics.Genotype.,
                 Batch=as.character(Parameter.Value.Batch.),
                 QC=Parameter.Value.QC.,
-                Sample=Extract.Name.1) %>%
+                Sample=Comment.Extract.Name.) %>%
[…]

### 2025-02-13 paper_figures.Rmd
take out SM samples

@@
 
-```{r FigS4D, fig.width=2 * fig_scale, fig.height=1.7 * fig_scale}
+```{r FigS4D, fig.width=2.5 * fig_scale, fig.height=1.5 * fig_scale}
 subclass.cols <- c("frac_clusters_SA1","frac_clusters_SA2",
@@
 
-```{r FigS4E, fig.width=2.25 * fig_scale, fig.height=2 * fig_scale}
+```{r FigS4E, fig.width=2.5 * fig_scale, fig.height=1.75 * fig_scale}
 tmp <- FB.data.pooled %>%
@@
                 dataset=ifelse(grepl('FB',Donor),
-                               ifelse(Sample_donor %in% FB_train_donors,'training','testing'),
+                               ifelse(Sample_donor %in% gsub('_','',FB_train_donors),'training','testing'),
                                'validation')) %>%
@@
 
-hm1 <- Heatmap(t(FB.X),
-               col=colorRamp2(c(-2,0, 2), c("blue",'gray90', "red")),
-               cluster_columns=dend,
-               cluster_rows=TRUE,
-               show_row_names=TRUE,
-               show_column_names=TRUE,
-               show_row_dend=TRUE,
-               top_annotation=HeatmapAnnotation(df=FB.meta[,c('Diagnosis','Sex','Age',#'pct_B','pct_IgA','pct_IgG',
+hm_meta <- FB.meta[,c('Diagnosis','Sex','Age',
 'Donor','Timepoint'),drop=FALSE] %>% 
@@
   dplyr::select(-c(Donor,Timepoint)) %>%
-                                                  tibble::column_to_rownames('name'),
+  tibble::column_to_rownames('name')
+
+hm1 <- Heatmap(t(FB.X),
+               col=colorRamp2(c(-2,0, 2), c("blue",'gray90', "red")),
+               cluster_columns=dend,
+               cluster_rows=TRUE,
+               show_row_names=TRUE,
+               show_column_names=TRUE,
+               show_row_dend=TRUE,
+               top_annotation=HeatmapAnnotation(df=hm_meta,
                                                 show_annotation_name=TRUE,
@@
 
+<!-- plot a decision tree using the top features -->
+
+<!-- ```{r FigS4E, fig.width=2.5 * fig_scale, fig.height=1.5 * fig_scale} -->
+<!-- library(randomForest) -->
+<!-- library(rpart) -->
+<!-- library(rpart.plot) -->
+
+<!-- set.seed(2) -->
+<!-- tmp <- cbind(human.X[test.here | train.here,],  -->
+<!--              droplevels(human.meta[test.here | train.here, 'status', drop=FALSE]), -->
+<!--              hm_meta[test.here | train.here,'Class',drop=FALSE]) -->
+
+<!-- rf <- randomForest(tmp[,selected.cols], tmp$status) -->
+<!-- #pred <- predict(rf, type='response', as.matrix(human.X[test.here,])) -->
+
+<!-- # auc1 <- list() -->
+<!-- # auc2 <- list() -->
+<!-- # for (k in seq(2,10)) { -->
+<!-- #   vars <- importance(rf) %>%  -->
+<!-- #     as.data.frame() %>%  -->
+<!-- #     tibble::rownames_to_column('var') %>% -->
+<!-- #     dplyr::slice_max(MeanDecreaseGini, n=k) %>% -->
+<!-- #     dplyr::pull(var) -->
+<!-- #   # cv_model <- cv.glmnet(as.matrix(human.X[train.here,vars]),  -->
+<!-- #   #                       human.meta[train.here,'status']!='Control',  -->
+<!-- #   #                       alpha = 0, family='binomial') -->
+<!-- #  -->
+<!-- #   mod <- glmnet(as.matrix(human.X[train.here,vars]),  -->
+<!-- #                 human.meta[train.here,'status']!='Control',  -->
+<!-- #                 alpha = 0, family = "binomial", -->
+<!-- #                 lambda = cv_model$lambda.min) -->
+<!-- #  -->
+<!-- #   pred_1 <- predict(mod, type = "response", newx=as.matrix(human.X[test.here,vars])) -->
+<!-- #   auc1[[k]] <- as.numeric(roc(human.meta[test.here,'status']!='Control', pred_1)$auc) -->
+<!-- #    -->
+<!-- #   pred_2 <- predict(mod, type = "response", newx=as.matrix(human.X[validation,vars])) -->
+<!-- #   auc2[[k]] <- as.numeric(roc(human.meta[validation,'status']!='Control', pred_2)$auc) -->
+<!-- # }   -->
+
+<!-- # vars <- importance(rf) %>%  -->
+<!-- #     as.data.frame() %>%  -->
+<!-- #     tibble::rownames_to_column('var') %>% -->
+<!-- #     dplyr::slice_max(MeanDecreaseGini, n=5) %>% -->
+<!-- #     dplyr::pull(var) -->
+<!-- #  -->
+<!-- # dtree <- rpart(status ~ ., data = tmp[, c(vars, 'status')],   method = "class") -->
+<!-- #rpart.plot(dtree, type=5, extra=1, box.palette=diagnosis_colors[c('Control','CVID')], tweak=.5) -->
+
+<!-- # rf <- randomForest(tmp[,selected.cols], as.factor(tmp$Class)) -->
+<!-- # vars <- importance(rf) %>%  -->
+<!-- #     as.data.frame() %>%  -->
+<!-- #     tibble::rownames_to_column('var') %>% -->
+<!-- #     dplyr::slice_max(MeanDecreaseGini, n=5) %>% -->
+<!-- #     dplyr::pull(var) -->
+<!-- # dtree <- rpart(Class ~ ., data = tmp[, c(vars, 'Class')],   method = "class") -->
+<!-- #rpart.plot(dtree, type=5, extra=1, box.palette=list('Cluster1'='gray20','darkgoldenrod1','forestgreen'), tweak=.5) -->
+
+<!-- rpart_splits <- function(fit, digits = getOption("digits")) { -->
+<!--   splits <- fit$splits -->
+<!--   if (!is.null(splits)) { -->
+<!--     ff <- fit$frame -->
+<!--     is.leaf <- ff$var == "<leaf>" -->
+<!--     n <- nrow(splits) -->
+<!--     nn <- ff$ncompete + ff$nsurrogate + !is.leaf -->
+<!--     ix <- cumsum(c(1L, nn)) -->
+<!--     ix_prim <- unlist(mapply(ix, ix + c(ff$ncompete, 0), FUN = seq, SIMPLIFY = F)) -->
+<!--     type <- rep.int("surrogate", n) -->
+<!--     type[ix_prim[ix_prim <= n]] <- "primary" -->
+<!--     type[ix[ix <= n]] <- "main" -->
+<!--     left <- character(nrow(splits)) -->
+<!--     side <- splits[, 2L] -->
+<!--     for (i in seq_along(left)) { -->
+<!--       left[i] <- if (side[i] == -1L) -->
+<!--                    paste("<", format(signif(splits[i, 4L], digits))) -->
+<!--                  else if (side[i] == 1L) -->
+<!--                    paste(">=", format(signif(splits[i, 4L], digits))) -->
+<!--                  else { -->
+<!--                    catside <- fit$csplit[splits[i, 4L], 1:side[i]] -->
+<!--                    paste(c("L", "-", "R")[catside], collapse = "", sep = "") -->
+<!--                  } -->
+<!--     } -->
+<!--     cbind(data.frame(var = rownames(splits), -->
+<!--                      type = type, -->
+<!--                      node = rep(as.integer(row.names(ff)), times = nn), -->
+<!--                      ix = rep(seq_len(nrow(ff)), nn), -->
+<!--                      left = left), -->
+<!--           as.data.frame(splits, row.names = F)) -->
+<!--   } -->
+<!-- } -->
+
+<!-- train.here <- (FB.meta$dataset=='training') & !(FB.meta[,'Diagnosis'] %in% c('Subclass','Atypical'))  -->
+<!-- test.here <- (FB.meta$dataset=='testing') & !(FB.meta$Donor %in% FB_train_donors) -->
+
+<!-- FB.dat <- FB.data.pooled %>% -->
+<!--   dplyr::mutate(status=as.factor(ifelse(Diagnosis=='Control','Control','CVID'))) -->
+
+<!-- tmp <- lapply(selected.cols, function(col) { -->
+<!--   mod <- rpart(status ~ ., data=FB.dat[train.here,c(col,'status')], -->
+<!--                method='class') -->
+<!--   pred <- predict(mod, newdata=FB.dat[test.here, col, drop=FALSE], type='class') -->
+<!--   acc <- as.numeric(confusionMatrix(pred, FB.dat[test.here,'status'])$overall['Accuracy']) -->
+<!--   split <- rpart_splits(mod)[1, 'left'] -->
+<!--   c(col=col, acc=acc, split=split) -->
+<!-- }) -->
+
+<!-- ggplot(do.call(rbind, tmp) %>% -->
+<!--          as.data.frame() %>% -->
+<!--          dplyr::mutate(acc=as.numeric(acc)) %>% -->
+<!--          dplyr::left_join(columns, by='col') %>% -->
+<!--          #dplyr::group_by(category) %>% -->
+<!--          #dplyr::slice_max(acc, n=3, with_ties=FALSE) %>% -->
+<!--          #dplyr::ungroup() %>% -->
+<!--          dplyr::slice_max(acc, n=10, with_ties=FALSE) %>% -->
+<!--          dplyr::arrange(acc) %>% -->
+<!--          dplyr::mutate(col=factor(gsub("_downsampled","",col),  -->
+<!--                                   levels=gsub('_downsampled','',col))) %>% -->
+<!--          rowwise() %>% -->
+<!--          dplyr::mutate(spl=paste(str_split(split, ' ')[[1]][1], signif(as.numeric(str_split(split, ' ')[[1]][2]),3))), -->
+<!--        aes(x=col, y=acc, fill=category)) +  -->
+<!--   geom_col() +  -->
+<!--   geom_text(aes(x=col, y=.05, label=spl), size=2, hjust=0) + -->
+<!--   #geom_hline(yintercept=mean(unlist(sample.CV)), linewidth=.5, linetype='dashed') + -->
+<!--   my_theme() + -->
+<!--   coord_flip() + -->
+<!--   scale_fill_manual(values=cat_cols) + -->
+<!--   labs(y='accuracy', x='', title='prediction using single feature') + -->
+<!--   theme(legend.position='none') -->
+
+<!-- ``` -->
+
 show PCA loadings per category
@@
                               names(diagnosis_colors))) +
-  guides(color=guide_legend(override.aes = list(shape=c(rep(16,3),rep(15,7)),size=2), ncol=4)) +
+  guides(color=guide_legend(override.aes = list(shape=c(rep(16,3),rep(15,6)),size=2), ncol=4)) +
   theme(legend.key.size=unit(1.5,'mm'),
@@
 ```{r Fig4J, fig.width=5.25 * fig_scale, fig.height=1.7 * fig_scale}
-features.here <- c('frac_clusters_SG1',
+features.here <- c('frac_breaks_SA_SA',
                    'mean_cluster_size_downsampled',

### 2025-02-13 prepare_input.R

@@ -355,14 +355,14 @@ FB_sample_sheet <- read.csv('../sodar/2023_FB/a_2023_FB.txt',
 FB.data <- human_results %>%
   tibble::rownames_to_column('Sample') %>%
   dplyr::inner_join(FB_sample_sheet %>% dplyr::distinct(Donor, Diagnosis, Timepoint, Sample_donor, Sex, Age, Sample, QC, Batch), by='Sample') %>%
-  dplyr::filter(nreads_final >= 500, QC=='PASS') %>%
+  dplyr::filter(nreads_final >= 500, QC=='PASS', grepl('FB', Donor)) %>%
   dplyr::mutate_all(~replace_na(.,0)) %>%
   dplyr::select(tidyselect::any_of(c('Donor','Diagnosis','Timepoint','Sample_donor','Sex','Age','Sample','Batch',columns$col))) 
 write.csv(FB.data, 'data/FB_data.csv')
 
 FB.data.pooled <- human_results %>%
   tibble::rownames_to_column('Sample') %>%
-  dplyr::filter(grepl('pooled_',Sample)) %>%
+  dplyr::filter(grepl('pooled_FB',Sample)) %>%
   dplyr::mutate(Sample_donor=gsub('pooled_','',Sample)) %>%
   dplyr::inner_join(FB_sample_sheet %>% dplyr::distinct(Donor, Diagnosis, Timepoint, Sample_donor, Sex, Age, 
                                                         splenomegaly, lymphoproliferation, Tetanus, Diphteria, PnPS, lung,

### 2025-07-09 paper_figures.Rmd
first updates for revision

@@
 
+check reproducibility of clones in the cell switch experiment
+
+```{r FigS1X, fig.width=1.75 * fig_scale, fig.height=1.5 * fig_scale}
+meta_clustering <- lapply(c('21084','21085','21086','mixed'), 
+                          function(donor) 
+                            lapply(if (donor=='mixed') seq(1,5) else seq(1,3),
+                                   function(rep)
+                                     lapply(c(50000,100000), function(nc) 
+                                     {
+                                       fname <- paste0('../human_samples/20240307_hg38_results/cellswitch_combined/',
+                                                       donor,'_',rep,'_',format(nc,scientific=FALSE),'_meta_clustering.csv')
+                                       if (file.exists(fname)) 
+                                         read.csv(fname, header=1,row.names=1) %>%
+                                         dplyr::group_by(sample, meta_cluster) %>%
+                                         dplyr::summarise(nreads=n()) %>%
+                                         dplyr::ungroup() %>%
+                                         dplyr::group_by(meta_cluster) %>%
+                                         dplyr::mutate(n_samples=n_distinct(sample[nreads > 0 * sum(nreads)])) %>%
+                                         dplyr::ungroup() %>%
+                                         dplyr::group_by(sample) %>%
+                                         dplyr::mutate(cluster_size=cut(nreads/sum(nreads), c(0,.0001,.001,.01,.1,1), 
+                                                                        labels=c('<.01%','<.1%','<1%','<10%','>10%'))) %>%
+                                         dplyr::mutate(Donor=donor,
+                                                       ncells=nc,
+                                                       replicate=rep)
+                                     }) %>%
+                                     do.call(rbind, .)) %>%
+                            do.call(rbind, .)) %>%
+  do.call(rbind, .)
+
+ggplot(meta_clustering %>%
+         dplyr::group_by(Donor,replicate,ncells) %>%
+         dplyr::summarise(frac_clusters=mean(n_samples > 1)) %>%
+         dplyr::ungroup() %>% 
+         dplyr::distinct(Donor, ncells, frac_clusters) %>%
+         dplyr::mutate(mixing=ifelse(Donor=='mixed','inter-donor','intra-donor'),
+                       Donor=plyr::revalue(Donor, c('21084'='A',
+                                                    '21085'='B',
+                                                    '21086'='C')),
+                       ncells=factor(ifelse(ncells==50000,'50k','100k'),levels=c('50k','100k'))),
+         # dplyr::group_by(Donor, ncells) %>%
+         # dplyr::summarise(mean_frac_clusters=mean(frac_clusters),
+         #                  se_frac_clusters=sd(frac_clusters)/sqrt(n())),
+       aes(x=ncells,y=100*frac_clusters, fill=mixing)) + 
+  geom_boxplot(position=position_dodge(width=.8), outlier.shape=NA) +
+  geom_jitter(aes(color=Donor, group=mixing), position=position_jitterdodge(dodge.width = .8, jitter.width=.05), size=1) +
+  #geom_line(aes(group=Donor)) +
+  # geom_errorbar(aes(ymin=100*(mean_frac_clusters-se_frac_clusters),
+  #                   ymax=100*(mean_frac_clusters+se_frac_clusters)), width=0,
+  #               position=position_dodge(width=.8)) +
+  my_theme() +
+  scale_fill_manual(values=c('gray','lightcyan')) +
+  #scale_color_manual(values=c('gray','coral')) +
+  #scale_color_brewer(palette='Set1') +
+  scale_color_manual(values=setNames(brewer.pal(3,'Set1'),c('A','B','C')),na.value='gray20') +
+  #coord_cartesian(ylim=c(0,100)) +
+  labs(x='# of cells',y='% shared clusters in pair', fill='pair', color='donor')
+```
+
 # Figure 2 + supplements
@@
                   pval=dt(tstat,4)) %>%
+    dplyr::mutate(padj=p.adjust(pval, method='BH'),
+                  sig=ifelse(padj < .05, 'sig','ns')) %>%
     ggplot(aes(x=mean1,y=mean2)) +
     geom_abline(slope=1,intercept=0,color='gray',linewidth=.5) +
-    geom_point(aes(color=variable,shape=variable,size=ifelse(pval < .05,'sig','ns'))) + 
+    geom_point(aes(color=variable, shape=variable, size=sig, alpha=sig)) + 
     stat_cor(method = "pearson", aes(label=..r.label..), size=2.5, show.legend=FALSE) +
@@
                       ymax=mean2+se2,
-                      color=variable),
+                      color=variable,
+                      alpha=sig),
                   size=.25) +
@@
                        xmax=mean1+se1,
-                       color=variable),
+                       color=variable,
+                       alpha=sig),
                    size=.25) +
@@
     scale_size_manual(values=c('sig'=2,'ns'=.5),guide='none') +
+    scale_alpha_manual(values=c('sig'=1,'ns'=.25),guide='none') +
     guides(color=guide_legend(ncol=5),
@@
     dplyr::mutate(tstat=abs(mean2-mean1)/(sqrt(se2**2+se1**2)),
-                  pval=dt(tstat,4)) %>%
+                  pval=dt(tstat,4),
+                  padj=p.adjust(pval, method='BH'),
+                  sig=ifelse(padj < .05, 'sig','ns')) %>%
     ggplot(aes(x=mean1,y=mean2)) +
     geom_abline(slope=1,intercept=0,color='gray',linewidth=.5) +
-    geom_point(aes(color=variable,shape=variable,size=ifelse(pval < .05,'sig','ns'))) + 
+    geom_point(aes(color=variable,shape=variable,size=sig, alpha=sig)) + 
     stat_cor(method = "pearson", aes(label=..r.label..), size=2.5, show.legend=FALSE, label.y.npc = 'bottom') +
@@
     scale_size_manual(values=c('sig'=2,'ns'=.5),guide='none') +
+    scale_alpha_manual(values=c('sig'=1,'ns'=.2),guide='none') +
     guides(color=guide_legend(ncol=4),
@@
 library(gtools)
+library(rstatix)
 
@@
   stat_compare_means(ref.group='WT',method='wilcox.test',label = "p.signif", hide.ns=TRUE, size=2) +
+  #stat_pvalue_manual(pvals, label = "p.adj.signif", tip.length = 0.01) +
   ggh4x::facet_grid2(celltype~variable, switch='y', scales='free_y', independent='y', strip=ggh4x::strip_vanilla(clip='off')) +
@@
   labs(color='',linetype='')
+```
+
+show the ridge coefficients
+
+```{r FigS4X2, fig.width=2 * fig_scale, fig.height=2.5 * fig_scale}
+tmp <- data.frame(coef=coef(mod)[,'s0']) %>%
+  dplyr::filter(rowSums(abs(.)) > 0) %>%
+  dplyr::slice_max(rowSums(abs(.)),n=20) %>%
+  dplyr::arrange(rowSums(abs(.))) %>%
+  tibble::rownames_to_column('col') %>%
+  dplyr::filter(col %in% columns$col)
+
+col.order <- gsub('_downsampled|_switch','',tmp$col[order(tmp$coef)])
+
+tmp <- tmp %>%
+  dplyr::mutate(col=gsub('_downsampled|_switch','',col),
+                col=factor(col, levels=col.order))
+
+anno_grob_y <- grid::rectGrob(y=1:length(tmp$col), x=0, gp=gpar(col=NA, fill=col_cols[rev(levels(tmp$col))], alpha=.5))
+
+tmp %>%
+  dplyr::mutate(col=factor(col, levels=rev(levels(tmp$col)))) %>%
+  ggplot(aes(x=1, y=col, fill=coef, label=round(coef,2))) + 
+  geom_tile() +
+  scale_fill_gradient2(low='pink2',high='steelblue3', mid='white', breaks=c(-.05, 0, .05)) +
+  my_theme() +
+  coord_cartesian(clip='off') +
+  theme(axis.text.x=element_blank(),
+        axis.ticks.x = element_blank(),
+        legend.position='right',
+        legend.direction='vertical') +
+  annotation_custom(grob=anno_grob_y, ymin = 0, ymax = 1, xmin = -3.5, xmax = 4) +
+  labs(x='',y='') 
 ```
@@
 
+hm2 <- Heatmap(t(FB.X),
+               col=colorRamp2(c(-2,0, 2), c("blue",'gray90', "red")),
+               #cluster_columns=dend,
+               column_split = grepl('09|32', row.names(FB.X)), 
+               column_title = NULL,
+               cluster_rows=TRUE,
+               show_row_names=TRUE,
+               show_column_names=TRUE,
+               show_row_dend=TRUE,
+               top_annotation=HeatmapAnnotation(df=hm_meta,
+                                                show_annotation_name=TRUE,
+                                                show_legend=TRUE,
+                                                simple_anno_size=unit(2,'mm'),
+                                                annotation_legend_param=list(title_gp=gpar(fontsize=7),
+                                                                             labels_gp=gpar(fontsize=7),
+                                                                             legend_height = unit(5, "mm"),
+                                                                             legend_width = unit(3, "mm"),
+                                                                             grid_height = unit(3, "mm"),
+                                                                             grid_width = unit(3, "mm")),
+                                                annotation_name_gp=gpar(fontsize=7),
+                                                col=list('Diagnosis'=diagnosis_colors,
+                                                         'Class'=class_colors,
+                                                         'Age'=colorRamp2(c(20,50,80),c("yellow2","orange2","coral3")),
+                                                         'Sex'=c('male'='mistyrose2',
+                                                                 'female'='lightseagreen'))),
+               column_names_gp=gpar(fontsize=4),
+               row_names_gp =gpar(fontsize = 7),
+               row_dend_width = unit(7, "mm"),
+               row_split = columns %>% tibble::column_to_rownames('col') %>% .[selected.cols, 'category'], 
+               heatmap_legend_param=list(labels_gp=gpar(fontsize=7),
+                                         title_gp=gpar(fontsize=7),
+                                         at=c(-2,0,2),
+                                         grid_height=unit(4, 'mm'),
+                                         grid_width=unit(3, 'mm'),
+                                         legend_height=unit(5,'mm'),
+                                         legend_width=unit(3,'mm')),
+               row_title=NULL,
+               name='z-score')
+
 hm3 <- Heatmap(t(pca_hm[,c('PC1','PC2')]),
@@
 hm1 %v% hm3
+
+hm2 %v% hm3
+
 ```

### 2025-07-31 paper_figures.Rmd
new panels for revision

--- 2025-01-31 paper_figures.Rmd
@@
 library(grid)
-#library(DescTools)
+library(gtools)
 library(lme4)
+library(dendextend)
 ```
@@
 fig_scale <- 1.
-columns <- list('Diversity'=c('nclusters_final_downsampled',
-                              'nclusters_eff_downsampled',
-                              'mean_cluster_size_downsampled',
-                              'std_cluster_size_downsampled',
-                              'cluster_gini_downsampled',
-                              'cluster_entropy_downsampled',
-                              'cluster_inverse_simpson_downsampled',
-                              'top_clone_occupancy_downsampled',
-                              'big_clones_occupancy_downsampled'),
+columns <- list('Diversity'=c('clusters',
+                              'clusters_eff',
+                              'cluster_size_mean',
+                              'cluster_size_std',
+                              'cluster_gini',
+                              'cluster_entropy',
+                              'cluster_inverse_simpson',
+                              'occupancy_top_cluster',
+                              'occupancy_big_clusters'),
                 'Isotypes'=c('alpha_ratio_clusters',
-                             'frac_clusters_SA1','frac_clusters_SA2','frac_clusters_SE',
-                             'frac_clusters_SG1','frac_clusters_SG2','frac_clusters_SG3',
-                             'frac_clusters_SG4','frac_clusters_SM',
-                             'frac_clusters_Sa','frac_clusters_Sg3',
-                             'frac_clusters_Sg2b','frac_clusters_Sg2c',
-                             'frac_clusters_Sm','frac_clusters_Sg1'),
-                'Structural'=c('insert_frequency',
-                               'frac_breaks_inversions','frac_breaks_duplications',
-                               'median_duplication_size','median_inversion_size'),
-                'Context'=c('homology_fw','homology_rv',
+                             'pct_clusters_SA1','pct_clusters_SA2','pct_clusters_SE',
+                             'pct_clusters_SG1','pct_clusters_SG2','pct_clusters_SG3',
+                             'pct_clusters_SG4','pct_clusters_SM',
+                             'pct_clusters_Sa','pct_clusters_Sg3',
+                             'pct_clusters_Sg2b','pct_clusters_Sg2c',
+                             'pct_clusters_Sm','pct_clusters_Sg1'),
+                'Structural'=c('pct_templated_inserts',
+                               'pct_inversions','pct_duplications',
+                               'duplication_size','inversion_size'),
+                'Context'=c('homology_score_fw','homology_score_rv',
                             'donor_score_S', 'donor_score_W', 
@@
                             'acceptor_complexity',
-                            'n_homology_switch', 
-                            'n_untemplated_switch',
-                            'frac_blunt'),
-                'Breaks'=c('mean_length',
-                           'std_length',
-                           'mean_GC',
-                           'std_GC',
-                           'mean_length_SA',
-                           'mean_length_SG',
-                           'std_length_SA',
-                           'std_length_SG',
-                           'frac_breaks_single',
-                           'frac_breaks_sequential',
-                           'frac_breaks_intra',
-                           'mean_intra_break_size',
-                           'std_intra_break_size',
-                           'spread_SM','spread_SA','spread_SG',
-                           'spread_SA1','spread_SA2',
-                           'spread_SG1','spread_SG2',
-                           'spread_SG2B','spread_SG2C',
-                           'spread_SG3','spread_SG4',
-                           'frac_breaks_SM_SM','frac_breaks_SM_SG',
-                           'frac_breaks_SG_SG','frac_breaks_SM_SA',
-                           'frac_breaks_SG_SA','frac_breaks_SA_SA'),
-                'Context_isotype'=c('homology_fw_SM_SA','homology_rv_SM_SA',
-                                    'homology_fw_SM_SG','homology_rv_SM_SG',
+                            'homology', 
+                            'untemplated_inserts',
+                            'pct_blunt'),
+                'Breaks'=c('read_length_mean',
+                           'read_length_std',
+                           'GC_content_mean',
+                           'GC_content_std',
+                           'read_length_SA_mean',
+                           'read_length_SG_mean',
+                           'read_length_SA_std',
+                           'read_length_SG_std',
+                           'pct_direct_switch',
+                           'pct_sequential_switch',
+                           'pct_intraswitch_deletion',
+                           'intraswitch_size_mean',
+                           'intraswitch_size_std',
+                           'break_dispersion_SM','break_dispersion_SA','break_dispersion_SG',
+                           'break_dispersion_SA1','break_dispersion_SA2',
+                           'break_dispersion_SG1','break_dispersion_SG2',
+                           'break_dispersion_SG2B','break_dispersion_SG2C',
+                           'break_dispersion_SG3','break_dispersion_SG4',
+                           'pct_breaks_SM_SM','pct_breaks_SM_SG',
+                           'pct_breaks_SG_SG','pct_breaks_SM_SA',
+                           'pct_breaks_SG_SA','pct_breaks_SA_SA'),
+                'Context_isotype'=c('homology_score_fw_SM_SA','homology_score_rv_SM_SA',
+                                    'homology_score_fw_SM_SG','homology_score_rv_SM_SG',
                                     'donor_score_GAGCT_SM_SG','donor_score_GGGST_SM_SA',
@@
                                     'acceptor_complexity_SM_SG',
-                                    'n_homology_switch_SA', 
-                                    'n_homology_switch_SG', 
-                                    'n_untemplated_switch_SA',
-                                    'n_untemplated_switch_SG',
-                                    'frac_blunt_SA',
-                                    'frac_blunt_SG'),
-                'Diversity_raw'=c('nclusters_final',
-                                  'nclusters_eff',
-                                  'mean_cluster_size',
-                                  'std_cluster_size',
-                                  'cluster_gini',
-                                  'cluster_entropy',
-                                  'cluster_inverse_simpson',
-                                  'top_clone_occupancy',
-                                  'big_clones_occupancy'),
-                'Variants'=c('frac_variants_germline',
+                                    'homology_SA', 
+                                    'homology_SG', 
+                                    'untemplated_inserts_SA',
+                                    'untemplated_inserts_SG',
+                                    'pct_blunt_SA',
+                                    'pct_blunt_SG'),
+                'Diversity_raw'=c('clusters_raw',
+                                  'clusters_eff_raw',
+                                  'cluster_size_mean_raw',
+                                  'cluster_size_std_raw',
+                                  'cluster_gini_raw',
+                                  'cluster_entropy_raw',
+                                  'cluster_inverse_simpson_raw',
+                                  'occupancy_top_cluster_raw',
+                                  'occupancy_big_clusters_raw'),
+                'Variants'=c('somatic_variants',
                              'frac_variants_transitions',
@@
   dplyr::mutate(category=factor(gsub('[0-9]*$','',category), 
-                                levels=c('Library',
-                                         'Isotypes',
+                                levels=c('Isotypes',
                                          'Diversity',
@@
                                          'Variants')))
-
 cat_cols <- c('Structural'='#FD9E40',
@@
 tissue_colors <- c('Splenocytes'='orange2','Lymph Node'='orangered3','CH12'='forestgreen')
-var_colors <- c('Batch'='palegreen2','Diagnosis'='tan2','Donor'='firebrick4','Sequencing'='slategray1','unexplained'='gray80',
-                'Sex'='steelblue2','Age'='darkseagreen4')
+var_colors <- c('Batch'='lavenderblush2','Diagnosis'='tan2','Donor'='firebrick4','Sequencing'='slategray1','unexplained'='gray80',
+                'Sex'='steelblue2','Age'='lightgoldenrod1','Treatment'='saddlebrown','IV'='navy')
 
@@
 
-group_colors <- c('SmB+CD21n'='gray20',
-                  'SmB-CD21lo'='orange3',
-                  'SmB-CD21n'='lightseagreen')
-
 col_cols <- columns %>% 
   dplyr::mutate(color=cat_cols[as.character(category)]) %>%
-  dplyr::mutate(col=gsub('_downsampled|_switch','',col)) %>%
   dplyr::pull(color, col)
@@
          dplyr::mutate(nreads=factor(nreads, levels=c('10k','20k','50k','100k'))), 
-       aes(x=nclones, y=nclusters_final, color=nreads)) + 
+       aes(x=nclones, y=clusters_raw, color=nreads)) + 
   geom_abline(slope=1,intercept=0, size=.5, color='gray80') +
@@
                                              '21086'='C'))),
-       aes(x=ncells, y=nclusters_final)) + 
+       aes(x=ncells, y=clusters_raw)) + 
   geom_abline(slope=1,intercept=0, size=.5, color='gray80') +
@@
 
+check reproducibility of clones in the cell switch experiment
+
+```{r Fig1F, fig.width=1.4 * fig_scale, fig.height=1.4 * fig_scale}
+meta_clustering <- read.csv('data/cellswitch_meta_clustering.csv', header=1, row.names=1)
+
+ggplot(meta_clustering %>% 
+         dplyr::mutate(ncells=factor(ncells, levels=c('50k','100k')),
+                       Donor=ifelse(Donor=='mixed','mixed',paste0(Donor,'-',Donor)),
+                       mixing=gsub('-donor','',mixing)), 
+       aes(x=ncells,y=100*frac_clusters, fill=mixing)) + 
+  geom_boxplot(position=position_dodge(width=.8), outlier.shape=NA) +
+  geom_jitter(aes(color=Donor, group=mixing), position=position_jitterdodge(dodge.width = .8, jitter.width=.05), size=1) +
+  my_theme() +
+  #theme(legend.position='bottom') +
+  #coord_cartesian(ylim=c(15,40)) +
+  scale_fill_manual(values=c('gray','white'), guide=FALSE) +
+  scale_color_manual(values=setNames(c(brewer.pal(3,'Set1'),'gray20'),c('A-A','B-B','C-C','mixed')),na.value='gray20') +
+  labs(x='# cells',y='% shared clusters in pair', fill='pair', color='Donors')
+```
+
 show VDJ results: isotype frequencies
 
-```{r Fig1F, fig.width=2.4 * fig_scale, fig.height=1.4 * fig_scale}
+```{r Fig1G, fig.width=1.4 * fig_scale, fig.height=1.4 * fig_scale}
 # bulk VDJ results for Fig. 1
 VDJ_results <- read.csv('data/VDJ_results.csv',header=1, row.names=1)
+
+get_gini <- function(x) {
+  # calculate Gini coefficient of array x; see https://stackoverflow.com/questions/48999542/more-efficient-weighted-gini-coefficient-in-python
+  sorted_x <- sort(x)
+  n <- length(x)
+  cumx <- cumsum(sorted_x)
+  (n + 1 - 2 * sum(cumx) / cumx[n]) / n
+}
+
+get_inverse_simpson <- function(x) {
+  rel_x <- x/sum(x)
+  sum(1 / rel_x**2)
+}
 
@@
   dplyr::group_by(Donor,ncells,Sample) %>%
-  dplyr::summarise(nclusters_final=n_distinct(clone),
+  dplyr::summarise(clusters_raw=n_distinct(clone),
+                   cluster_gini_raw=get_gini(uniqueMoleculeCount),
+                   cluster_inverse_simpson_raw=get_inverse_simpson(uniqueMoleculeCount),
                    alpha_ratio_clusters=mean(grepl('IGHA',allCHitsWithScore))) %>%
@@
   dplyr::select(Donor,ncells,
-                nclusters_final,
+                clusters_raw,
+                cluster_gini_raw,
+                cluster_inverse_simpson_raw,
                 alpha_ratio_clusters) %>%
@@
 vdj_df <- rbind(swibrid_stats[,vdj_cols],vdj_stats[,vdj_cols]) %>%
-  gather(metric,value, c(nclusters_final,alpha_ratio_clusters)) %>%
+  gather(metric,value, c(clusters_raw,alpha_ratio_clusters,cluster_gini_raw,cluster_inverse_simpson_raw)) %>%
   dplyr::mutate(ncells=factor(as.character(as.integer(ncells)), levels=c('50000','100000')))
-
-# vdj_df %>% 
-#   dplyr::filter(metric=='nclusters_final') %>%
-#   ggplot(aes(x=factor(gsub('000$','',ncells), levels=c('50','100')), y=value)) +
-#   geom_boxplot(outlier.shape=NA, lwd=.5) +
-#   geom_point(aes(color=gsub('Donor ','',Donor)),size=.5) +
-#   facet_wrap(~method, ncol=2, scales='free_y') + 
-#   my_theme() +
-#   scale_color_brewer(palette='Set1') +
-#   theme(strip.clip='off',
-#         legend.position='none',
-#         legend.background =element_rect(fill=NA)) +
-#   labs(y='# Clones',x='# Cells / 1000',color='')
 
@@
   labs(y='Fraction IGHA SWIBRID',x='Fraction IGHA VDJ',shape='# cells')
+```
+
+```{r FigS1X1, fig.width=1.4 * fig_scale, fig.height=1.4 * fig_scale}
+vdj_df %>% 
+  dplyr::mutate(ncells=factor(gsub('000$','k',ncells), levels=c('50k','100k')),
+                Donor=gsub('Donor ','',Donor)) %>%
+  dplyr::filter(metric=='cluster_gini_raw') %>%
+  dplyr::group_by(ncells, Donor, method) %>%
+  dplyr::summarise(mean_gini=mean(value),
+                   se_gini=sd(value)/sqrt(length(value))) %>%
+  pivot_wider(id_cols=c(ncells,Donor),names_from=method,values_from=c('mean_gini','se_gini')) %>%
+  ggplot(aes(x=mean_gini_VDJ, y=mean_gini_SWIBRID,
+             xmin=mean_gini_VDJ-se_gini_VDJ,
+             xmax=mean_gini_VDJ+se_gini_VDJ,
+             ymin=mean_gini_SWIBRID-se_gini_SWIBRID,
+             ymax=mean_gini_SWIBRID+se_gini_SWIBRID)) +
+  geom_point(aes(color=Donor,shape=ncells),size=1) +
+  geom_errorbar(linewidth=.25) +
+  geom_errorbarh(linewidth=.25) +
+  #scale_x_continuous(breaks=c(.6,.65)) +
+  my_theme() +
+  scale_color_brewer(palette='Set1') +
+  theme(strip.background=element_blank()) +
+  labs(y='Gini coefficient SWIBRID',x='Gini coefficient VDJ',shape='# cells')
+```
+
+```{r FigS1X2, fig.width=2 * fig_scale, fig.height=1.4 * fig_scale}
+library(ggpmisc)
+PCR.data <- read.csv('data/PCR_data.csv',header=1,row.names=1)
+ggplot(PCR.data, aes(x=as.integer(ncycles), y=log10(reads))) + 
+  geom_point(aes(color=donor)) + 
+  geom_smooth(method='lm', color='black', linewidth=.5) +
+  geom_abline(slope=log10(2), intercept=-1.5, linewidth=.5, linetype='dashed') +
+  my_theme() + 
+  scale_color_brewer(palette='Set2') +
+  scale_y_continuous(breaks = log10(c(1, 10, 100, 1000,10000, 100000)),
+                     label = c(1, 10, 100, 1000, 10000, 100000)) +
+  labs(x='# of cycles',y='# of reads')
 ```
@@
 
-cols.here <- c('cluster_entropy_downsampled',
-               'cluster_gini_downsampled',
-               'mean_cluster_size_downsampled',
-               'std_cluster_size_downsampled',
-               'top_clone_occupancy_downsampled',
+cols.here <- c('cluster_entropy',
+               'cluster_gini',
+               'cluster_size_mean',
+               'cluster_size_std',
+               'occupancy_top_cluster',
                'alpha_ratio_clusters',
-               'frac_clusters_SA1',
-               'frac_clusters_SA2',
-               'frac_clusters_SG1',
-               'frac_clusters_SG2',
-               'insert_frequency',
-               'frac_breaks_inversions',
-               'frac_breaks_duplications',
-               'median_duplication_size',
-               'median_inversion_size',
-               'homology_fw',
+               'pct_clusters_SA1',
+               'pct_clusters_SA2',
+               'pct_clusters_SG1',
+               'pct_clusters_SG2',
+               'pct_templated_inserts',
+               'pct_inversions',
+               'pct_duplications',
+               'duplication_size',
+               'inversion_size',
+               'homology_score_fw',
                'donor_score_WGCW',
                'acceptor_complexity',
-               'n_homology_switch',
-               'n_untemplated_switch',
-               'frac_breaks_single',
-               'frac_breaks_sequential',
-               'frac_breaks_intra',
-               'spread_SM',
-               'spread_SA')
+               'homology',
+               'untemplated_inserts',
+               'pct_direct_switch',
+               'pct_sequential_switch',
+               'pct_intraswitch_deletion',
+               'break_dispersion_SM',
+               'break_dispersion_SA')
 
@@
   dplyr::filter(col %in% cols.here) %>%
-  dplyr::mutate(col=factor(gsub('_downsampled|_switch','',col), levels=gsub('_downsampled|_switch','',cols.here))) %>%
+  dplyr::mutate(col=factor(col, levels=cols.here)) %>%
   ggplot(aes(x=col, y=100*variance, fill=variable)) + 
@@
 
-```{r Fig2D, fig.width=1.75 * fig_scale,fig.height=1.7 * fig_scale}
+```{r Fig2D, fig.width=1.5 * fig_scale,fig.height=1.75 * fig_scale}
 donor_cutoff <- .25
@@
 
-```{r Fig2E, fig.width=.9 * fig_scale, fig.height=1.7 * fig_scale}
-age.cols <- HD.var %>%
-  dplyr::left_join(columns, by='col') %>%
-  spread(variable, variance) %>%
-  dplyr::filter(!(category %in% c('Library','Diversity_raw','Context_isotype'))) %>%
-  dplyr::slice_max(Age, n=1) %>%
-  dplyr::pull(col)
-age.cols <- 'top_clone_occupancy_downsampled'
-
-sex.cols <- HD.var %>%
-  dplyr::left_join(columns, by='col') %>%
-  spread(variable, variance) %>%
-  dplyr::filter(!(category %in% c('Library','Diversity_raw','Context_isotype'))) %>%
-  dplyr::slice_max(Sex, n=1) %>%
-  dplyr::pull(col)
-sex.cols <- 'spread_SG1'
-
+```{r Fig2E, fig.width=.75 * fig_scale, fig.height=1.5 * fig_scale}
+col <- 'break_dispersion_SG1'
 HD.data %>%
-  dplyr::filter(feature %in% sex.cols, Cohort=='C1') %>%
-  dplyr::mutate(feature=factor(gsub('_downsampled|_switch','',feature),
-                               levels=gsub('_downsampled|_switch','',sex.cols))) %>% 
+  dplyr::filter(feature == col, Cohort=='C1') %>%
   ggplot(aes(x=str_to_title(Sex), y=value)) + 
@@
                      method='wilcox.test',label = "p.signif", hide.ns=TRUE, size=2.5) +
-  facet_wrap(~feature, scales='free', ncol=8, switch='y') + 
   my_theme() + 
@@
         axis.text.x=element_text(angle=45, hjust=1, vjust=1)) +
-  labs(x='',y='')
-
+  labs(x='',y=col)
+
+col <- 'occupancy_top_cluster'
 HD.data %>%
-  dplyr::filter(feature %in% age.cols, Cohort=='C1') %>%
-  dplyr::mutate(feature=factor(gsub('_downsampled|_switch','',feature),
-                               levels=gsub('_downsampled|_switch','',age.cols))) %>% 
+  dplyr::filter(feature==col, Cohort=='C1') %>%
   ggplot(aes(x=Age, y=value)) + 
@@
                      method='wilcox.test',label = "p.signif", hide.ns=TRUE, size=2.5) +
-  facet_wrap(~feature, scales='free', ncol=8, switch='y') + 
   my_theme() +
@@
         axis.text.x=element_text(angle=45, hjust=1, vjust=1)) +
-  labs(x='',y='')
+  labs(x='',y=col)
 ```
@@
 R <- cor(FB.X[,selected.cols], use='na.or.complete')
-io <- gsub('_downsampled|_switch','',rownames(R)[order.hclust(hclust(dist(R)))])
+io <- rownames(R)[order.hclust(hclust(dist(R)))]
 
@@
   dplyr::select(-category)  %>%
-  dplyr::mutate(var1=factor(gsub('_downsampled|_switch','',var1), levels=io),
-                var2=factor(gsub('_downsampled|_switch','',var2), levels=io),
+  dplyr::mutate(var1=factor(var1, levels=io),
+                var2=factor(var2, levels=io),
                 i1=as.vector(sapply(var1, function(x) which(io==x))),
@@
   annotation_custom(grob=anno_grob_y, xmin = 0, xmax = 1, ymin = length(io) + 10, ymax = length(io) + 28) +
-  annotation_custom(grob=anno_grob_x, xmin = -9, xmax = 8, ymin = 0, ymax = 1)
+  annotation_custom(grob=anno_grob_x, xmin = -8, xmax = 9, ymin = 0, ymax = 1)
 
@@
 ```{r Fig2C, fig.width=4.25 * fig_scale, fig.height=1.5 * fig_scale}
-library(gtools)
-cols.here.human <- c('insert_frequency',
-                     'n_untemplated_switch',
-                     'frac_breaks_sequential',
-                     'frac_breaks_intra',
-                     'cluster_gini_downsampled',
-                     'frac_clusters_SG1')
+cols.here.human <- c('pct_templated_inserts',
+                     'untemplated_inserts',
+                     'pct_sequential_switch',
+                     'pct_intraswitch_deletion',
+                     'cluster_gini',
+                     'pct_clusters_SG1')
 
@@
   dplyr::left_join(columns, by=c('feature'='col')) %>%
-  dplyr::mutate(feature=factor(gsub('_downsampled|_switch','',feature), 
-                               levels=gsub('_downsampled|_switch','',cols.here.human))) %>%
+  dplyr::mutate(feature=factor(feature, levels=cols.here.human)) %>%
   dplyr::filter(category %in% c('Diversity','Isotypes','Breaks','Context','Structural'))
@@
 ```{r Fig2F, fig.width=4. * fig_scale, fig.height=1.75 * fig_scale}
-library(gtools)
 mouse.data <- read.csv('data/mouse_data.csv',header=1, row.names=1)  %>%
@@
 
-cols.here.mouse <- c('frac_breaks_inversions',
-                     'frac_breaks_intra',
-                     'mean_cluster_size_downsampled')
+cols.here.mouse <- c('pct_inversions',
+                     'pct_intraswitch_deletion',
+                     'cluster_size_mean')
 
@@
   dplyr::left_join(columns, by=c('variable'='col')) %>%
-  dplyr::mutate(variable=factor(gsub('_downsampled|_switch','',variable), 
-                                levels=gsub('_downsampled|_switch','',cols.here.mouse))) %>%
+  dplyr::mutate(variable=factor(variable, levels=cols.here.mouse)) %>%
   dplyr::filter(category %in% c('Diversity','Isotypes','Breaks','Context','Structural'))
@@
   coord_cartesian(clip='off') +
-  #annotate('text',x=1.5,y=Inf,label='SWIBRID', size=2) +
-  #annotate('text',x=5.5,y=Inf,label='Vincendeau et al.', size=2) +
   facet_wrap(~variable, ncol=3, switch='y', scales='free_y') +
@@
 ```{r FigS2K, fig.width=1.5 * fig_scale, fig.height=1.75 * fig_scale}
-cols.here.mouse <- c('mean_length_SG')
+cols.here.mouse <- c('read_length_SG_mean')
 cohort.data <- rbind(mouse.data %>%
@@
   dplyr::left_join(columns, by=c('variable'='col')) %>%
-  dplyr::mutate(variable=factor(gsub('_downsampled|_switch','',variable), 
-                                levels=gsub('_downsampled|_switch','',cols.here.mouse))) %>%
   dplyr::filter(category %in% c('Diversity','Isotypes','Breaks','Context','Structural'))
@@
                                  dplyr::filter(col %in% selected.cols) %>% 
-                                 dplyr::pull(col)),
-                feature_name=gsub('_downsampled|_switch','',feature)) %>% #,' (R=',signif(R,2),')')) %>%
+                                 dplyr::pull(col))) %>%
   dplyr::arrange(feature)
@@
 cols_shapes <- tmp %>% 
-  dplyr::distinct(category, feature, feature_name) %>%
+  dplyr::distinct(category, feature) %>%
   dplyr::mutate(var_col=cat_cols[as.character(category)]) %>%
@@
   geom_point(aes(color=feature,shape=feature),size=1) + 
-  #geom_text_repel(aes(label=feature), size=1) +
   my_theme() +
@@
                      labels=cols_shapes %>%
-                       dplyr::pull(feature_name)) +
+                       dplyr::pull(feature)) +
   scale_shape_manual(values=cols_shapes %>%
@@
                      labels=cols_shapes %>%
-                       dplyr::pull(feature_name)) +
+                       dplyr::pull(feature)) +
   guides(color=guide_legend(ncol=2),
@@
 pcarot <- data.frame(coef=pca$rot[,'PC1'],
-                     col=gsub('_downsampled|_switch','',row.names(pca$rot)))  %>%
+                     col=row.names(pca$rot))  %>%
   dplyr::arrange(coef) %>%
@@
   dplyr::distinct(category, variable) %>%
-  dplyr::mutate(var_col=cat_cols[as.character(category)],
-                variable_name=gsub('_downsampled','',variable)) %>%
+  dplyr::mutate(var_col=cat_cols[as.character(category)]) %>%
   dplyr::group_by(category) %>%
@@
                        labels=cols_shapes %>%
-                         dplyr::pull(variable_name)) +
+                         dplyr::pull(variable)) +
     scale_shape_manual(values=cols_shapes %>%
@@
                        labels=cols_shapes %>%
-                         dplyr::pull(variable_name)) +
+                         dplyr::pull(variable)) +
     scale_size_manual(values=c('sig'=2,'ns'=.5),guide='none') +
+    scale_alpha_manual(values=c('sig'=1,'ns'=.25),guide='none') +
     guides(color=guide_legend(ncol=5),
@@
 ```
+
+<!-- the PCR cycles experiment -->
+
+<!-- ```{r PCR_cycles} -->
+<!-- PCR.data <- read.csv('../human_samples/20250703_PCR_cycles/output/summary/20250703_PCR_cycles_stats.csv',header=1,row.names=1) %>% -->
+<!--   tibble::rownames_to_column('sample') %>% -->
+<!--   dplyr::mutate(donor=gsub('[0-9]*_(P[0-9])x([0-9]*)','\\1',sample), -->
+<!--                 ncycles=gsub('[0-9]*_(P[0-9])x([0-9]*)','\\2',sample))  -->
+
+<!-- PCR.X <- PCR.data[,c('sample',selected.cols)] %>% -->
+<!--   dplyr::mutate_if(is.numeric,scale) %>% -->
+<!--   tibble::column_to_rownames('sample') -->
+
+<!-- PCR.meta <- PCR.data[,c('ncycles','donor')] -->
+
+<!-- cols.here <- c('reads','clusters_raw','PCR_bias_length','PCR_bias_GC') -->
+<!-- PCR.data %>% -->
+<!--   gather(variable, value, cols.here) %>% -->
+<!--   dplyr::select(variable, value, ncycles, donor) %>% -->
+<!--   ggplot(aes(x=ncycles, y=value, fill=ncycles)) +  -->
+<!--   geom_boxplot(position=position_dodge(width=.75), width=.7, outlier.shape=NA, lwd=.25) +  -->
+<!--   geom_point(aes(color=donor), position=position_jitterdodge(jitter.width=.2, dodge.width=.75, jitter.height = 0), size=.5) +  -->
+<!--   my_theme() + -->
+<!--   coord_cartesian(clip='off') + -->
+<!--   facet_wrap(~variable, switch='y', scales='free_y') + -->
+<!--   theme(axis.text.x=element_text(angle=45, hjust=1, vjust=1), -->
+<!--         legend.position='none') + -->
+<!--   labs(x='',y='', fill='')  -->
+
+<!-- take <- (rowSums(is.na(PCR.X))==0) &  (PCR.meta$ncycles %in% c('15','20','25')) -->
+<!-- pca <- prcomp(PCR.X[take,]) -->
+<!-- pcadat <- cbind(pca$x,PCR.meta[take,]) -->
+
+<!-- ggplot(pcadat,aes(x=PC1,y=PC2)) + -->
+<!--   geom_point(aes(color=donor, shape=ncycles),size=3) + -->
+<!--   my_theme() + -->
+<!--   guides(color=guide_legend(override.aes = list(shape=15,size=2), ncol=3)) + -->
+<!--   theme(aspect.ratio=pca$sdev[2]**2/pca$sdev[1]**2, -->
+<!--         legend.key.size=unit(1.5,'mm'), -->
+<!--         legend.key.spacing.y=unit(0,'mm'), -->
+<!--         legend.position='bottom', -->
+<!--         legend.box='horizontal') + -->
+<!--   labs(x=paste0('PC 1 (',100*signif(pca$sdev[1]**2/sum(pca$sdev**2),3),'%)'), -->
+<!--        y=paste0('PC 2 (',100*signif(pca$sdev[2]**2/sum(pca$sdev**2),3),'%)'), -->
+<!--        color='',shape='',fill='') -->
+<!-- ``` -->
+
+<!-- ```{r cell_switch} -->
+<!-- cellswitch_sample_sheet <- read.csv('../sodar/2020_CellSwitch/a_2020_CellSwitch.txt', -->
+<!--                             sep='\t',header=1) %>% -->
+<!--   dplyr::left_join(read.csv('../sodar/2020_CellSwitch/s_2020_CellSwitch.txt',sep='\t',header=1), -->
+<!--                    by='Sample.Name') %>% -->
+<!--   dplyr::mutate(Donor=as.character(Source.Name), -->
+<!--                 sample=Sample.Name, -->
+<!--                 ncells=Parameter.Value.Cell.Number., -->
+<!--                 Batch=as.character(Parameter.Value.Batch.), -->
+<!--                 QC=as.character(Parameter.Value.QC.)) %>% -->
+<!--   dplyr::select(Donor,sample,ncells,Batch,QC) %>% -->
+<!--   dplyr::filter(Donor %in% c('21084','21085','21086')) -->
+
+<!-- cellswitch.data <- read.csv('../human_samples/20240307_hg38_results/output/summary/20240307_human_samples_stats.csv', -->
+<!--                           header=1,row.names=1) %>% -->
+<!--   tibble::rownames_to_column('sample') %>% -->
+<!--   dplyr::inner_join(cellswitch_sample_sheet, by='sample') %>% -->
+<!--   dplyr::filter(reads >= 500, QC=='PASS') %>% -->
+<!--   dplyr::mutate_all(~replace_na(.,0)) %>% -->
+<!--   dplyr::mutate(Donor=plyr::revalue(Donor, c('21084'='A', -->
+<!--                                              '21085'='B', -->
+<!--                                              '21086'='C'))) %>% -->
+<!--   dplyr::select(Donor,ncells,Batch,sample,selected.cols) -->
+
+<!-- cellswitch.X <- cellswitch.data[,c('sample',selected.cols)] %>% -->
+<!--   dplyr::mutate_if(is.numeric,scale) %>% -->
+<!--   tibble::column_to_rownames('sample') -->
+
+<!-- cellswitch.meta <- cellswitch.data[,c('Donor','ncells')] -->
+
+<!-- take <- (rowSums(is.na(cellswitch.X))==0) #& (cellswitch.meta$ncells > 10000) -->
+<!-- pca <- prcomp(cellswitch.X[take,]) -->
+<!-- pcadat <- cbind(pca$x,cellswitch.meta[take,]) -->
+
+<!-- ggplot(pcadat,aes(x=PC1,y=PC2)) + -->
+<!--   geom_point(aes(fill=ncells),size=3,pch=21) + -->
+<!--   geom_line(aes(group=Donor,color=Donor)) + -->
+<!--   my_theme() + -->
+<!--   theme(aspect.ratio=pca$sdev[2]**2/pca$sdev[1]**2, -->
+<!--         legend.key.size=unit(1.5,'mm'), -->
+<!--         legend.key.spacing.y=unit(0,'mm'), -->
+<!--         legend.position='right') + -->
+<!--   scale_fill_distiller(palette='Spectral',limits=c(100,200000), trans='log', breaks=c(1000,10000,100000)) + -->
+<!--   labs(x=paste0('PC 1 (',100*signif(pca$sdev[1]**2/sum(pca$sdev**2),3),'%)'), -->
+<!--        y=paste0('PC 2 (',100*signif(pca$sdev[2]**2/sum(pca$sdev**2),3),'%)'), -->
+<!--        color='Donor',fill='# cells') -->
+
+<!-- take <- (rowSums(is.na(cellswitch.X))==0) & (cellswitch.meta$ncells > 10000) -->
+<!-- pca <- prcomp(cellswitch.X[take,]) -->
+<!-- pcadat <- cbind(pca$x,cellswitch.meta[take,]) -->
+
+<!-- ggplot(pcadat,aes(x=PC1,y=PC2)) + -->
+<!--   geom_point(aes(fill=ncells,color=Donor),size=3,pch=21) + -->
+<!--   geom_line(aes(group=Donor,color=Donor)) + -->
+<!--   my_theme() + -->
+<!--   theme(aspect.ratio=pca$sdev[2]**2/pca$sdev[1]**2, -->
+<!--         legend.key.size=unit(1.5,'mm'), -->
+<!--         legend.key.spacing.y=unit(0,'mm'), -->
+<!--         legend.position='right') + -->
+<!--   scale_fill_distiller(palette='Spectral',limits=c(100,200000), trans='log', breaks=c(1000,10000,100000)) + -->
+<!--   labs(x=paste0('PC 1 (',100*signif(pca$sdev[1]**2/sum(pca$sdev**2),3),'%)'), -->
+<!--        y=paste0('PC 2 (',100*signif(pca$sdev[2]**2/sum(pca$sdev**2),3),'%)'), -->
+<!--        color='Donor',fill='# cells') -->
+<!-- ``` -->
 
@@
                        labels=cols_shapes %>%
-                         dplyr::pull(variable_name)) +
+                         dplyr::pull(variable)) +
     scale_size_manual(values=c('sig'=2,'ns'=.5),guide='none') +
+    scale_alpha_manual(values=c('sig'=1,'ns'=.2),guide='none') +
     guides(color=guide_legend(ncol=4),
@@
 ```{r FigS3C, fig.width=3.75 * fig_scale, fig.height=2.5 * fig_scale}
-library(gtools)
-
-cols.here.HTGTS <- c('n_homology_switch',
-                     'frac_breaks_inversions',
-                     'cluster_entropy',
-                     'mean_cluster_size',
+cols.here.HTGTS <- c('homology',
+                     'pct_inversions',
                      'acceptor_score_W',
@@
   dplyr::left_join(columns, by=c('variable'='col')) %>%
-  dplyr::mutate(variable=factor(gsub('_downsampled|_switch','',variable), 
-                                levels=gsub('_downsampled|_switch','',cols.here.HTGTS))) %>%
+  dplyr::mutate(variable=factor(variable, levels=cols.here.HTGTS)) %>%
   dplyr::filter(category %in% c('Diversity','Isotypes','Breaks','Context','Structural')) %>%
@@
   scale_fill_manual(values=genotype_colors) +
-  #scale_x_discrete(breaks=c('lymph node','splenocytes','WT','Xlf','Shld1','Shld1_Xlf'),
-  #                 limits=c('lymph node','splenocytes','','WT','Xlf','Shld1','Shld1_Xlf'), drop=FALSE) +
   labs(x='',y='', fill='') 
@@
           ncol=3)
-
-# p3 <- ggplot(pcadat,aes(x=PC3,y=PC4, color=material)) +
-#   geom_point(aes(shape=Genotype),size=1) +
-#   geom_polygon(data = pcadat %>%
-#                  dplyr::group_by(material) %>%
-#                  dplyr::slice(chull(PC3, PC4)), alpha = 0.25, aes(fill=material)) +
-#   scale_fill_manual(values=tissue_colors) +
-#   scale_color_manual(values=tissue_colors) +
-#   my_theme() +
-#   labs(x=paste0('PC 3 (',100*signif(pca$sdev[3]**2/sum(pca$sdev**2),3),'%)'),
-#        y=paste0('PC 4 (',100*signif(pca$sdev[4]**2/sum(pca$sdev**2),3),'%)'),
-#        color='',shape='',fill='', title='all switch regions') 
-# 
-# p4 <- ggplot(pcadat.noSg,aes(x=PC3,y=PC4, color=material)) +
-#   geom_point(aes(shape=Genotype),size=1) +
-#   geom_polygon(data = pcadat.noSg %>%
-#                  dplyr::group_by(material) %>%
-#                  dplyr::slice(chull(PC3, PC4)), alpha = 0.25, aes(fill=material)) +
-#   my_theme() +
-#   scale_fill_manual(values=tissue_colors) +
-#   scale_color_manual(values=tissue_colors) +
-#   labs(x=paste0('PC 3 (',100*signif(pca.noSg$sdev[3]**2/sum(pca.noSg$sdev**2),3),'%)'),
-#        y=paste0('PC 4 (',100*signif(pca.noSg$sdev[4]**2/sum(pca.noSg$sdev**2),3),'%)'),
-#        color='',shape='',fill='', title='only Sm + Sa') 
-# 
-# plot_grid(p3 + theme(legend.position='none'),
-#           p4 + theme(legend.position='none'),
-#           get_legend(p3 + theme(legend.box='vertical')),
-#           rel_widths=c(.4,.4,.2),
-#           ncol=3)
 ```
@@
 ```{r Fig3D,fig.width=3.5 * fig_scale,fig.height=2.5 * fig_scale}
-CH12.features <- c('frac_breaks_single',
-                   'mean_length',
-                   'frac_breaks_duplications',
-                   'frac_breaks_inversions',
-                   'insert_frequency',
-                   'mean_cluster_size_downsampled')
+CH12.features <- c('pct_direct_switch',
+                   'read_length_mean',
+                   'pct_duplications',
+                   'pct_inversions',
+                   'pct_templated_inserts',
+                   'cluster_size_mean')
 
@@
   gather(col,value,-c(Genotype,ng,Genotype_new)) %>% 
-  dplyr::mutate(col=factor(gsub("_downsampled|_switch",'',col),
-                           levels=gsub("_downsampled|_switch","",CH12.features)))
+  dplyr::mutate(col=factor(col, levels=CH12.features))
 
@@
 CH12.cols <- setdiff(grep('SG|Sg',intersect(selected.cols.mouse,colnames(CH12.data)),value=TRUE,invert=TRUE),
-                     c('frac_breaks_sequential','frac_clusters_Sm'))
+                     c('pct_sequential_switch','pct_clusters_Sm'))
 CH12.X <- CH12.data[CH12.samples,CH12.cols] %>%
@@
   dplyr::pull(sample)
-colnames(CH12.X) <- gsub('_switch|_downsampled','',colnames(CH12.X))
 
@@
 
-hm1 <- Heatmap(t(CH12.X[samples,gsub('_switch|_downsampled','',CH12.cols)]),
+hm1 <- Heatmap(t(CH12.X[samples,CH12.cols]),
                col=colorRamp2(c(-2,0, 2), c("blue",'gray90', "red")),
@@
 
-draw(hm1, heatmap_legend_side = "bottom", annotation_legend_side = "bottom")#, legend_grouping="original")
+draw(hm1, heatmap_legend_side = "bottom", annotation_legend_side = "bottom")
 
@@
 ```{r Fig3G, fig.width=1.25 * fig_scale, fig.height=3.5 * fig_scale}
-train_batches <- c('20220425','20221118','20221212','20230120','20230213')
+train_batches <- c('20220`4`25','20221118','20221212','20230120','20230213')
 test_batches <- '20240830'
@@
 CH12.X <- CH12.data[, CH12.cols] %>%
-  # dplyr::mutate_if(is.numeric, function(x) (x-mean(x[train]))/sd(x[train])) 
   dplyr::mutate_if(is.numeric, function(x) { 
@@
   })
-#CH12.X.test <- CH12.data[test, CH12.cols] %>%
-#  dplyr::mutate_if(is.numeric, scale) #function(x) (x-mean(x[train]))/sd(x[train]))
 
@@
   my_theme() +
-  #scale_y_discrete(labels=ifelse(grepl('\\.20|\\.27|\\.14|\\.01|\\.04',tmp$sample),"*","")) +
   theme(legend.position='none',
@@
         axis.text.x=element_text(angle=90, hjust=1, vjust=.5)) +
-  labs(x='',y='36 new samples') # (* 25ng IL4)')
-```
+  labs(x='',y='36 new samples') 
+```
+
+(here we have `r sum(train)` samples in the training data and `r sum(test)` in the testing)
 
@@
 
-col.order <- gsub('_downsampled|_switch','',tmp$col[order.hclust(hclust(dist(tmp[,2:6])))])
+col.order <- tmp$col[order.hclust(hclust(dist(tmp[,2:6])))]
 
 tmp <- tmp %>%
-  dplyr::mutate(col=gsub('_downsampled|_switch','',col),
-                col=factor(col, levels=col.order))
+  dplyr::mutate(col=factor(col, levels=col.order))
 
@@
         legend.direction='horizontal') +
-  annotation_custom(grob=anno_grob_y, ymin = 0, ymax = 1, xmin = -4, xmax = 4.5) +
+  annotation_custom(grob=anno_grob_y, ymin = 0, ymax = 1, xmin = -4., xmax = 4.5) +
   labs(x='',y='') 
@@
 
-explained variance for FB cohort (make FB_13/21 "Atypical")
-
-```{r Fig4B,fig.width=1 * fig_scale, fig.height=1.7 * fig_scale}
+explained variance for FB cohort
+
+```{r Fig4B,fig.width=1.25 * fig_scale, fig.height=1.7 * fig_scale}
 FB.data <- read.csv('data/FB_data.csv', header=1, row.names=1) %>%
@@
   dplyr::filter(grepl('2023',Batch)) %>%
-  dplyr::select(Donor,Batch,Sex,Age,Diagnosis,Timepoint)
+  dplyr::select(Donor,Batch,Sex,Age,Diagnosis,Timepoint,Treatment,IV)
 
 FB.var <- fitExtractVarPartModel(t(FB.X[,colSums(is.na(FB.X))==0]), 
-                                 ~ (1 | Donor) + (1 | Diagnosis) + (1 | Sex) + Age + (1 | Batch), FB.meta) %>%
+                                 ~ (1 | Donor) + (1 | Diagnosis) + (1 | Sex) + Age + (1 | Batch) + (1 | Treatment) + (1 | IV),
+                                 FB.meta) %>%
   as.data.frame() %>%
@@
 FB.var %>%  dplyr::left_join(columns, by='col') %>%
-  dplyr::mutate(variable=factor(variable,levels=c('Donor','Diagnosis','Sex','Age','Batch','Residuals'))) %>%
+  dplyr::mutate(variable=factor(variable,levels=c('Donor','Diagnosis','Sex','Age','Batch','Treatment','IV',
+                                                  'Complication','Residuals'))) %>%
   dplyr::filter(variable!='Residuals', category %in% c('Isotypes','Diversity','Structural','Context','Breaks','Variants')) %>%
@@
   geom_jitter(position=position_jitter(width=.25), size=.25) +
+  #geom_jitter(aes(color=ifelse(col %in% selected.cols, 'selected','unused')), position=position_jitter(width=.25), size=.25) +
+  #scale_color_manual(values=c('unused'='gray','selected'='black')) +
   my_theme() +
@@
 FB.var %>%  dplyr::left_join(columns, by='col') %>%
-  dplyr::mutate(variable=factor(variable, levels=c('Donor','Diagnosis','Sex','Age','Batch','Residuals'))) %>%
+  dplyr::mutate(variable=factor(variable, levels=c('Donor','Diagnosis','Sex','Age','Batch','Treatment','IV','Residuals'))) %>%
   dplyr::filter(variable!='Residuals', category %in% c('Isotypes','Diversity','Structural','Context','Breaks','Variants')) %>%
@@
         legend.box='horizontal') +
-  guides(fill=guide_legend(override.aes = list(shape=22))) +
+  guides(fill=guide_legend(override.aes = list(shape=22),ncol=7)) +
   scale_fill_manual(values=var_colors) +
@@
 FB.data.pooled <- read.csv('data/FB_data_pooled.csv', header=1, row.names=1) %>%
-  dplyr::mutate(num_variants_somatic = num_variants * (1 - frac_variants_germline),
-                status=ifelse(Diagnosis %in% c('Control','Subclass','Atypical'), Diagnosis, 'CVID')) %>%
+  dplyr::mutate(status=ifelse(Diagnosis %in% c('Control','Subclass','Atypical'), Diagnosis, 'CVID'),
+                Treatment=ifelse((Donor=='FB_35') | (Treatment==''),'none',Treatment),
+                IV=ifelse(IV=='','none',IV)) %>%
   tibble::column_to_rownames('Sample')
@@
 
-```{r FigS4D, fig.width=2 * fig_scale, fig.height=1.7 * fig_scale}
-subclass.cols <- c("frac_clusters_SA1","frac_clusters_SA2",
-                   "frac_clusters_SG1","frac_clusters_SG2","frac_clusters_SG3","frac_clusters_SG4")
+```{r FigS4D, fig.width=2.5 * fig_scale, fig.height=1.5 * fig_scale}
+subclass.cols <- c("pct_clusters_SA1","pct_clusters_SA2",
+                   "pct_clusters_SG1","pct_clusters_SG2","pct_clusters_SG3","pct_clusters_SG4")
 
@@
   gather(feature, frac_clusters, subclass.cols) %>%
-  dplyr::mutate(feature=factor(gsub('frac_clusters_','',feature),
-                               levels=gsub('frac_clusters_','',subclass.cols)))
+  dplyr::mutate(feature=factor(gsub('pct_clusters_','',feature),
+                               levels=gsub('pct_clusters_','',subclass.cols)))
 
@@
 
-```{r FigS4E, fig.width=2.25 * fig_scale, fig.height=2 * fig_scale}
+```{r FigS4E, fig.width=2.5 * fig_scale, fig.height=1.5 * fig_scale}
 tmp <- FB.data.pooled %>%
@@
   geom_abline(slope=1, intercept=0, size=.5) +
-  #stat_cor(method = "pearson", aes(label=..r.label..), size=2.5, show.legend=FALSE, label.x=.75, label.y=c(.4, .33)) +
   geom_point() + 
@@
 ```{r Fig4C, fig.width=4 * fig_scale, fig.height=2.7 * fig_scale}
-FB.features <- c('frac_clusters_SG2',
-                 'frac_clusters_SG3',
-                 'mean_length',
-                 'frac_breaks_single',
-                 'n_homology_switch',
-                 'frac_breaks_inversions',
-                 'mean_cluster_size_downsampled',
-                 'num_variants_somatic')
+FB.features <- c('pct_clusters_SG2',
+                 'pct_clusters_SG3',
+                 'read_length_mean',
+                 'pct_direct_switch',
+                 'homology',
+                 'pct_inversions',
+                 'cluster_size_mean',
+                 'somatic_variants')
 
@@
   gather(col,value,-c(Diagnosis_simple,ng,Diagnosis_new,Donor,Diagnosis)) %>% 
-  dplyr::mutate(col=factor(gsub("_downsampled|_switch",'',col),
-                           levels=gsub("_downsampled|_switch","",FB.features)))
+  dplyr::mutate(col=factor(col, levels=FB.features))
 
@@
   geom_point(aes(color=Diagnosis,alpha=dataset, shape=dataset), size=1) +
-  geom_line(aes(group=Donor, color=Diagnosis,alpha=dataset, shape=dataset)) +
+  geom_line(aes(group=Donor, color=Diagnosis,shape=dataset)) +
   my_theme() +
@@
                   clip='off') +
-  labs(x='PC 1', #paste0('PC 1 (',100*signif(pca$sdev[1]**2/sum(pca$sdev**2),3),'%)'),
-       y='PC 2', #paste0('PC 2 (',100*signif(pca$sdev[2]**2/sum(pca$sdev**2),3),'%)'),
+  labs(x='PC 1',
+       y='PC 2',
        color='',shape='',fill='',alpha='')
@@
                   clip='off') +
-  labs(x='PC 1', #paste0('PC 1 (',100*signif(pca$sdev[1]**2/sum(pca$sdev**2),3),'%)'),
-       y='PC 2', #paste0('PC 2 (',100*signif(pca$sdev[2]**2/sum(pca$sdev**2),3),'%)'),
+  labs(x='PC 1', 
+       y='PC 2', 
        color='',shape='',fill='',alpha='')
@@
   labs(color='',linetype='')
+```
+
+here we have `r sum(train.here)` samples in the training data set, `r sum(test.here)` in the testing dataset, and `r sum(validation)` in the validation cohort
+
+show the ridge coefficients
+
+```{r FigS4H, fig.width=2.5 * fig_scale, fig.height=1.5 * fig_scale}
+tmp <- data.frame(coef=coef(mod)[,'s0']) %>%
+  dplyr::filter(rowSums(abs(.)) > 0) %>%
+  dplyr::slice_max(rowSums(abs(.)),n=20) %>%
+  dplyr::arrange(rowSums(abs(.))) %>%
+  tibble::rownames_to_column('col') %>%
+  dplyr::filter(col %in% columns$col)
+
+col.order <- tmp$col[order(tmp$coef)]
+
+tmp <- tmp %>%
+  dplyr::mutate(col=factor(col, levels=col.order))
+
+anno_grob_x <- grid::rectGrob(x=1:length(tmp$col), y=0, gp=gpar(col=NA, fill=col_cols[rev(levels(tmp$col))], alpha=.5))
+
+tmp %>%
+  dplyr::mutate(col=factor(col, levels=rev(levels(tmp$col)))) %>%
+  ggplot(aes(x=col, y=1, fill=coef, label=round(coef,2))) + 
+  geom_tile() +
+  scale_fill_gradient2(low='pink2',high='steelblue3', mid='white', breaks=c(-.05, 0, .05), labels=c('-.05 (HD)','0','+.05 (CVID)')) +
+  my_theme() +
+  coord_cartesian(clip='off') +
+  theme(axis.text.y=element_blank(),
+        axis.ticks.y = element_blank(),
+        axis.text.x=element_text(angle=90, hjust=1, vjust=.5),
+        legend.position='right',
+        legend.direction='vertical') +
+  annotation_custom(grob=anno_grob_x, xmin = 0, xmax = 1, ymin = -2.8, ymax = 3) +
+  labs(x='',y='') 
 ```
@@
       }) %>%
-    #dplyr::mutate_if(is.numeric, scale) %>%
     as.matrix()
@@
   ggplot(aes(x=1,y=acc)) +
-  #geom_boxplot(outlier.shape=NA, lwd=.5) + 
   geom_jitter(width=.1, height=0, size=.25) + 
@@
   dplyr::pull(name)
-colnames(FB.X) <- gsub('_downsampled|_switch','',colnames(FB.X))
 pca_hm <- pcadat.combined[gsub('pooled_(FB)_([0-9]*)','\\1\\2',FB.samples), c('PC1','PC2','dataset','Donor')]
@@
                 dataset=ifelse(grepl('FB',Donor),
-                               ifelse(Sample_donor %in% FB_train_donors,'training','testing'),
+                               ifelse(Sample_donor %in% gsub('_','',FB_train_donors),'training','testing'),
                                'validation')) %>%
-  dplyr::select(Donor,Sample_donor,Diagnosis,dataset,Sex,Age,Timepoint,
-                pct_B,pct_IgA,pct_IgG)
+  dplyr::select(Donor,Sample_donor,Diagnosis,dataset,Sex,Age,Timepoint,Complication)
 
 dend <- hclust(dist(FB.X))
-# dend <- reorder(as.dendrogram(dend),wts=order(match(FB.meta %>%
-#                                                       dplyr::arrange(diagnosis) %>%
-#                                                       dplyr::mutate(Sample_donor=gsub('pooled_FB_','',Sample_donor)) %>%
-#                                                       dplyr::pull(Sample_donor),
-#                                                     row.names(FB.X))),agglo.FUN=mean)
 
@@
                   'Cluster3'=cat_cols[['Diversity']])
+
+hm_meta <- FB.meta[,c('Diagnosis','Sex','Age','Donor','Timepoint' ,'Complication'),drop=FALSE] %>% 
+  tibble::remove_rownames() %>%
+  dplyr::mutate(name=gsub('t','',gsub('FB_','',paste0(Donor,'_',Timepoint))),
+                Diagnosis=factor(Diagnosis,
+                                 levels=c('Control','Subclass','Atypical',
+                                          'CVID','ICOS','Kabuki',
+                                          'AID','IgG4def','NFKB2','PI3K','TNFRSF13B')),
+                Complication=factor(Complication, levels=c('NA','io','c'))) %>%
+  dplyr::left_join(class_df,by='name') %>%
+  dplyr::select(-c(Donor,Timepoint)) %>%
+  tibble::column_to_rownames('name')
 
@@
                show_row_dend=TRUE,
-               top_annotation=HeatmapAnnotation(df=FB.meta[,c('Diagnosis','Sex','Age',#'pct_B','pct_IgA','pct_IgG',
-                                                              'Donor','Timepoint'),drop=FALSE] %>% 
-                                                  tibble::remove_rownames() %>%
-                                                  dplyr::mutate(name=gsub('t','',gsub('FB_','',paste0(Donor,'_',Timepoint))),
-                                                                Diagnosis=factor(Diagnosis,
-                                                                                 levels=c('Control','Subclass','Atypical',
-                                                                                          'CVID','ICOS','Kabuki',
-                                                                                          'AID','IgG4def','NFKB2','PI3K','TNFRSF13B'))) %>%
-                                                  dplyr::left_join(class_df,by='name') %>%
-                                                  dplyr::select(-c(Donor,Timepoint)) %>%
-                                                  tibble::column_to_rownames('name'),
+               top_annotation=HeatmapAnnotation(df=hm_meta,
+                                                show_annotation_name=TRUE,
+                                                show_legend=TRUE,
+                                                simple_anno_size=unit(2,'mm'),
+                                                annotation_legend_param=list(title_gp=gpar(fontsize=7),
+                                                                             labels_gp=gpar(fontsize=7),
+                                                                             legend_height = unit(5, "mm"),
+                                                                             legend_width = unit(3, "mm"),
+                                                                             grid_height = unit(3, "mm"),
+                                                                             grid_width = unit(3, "mm")),
+                                                annotation_name_gp=gpar(fontsize=7),
+                                                col=list('Diagnosis'=diagnosis_colors,
+                                                         'Class'=class_colors,
+                                                         'Age'=colorRamp2(c(20,50,80),c("yellow2","orange2","coral3")),
+                                                         'Sex'=c('male'='mistyrose2',
+                                                                 'female'='lightseagreen'),
+                                                         'Complication'=c('NA'='gray80',
+                                                                          'io'='blue3',
+                                                                          'c'='red3'))),
+               column_names_gp=gpar(fontsize=4),
+               row_names_gp =gpar(fontsize = 7),
+               row_dend_width = unit(7, "mm"),
+               row_split = columns %>% tibble::column_to_rownames('col') %>% .[selected.cols, 'category'], 
+               heatmap_legend_param=list(labels_gp=gpar(fontsize=7),
+                                         title_gp=gpar(fontsize=7),
+                                         at=c(-2,0,2),
+                                         grid_height=unit(4, 'mm'),
+                                         grid_width=unit(3, 'mm'),
+                                         legend_height=unit(5,'mm'),
+                                         legend_width=unit(3,'mm')),
+               row_title=NULL,
+               name='z-score')
+
+hm2 <- Heatmap(t(FB.X),
+               col=colorRamp2(c(-2,0, 2), c("blue",'gray90', "red")),
+               column_split = grepl('09|32', row.names(FB.X)), 
+               column_title = NULL,
+               cluster_rows=TRUE,
+               show_row_names=TRUE,
+               show_column_names=TRUE,
+               show_row_dend=TRUE,
+               top_annotation=HeatmapAnnotation(df=hm_meta,
                                                 show_annotation_name=TRUE,
@@
 
-```{r FigS4C, fig.width=3 * fig_scale, fig.height=4 * fig_scale}
+```{r FigS4C, fig.width=2.5 * fig_scale, fig.height=4 * fig_scale}
 tmp <- data.frame(pca$rotation) %>%
@@
                              dplyr::filter(col %in% selected.cols) %>% 
-                             dplyr::pull(col)),
-                col_name=gsub("_downsampled|_switch",'',col))
+                             dplyr::pull(col)))
 
 cols_shapes <- tmp %>% 
-  dplyr::distinct(category, col,col_name) %>%
+  dplyr::distinct(category, col) %>%
   dplyr::mutate(var_col=cat_cols[as.character(category)]) %>%
@@
   geom_point(size=1) +
-  #geom_text_repel(aes(label=col), size=1) +
   my_theme() +
@@
                      labels=cols_shapes %>%
-                       dplyr::pull(col_name)) +
+                       dplyr::pull(col)) +
   scale_shape_manual(values=cols_shapes %>%
@@
                      labels=cols_shapes %>%
-                       dplyr::pull(col_name)) +
+                       dplyr::pull(col)) +
   theme(legend.key.size=unit(1.5,'mm'),
@@
        color='',shape='',fill='',alpha='')
-
-# ggplot(tmp,aes(x=PC1,y=PC2, color=col, shape=col)) +
-#   geom_point(size=1) +
-#   #geom_text_repel(aes(label=col), size=1) +
-#   my_theme() +
-#   scale_color_manual(values=cols_shapes %>%
-#                        dplyr::pull(var_col,col),
-#                      labels=cols_shapes %>%
-#                        dplyr::pull(col_name)) +
-#   scale_shape_manual(values=cols_shapes %>%
-#                        dplyr::pull(var_ind,col),
-#                      labels=cols_shapes %>%
-#                        dplyr::pull(col_name)) +
-#   theme(legend.key.size=unit(1.5,'mm'),
-#         aspect.ratio=pca$sdev[2]**2/pca$sdev[1]**2,
-#         legend.spacing=unit(0,'mm'),
-#         legend.key.spacing.y=unit(0,'mm'),
-#         legend.position='bottom',
-#         legend.box='horizontal') +
-#   guides(color=guide_legend(ncol=2),
-#          shape=guide_legend(ncol=2)) +
-#   labs(x='PC 1', y='PC 2',
-#        color='',shape='',fill='',alpha='')
 ```
@@
         aspect.ratio=pca$sdev[2]**2/pca$sdev[1]**2) +
-  labs(x='PC 1', #paste0('PC 1 (',100*signif(pca$sdev[1]**2/sum(pca$sdev**2),3),'%)'),
-       y='PC 2', #paste0('PC 2 (',100*signif(pca$sdev[2]**2/sum(pca$sdev**2),3),'%)'),
+  labs(x='PC 1', 
+       y='PC 2', 
        color='')
@@
 ```{r Fig4J, fig.width=5.25 * fig_scale, fig.height=1.7 * fig_scale}
-features.here <- c('frac_clusters_SG1',
-                   'mean_cluster_size_downsampled',
-                   'n_homology_switch',
-                   'n_untemplated_switch',
-                   'frac_breaks_intra',
-                   'insert_frequency')
+features.here <- c('cluster_size_mean',
+                   'pct_intraswitch_deletion',
+                   'homology',
+                   'untemplated_inserts',
+                   'pct_templated_inserts')
 
@@
   gather(col,value,features.here) %>%
-  dplyr::mutate(col=factor(gsub("_downsampled|_switch",'',col),
-                           levels=gsub("_downsampled|_switch","",features.here)))
+  dplyr::mutate(col=factor(col, levels=features.here))
 
@@
   geom_jitter(width=.1,height=0,size=.5) +
-  #annotate('text',x=2,y=Inf,label='', size=2) +
-  #annotate('text',x=6,y=Inf,label='validation', size=2) +
-  # stat_compare_means(comparisons=list(c('Cluster1','Cluster3'),
-  #                                     c('Cluster1','Cluster2'),
-  #                                     c('Cluster2','Cluster3')),
-  #                    method='wilcox.test',label = "p.signif", size=2, vjust=.5) +
-  # stat_compare_means(comparisons=list(c('Control','ATM'),
-  #                                     c('Control','BRCA1')),
-  #                    method='wilcox.test',label = "p.signif", size=2, vjust=.5) +
   stat_compare_means(comparisons=list(c(1,2),c(1,3),c(2,3)),
-                     method='wilcox.test',label = "p.signif", size=2, vjust=.5) +
+                     method='wilcox.test',label = "p.signif", size=2, vjust=.5, hide.ns=TRUE) +
   stat_compare_means(comparisons=list(c(5,6),c(5,7),c(6,7)),
-                     method='wilcox.test',label = "p.signif", size=2, vjust=.5) +
+                     method='wilcox.test',label = "p.signif", size=2, vjust=.5, hide.ns=TRUE) +
   facet_wrap(.~col,ncol=length(features.here),scales='free_y') +
@@
                 PnPS=ifelse(status=='Control','Positive',str_to_title(PnPS))) %>%
-  gather(feature, value, c('frac_clusters_SA1','frac_clusters_SG3')) %>%
+  gather(feature, value, c('pct_clusters_SA1','pct_clusters_SG3')) %>%
   gather(vacc, vacc_status, c('Tetanus','Diphteria','PnPS')) %>%
@@
         axis.text.x=element_text(angle=45, hjust=1, vjust=1))
-
-# for (x in c("splenomegaly",
-#             "lymphoproliferation",
-#             "lung")) {
-#   for (f in selected.cols) {
-#     data <- FB.data.pooled[FB.samples,c(x,f,'status')] %>%
-#       dplyr::filter(status!='control') %>%
-#       tidyr::drop_na()
-#     form <- as.formula(paste0(x,'!="" ~ ',f))
-#     res <- glm(form, family='binomial', data=data) %>% summary()
-#     pvals[[paste0(x,'_',f)]] <- data.frame(pval=coef(res)[2,4],
-#                                            vacc=x,
-#                                            col=f)
-#   }
-# }
-
-# FB.data.pooled[FB.samples, ] %>%
-#   gather(feature, value, c('frac_clusters_SG3','alpha_ratio_clusters')) %>%
-#   gather(disease, disease_status, c('splenomegaly','lymphoproliferation','lung')) %>%
-#   dplyr::mutate(disease_status=factor(ifelse(status=='control','control',
-#                                              ifelse(disease_status=='','absent','present')),
-#                                       levels=c('control','absent','present'))) %>%
-#   ggplot(aes(x=disease_status, y=value, fill=disease_status)) +
-#   geom_boxplot(outlier.shape=NA) +
-#   geom_jitter(aes(color=status),width=.1, size=.5) +
-#   stat_compare_means(comparisons=list(c('absent','present'),
-#                                       c('absent','control')),
-#                      method='wilcox.test', label = "p.signif", hide.ns=TRUE, size=2, vjust=0) +
-#   facet_grid(feature ~ disease, switch='y', scales='free_y') + 
-#   my_theme() +
-#   scale_fill_manual(values=c('present'='red3','absent'='gray')) +
-#   scale_color_manual(values=diagnosis_colors) +
-#   labs(x='',y='') +
-#   coord_cartesian(clip='off') +
-#   theme(legend.position='none',
-#         strip.clip='off',
-#         axis.text.x=element_text(angle=45, hjust=1, vjust=1))
+```
+
+show top examples for features related to Treatment or IV
+
+```{r FigS4X1, fig.width=1.5 * fig_scale, fig.height=1.5 * fig_scale}
+FB.data.pooled %>%
+  dplyr::mutate(IV=factor(IV,levels=c('none','ivig','scIg'))) %>%
+  ggplot(aes(x=IV, y=cluster_entropy)) + 
+  geom_boxplot(outlier.shape=NA, lwd=.5) +
+  geom_jitter(aes(color=status), width=.1, size=.5) + 
+  stat_compare_means(comparisons=list(c('none','ivig'), c('none','scIg'),c('ivig','scIg')),
+                     method='wilcox.test',label = "p.signif", hide.ns=TRUE, size=2.5) +
+  my_theme() + 
+  coord_cartesian(clip='off') +
+  scale_color_manual(values=diagnosis_colors) +
+  theme(legend.position='none',
+        axis.text.x=element_text(angle=45, hjust=1, vjust=1)) +
+  labs(x='')
+
+FB.data.pooled %>%
+  dplyr::mutate(Treatment=factor(Treatment,levels=c('none','cortison','infliximab','prednison','sirolimus','steroid','trixomazl'))) %>%
+  ggplot(aes(x=Treatment, y=occupancy_top_cluster)) + 
+  geom_boxplot(outlier.shape=NA, lwd=.5) +
+  geom_jitter(aes(color=status), width=.1, size=.5) + 
+  my_theme() + 
+  coord_cartesian(clip='off') +
+  scale_color_manual(values=diagnosis_colors) +
+  theme(legend.position='none',
+        axis.text.x=element_text(angle=45, hjust=1, vjust=1)) +
+  labs(x='')
 ```
@@
   dplyr::group_by(category) %>%
-  dplyr::mutate(#value=(value-mean(value[Diagnosis=='control'])/sd(value[Diagnosis=='control'])),
-                name=gsub('t','',gsub('FB_','',paste0(Donor,'_',Timepoint)))) %>%
+  dplyr::mutate(name=gsub('t','',gsub('FB_','',paste0(Donor,'_',Timepoint)))) %>%
   dplyr::select(name,category,value) %>%

### 2025-07-31 prepare_input.R

@@ -2,26 +2,28 @@ library(dplyr)
 library(tidyverse)
 library(scales)
 
-columns <- list('Diversity'=c('nclusters_final_downsampled',
-                              'nclusters_eff_downsampled',
-                              'mean_cluster_size_downsampled',
-                              'std_cluster_size_downsampled',
-                              'cluster_gini_downsampled',
-                              'cluster_entropy_downsampled',
-                              'cluster_inverse_simpson_downsampled',
-                              'top_clone_occupancy_downsampled',
-                              'big_clones_occupancy_downsampled'),
+set.seed(1)
+
+columns <- list('Diversity'=c('clusters',
+                              'clusters_eff',
+                              'cluster_size_mean',
+                              'cluster_size_std',
+                              'cluster_gini',
+                              'cluster_entropy',
+                              'cluster_inverse_simpson',
+                              'occupancy_top_cluster',
+                              'occupancy_big_clusters'),
                 'Isotypes'=c('alpha_ratio_clusters',
-                             'frac_clusters_SA1','frac_clusters_SA2','frac_clusters_SE',
-                             'frac_clusters_SG1','frac_clusters_SG2','frac_clusters_SG3',
-                             'frac_clusters_SG4','frac_clusters_SM',
-                             'frac_clusters_Sa','frac_clusters_Sg3',
-                             'frac_clusters_Sg2b','frac_clusters_Sg2c',
-                             'frac_clusters_Sm','frac_clusters_Sg1'),
-                'Structural'=c('insert_frequency',
-                               'frac_breaks_inversions','frac_breaks_duplications',
-                               'median_duplication_size','median_inversion_size'),
-                'Context'=c('homology_fw','homology_rv',
+                             'pct_clusters_SA1','pct_clusters_SA2','pct_clusters_SE',
+                             'pct_clusters_SG1','pct_clusters_SG2','pct_clusters_SG3',
+                             'pct_clusters_SG4','pct_clusters_SM',
+                             'pct_clusters_Sa','pct_clusters_Sg3',
+                             'pct_clusters_Sg2b','pct_clusters_Sg2c',
+                             'pct_clusters_Sm','pct_clusters_Sg1'),
+                'Structural'=c('pct_templated_inserts',
+                               'pct_inversions','pct_duplications',
[…]

### 2025-07-31 swibrid_runs/human/meta_clustering.py

@@ -1,0 +1,79 @@
+import numpy as np
+import pandas as pd
+import fastcluster
+import glob
+import scipy.sparse
+import os
+from swibrid.utils import remove_gaps
+from argparse import ArgumentParser
+
+parser=ArgumentParser()
+parser.add_argument('-d','--donor')
+parser.add_argument('-r','--replicate', default=3)
+parser.add_argument('-s','--seed', type=int)
+parser.add_argument('-n','--ncells',type=int)
+parser.add_argument('-o','--out')
+parser.add_argument('--samples')
+args=parser.parse_args()
+
+study=pd.read_csv('../../sodar/2020_CellSwitch/s_2020_CellSwitch.txt', sep='\t',header=0)
+assay=pd.read_csv('../../sodar/2020_CellSwitch/a_2020_CellSwitch.txt', sep='\t',header=0)
+QC=assay.set_index('Sample Name')['Parameter Value[QC]']
+
+if args.samples:
+    samples=args.samples.split(',')
+else:
+    ncells=args.ncells
+    if args.donor and args.donor!='mixed':
+        donor=args.donor
+        if args.seed is not None:
+            seed=args.seed
+            np.random.seed(seed)
+            samples=study['Sample Name'][(study['Source Name'].astype(str)==donor).values & (QC[study['Sample Name']] == 'PASS').values & (study['Parameter Value[Cell Number]']==ncells).values].sample(2).values
+        else:
+            samples=study['Sample Name'][(study['Source Name'].astype(str)==donor).values & (QC[study['Sample Name']] == 'PASS').values & (study['Parameter Value[Cell Number]']==ncells).values].values
+    else:
+        seed=args.seed
+        np.random.seed(seed)
+        samples=[study['Sample Name'][(study['Source Name'].astype(str)==donor).values & (QC[study['Sample Name']] == 'PASS').values & (study['Parameter Value[Cell Number]']==ncells).values].sample(1).values[0] for donor in np.random.choice(['21084','21085','21086'], size=2)]
+    
+nsamples=len(samples)
+print(f"meta-clustering for samples " + ",".join(samples))
+
+msa=dict((sample,scipy.sparse.load_npz(f'pipeline/{sample}/{sample}_msa.npz')) for sample in samples)
[…]

### 2025-07-31 swibrid_runs/human/meta_clustering.sh

@@ -1,0 +1,12 @@
+#!/bin/bash -e
+
+# use this to cluster two random samples of donor ${donor} with ${ncells} cells
+#python meta_clustering.py -s ${seed} -d ${donor} -n ${ncells} -o meta_clustering/${donor}_${seed}_${ncells}_meta_clustering.csv
+# use this to cluster all replicates of donor ${donro} with ${ncells} cells
+#python meta_clustering.py -d ${donor} -n ${ncells} -o meta_clustering/${donor}_${ncells}_meta_clustering.csv
+# use this to cluster minION and pacBio samples of HD19005
+#python meta_clustering.py --samples 20230313_HD19005,20230504_HD19005 -o meta_clustering/HD19005_meta_clustering.csv
+# use this to cluster replicates with different PCR cycle numbers
+#python meta_clustering.py --samples 20250703_${donor}x15,20250703_${donor}x20,20250703_${donor}x25 -o meta_clustering/${donor}_meta_clustering.csv
+
+

### 2025-08-01 paper_figures.Rmd
update for revision

@@
 ```{r Fig3G, fig.width=1.25 * fig_scale, fig.height=3.5 * fig_scale}
-train_batches <- c('20220`4`25','20221118','20221212','20230120','20230213')
+train_batches <- c('20220425','20221118','20221212','20230120','20230213')
 test_batches <- '20240830'

### 2025-10-22 paper_figures.Rmd
re-order source data

@@
 library(dendextend)
+library(openxlsx)
 ```
@@
 # simulation data for Fig. 1
-synthetic.data <- read.csv('data/synthetic_data.csv',header=1, row.names=1)
+#synthetic.data <- read.csv('data/synthetic_data.csv',header=1, row.names=1)
+synthetic.data <- read.xlsx('data/source_data.xlsx',sheet='synthetic data')
 ggplot(synthetic.data %>% 
@@
 # cell number experiment for  Fig. 1
-cellswitch.data <- read.csv('data/cellswitch_data.csv',header=1, row.names=1)
+cellswitch.data <- read.xlsx('data/source_data.xlsx',sheet='cell numbers')
 
@@
 ```{r Fig1F, fig.width=1.4 * fig_scale, fig.height=1.4 * fig_scale}
-meta_clustering <- read.csv('data/cellswitch_meta_clustering.csv', header=1, row.names=1)
+meta_clustering <- read.xlsx('data/source_data.xlsx', sheet='meta-clustering')
 
@@
   my_theme() +
-  #theme(legend.position='bottom') +
-  #coord_cartesian(ylim=c(15,40)) +
   scale_fill_manual(values=c('gray','white'), guide=FALSE) +
@@
 # bulk VDJ results for Fig. 1
-VDJ_results <- read.csv('data/VDJ_results.csv',header=1, row.names=1)
+VDJ_results <- read.xlsx('data/source_data.xlsx',sheet='VDJ data')
 
@@
   labs(y='Fraction IGHA SWIBRID',x='Fraction IGHA VDJ',shape='# cells')
-```
-
-```{r FigS1X1, fig.width=1.4 * fig_scale, fig.height=1.4 * fig_scale}
-vdj_df %>% 
-  dplyr::mutate(ncells=factor(gsub('000$','k',ncells), levels=c('50k','100k')),
-                Donor=gsub('Donor ','',Donor)) %>%
-  dplyr::filter(metric=='cluster_gini_raw') %>%
-  dplyr::group_by(ncells, Donor, method) %>%
-  dplyr::summarise(mean_gini=mean(value),
-                   se_gini=sd(value)/sqrt(length(value))) %>%
-  pivot_wider(id_cols=c(ncells,Donor),names_from=method,values_from=c('mean_gini','se_gini')) %>%
-  ggplot(aes(x=mean_gini_VDJ, y=mean_gini_SWIBRID,
-             xmin=mean_gini_VDJ-se_gini_VDJ,
-             xmax=mean_gini_VDJ+se_gini_VDJ,
-             ymin=mean_gini_SWIBRID-se_gini_SWIBRID,
-             ymax=mean_gini_SWIBRID+se_gini_SWIBRID)) +
-  geom_point(aes(color=Donor,shape=ncells),size=1) +
-  geom_errorbar(linewidth=.25) +
-  geom_errorbarh(linewidth=.25) +
-  #scale_x_continuous(breaks=c(.6,.65)) +
-  my_theme() +
-  scale_color_brewer(palette='Set1') +
-  theme(strip.background=element_blank()) +
-  labs(y='Gini coefficient SWIBRID',x='Gini coefficient VDJ',shape='# cells')
-```
-
-```{r FigS1X2, fig.width=2 * fig_scale, fig.height=1.4 * fig_scale}
-library(ggpmisc)
-PCR.data <- read.csv('data/PCR_data.csv',header=1,row.names=1)
-ggplot(PCR.data, aes(x=as.integer(ncycles), y=log10(reads))) + 
-  geom_point(aes(color=donor)) + 
-  geom_smooth(method='lm', color='black', linewidth=.5) +
-  geom_abline(slope=log10(2), intercept=-1.5, linewidth=.5, linetype='dashed') +
-  my_theme() + 
-  scale_color_brewer(palette='Set2') +
-  scale_y_continuous(breaks = log10(c(1, 10, 100, 1000,10000, 100000)),
-                     label = c(1, 10, 100, 1000, 10000, 100000)) +
-  labs(x='# of cycles',y='# of reads')
 ```
@@
 # HD data for Fig. 2
-HD.data <- read.csv('data/HD_data.csv', header=1, row.names=1)
+HD.data <- read.xlsx('data/source_data.xlsx', sheet='HD data')
 
@@
 
-```{r Fig2D, fig.width=1.5 * fig_scale,fig.height=1.75 * fig_scale}
+```{r Fig2C, fig.width=1.5 * fig_scale,fig.height=1.75 * fig_scale}
 donor_cutoff <- .25
@@
 
-```{r FigS2J_1, fig.width=1.4 * fig_scale, fig.height=1.7 * fig_scale}
+```{r FigS2I_1, fig.width=1.4 * fig_scale, fig.height=1.7 * fig_scale}
 tmp <- HD.var %>%  
@@
 
-```{r FigS2J_2, fig.width=1.4 * fig_scale, fig.height=1.7 * fig_scale}
+```{r FigS2I_2, fig.width=1.4 * fig_scale, fig.height=1.7 * fig_scale}
 tmp <- HD.var %>%  
@@
 
-```{r FigS2J_3, fig.width=1.4 * fig_scale,fig.height=1.7 * fig_scale}
-HD.data.br <- read.csv('data/HD_data_by_reads.csv', header=1, row.names=1)
+```{r FigS2I_3, fig.width=1.4 * fig_scale,fig.height=1.7 * fig_scale}
+HD.data.br <- read.xlsx('data/source_data.xlsx', sheet='HD data by reads')
 
@@
 
-```{r Fig2C, fig.width=4.25 * fig_scale, fig.height=1.5 * fig_scale}
+```{r Fig2D, fig.width=4.25 * fig_scale, fig.height=1.5 * fig_scale}
 cols.here.human <- c('pct_templated_inserts',
@@
 ```{r Fig2F, fig.width=4. * fig_scale, fig.height=1.75 * fig_scale}
-mouse.data <- read.csv('data/mouse_data.csv',header=1, row.names=1)  %>%
+mouse.data <- read.xlsx('data/source_data.xlsx', sheet='mouse tissues') %>%
   tibble::column_to_rownames('Sample')
-vincendeau.data <- read.csv('data/vincendeau_data.csv', header=1, row.names=1)  %>%
+vincendeau.data <- read.xlsx('data/source_data.xlsx', sheet='Vincendeau data') %>%
   tibble::column_to_rownames('Sample')
@@
 
-```{r FigS2K, fig.width=1.5 * fig_scale, fig.height=1.75 * fig_scale}
+```{r FigS2J, fig.width=1.5 * fig_scale, fig.height=1.75 * fig_scale}
 cols.here.mouse <- c('read_length_SG_mean')
@@
     dplyr::mutate(tstat=abs(mean2-mean1)/(sqrt(se2**2+se1**2)),
-                  pval=dt(tstat,4)) %>%
+                  pval=2*pt(-abs(tstat),4)) %>%
     dplyr::mutate(padj=p.adjust(pval, method='BH'),
@@
 ```
-
-<!-- the PCR cycles experiment -->
-
-<!-- ```{r PCR_cycles} -->
-<!-- PCR.data <- read.csv('../human_samples/20250703_PCR_cycles/output/summary/20250703_PCR_cycles_stats.csv',header=1,row.names=1) %>% -->
-<!--   tibble::rownames_to_column('sample') %>% -->
-<!--   dplyr::mutate(donor=gsub('[0-9]*_(P[0-9])x([0-9]*)','\\1',sample), -->
-<!--                 ncycles=gsub('[0-9]*_(P[0-9])x([0-9]*)','\\2',sample))  -->
-
-<!-- PCR.X <- PCR.data[,c('sample',selected.cols)] %>% -->
-<!--   dplyr::mutate_if(is.numeric,scale) %>% -->
-<!--   tibble::column_to_rownames('sample') -->
-
-<!-- PCR.meta <- PCR.data[,c('ncycles','donor')] -->
-
-<!-- cols.here <- c('reads','clusters_raw','PCR_bias_length','PCR_bias_GC') -->
-<!-- PCR.data %>% -->
-<!--   gather(variable, value, cols.here) %>% -->
-<!--   dplyr::select(variable, value, ncycles, donor) %>% -->
-<!--   ggplot(aes(x=ncycles, y=value, fill=ncycles)) +  -->
-<!--   geom_boxplot(position=position_dodge(width=.75), width=.7, outlier.shape=NA, lwd=.25) +  -->
-<!--   geom_point(aes(color=donor), position=position_jitterdodge(jitter.width=.2, dodge.width=.75, jitter.height = 0), size=.5) +  -->
-<!--   my_theme() + -->
-<!--   coord_cartesian(clip='off') + -->
-<!--   facet_wrap(~variable, switch='y', scales='free_y') + -->
-<!--   theme(axis.text.x=element_text(angle=45, hjust=1, vjust=1), -->
-<!--         legend.position='none') + -->
-<!--   labs(x='',y='', fill='')  -->
-
-<!-- take <- (rowSums(is.na(PCR.X))==0) &  (PCR.meta$ncycles %in% c('15','20','25')) -->
-<!-- pca <- prcomp(PCR.X[take,]) -->
-<!-- pcadat <- cbind(pca$x,PCR.meta[take,]) -->
-
-<!-- ggplot(pcadat,aes(x=PC1,y=PC2)) + -->
-<!--   geom_point(aes(color=donor, shape=ncycles),size=3) + -->
-<!--   my_theme() + -->
-<!--   guides(color=guide_legend(override.aes = list(shape=15,size=2), ncol=3)) + -->
-<!--   theme(aspect.ratio=pca$sdev[2]**2/pca$sdev[1]**2, -->
-<!--         legend.key.size=unit(1.5,'mm'), -->
-<!--         legend.key.spacing.y=unit(0,'mm'), -->
-<!--         legend.position='bottom', -->
-<!--         legend.box='horizontal') + -->
-<!--   labs(x=paste0('PC 1 (',100*signif(pca$sdev[1]**2/sum(pca$sdev**2),3),'%)'), -->
-<!--        y=paste0('PC 2 (',100*signif(pca$sdev[2]**2/sum(pca$sdev**2),3),'%)'), -->
-<!--        color='',shape='',fill='') -->
-<!-- ``` -->
-
-<!-- ```{r cell_switch} -->
-<!-- cellswitch_sample_sheet <- read.csv('../sodar/2020_CellSwitch/a_2020_CellSwitch.txt', -->
-<!--                             sep='\t',header=1) %>% -->
-<!--   dplyr::left_join(read.csv('../sodar/2020_CellSwitch/s_2020_CellSwitch.txt',sep='\t',header=1), -->
-<!--                    by='Sample.Name') %>% -->
-<!--   dplyr::mutate(Donor=as.character(Source.Name), -->
-<!--                 sample=Sample.Name, -->
-<!--                 ncells=Parameter.Value.Cell.Number., -->
-<!--                 Batch=as.character(Parameter.Value.Batch.), -->
-<!--                 QC=as.character(Parameter.Value.QC.)) %>% -->
-<!--   dplyr::select(Donor,sample,ncells,Batch,QC) %>% -->
-<!--   dplyr::filter(Donor %in% c('21084','21085','21086')) -->
-
-<!-- cellswitch.data <- read.csv('../human_samples/20240307_hg38_results/output/summary/20240307_human_samples_stats.csv', -->
-<!--                           header=1,row.names=1) %>% -->
-<!--   tibble::rownames_to_column('sample') %>% -->
-<!--   dplyr::inner_join(cellswitch_sample_sheet, by='sample') %>% -->
-<!--   dplyr::filter(reads >= 500, QC=='PASS') %>% -->
-<!--   dplyr::mutate_all(~replace_na(.,0)) %>% -->
-<!--   dplyr::mutate(Donor=plyr::revalue(Donor, c('21084'='A', -->
-<!--                                              '21085'='B', -->
-<!--                                              '21086'='C'))) %>% -->
-<!--   dplyr::select(Donor,ncells,Batch,sample,selected.cols) -->
-
-<!-- cellswitch.X <- cellswitch.data[,c('sample',selected.cols)] %>% -->
-<!--   dplyr::mutate_if(is.numeric,scale) %>% -->
-<!--   tibble::column_to_rownames('sample') -->
-
-<!-- cellswitch.meta <- cellswitch.data[,c('Donor','ncells')] -->
-
-<!-- take <- (rowSums(is.na(cellswitch.X))==0) #& (cellswitch.meta$ncells > 10000) -->
-<!-- pca <- prcomp(cellswitch.X[take,]) -->
-<!-- pcadat <- cbind(pca$x,cellswitch.meta[take,]) -->
-
-<!-- ggplot(pcadat,aes(x=PC1,y=PC2)) + -->
-<!--   geom_point(aes(fill=ncells),size=3,pch=21) + -->
-<!--   geom_line(aes(group=Donor,color=Donor)) + -->
-<!--   my_theme() + -->
-<!--   theme(aspect.ratio=pca$sdev[2]**2/pca$sdev[1]**2, -->
-<!--         legend.key.size=unit(1.5,'mm'), -->
-<!--         legend.key.spacing.y=unit(0,'mm'), -->
-<!--         legend.position='right') + -->
-<!--   scale_fill_distiller(palette='Spectral',limits=c(100,200000), trans='log', breaks=c(1000,10000,100000)) + -->
-<!--   labs(x=paste0('PC 1 (',100*signif(pca$sdev[1]**2/sum(pca$sdev**2),3),'%)'), -->
-<!--        y=paste0('PC 2 (',100*signif(pca$sdev[2]**2/sum(pca$sdev**2),3),'%)'), -->
-<!--        color='Donor',fill='# cells') -->
-
-<!-- take <- (rowSums(is.na(cellswitch.X))==0) & (cellswitch.meta$ncells > 10000) -->
-<!-- pca <- prcomp(cellswitch.X[take,]) -->
-<!-- pcadat <- cbind(pca$x,cellswitch.meta[take,]) -->
-
-<!-- ggplot(pcadat,aes(x=PC1,y=PC2)) + -->
-<!--   geom_point(aes(fill=ncells,color=Donor),size=3,pch=21) + -->
-<!--   geom_line(aes(group=Donor,color=Donor)) + -->
-<!--   my_theme() + -->
-<!--   theme(aspect.ratio=pca$sdev[2]**2/pca$sdev[1]**2, -->
-<!--         legend.key.size=unit(1.5,'mm'), -->
-<!--         legend.key.spacing.y=unit(0,'mm'), -->
-<!--         legend.position='right') + -->
-<!--   scale_fill_distiller(palette='Spectral',limits=c(100,200000), trans='log', breaks=c(1000,10000,100000)) + -->
-<!--   labs(x=paste0('PC 1 (',100*signif(pca$sdev[1]**2/sum(pca$sdev**2),3),'%)'), -->
-<!--        y=paste0('PC 2 (',100*signif(pca$sdev[2]**2/sum(pca$sdev**2),3),'%)'), -->
-<!--        color='Donor',fill='# cells') -->
-<!-- ``` -->
-
 # Figure 3 + supplement (CH12)
@@
 ```{r FigS3A, fig.width=5.5 * fig_scale, fig.height=4.5 * fig_scale}
-HTGTS.data <- read.csv("data/HTGTS_data.csv", header=1, row.names=1) %>%
+HTGTS.data <- read.xlsx("data/source_data.xlsx", sheet='HTGTS data') %>%
   tibble::column_to_rownames('Sample')
@@
     dplyr::mutate(tstat=abs(mean2-mean1)/(sqrt(se2**2+se1**2)),
-                  pval=dt(tstat,4),
+                  pval=2*pt(-abs(tstat),4),
                   padj=p.adjust(pval, method='BH'),
@@
 ```{r FigS3D, fig.width=3.25 * fig_scale, fig.height=1.25 * fig_scale}
-read.csv('data/HTGTS_homology.csv',header=1, row.names=1) %>%
+read.xlsx('data/source_data.xlsx', sheet='HTGTS homology') %>%
   dplyr::mutate(Genotype=factor(Genotype, levels=c('WT','Atm','Lig4','Trp53bp1')),
-                celltype=gsub('spleen','Spleen',celltype)) %>%
+                celltype=gsub('spleen','Spleen',celltype),
+                n_homology=as.numeric(n_homology)) %>%
   dplyr::filter(xor(isotype=='Sa', celltype=='Spleen')) %>%
@@
 ```{r Fig3B, fig.width=3.5 * fig_scale, fig.height=1.75 * fig_scale}
-CH12.data <- read.csv('data/CH12_data.csv',header=1, row.names=1) %>%
+CH12.data <- read.xlsx('data/source_data.xlsx',sheet='CH12 data') %>%
   tibble::column_to_rownames('Sample')
-CH12.data.noSg <- read.csv('data/CH12_data_noSg.csv', header=1, row.names=1) %>%
+CH12.data.noSg <- read.xlsx('data/source_data.xlsx',sheet='CH12 data (no Sg)') %>%
   tibble::column_to_rownames('Sample')
-mouse.data.noSg <- read.csv('data/mouse_data_noSg.csv', header=1, row.names=1)
+mouse.data.noSg <- read.xlsx('data/source_data.xlsx',sheet='mouse tissues (no Sg)')
 
@@
 ```{r Fig4B,fig.width=1.25 * fig_scale, fig.height=1.7 * fig_scale}
-FB.data <- read.csv('data/FB_data.csv', header=1, row.names=1) %>%
+FB.data <- read.xlsx('data/source_data.xlsx', sheet='CVID data') %>%
   dplyr::mutate(Batch=as.character(Batch)) %>%
@@
   geom_jitter(position=position_jitter(width=.25), size=.25) +
-  #geom_jitter(aes(color=ifelse(col %in% selected.cols, 'selected','unused')), position=position_jitter(width=.25), size=.25) +
-  #scale_color_manual(values=c('unused'='gray','selected'='black')) +
   my_theme() +
@@
 ```{r Fig4D, fig.width=2.25 * fig_scale, fig.height=2 * fig_scale} 
-FB.data.pooled <- read.csv('data/FB_data_pooled.csv', header=1, row.names=1) %>%
+FB.data.pooled <- read.xlsx('data/source_data.xlsx',sheet='CVID data pooled') %>%
   dplyr::mutate(status=ifelse(Diagnosis %in% c('Control','Subclass','Atypical'), Diagnosis, 'CVID'),
@@
   tibble::column_to_rownames('Sample')
-QP.data <- read.csv('data/QP_data.csv', header=1, row.names=1) %>%
+QP.data <- read.xlsx('data/source_data.xlsx', sheet='DNA repair data') %>%
   dplyr::mutate(Diagnosis=plyr::revalue(Diagnosis,c('control'='Control'))) %>%
@@
 
-```{r FigS4H, fig.width=2.5 * fig_scale, fig.height=1.5 * fig_scale}
+```{r FigS4G, fig.width=2.5 * fig_scale, fig.height=1.5 * fig_scale}
 tmp <- data.frame(coef=coef(mod)[,'s0']) %>%
@@
 
-```{r FigS4G, fig.width=2.5 * fig_scale, fig.height=2 * fig_scale}
+```{r FigS4H, fig.width=2.5 * fig_scale, fig.height=2 * fig_scale}
 ggplot(pca_hm %>% tibble::rownames_to_column('name') %>%
@@
         axis.text.x=element_text(angle=45, hjust=1, vjust=1))
-```
-
-show top examples for features related to Treatment or IV
-
-```{r FigS4X1, fig.width=1.5 * fig_scale, fig.height=1.5 * fig_scale}
-FB.data.pooled %>%
-  dplyr::mutate(IV=factor(IV,levels=c('none','ivig','scIg'))) %>%
-  ggplot(aes(x=IV, y=cluster_entropy)) + 
-  geom_boxplot(outlier.shape=NA, lwd=.5) +
-  geom_jitter(aes(color=status), width=.1, size=.5) + 
-  stat_compare_means(comparisons=list(c('none','ivig'), c('none','scIg'),c('ivig','scIg')),
-                     method='wilcox.test',label = "p.signif", hide.ns=TRUE, size=2.5) +
-  my_theme() + 
-  coord_cartesian(clip='off') +
-  scale_color_manual(values=diagnosis_colors) +
-  theme(legend.position='none',
-        axis.text.x=element_text(angle=45, hjust=1, vjust=1)) +
-  labs(x='')
-
-FB.data.pooled %>%
-  dplyr::mutate(Treatment=factor(Treatment,levels=c('none','cortison','infliximab','prednison','sirolimus','steroid','trixomazl'))) %>%
-  ggplot(aes(x=Treatment, y=occupancy_top_cluster)) + 
-  geom_boxplot(outlier.shape=NA, lwd=.5) +
-  geom_jitter(aes(color=status), width=.1, size=.5) + 
-  my_theme() + 
-  coord_cartesian(clip='off') +
-  scale_color_manual(values=diagnosis_colors) +
-  theme(legend.position='none',
-        axis.text.x=element_text(angle=45, hjust=1, vjust=1)) +
-  labs(x='')
 ```

### 2025-10-22 prepare_input.R

@@ -1,6 +1,7 @@
 library(dplyr)
 library(tidyverse)
 library(scales)
+library(openxlsx)
 
 set.seed(1)
 
@@ -102,6 +103,46 @@ columns <- list('Diversity'=c('clusters',
                                          'Breaks',
                                          'Variants')))
 
+wb <- createWorkbook()
+addWorksheet(wb, sheetName='description')
+addWorksheet(wb, sheetName='synthetic data')
+addWorksheet(wb, sheetName='cell numbers')
+addWorksheet(wb, sheetName='meta-clustering')
+addWorksheet(wb, sheetName='VDJ data')
+addWorksheet(wb, sheetName='HD data')
+addWorksheet(wb, sheetName='HD data by reads')
+addWorksheet(wb, sheetName='Vincendeau data')
+addWorksheet(wb, sheetName='HTGTS data')
+addWorksheet(wb, sheetName='HTGTS homology')
+addWorksheet(wb, sheetName='mouse tissues')
+addWorksheet(wb, sheetName='mouse tissues (no Sg)')
+addWorksheet(wb, sheetName='CH12 data')
+addWorksheet(wb, sheetName='CH12 data (no Sg)')
+addWorksheet(wb, sheetName='CVID data')
+addWorksheet(wb, sheetName='CVID data pooled')
+addWorksheet(wb, sheetName='DNA repair data')
+
+description <- data.frame(
+  rbind(c('synthetic data','data from simulation experiments','1D'),
+        c('cell numbers','data from runs with different cell numbers','1E, 1F, 1G'),
+        c('meta-clustering','meta-clustering of cell number data','1F'),
+        c('VDJ data','data from bulk VDJ sequencing','1G'),
+        c('HD data','data for cohort C1+C2','2B, 2C, 2D, S2F, S2G, S2I'),
+        c('HD data by reads','data for cohort C1+C2 (no clustering)','S2I'),
+        c('CVID data','data for CVID cohort + controls','2D, 4B, S4A'),
+        c('Vincendeau data','data from Vincendeau et al.','2F, 2G, 2H, S2J'),
+        c('HTGTS data','re-processed HTGTS data','S3A-C'),
+        c('HTGTS homology','homology results for HTGTS data','S3D'),
[…]

### 2025-10-22 swibrid_runs/human/plot_clustering.sh

@@ -1,6 +1,6 @@
 #!/bin/bash -e
 
-for sample in 20231211_rFP_01_downsampled 20250113_SKA1_downsampled 20211122_85_n500_1_downsampled; do
+for sample in 20231211_rFP_01_downsampled 20250113_SKA1_downsampled 20211122_85_n500_1_downsampled 20210422_plasmid_downsampled; do
 	swibrid plot_clustering --msa pipeline/${sample}/${sample}_msa.npz --figure ${sample}_reads_by_cluster.png --linkage pipeline/${sample}/${sample}_linkage.npz --info input/${sample}_info.csv  --clustering_results pipeline/${sample}/${sample}_clustering.csv --switch_annotation hg38_switch_regions.bed --switch_coords "chr14:105583000-105872000:-" --color_by cluster --clustering_stats pipeline/${sample}/${sample}_cluster_stats.csv --sample ${sample} --dpi 500 --fig_width 4 --fig_height 1.25 --cmax 0.5 --linkage_border .2 --cutoff 0.01 --cmax 0.1 --no_x_ticks --omit_scale_bar --dpi 600
 done
 

### 2025-11-13 paper_figures.Rmd
create source data file

@@
 author: "Benedikt Obermayer"
-date: "2024/12/19"
+date: "2025/11/11"
 output: 
@@
 }
+
+wb <- createWorkbook()
+addWorksheet(wb, sheetName='Fig1D')
+addWorksheet(wb, sheetName='Fig1E')
+addWorksheet(wb, sheetName='Fig1F')
+addWorksheet(wb, sheetName='Fig1G')
+addWorksheet(wb, sheetName='Fig2B')
+addWorksheet(wb, sheetName='Fig2C')
+addWorksheet(wb, sheetName='Fig2D')
+addWorksheet(wb, sheetName='Fig2E')
+addWorksheet(wb, sheetName='Fig2F')
+addWorksheet(wb, sheetName='Fig2G')
+addWorksheet(wb, sheetName='Fig2H')
+addWorksheet(wb, sheetName='Fig3B')
+addWorksheet(wb, sheetName='Fig3C')
+addWorksheet(wb, sheetName='Fig3D')
+addWorksheet(wb, sheetName='Fig3E')
+addWorksheet(wb, sheetName='Fig3F')
+addWorksheet(wb, sheetName='Fig3G')
+addWorksheet(wb, sheetName='Fig4B')
+addWorksheet(wb, sheetName='Fig4C')
+addWorksheet(wb, sheetName='Fig4D')
+addWorksheet(wb, sheetName='Fig4E')
+addWorksheet(wb, sheetName='Fig4F')
+addWorksheet(wb, sheetName='Fig4G')
+addWorksheet(wb, sheetName='Fig4H')
+addWorksheet(wb, sheetName='Fig4J')
+addWorksheet(wb, sheetName='FigS2F')
+addWorksheet(wb, sheetName='FigS2G')
+addWorksheet(wb, sheetName='FigS2H')
+addWorksheet(wb, sheetName='FigS2I')
+addWorksheet(wb, sheetName='FigS2J')
+addWorksheet(wb, sheetName='FigS3A')
+addWorksheet(wb, sheetName='FigS3B')
+addWorksheet(wb, sheetName='FigS3C')
+addWorksheet(wb, sheetName='FigS3D')
+addWorksheet(wb, sheetName='FigS4A')
+addWorksheet(wb, sheetName='FigS4B')
+addWorksheet(wb, sheetName='FigS4C')
+addWorksheet(wb, sheetName='FigS4D')
+addWorksheet(wb, sheetName='FigS4E')
+addWorksheet(wb, sheetName='FigS4F')
+addWorksheet(wb, sheetName='FigS4G')
+addWorksheet(wb, sheetName='FigS4H')
 ```
@@
 #synthetic.data <- read.csv('data/synthetic_data.csv',header=1, row.names=1)
-synthetic.data <- read.xlsx('data/source_data.xlsx',sheet='synthetic data')
-ggplot(synthetic.data %>% 
-         dplyr::mutate(nreads=factor(nreads, levels=c('10k','20k','50k','100k'))), 
-       aes(x=nclones, y=clusters_raw, color=nreads)) + 
+synthetic.data <- read.xlsx('data/input_data.xlsx',sheet='synthetic data')
+dat <- synthetic.data %>% 
+         dplyr::mutate(nreads=factor(nreads, levels=c('10k','20k','50k','100k'))) %>%
+  dplyr::select(nclones, clusters_raw, nreads)
+ggplot(dat, aes(x=nclones, y=clusters_raw, color=nreads)) + 
   geom_abline(slope=1,intercept=0, size=.5, color='gray80') +
@@
   labs(x='# Clones', y='# Filtered clusters', color='# Reads')
+
+writeData(wb, 'Fig1D', as.data.frame(dat))
 ```
@@
 # cell number experiment for  Fig. 1
-cellswitch.data <- read.xlsx('data/source_data.xlsx',sheet='cell numbers')
-
-ggplot(cellswitch.data %>%
+cellswitch.data <- read.xlsx('data/input_data.xlsx',sheet='cell numbers')
+dat <- cellswitch.data %>%
          dplyr::mutate(Donor=plyr::revalue(Donor,
@@
                                              '21085'='B',
-                                             '21086'='C'))),
-       aes(x=ncells, y=clusters_raw)) + 
+                                             '21086'='C'))) %>%
+  dplyr::select(ncells, clusters_raw, Donor)
+ggplot(dat, aes(x=ncells, y=clusters_raw)) + 
   geom_abline(slope=1,intercept=0, size=.5, color='gray80') +
@@
   labs(x='# Input cells', y='# Filtered clusters', color='Donor')
+
+writeData(wb, 'Fig1E', as.data.frame(dat))
 ```
@@
 ```{r Fig1F, fig.width=1.4 * fig_scale, fig.height=1.4 * fig_scale}
-meta_clustering <- read.xlsx('data/source_data.xlsx', sheet='meta-clustering')
-
-ggplot(meta_clustering %>% 
+meta_clustering <- read.xlsx('data/input_data.xlsx', sheet='meta-clustering')
+dat <- meta_clustering %>% 
          dplyr::mutate(ncells=factor(ncells, levels=c('50k','100k')),
                        Donor=ifelse(Donor=='mixed','mixed',paste0(Donor,'-',Donor)),
-                       mixing=gsub('-donor','',mixing)), 
-       aes(x=ncells,y=100*frac_clusters, fill=mixing)) + 
+                       mixing=gsub('-donor','',mixing)) %>%
+  dplyr::select(ncells, frac_clusters, mixing, Donor)
+ggplot(dat, aes(x=ncells,y=100*frac_clusters, fill=mixing)) + 
   geom_boxplot(position=position_dodge(width=.8), outlier.shape=NA) +
@@
   labs(x='# cells',y='% shared clusters in pair', fill='pair', color='Donors')
+
+writeData(wb, 'Fig1F', as.data.frame(dat))
 ```
@@
 # bulk VDJ results for Fig. 1
-VDJ_results <- read.xlsx('data/source_data.xlsx',sheet='VDJ data')
+VDJ_results <- read.xlsx('data/input_data.xlsx',sheet='VDJ data')
 
@@
 
-vdj_df %>% 
+dat <- vdj_df %>% 
   dplyr::mutate(ncells=factor(gsub('000$','k',ncells), levels=c('50k','100k')),
@@
   pivot_wider(id_cols=c(ncells,Donor),names_from=method,values_from=c('mean_alpha_ratio','se_alpha_ratio')) %>%
-  ggplot(aes(x=mean_alpha_ratio_VDJ, y=mean_alpha_ratio_SWIBRID,
+  dplyr::select(mean_alpha_ratio_VDJ,
+                se_alpha_ratio_VDJ,
+                mean_alpha_ratio_SWIBRID,
+                se_alpha_ratio_SWIBRID,
+                Donor, ncells)
+
+ggplot(dat, aes(x=mean_alpha_ratio_VDJ, y=mean_alpha_ratio_SWIBRID,
              xmin=mean_alpha_ratio_VDJ-se_alpha_ratio_VDJ,
@@
   labs(y='Fraction IGHA SWIBRID',x='Fraction IGHA VDJ',shape='# cells')
+
+writeData(wb, 'Fig1G', as.data.frame(dat))
 ```
@@
 # HD data for Fig. 2
-HD.data <- read.xlsx('data/source_data.xlsx', sheet='HD data')
+HD.data <- read.xlsx('data/input_data.xlsx', sheet='HD data')
 
@@
 
-HD.var %>%  dplyr::left_join(columns, by='col') %>%
+dat <- HD.var %>%  
+  dplyr::left_join(columns, by='col') %>%
   dplyr::mutate(variable=factor(gsub('Residuals','unexplained',variable),
@@
   dplyr::mutate(col=factor(col, levels=cols.here)) %>%
-  ggplot(aes(x=col, y=100*variance, fill=variable)) + 
+  dplyr::select(col, variance, variable, category)
+
+ggplot(dat, aes(x=col, y=100*variance, fill=variable)) + 
   geom_col(width=.8) +
@@
   labs(x='', y='% Explained variance', fill='')
+
+writeData(wb, 'Fig2B', as.data.frame(dat))
 ```
@@
 
-HD.var %>%  dplyr::left_join(columns, by='col') %>%
+dat <- HD.var %>%  
+  dplyr::left_join(columns, by='col') %>%
   dplyr::mutate(variable=factor(gsub('Residuals','unexplained',variable),
                                 levels=c('Sequencing','Batch','Donor','Sex','Age'))) %>%
-  dplyr::filter(variable!='unexplained', category %in% c('Isotypes','Diversity','Structural','Context','Breaks','Variants')) %>%
-  ggplot(aes(x=variable, y=100*variance)) + 
+  dplyr::filter(variable!='unexplained', category %in% c('Isotypes','Diversity','Structural','Context','Breaks','Variants'))
+
+ggplot(dat, aes(x=variable, y=100*variance)) + 
   geom_boxplot(aes(fill=variable), outlier.shape=NA, lwd=.5, legend=FALSE) + 
@@
   labs(x='', y='% Explained variance', fill='', color='')
+
+writeData(wb, 'Fig2C', as.data.frame(dat))
 ```
@@
 ```{r FigS2F,fig.width=3.5 * fig_scale,fig.height=1.7 * fig_scale}
-HD.var %>%  
+dat <- HD.var %>%  
   dplyr::left_join(columns, by='col') %>%
@@
   dplyr::filter(variable!='Residuals', category %in% c('Isotypes','Diversity','Structural','Context','Breaks','Variants')) %>%
-  ggplot(aes(x=category, y=100*variance, fill=variable)) + 
+  dplyr::select(category, variance, variable, col)
+ggplot(dat, aes(x=category, y=100*variance, fill=variable)) + 
   geom_boxplot(position=position_dodge(width=.8), outlier.shape=NA, lwd=.5) + 
@@
   labs(x='', y='% Explained variance', fill='')
+
+writeData(wb, 'FigS2F', as.data.frame(dat))
 ```
@@
 ```{r FigS2I_1, fig.width=1.4 * fig_scale, fig.height=1.7 * fig_scale}
-tmp <- HD.var %>%  
+dat1 <- HD.var %>%  
   dplyr::left_join(columns, by='col') %>%
@@
 
-pvals <- tmp %>%
+pvals <- dat1 %>%
   dplyr::group_by(variable) %>%
@@
 
-ggplot(tmp,aes(x=variable, y=100*variance, fill=category)) + 
+ggplot(dat1,aes(x=variable, y=100*variance, fill=category)) + 
   geom_boxplot(position=position_dodge(width=.8), outlier.shape=NA, lwd=.5) + 
@@
 ```{r FigS2I_2, fig.width=1.4 * fig_scale, fig.height=1.7 * fig_scale}
-tmp <- HD.var %>%  
+dat2 <- HD.var %>%  
   dplyr::left_join(columns, by='col') %>%
@@
 
-pvals <- tmp %>% 
+pvals <- dat2 %>% 
   dplyr::group_by(variable) %>%
@@
 
-ggplot(tmp, aes(x=variable, y=100*variance, fill=category)) + 
+ggplot(dat2, aes(x=variable, y=100*variance, fill=category)) + 
   geom_boxplot(position=position_dodge(width=.8), outlier.shape=NA, lwd=.5) + 
@@
 ```{r FigS2I_3, fig.width=1.4 * fig_scale,fig.height=1.7 * fig_scale}
-HD.data.br <- read.xlsx('data/source_data.xlsx', sheet='HD data by reads')
+HD.data.br <- read.xlsx('data/input_data.xlsx', sheet='HD data by reads')
 
@@
 
-tmp <- rbind(HD.var %>% 
+dat3 <- rbind(HD.var %>% 
                dplyr::mutate(averaging='Clusters'), 
@@
                 variable %in% c("Donor", "Batch", "Sequencing")) %>% 
-  dplyr::arrange(col,averaging,variable)
-
-pvals <- tmp %>%
+  dplyr::mutate(category=averaging) %>%
+  dplyr::select(col, variable, variance, category) %>%
+  dplyr::arrange(col,category,variable)
+
+pvals <- dat3 %>%
   dplyr::group_by(variable) %>%
-  dplyr::do(wc=wilcox.test(variance ~ averaging, paired=TRUE, data=., exact=FALSE)) %>%
+  dplyr::do(wc=wilcox.test(variance ~ category, paired=TRUE, data=., exact=FALSE)) %>%
   dplyr::summarise(variable, p=wc$p.value)
 
-ggplot(tmp, aes(x=variable, y=100*variance, fill=averaging)) + 
+ggplot(dat3, aes(x=variable, y=100*variance, fill=category)) + 
   geom_boxplot(position=position_dodge(width=.8), outlier.shape=NA, lwd=.5) + 
@@
   labs(x='', y='% Explained variance', fill='Averages over')
+
+writeData(wb, 'FigS2I', rbind(dat1 %>% dplyr::mutate(panel='FigS2I_1'), 
+                              dat2 %>% dplyr::mutate(panel='FigS2I_2'), 
+                              dat3 %>% dplyr::mutate(panel='FigS2I_3')) %>% as.data.frame())
 ```
@@
 ```{r Fig2E, fig.width=.75 * fig_scale, fig.height=1.5 * fig_scale}
-col <- 'break_dispersion_SG1'
-HD.data %>%
-  dplyr::filter(feature == col, Cohort=='C1') %>%
+dat <- HD.data %>%
+  dplyr::filter(feature %in% c('break_dispersion_SG1','occupancy_top_cluster'), Cohort=='C1') %>%
+  dplyr::select(Age, value, feature, Sex)
+
+dat %>% dplyr::filter(feature == 'break_dispersion_SG1') %>%
   ggplot(aes(x=str_to_title(Sex), y=value)) + 
@@
         axis.text.x=element_text(angle=45, hjust=1, vjust=1)) +
-  labs(x='',y=col)
-
-col <- 'occupancy_top_cluster'
-HD.data %>%
-  dplyr::filter(feature==col, Cohort=='C1') %>%
+  labs(x='',y='break_dispersion_SG1')
+
+dat %>% dplyr::filter(feature == 'occupancy_top_cluster') %>%
   ggplot(aes(x=Age, y=value)) + 
@@
         axis.text.x=element_text(angle=45, hjust=1, vjust=1)) +
-  labs(x='',y=col)
+  labs(x='',y='occupancy_top_cluster')
+
+writeData(wb, 'Fig2E', as.data.frame(dat))
 ```
@@
 
-tmp <- as.data.frame(R) %>% 
+dat <- as.data.frame(R) %>% 
   tibble::rownames_to_column('var2') %>%
@@
                 i1=as.vector(sapply(var1, function(x) which(io==x))),
-                i2=as.vector(sapply(var2, function(x) which(io==x))))
+                i2=as.vector(sapply(var2, function(x) which(io==x)))) %>%
+  dplyr::filter(i1 <= i2) %>%
+  dplyr::select(var1, var2, R)
 
@@
 
-ggplot(tmp %>%
-  dplyr::filter(i1 <= i2),
-  aes(x=var1,y=var2,fill=R)) + 
+ggplot(dat, aes(x=var1,y=var2,fill=R)) + 
   geom_tile() +
@@
 
+writeData(wb, 'FigS2H', as.data.frame(dat)) 
 ```
@@
 
-ggplot(cohort.data, aes(x=Cohort, y=value)) + 
+dat <- cohort.data %>%
+  dplyr::select(Cohort, value, Donor, feature, category)
+ggplot(dat, aes(x=Cohort, y=value)) + 
   geom_boxplot(aes(group=Cohort), position=position_dodge(width=.75), width=.7, outlier.shape=NA, lwd=.25, fill='gray80') + 
-  geom_pointrange(data=cohort.data %>%
+  geom_pointrange(data=dat %>%
                     dplyr::group_by(Donor,feature,category,Cohort) %>%
@@
   labs(x='',y='')
+
+writeData(wb, 'Fig2D', as.data.frame(dat))
 ```
@@
 ```{r Fig2F, fig.width=4. * fig_scale, fig.height=1.75 * fig_scale}
-mouse.data <- read.xlsx('data/source_data.xlsx', sheet='mouse tissues') %>%
+mouse.data <- read.xlsx('data/input_data.xlsx', sheet='mouse tissues') %>%
   tibble::column_to_rownames('Sample')
-vincendeau.data <- read.xlsx('data/source_data.xlsx', sheet='Vincendeau data') %>%
+vincendeau.data <- read.xlsx('data/input_data.xlsx', sheet='Vincendeau data') %>%
   tibble::column_to_rownames('Sample')
@@
 
-cohort.data %>%
-  ggplot(aes(x=Cohort, y=value, fill=species)) + 
+dat <- cohort.data %>%
+  dplyr::select(Cohort, value, species, variable)
+
+ggplot(dat, aes(x=Cohort, y=value, fill=species)) + 
   geom_boxplot(position=position_dodge(width=.75), width=.7, outlier.shape=NA, lwd=.25) + 
@@
   labs(x='',y='', fill='')
+
+writeData(wb, 'Fig2F', as.data.frame(dat))
 ```
@@
 cols.here.mouse <- c('read_length_SG_mean')
-cohort.data <- rbind(mouse.data %>%
+dat <- rbind(mouse.data %>%
                        dplyr::mutate(species='mouse tissue',
@@
   dplyr::left_join(columns, by=c('variable'='col')) %>%
-  dplyr::filter(category %in% c('Diversity','Isotypes','Breaks','Context','Structural'))
-
-cohort.data %>%
-  ggplot(aes(x=Cohort, y=value, fill=species)) + 
+  dplyr::filter(category %in% c('Diversity','Isotypes','Breaks','Context','Structural')) %>%
+  dplyr::select(Cohort, value, species, variable)
+
+ggplot(dat, aes(x=Cohort, y=value, fill=species)) + 
   geom_boxplot(position=position_dodge(width=.75), width=.7, outlier.shape=NA, lwd=.25) + 
@@
   labs(x='',y='', fill='')
+
+writeData(wb, 'FigS2J', as.data.frame(dat))
 ```
@@
 ```{r FigS2G, fig.width=2.75 * fig_scale, fig.height=5 * fig_scale}
-tmp <- HD.scaled %>%
+dat <- HD.scaled %>%
   dplyr::filter(feature %in% selected.cols, Batch=='20230313') %>%
@@
                                  dplyr::pull(col))) %>%
+  dplyr::select(minION, pacBio, feature, category) %>%
   dplyr::arrange(feature)
 
-cols_shapes <- tmp %>% 
+cols_shapes <- dat %>% 
   dplyr::distinct(category, feature) %>%
@@
   dplyr::mutate(var_ind=1:n()) %>%
-  dplyr::mutate(feature=factor(feature, levels=levels(tmp$feature))) %>% 
+  dplyr::mutate(feature=factor(feature, levels=levels(dat$feature))) %>% 
   dplyr::arrange(feature)
 
-ggplot(tmp,aes(x=minION,y=pacBio)) +
+ggplot(dat,aes(x=minION,y=pacBio)) +
   geom_abline(slope=1,intercept=0,color='gray',linewidth=.5) +
@@
         legend.key.spacing.y=unit(0, 'mm'))
+
+writeData(wb, 'FigS2G', as.data.frame(dat))
 ```
@@
 
-ggplot(pcadat,aes(x=PC1,y=PC2, color=Genotype)) +
+dat <- pcadat %>%
+  dplyr::select(PC1, PC2, Genotype)
+
+ggplot(dat,aes(x=PC1,y=PC2, color=Genotype)) +
   geom_point(size=1) +
-  geom_polygon(data = pcadat %>%
+  geom_polygon(data = dat %>%
                  dplyr::group_by(Genotype) %>%
@@
   scale_fill_manual(values=genotype_colors)
+
+writeData(wb, 'Fig2G', as.data.frame(dat), rowNames=TRUE)
 ```
@@
 
+dat <- tmp %>%
+  dplyr::select(-c(var_col,category))
+
 plots <- lapply(comparisons, function(x) 
-  tmp %>%
+  dat %>%
     dplyr::mutate_(mean1=paste0('mean_',x[[1]]),
@@
           get_legend(plots[[1]] + theme(legend.margin=margin(0,0,0,0,'mm'))), ncol=1, rel_heights=c(.65,.35))
+
+writeData(wb, 'Fig2H', as.data.frame(dat))
 ```
@@
 ```{r FigS3A, fig.width=5.5 * fig_scale, fig.height=4.5 * fig_scale}
-HTGTS.data <- read.xlsx("data/source_data.xlsx", sheet='HTGTS data') %>%
+HTGTS.data <- read.xlsx("data/input_data.xlsx", sheet='HTGTS data') %>%
   tibble::column_to_rownames('Sample')
@@
 
-tmp <- HTGTS.data %>%
+dat <- HTGTS.data %>%
   dplyr::mutate(celltype=gsub('spleen','Spleen',celltype)) %>%
@@
 
-cols_shapes <- tmp %>% 
+cols_shapes <- dat %>% 
   dplyr::distinct(category, variable) %>%
@@
   dplyr::mutate(var_ind=1:n()) %>%
-  dplyr::mutate(variable=factor(variable, levels=levels(tmp$variable))) %>% 
+  dplyr::mutate(variable=factor(variable, levels=levels(dat$variable))) %>% 
   dplyr::arrange(variable)
@@
 plots <- lapply(comparisons, function(x) 
-  tmp %>%
+  dat %>%
     dplyr::mutate_(mean1=paste0('mean_',x[[1]],'_',x[[2]]),
@@
           get_legend(plots[[1]] + theme(legend.margin=margin(0,0,0,0,'mm'))), ncol=1, rel_heights=c(.65,.35))
+
+writeData(wb, 'FigS3A', dat %>% dplyr::select(-c(category, var_col)) %>% as.data.frame())
 ```
@@
 
-HTGTS.data %>%
+dat <- HTGTS.data %>%
   dplyr::mutate(celltype=factor(gsub('spleen','Spleen',celltype), levels=c('CH12','Spleen')),
@@
   dplyr::filter(category %in% c('Diversity','Isotypes','Breaks','Context','Structural')) %>%
-  ggplot(aes(x=Genotype, y=value, fill=Genotype)) + 
+  dplyr::select(Genotype, value, celltype, variable)
+
+ggplot(dat, aes(x=Genotype, y=value, fill=Genotype)) + 
   geom_boxplot(position=position_dodge(width=.75), width=.7, outlier.shape=NA, lwd=.25) + 
@@
   labs(x='',y='', fill='')
+
+writeData(wb, 'FigS3C', as.data.frame(dat))
 ```
@@
 ```{r FigS3D, fig.width=3.25 * fig_scale, fig.height=1.25 * fig_scale}
-read.xlsx('data/source_data.xlsx', sheet='HTGTS homology') %>%
+dat <- read.xlsx('data/input_data.xlsx', sheet='HTGTS homology') %>%
   dplyr::mutate(Genotype=factor(Genotype, levels=c('WT','Atm','Lig4','Trp53bp1')),
@@
                    se_frac=sd(frac)/sqrt(n())) %>%
-  ggplot(aes(x=n_homology, y=mean_frac, color=Genotype)) +
+  dplyr::select(n_homology, mean_frac, se_frac, Genotype, celltype)
+
+ggplot(dat, aes(x=n_homology, y=mean_frac, color=Genotype)) +
   geom_line(aes(group=Genotype)) +
@@
   coord_cartesian(clip='off')
+
+writeData(wb, 'FigS3D', as.data.frame(dat))
 ```
@@
 
-p1 <- ggplot(pcadat.CH12,aes(x=PC1,y=PC2, color=Genotype)) +
+dat <- rbind(pcadat.CH12 %>% dplyr::mutate(material='CH12') %>%
+               dplyr::select(PC1, PC2, Genotype, material),
+             pcadat.spleen %>% dplyr::mutate(material='spleen') %>%
+               dplyr::select(PC1, PC2, Genotype, material))
+
+p1 <- ggplot(dat %>% dplyr::filter(material=='CH12'),aes(x=PC1,y=PC2, color=Genotype)) +
   geom_point(size=1) +
-  geom_polygon(data = pcadat.CH12 %>%
+  geom_polygon(data = dat %>% dplyr::filter(material=='CH12') %>%
                  dplyr::group_by(Genotype) %>%
@@
 
-p2 <- ggplot(pcadat.spleen,aes(x=PC1,y=PC2, color=Genotype)) +
+p2 <- ggplot(dat %>% dplyr::filter(material=='spleen'),aes(x=PC1,y=PC2, color=Genotype)) +
   geom_point(size=1) +
-  geom_polygon(data = pcadat.spleen %>%
+  geom_polygon(data = dat %>% dplyr::filter(material=='spleen') %>%
                  dplyr::group_by(Genotype) %>%
@@
 plot_grid(p1,p2, ncol=1, align='vh')
+
+writeData(wb, 'FigS3B', as.data.frame(dat), rowNames=TRUE)
 ```
@@
 ```{r Fig3B, fig.width=3.5 * fig_scale, fig.height=1.75 * fig_scale}
-CH12.data <- read.xlsx('data/source_data.xlsx',sheet='CH12 data') %>%
+CH12.data <- read.xlsx('data/input_data.xlsx',sheet='CH12 data') %>%
   tibble::column_to_rownames('Sample')
-CH12.data.noSg <- read.xlsx('data/source_data.xlsx',sheet='CH12 data (no Sg)') %>%
+CH12.data.noSg <- read.xlsx('data/input_data.xlsx',sheet='CH12 data (no Sg)') %>%
   tibble::column_to_rownames('Sample')
-mouse.data.noSg <- read.xlsx('data/source_data.xlsx',sheet='mouse tissues (no Sg)')
+mouse.data.noSg <- read.xlsx('data/input_data.xlsx',sheet='mouse tissues (no Sg)')
 
@@
 
-p1 <- ggplot(pcadat,aes(x=PC1,y=PC2, color=material)) +
-  geom_point(aes(shape=Genotype),size=1) +
-  geom_polygon(data = pcadat %>%
-                 dplyr::group_by(material) %>%
-                 dplyr::slice(chull(PC1, PC2)), alpha = 0.25, aes(fill=material)) +
-  my_theme() +
-  scale_fill_manual(values=tissue_colors) +
-  scale_color_manual(values=tissue_colors) +
-  labs(x=paste0('PC 1 (',100*signif(pca$sdev[1]**2/sum(pca$sdev**2),3),'%)'),
-       y=paste0('PC 2 (',100*signif(pca$sdev[2]**2/sum(pca$sdev**2),3),'%)'),
-       color='',shape='',fill='', title='all switch regions')
-
 CH12.take.noSg <- rep(TRUE, dim(CH12.data.noSg)[1]) 
@@
 
-p2 <- ggplot(pcadat.noSg,aes(x=PC1,y=PC2, color=material)) +
+dat <- rbind(pcadat %>%
+               dplyr::mutate(set='all regions')  %>%
+               dplyr::select(PC1, PC2, material, Genotype, set),
+             pcadat.noSg %>%
+               dplyr::mutate(set='no Sg') %>%
+               dplyr::select(PC1, PC2, material, Genotype, set))
+
+p1 <- ggplot(dat %>% dplyr::filter(set=='all regions'),aes(x=PC1,y=PC2, color=material)) +
+  geom_point(aes(shape=Genotype),size=1) +
+  geom_polygon(data = dat %>%
+                 dplyr::filter(set=='all regions') %>%
+                 dplyr::group_by(material) %>%
+                 dplyr::slice(chull(PC1, PC2)), alpha = 0.25, aes(fill=material)) +
+  my_theme() +
+  scale_fill_manual(values=tissue_colors) +
+  scale_color_manual(values=tissue_colors) +
+  labs(x=paste0('PC 1 (',100*signif(pca$sdev[1]**2/sum(pca$sdev**2),3),'%)'),
+       y=paste0('PC 2 (',100*signif(pca$sdev[2]**2/sum(pca$sdev**2),3),'%)'),
+       color='',shape='',fill='', title='all switch regions')
+
+p2 <- ggplot(dat %>% dplyr::filter(set=='no Sg'),aes(x=PC1,y=PC2, color=material)) +
   geom_point(aes(shape=Genotype), size=1) +
-  geom_polygon(data = pcadat.noSg %>%
+  geom_polygon(data = dat %>%
+                 dplyr::filter(set=='no Sg') %>%
                  dplyr::group_by(material) %>%
@@
           ncol=3)
+
+writeData(wb, 'Fig3B', as.data.frame(dat), rowNames=TRUE)
 ```
@@
 
-tmp <- CH12.data %>% 
+dat <- CH12.data %>% 
   dplyr::filter(Batch != '20240830') %>%
@@
   gather(col,value,-c(Genotype,ng,Genotype_new)) %>% 
-  dplyr::mutate(col=factor(col, levels=CH12.features))
-
-ggplot(tmp, aes(x=Genotype,y=value,fill=Genotype)) + 
+  dplyr::mutate(col=factor(col, levels=CH12.features)) %>%
+  dplyr::select(Genotype, value, col)
+
+ggplot(dat, aes(x=Genotype,y=value,fill=Genotype)) + 
   geom_boxplot(outlier.shape=NA, lwd=.5) + 
@@
   labs(x='',y='', color='# Reads')
+
+writeData(wb, 'Fig3D', as.data.frame(dat))
 ```
@@
 
-ggplot(pcadat,aes(x=PC1,y=PC2, color=Genotype)) +
+dat <- pcadat %>%
+  dplyr::select(PC1, PC2, Genotype)
+
+ggplot(dat,aes(x=PC1,y=PC2, color=Genotype)) +
   geom_point(size=2) +
-  geom_polygon(data = pcadat %>%
+  geom_polygon(data = dat %>%
                  dplyr::group_by(Genotype) %>%
@@
   scale_color_manual(values=genotype_colors)
+
+writeData(wb, 'Fig3E', as.data.frame(dat), rowNames=TRUE)
 ```
@@
 
+dat <- rbind(t(CH12.meta[samples, c('Genotype','Batch')]),
+             t(CH12.X[samples,CH12.cols]))
+
+writeData(wb, 'Fig3C', as.data.frame(dat), rowNames=TRUE)
 ```
@@
 
-tmp <- as.data.frame(predict(mod, type = "response", 
+dat <- as.data.frame(predict(mod, type = "response", 
                               newx=as.matrix(CH12.X[test,]))) %>%
@@
   dplyr::mutate(sample=factor(gsub('20240830_','',sample),
-                              levels=gsub('20240830_','',unique(sample)))) 
-ggplot(tmp, aes(x=prediction,y=sample,
+                              levels=gsub('20240830_','',unique(sample)))) %>%
+  dplyr::select(prediction, sample, p, c)
+
+ggplot(dat, aes(x=prediction,y=sample,
                 label=round(p,2),fill=p,color=c)) + 
   geom_tile() + 
-  geom_text(data=. %>% dplyr::filter(p > .25),
+  geom_text(data=dat %>% dplyr::filter(p > .25),
             size=2) +
@@
   labs(x='',y='36 new samples')
+
+writeData(wb, 'Fig3G', as.data.frame(dat))
 ```
@@
 
-tmp %>%
+dat <- tmp %>%
   gather(Genotype,coef,-col) %>%
@@
                 col=factor(col, levels=rev(levels(tmp$col)))) %>%
-  ggplot(aes(x=Genotype, y=col, fill=coef, label=round(coef,2))) + 
+  dplyr::select(Genotype, col, coef)
+
+ggplot(dat, aes(x=Genotype, y=col, fill=coef, label=round(coef,2))) + 
   geom_tile() +
@@
   labs(x='',y='')
+
+writeData(wb, 'Fig3F', as.data.frame(dat))
 ```
@@
 ```{r Fig4B,fig.width=1.25 * fig_scale, fig.height=1.7 * fig_scale}
-FB.data <- read.xlsx('data/source_data.xlsx', sheet='CVID data') %>%
+FB.data <- read.xlsx('data/input_data.xlsx', sheet='CVID data') %>%
   dplyr::mutate(Batch=as.character(Batch)) %>%
@@
 
-FB.var %>%  dplyr::left_join(columns, by='col') %>%
+dat <- FB.var %>%  dplyr::left_join(columns, by='col') %>%
   dplyr::mutate(variable=factor(variable,levels=c('Donor','Diagnosis','Sex','Age','Batch','Treatment','IV',
@@
   dplyr::filter(variable!='Residuals', category %in% c('Isotypes','Diversity','Structural','Context','Breaks','Variants')) %>%
-  ggplot(aes(x=variable, y=100*variance)) + 
+  dplyr::select(variable, variance, col)
+
+ggplot(dat, aes(x=variable, y=100*variance)) + 
   geom_boxplot(aes(fill=variable), outlier.shape=NA, lwd=.5, legend=FALSE) + 
@@
   labs(x='', y='% Explained variance', fill='', color='')
+
+writeData(wb, 'Fig4B', as.data.frame(dat)) 
 ```
@@
 ```{r FigS4A,fig.width=3.5 * fig_scale,fig.height=1.7 * fig_scale}
-FB.var %>%  dplyr::left_join(columns, by='col') %>%
+dat <- FB.var %>%  dplyr::left_join(columns, by='col') %>%
   dplyr::mutate(variable=factor(variable, levels=c('Donor','Diagnosis','Sex','Age','Batch','Treatment','IV','Residuals'))) %>%
   dplyr::filter(variable!='Residuals', category %in% c('Isotypes','Diversity','Structural','Context','Breaks','Variants')) %>%
-  ggplot(aes(x=category, y=100*variance, fill=variable)) + 
+  dplyr::select(category, variance, variable, col)
+ggplot(dat, aes(x=category, y=100*variance, fill=variable)) + 
   geom_boxplot(position=position_dodge(width=.8), outlier.shape=NA, lwd=.5) + 
@@
   labs(x='', y='% Explained variance', fill='')
+
+writeData(wb, 'FigS4A', as.data.frame(dat))
 ```
@@
 ```{r Fig4D, fig.width=2.25 * fig_scale, fig.height=2 * fig_scale} 
-FB.data.pooled <- read.xlsx('data/source_data.xlsx',sheet='CVID data pooled') %>%
+FB.data.pooled <- read.xlsx('data/input_data.xlsx',sheet='CVID data pooled') %>%
   dplyr::mutate(status=ifelse(Diagnosis %in% c('Control','Subclass','Atypical'), Diagnosis, 'CVID'),
@@
   tibble::column_to_rownames('Sample')
-QP.data <- read.xlsx('data/source_data.xlsx', sheet='DNA repair data') %>%
+QP.data <- read.xlsx('data/input_data.xlsx', sheet='DNA repair data') %>%
   dplyr::mutate(Diagnosis=plyr::revalue(Diagnosis,c('control'='Control'))) %>%
@@
 
-ggplot(pcadat,aes(x=PC1,y=PC2)) +
+dat <- pcadat %>% 
+  dplyr::select(PC1, PC2, Diagnosis, Donor, status)
+
+ggplot(dat,aes(x=PC1,y=PC2)) +
   geom_point(aes(color=Diagnosis),size=1) +
   geom_line(aes(group=Donor, color=Diagnosis)) +
-  geom_polygon(data = pcadat %>%
+  geom_polygon(data = dat %>%
                  dplyr::group_by(status) %>%
@@
        color='',shape='',fill='')
+
+writeData(wb, 'Fig4D', as.data.frame(dat), rowNames=TRUE)
 ```
@@
 
-tmp <- FB.data.pooled %>%
+dat <- FB.data.pooled %>%
   dplyr::filter(Sample_donor %in% FB_train_donors, Diagnosis %in% c('Control','Subclass')) %>%
@@
   dplyr::mutate(feature=factor(gsub('pct_clusters_','',feature),
-                               levels=gsub('pct_clusters_','',subclass.cols)))
-
-pvals <- tmp %>%
+                               levels=gsub('pct_clusters_','',subclass.cols))) %>%
+  dplyr::select(feature, Diagnosis, frac_clusters, Donor)
+
+pvals <- dat %>%
   dplyr::group_by(feature) %>%
@@
 
-ggplot(tmp, aes(x=feature, y=frac_clusters)) +
+ggplot(dat, aes(x=feature, y=frac_clusters)) +
   geom_boxplot(aes(color=Diagnosis), position=position_dodge(width=.8),width=.7,outlier.shape=NA) +
   geom_point(aes(color=Diagnosis), position=position_dodge(width=.8), size=.5) +
-  geom_line(data=tmp %>% dplyr::filter(Diagnosis=='Control'),
+  geom_line(data=dat %>% dplyr::filter(Diagnosis=='Control'),
             aes(group=Donor), position=position_nudge(x=-.2), alpha=.25, size=.25) +
-  geom_line(data=tmp %>% dplyr::filter(Diagnosis=='Ssubclass'),
+  geom_line(data=dat %>% dplyr::filter(Diagnosis=='Ssubclass'),
             aes(group=Donor), position=position_nudge(x=+.2), alpha=.25, size=.25) +
@@
   labs(color='',x='', y='Fraction clusters')
+
+writeData(wb, 'FigS4D', as.data.frame(dat))
 ```
@@
   dplyr::mutate(status_R=paste0(status," (R=",signif(cor(alpha_ratio_FACS,alpha_ratio_clusters,use='na.or.complete'),2),")"))
-ggplot(tmp, aes(x=alpha_ratio_FACS, y=alpha_ratio_clusters, color=status)) + 
+
+dat <- tmp %>%
+  dplyr::select(alpha_ratio_FACS, alpha_ratio_clusters, status, Donor)
+
+ggplot(dat, aes(x=alpha_ratio_FACS, y=alpha_ratio_clusters, color=status)) + 
   geom_abline(slope=1, intercept=0, size=.5) +
@@
         legend.title=element_blank())
+
+writeData(wb, 'FigS4E', as.data.frame(dat))
 ```
@@
 
-tmp <- FB.data.pooled[FB.data.pooled$Sample_donor %in% FB_train_donors,] %>% 
+dat <- FB.data.pooled[FB.data.pooled$Sample_donor %in% FB_train_donors,] %>% 
   dplyr::select(Donor,Diagnosis,FB.features) %>%
@@
   gather(col,value,-c(Diagnosis_simple,ng,Diagnosis_new,Donor,Diagnosis)) %>% 
-  dplyr::mutate(col=factor(col, levels=FB.features))
-
-ggplot(tmp, aes(x=Diagnosis_simple,y=value,fill=Diagnosis_simple)) + 
+  dplyr::mutate(col=factor(col, levels=FB.features)) %>%
+  dplyr::select(Diagnosis_simple, value, col,Donor)
+
+ggplot(dat, aes(x=Diagnosis_simple,y=value,fill=Diagnosis_simple)) + 
   geom_boxplot(outlier.shape=NA, lwd=.5) + 
@@
   labs(x='',y='', color='# Reads')
+
+writeData(wb, 'Fig4C', as.data.frame(dat))
 ```
@@
 
-ggplot(pcadat.combined %>% dplyr::filter(dataset!='validation'),
-       aes(x=PC1,y=PC2)) +
-  geom_polygon(data = pcadat.combined %>% dplyr::filter(dataset=='training') %>%
+dat <- pcadat.combined %>%
+  dplyr::filter(dataset!='validation') %>%
+  dplyr::select(PC1, PC2, dataset, status, Diagnosis, Donor)
+
+ggplot(dat, aes(x=PC1,y=PC2)) +
+  geom_polygon(data = dat %>% dplyr::filter(dataset=='training') %>%
                  dplyr::group_by(status) %>%
@@
        color='',shape='',fill='',alpha='')
+
+writeData(wb, 'Fig4E', as.data.frame(dat), rowNames=TRUE)
 ```
@@
 ```{r Fig4F, fig.width=2.25 * fig_scale, fig.height=2 * fig_scale}
-ggplot(pcadat.combined %>% dplyr::filter(dataset!='testing'),
-       aes(x=PC1,y=PC2)) +
-  geom_polygon(data = pcadat.combined %>% dplyr::filter(dataset=='training') %>%
+dat <- pcadat.combined %>%
+  dplyr::filter(dataset!='testing') %>%
+  dplyr::select(PC1, PC2, dataset, status, Diagnosis, Donor)
+
+ggplot(dat, aes(x=PC1,y=PC2)) +
+  geom_polygon(data = dat %>% dplyr::filter(dataset=='training') %>%
                  dplyr::group_by(status) %>%
@@
        color='',shape='',fill='',alpha='')
+
+writeData(wb, 'Fig4F', as.data.frame(dat), rowNames=TRUE)
 ```
@@
 
-FB_res <- data.frame(specificity=c(ROC_1$specificities,ROC_2$specificities),
+dat <- data.frame(specificity=c(ROC_1$specificities,ROC_2$specificities),
                      sensitivity=c(ROC_1$sensitivities,ROC_2$sensitivities),
@@
   dplyr::mutate(model=factor(model, levels=unique(model))) %>%
-  dplyr::arrange(model,sensitivity)
-
-ggplot(FB_res, aes(x=specificity,y=sensitivity,color=model)) + 
+  dplyr::arrange(model,sensitivity)  %>%
+  dplyr::select(specificity, sensitivity, model)
+
+ggplot(dat, aes(x=specificity,y=sensitivity,color=model)) + 
   geom_step() +
@@
   labs(color='',linetype='')
+
+writeData(wb, 'Fig4G', as.data.frame(dat))
 ```
@@
 
-tmp %>%
+dat <- tmp %>%
   dplyr::mutate(col=factor(col, levels=rev(levels(tmp$col)))) %>%
-  ggplot(aes(x=col, y=1, fill=coef, label=round(coef,2))) + 
+  dplyr::select(col, coef)
+
+ggplot(dat, aes(x=col, y=1, fill=coef, label=round(coef,2))) + 
   geom_tile() +
@@
   labs(x='',y='')
+
+writeData(wb, 'FigS4G', as.data.frame(dat))
 ```
@@
 
-hm_meta <- FB.meta[,c('Diagnosis','Sex','Age','Donor','Timepoint' ,'Complication'),drop=FALSE] %>% 
+hm_meta <- FB.meta[,c('Diagnosis','Sex','Age','Donor','Timepoint'),drop=FALSE] %>% 
   tibble::remove_rownames() %>%
@@
                                           'CVID','ICOS','Kabuki',
-                                          'AID','IgG4def','NFKB2','PI3K','TNFRSF13B')),
-                Complication=factor(Complication, levels=c('NA','io','c'))) %>%
+                                          'AID','IgG4def','NFKB2','PI3K','TNFRSF13B'))) %>%
   dplyr::left_join(class_df,by='name') %>%
@@
                cluster_columns=dend,
-               cluster_rows=TRUE,
-               show_row_names=TRUE,
-               show_column_names=TRUE,
-               show_row_dend=TRUE,
-               top_annotation=HeatmapAnnotation(df=hm_meta,
-                                                show_annotation_name=TRUE,
-                                                show_legend=TRUE,
-                                                simple_anno_size=unit(2,'mm'),
-                                                annotation_legend_param=list(title_gp=gpar(fontsize=7),
-                                                                             labels_gp=gpar(fontsize=7),
-                                                                             legend_height = unit(5, "mm"),
-                                                                             legend_width = unit(3, "mm"),
-                                                                             grid_height = unit(3, "mm"),
-                                                                             grid_width = unit(3, "mm")),
-                                                annotation_name_gp=gpar(fontsize=7),
-                                                col=list('Diagnosis'=diagnosis_colors,
-                                                         'Class'=class_colors,
-                                                         'Age'=colorRamp2(c(20,50,80),c("yellow2","orange2","coral3")),
-                                                         'Sex'=c('male'='mistyrose2',
-                                                                 'female'='lightseagreen'),
-                                                         'Complication'=c('NA'='gray80',
-                                                                          'io'='blue3',
-                                                                          'c'='red3'))),
-               column_names_gp=gpar(fontsize=4),
-               row_names_gp =gpar(fontsize = 7),
-               row_dend_width = unit(7, "mm"),
-               row_split = columns %>% tibble::column_to_rownames('col') %>% .[selected.cols, 'category'], 
-               heatmap_legend_param=list(labels_gp=gpar(fontsize=7),
-                                         title_gp=gpar(fontsize=7),
-                                         at=c(-2,0,2),
-                                         grid_height=unit(4, 'mm'),
-                                         grid_width=unit(3, 'mm'),
-                                         legend_height=unit(5,'mm'),
-                                         legend_width=unit(3,'mm')),
-               row_title=NULL,
-               name='z-score')
-
-hm2 <- Heatmap(t(FB.X),
-               col=colorRamp2(c(-2,0, 2), c("blue",'gray90', "red")),
-               column_split = grepl('09|32', row.names(FB.X)), 
-               column_title = NULL,
                cluster_rows=TRUE,
@@
 
+# hm2 <- Heatmap(t(FB.X),
+#                col=colorRamp2(c(-2,0, 2), c("blue",'gray90', "red")),
+#                column_split = grepl('09|32', row.names(FB.X)), 
+#                column_title = NULL,
+#                cluster_rows=TRUE,
+#                show_row_names=TRUE,
+#                show_column_names=TRUE,
+#                show_row_dend=TRUE,
+#                top_annotation=HeatmapAnnotation(df=hm_meta,
+#                                                 show_annotation_name=TRUE,
+#                                                 show_legend=TRUE,
+#                                                 simple_anno_size=unit(2,'mm'),
+#                                                 annotation_legend_param=list(title_gp=gpar(fontsize=7),
+#                                                                              labels_gp=gpar(fontsize=7),
+#                                                                              legend_height = unit(5, "mm"),
+#                                                                              legend_width = unit(3, "mm"),
+#                                                                              grid_height = unit(3, "mm"),
+#                                                                              grid_width = unit(3, "mm")),
+#                                                 annotation_name_gp=gpar(fontsize=7),
+#                                                 col=list('Diagnosis'=diagnosis_colors,
+#                                                          'Class'=class_colors,
+#                                                          'Age'=colorRamp2(c(20,50,80),c("yellow2","orange2","coral3")),
+#                                                          'Sex'=c('male'='mistyrose2',
+#                                                                  'female'='lightseagreen'))),
+#                column_names_gp=gpar(fontsize=4),
+#                row_names_gp =gpar(fontsize = 7),
+#                row_dend_width = unit(7, "mm"),
+#                row_split = columns %>% tibble::column_to_rownames('col') %>% .[selected.cols, 'category'], 
+#                heatmap_legend_param=list(labels_gp=gpar(fontsize=7),
+#                                          title_gp=gpar(fontsize=7),
+#                                          at=c(-2,0,2),
+#                                          grid_height=unit(4, 'mm'),
+#                                          grid_width=unit(3, 'mm'),
+#                                          legend_height=unit(5,'mm'),
+#                                          legend_width=unit(3,'mm')),
+#                row_title=NULL,
+#                name='z-score')
+
 hm3 <- Heatmap(t(pca_hm[,c('PC1','PC2')]),
@@
 
-hm2 %v% hm3
-
+#hm2 %v% hm3
+
+dat <- rbind(t(hm_meta),
+             t(FB.X),
+             t(pca_hm[,c('PC1','PC2')]))
+
+writeData(wb, 'Fig4H', as.data.frame(dat), rowNames=TRUE)
 ```
@@
 
-ggplot(tmp,aes(x=abs(PC1),y=abs(PC2), color=col, shape=col)) +
+dat <- tmp %>%
+  dplyr::select(PC1, PC2, col)
+
+ggplot(dat,aes(x=abs(PC1),y=abs(PC2), color=col, shape=col)) +
   geom_point(size=1) +
@@
        color='',shape='',fill='',alpha='')
+
+writeData(wb, 'FigS4C', as.data.frame(dat))
 ```
@@
 ```{r FigS4H, fig.width=2.5 * fig_scale, fig.height=2 * fig_scale}
-ggplot(pca_hm %>% tibble::rownames_to_column('name') %>%
+dat <- pca_hm %>% tibble::rownames_to_column('name') %>%
   dplyr::filter(dataset!='validation') %>%
-         dplyr::left_join(class_df, by='name'), 
-       aes(x=PC1,y=PC2, color=Class, shape=dataset)) +
+  dplyr::left_join(class_df, by='name') %>%
+  dplyr::select(PC1, PC2, Class, dataset, Donor)
+
+ggplot(dat, aes(x=PC1,y=PC2, color=Class, shape=dataset)) +
   geom_point(size=1) +
@@
        color='')
+
+writeData(wb, 'FigS4H', as.data.frame(dat))
 ```
@@
 
-tmp <- cbind(rbind(FB.data.pooled[,features.here],
+dat <- cbind(rbind(FB.data.pooled[,features.here],
                    QP.data[,features.here]),
@@
   gather(col,value,features.here) %>%
-  dplyr::mutate(col=factor(col, levels=features.here))
-
-ggplot(tmp, aes(x=ifelse(grepl('Cluster',group),as.numeric(group),1+as.numeric(group)),
+  dplyr::mutate(col=factor(col, levels=features.here)) %>%
+  dplyr::select(group, value, col, Donor)
+
+ggplot(dat, aes(x=ifelse(grepl('Cluster',group),as.numeric(group),1+as.numeric(group)),
                 y=value,fill=group)) +
@@
   labs(x='',y='', color='Diagnosis')
+
+writeData(wb, 'Fig4J', as.data.frame(dat))
 ```
@@
 
-FB.data.pooled[FB.samples, ] %>%
+dat <- FB.data.pooled[FB.samples, ] %>%
   dplyr::mutate(Tetanus=ifelse(status=='Control','Positive',str_to_title(Tetanus)),
@@
                                    c('Negative','Positive','No data'))) %>%
-  ggplot(aes(x=vacc_status, y=value, fill=vacc_status)) +
+  dplyr::select(vacc_status, value, feature, vacc, status)
+
+ggplot(dat, aes(x=vacc_status, y=value, fill=vacc_status)) +
   geom_boxplot(outlier.shape=NA) +
@@
         axis.text.x=element_text(angle=45, hjust=1, vjust=1))
+
+writeData(wb, 'FigS4B', as.data.frame(dat))
 ```
@@
 hm2
+
+dat <- rbind(t(FB.meta.aggregated[samples,]),
+             t(FB.X.aggregated[samples,]))
+
+writeData(wb, 'FigS4F', as.data.frame(dat), rowNames=TRUE)
+```
+
+```{r save}
+saveWorkbook(wb, file = "data/source_data_file.xlsx", overwrite = TRUE)
 ```

### 2025-11-13 prepare_input.R

@@ -521,4 +521,4 @@ VDJ_results <- lapply(vdj_samples, function(sample)
   dplyr::select(cloneId,readCount,uniqueMoleculeCount,allVHitsWithScore,allCHitsWithScore,aaSeqCDR3,Sample,Donor,ncells)
 writeData(wb, 'VDJ data', VDJ_results)
 
-saveWorkbook(wb, file = "data/source_data.xlsx", overwrite = TRUE)
+saveWorkbook(wb, file = "data/input_data.xlsx", overwrite = TRUE)
